---
title: "cfu_counts"
output: html_document
date: "2024-07-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggpubr)
library(ggplot2)
library(qiime2R)
library(tidyverse)
library(ggh4x)
library(viridis)
library(ggrepel)
library(cowplot)
library(grid)
library(gridExtra)
```
**Functions**
```{r}
## plot!!
make_plot <- function(input_table,
                      wanted_experiment,
                      wanted_vendor,
                      x_axis,
                      y_axis,
                      alpha_value,
                      color_value,
                      shape_value,
                      point_size,
                      y_labels,
                      color_legend_title,
                      color_labels,
                      alpha_legend_title,
                      alpha_labels,
                      shape_legend_title,
                      shape_labels,
                      x_name,
                      y_name,
                      wanted_title,
                      wanted_subtitle){
  ifelse(class(wanted_experiment) == "character", 
         filt_input_table <- input_table %>% 
                        mutate(microbe_presence = as.factor(microbe_presence)) %>% 
                        na.omit() %>% 
                        filter(experiment == wanted_experiment,
                               vendor == wanted_vendor),
         filt_input_table <- input_table %>% 
                        mutate(microbe_presence = as.factor(microbe_presence)) %>% 
                        na.omit() %>% 
                        filter(vendor == wanted_vendor))
  
  plot <- filt_input_table %>% 
            ggplot(aes(x = .data[[x_axis]], y = .data[[y_axis]])) +
            theme_bw(base_size = 20) +
            theme(axis.text.x = element_blank()) +
            geom_point(aes(color = .data[[color_value]], shape = .data[[shape_value]],
                           alpha = .data[[alpha_value]]), 
                       size = point_size,
                       stroke = 2) +
            scale_y_discrete(labels = y_labels) +
            scale_color_viridis(option = 'H', 
                               name = color_legend_title,
                               discrete = TRUE,
                               labels = color_labels) +
            scale_shape_manual(values = c(4, 4, 16, 16),
                               name = shape_legend_title,
                               labels = shape_labels) +
            scale_alpha_manual(values = c(0.1, 1, 0.1, 1),
                               name = alpha_legend_title,
                               labels = alpha_labels) +
            labs(x = x_name,
                 y = y_name, 
                 title = wanted_title, 
                 subtitle = wanted_subtitle)
  return(plot)
}

## arranging plots in same image via cowplot
put_plots_together <- function(plot1,
                               plot2,
                               y_axis_label){
  legend <- cowplot::get_legend(plot1)
  pre_plot <- plot_grid(plot1 +
                          xlab(' ') +
                          ylab(' ') +
                          theme(legend.position = 'none'),
                        plot2 +
                          ggtitle(' ') +
                          ylab(' ') +
                          theme(legend.position = 'none'),
                        ncol = 1)
  plots_together <- plot_grid(pre_plot, legend,
                              rel_widths = c(1, 0.5))
  plots_together_y <- textGrob(y_axis_label,
                               gp = gpar(fontsize = 20),
                               rot = 90)
  plots_together <- grid.arrange(arrangeGrob(plots_together, left = plots_together_y))
  
  return(plots_together)
}

## making a count table for the results
make_count_table <- function(input_table,
                             variable_col){
  ## survival counts
  survival_counts <- input_table %>% 
                       group_by(.data[[variable_col]], status) %>% 
                       tally()
  ## culture counts
  culture_counts <- input_table %>% 
                      group_by(.data[[variable_col]], pos_culture) %>% 
                      na.omit() %>% 
                      tally()
  ## putting the two together in the same table 
  comb_counts <- survival_counts %>% 
                        spread(status, n) %>% 
                        left_join(culture_counts, by = variable_col) %>% 
                        spread(pos_culture, n) %>% 
                        rename('neg_culture' = 'FALSE',
                               'pos_culture' = 'TRUE',
                               'variable' = variable_col)
  return(comb_counts)
}
```

**File Paths**
```{r}
metadata_fp <- '../data/misc/proc_newExp_d15-d3_metadata.tsv'
cfu_count_fp <- '../data/misc/combined_CDD_cfuCounts.txt'
surv_status_fp <- '../data/misc/survival_status.tsv'
cult_genusAbun_fp <- '../data/misc/culture_genusAbun_table.tsv'

location_y_labs <- c('Blood',
                     'Liver',
                     'Spleen')
diet_fill_labs <- c('Chow',
                    'HFt/HFb',
                    'HFt/LFb',
                    'LFt/HFb',
                    'LFt/LFb')
short_diet_fill_labs <- c('Chow',
                          'LFt/LFb')
presAbs_labs <- c('Absent',
                  'Present')
surv_labs <- c("Didn't Survive",
               'Survived')
surv_presAbs_title <- 'Survival and\nMicrobe Status'
surv_presAbs_labs <- c("Didn't Survive, Absent",
                       "Didn't Survive, Present",
                       "Survived, Absent",
                       "Survived, Present")
diet_labs <- c('Chow',
               'High Fat / High Fiber',
               'High Fat / Low Fiber',
               'Low Fat / High Fiber',
               'Low Fat / Low Fiber')
names(diet_labs) <- c('Chow',
                      'HF/HF',
                      'HF/LF',
                      'LF/HF',
                      'LF/LF')

diet_order_labs <- c('High Fat / Low Fiber',
                     'Low Fat / Low Fiber',
                     'High Fat / High Fiber',
                     'Low Fat / High Fiber',
                     'Chow')
names(diet_order_labs) <- c(1,
                            2, 
                            3,
                            4,
                            5)

vendor_labs <- c('Charles River',
                 'Taconic')
names(vendor_labs) <- c('charles_river',
                        'taconic')
```

**Reading in Files**
```{r}
meta <- read_tsv(metadata_fp)
cfu_count_table <- read_tsv(cfu_count_fp)
surv_status <- read_tsv(surv_status_fp)
cult_genusAbun_table <- read_tsv(cult_genusAbun_fp)
```
**Wrangling the positive culture and survival results**
```{r}
## cfu counts table
cfu_count_inter <- cfu_count_table %>% 
  select(mouse_id, presence_spleen, presence_liver, presence_blood) %>% 
  gather(-mouse_id, key = 'location', value = 'microbe_presence') %>% 
  mutate(location = ifelse(location == 'presence_spleen', 'spleen', location),
         location = ifelse(location == 'presence_liver', 'liver', location),
         location = ifelse(location == 'presence_blood', 'blood', location),
         microbe_presence = ifelse(microbe_presence == TRUE, 1, 0))

proc_cfu_count <- cfu_count_table %>% 
  select(!c('presence_spleen', 'presence_liver', 'presence_blood',
            'CFUs.g_tissueSpleen', 'CFUs.g_tissueLiver', 'CFUs.mL_blood')) %>% 
  separate_wider_delim(cols = 'mouse_id',
                       delim = '.',
                       names = c('takeout1',
                                 'vendor',
                                 'takeout2',
                                 'takeout3'),
                       cols_remove = FALSE) %>% 
  select(!c('takeout1', 'takeout2', 'takeout3')) %>% 
  left_join(cfu_count_inter, by = 'mouse_id') %>% 
  mutate(diet = ifelse(diet == 'LFHF', 'LF/HF', diet),
         diet = ifelse(diet == 'LFLF', 'LF/LF', diet),
         diet = ifelse(diet == 'HFLF', 'HF/LF', diet), 
         diet = ifelse(diet == 'HFHF', 'HF/HF', diet),
         vendor = ifelse(vendor == 'CR', 'charles_river', 'taconic')) %>% 
  rename('mouse_sex' = 'sex') %>% 
  left_join(surv_status, by = "mouse_id") %>% 
  mutate(status_microbePres = as.factor(paste(status, microbe_presence, sep = ","))) %>% 
  na.omit()
```

**Wrangling the taxonomy of positive cultures**
info on culture results:
- samples taken from 40 different mice
- 64 samples total from different locations 
```{r}
## pulling out mice/locations that have culture results that include enterococcus, escherichia-shigella,
## and proteus 
cult_genusAbun_table %>% 
  select(location, mouse_id, Genus, rel_abund) %>% 
  filter(Genus == 'Enterococcus' | Genus == 'Escherichia-Shigella' | Genus == 'Proteus',
         rel_abund < 0.9) %>% 
  spread(key = 'Genus', value = 'rel_abund', fill = 0) %>% 
  mutate(total_abun = (Enterococcus + `Escherichia-Shigella` + Proteus)) %>% 
  filter(total_abun >= 0.8) %>% 
  mutate(Genus = paste("enterococcus", "escherichiaShigella", "proteus", sep = "_"),
         sanity = paste(mouse_id, location, sep = "_")) %>% 
  rename(rel_abund = total_abun) %>% 
  select(mouse_id, location, Genus, sanity, rel_abund) -> big3_polymicro_table

## pulling out mice/locations with culture results that are 90% or above one microbe
cult_genusAbun_table %>%
    select(mouse_id, Genus, rel_abund, location) %>%
    group_by(mouse_id, Genus) %>%
    filter(rel_abund > 0.9) %>% 
  mutate(sanity = paste(mouse_id, location, sep = "_")) -> top_tax_table

append(top_tax_table$sanity,
      big3_polymicro_table$sanity) -> sanity_list

## pulling out the last mice/locations that are not included in either of the tables above 
cult_genusAbun_table %>% 
  group_by(mouse_id, location) %>% 
  filter(rel_abund == max(rel_abund)) %>% 
  mutate(sanity = paste(mouse_id, location, sep = "_")) %>% 
  filter(!(sanity %in% sanity_list)) %>% 
  select(mouse_id, Genus, rel_abund, location, sanity) -> missing_mice

## putting all tables together and more data wrangling 
proc_cult_genusAbun <- rbind(top_tax_table,
                             big3_polymicro_table,
                             missing_mice)

proc_cult_genusAbun <- proc_cult_genusAbun %>% 
  mutate(top_tax = ifelse(is.na(Genus), 'Unknown', Genus))
```

**Combining all tables together for the plot**
```{r}
## added staph to the other list bc there's only one sample with it 
other_list <- c('Unknown',
                'Micrococcus',
                'Oceanobacillus',
                'Dietzia',
                'Gemella',
                'Corynebacterium',
                'Bacillus',
                'Streptococcus',
                'Staphylococcus',
                'Cutibacterium',
                'Kocuria')


comb_cfu_genusAbun <- proc_cfu_count %>% 
  left_join(proc_cult_genusAbun, by = c("mouse_id", "location")) %>% 
  filter(experiment != 'CDD01')  %>% 
  mutate(top_tax = ifelse(is.na(top_tax), 'None', top_tax),
         top_tax = ifelse(top_tax %in% other_list, 'Other', top_tax),
         diet_order = case_when(
           diet == 'HF/LF' ~ 1,
           diet == 'LF/LF' ~ 2,
           diet == 'HF/HF' ~ 3,
           diet == 'LF/HF' ~ 4,
           diet == 'Chow' ~ 5
         ))
```

**Putting together color palettes**
color palette?
took the hexcodes from scale_color_brewers 'Paired' palette and put gray in the middle for the "None" category. 
```{r}
##RColorBrewer::brewer.pal(n = 8, name = 'Spectral')
my_spectral_palette <- c("#D53E4F",
                         "#F46D43",
                         "#FDAE61",
                         "#FEE08B",
                         "#E6F598",
                         "#ABDDA4",
                         "#66C2A5",
                         "#3288BD",
                         "#999999")

my_paired_palette <- c("#A6CEE3", 
                       "#1F78B4",
                       "#B2DF8A",
                       "#33A02C",
                       # "#FB9A99",
                       # "#E31A1C",
                       # "#FDBF6F",
                       # "#FF7F00",
                       "#CAB2D6",
                       "#6A3D9A",
                       "#999999")
```

plot
```{r, fig.width=22, fig.height=12}
comb_cfu_genusAbun %>% 
  ggplot(aes(x = mouse_id, y = location)) +
  theme_bw(base_size = 20) +
  theme(axis.text.x = element_blank()) +
  geom_point(aes(color = top_tax, shape = status_microbePres, alpha = status_microbePres), size = 6, 
             stroke = 2) +
  scale_y_discrete(labels = location_y_labs) +
  facet_nested_wrap(~diet_order + vendor,
                    scales = "free_x",
                    labeller = labeller(diet_order = diet_order_labs,
                                        vendor = vendor_labs),
                    nrow = 3) +
  scale_color_manual(breaks = c('Enterococcus',
                                'Escherichia-Shigella',
                                'Proteus',
                                'enterococcus_escherichiaShigella_proteus',
                                'Lactobacillus',
                                'Other',
                                'None'),
                     labels = c('Enterococcus',
                                'Escherichia-Shigella',
                                'Proteus',
                                'Enterococcus, Escherichia-Shigella,\nand Proteus',
                                'Lactobacillus',
                                'Other',
                                'None'),
                     values = my_paired_palette,
                     name = 'Genus') +
  scale_shape_manual(values = c(4, 4, 16, 16),
                     name = surv_presAbs_title,
                     labels = surv_presAbs_labs) +
  scale_alpha_manual(values = c(0.15, 1, 0.15, 1),
                     name = surv_presAbs_title,
                     labels = surv_presAbs_labs) +
  labs(x = 'Mouse',
       y = 'Location',
       title = 'All Experiments Culture Results') -> everything_plot

everything_plot
```

**CDD01 Plots**
```{r, fig.width=10, fig.height=5}
cdd01_cr_plot <- make_plot(input_table = proc_cfu_count,
                           wanted_experiment = 'CDD01',
                           wanted_vendor = 'charles_river',
                           x_axis = 'mouse_id',
                           y_axis = 'location',
                           alpha_value = 'status_microbePres',
                           color_value = 'diet',
                           shape_value = 'status_microbePres',
                           point_size = 6,
                           y_labels = location_y_labs,
                           color_legend_title = 'Diet',
                           color_labels = short_diet_fill_labs,
                           alpha_legend_title = surv_presAbs_title,
                           alpha_labels = surv_presAbs_labs,
                           shape_legend_title = surv_presAbs_title,
                           shape_labels = surv_presAbs_labs,
                           x_name = 'Mouse',
                           y_name = 'Location',
                           wanted_title = 'CDD01 Results',
                           wanted_subtitle = 'Charles River Mice')

cdd01_tc_plot <- make_plot(input_table = proc_cfu_count,
                           wanted_experiment = 'CDD01',
                           wanted_vendor = 'taconic',
                           x_axis = 'mouse_id',
                           y_axis = 'location',
                           alpha_value = 'status_microbePres',
                           color_value = 'diet',
                           shape_value = 'status_microbePres',
                           point_size = 6,
                           y_labels = location_y_labs,
                           color_legend_title = 'Diet',
                           color_labels = short_diet_fill_labs,
                           alpha_legend_title = surv_presAbs_title,
                           alpha_labels = surv_presAbs_labs,
                           shape_legend_title = surv_presAbs_title,
                           shape_labels = surv_presAbs_labs,
                           x_name = 'Mouse',
                           y_name = 'Location',
                           wanted_title = 'CDD01 Results',
                           wanted_subtitle = 'Taconic Mice')

cdd01_cr_plot
cdd01_tc_plot
```

**CDD02 Plots**
```{r, fig.width=10, fig.height=5}
cdd02_cr_plot <- make_plot(input_table = proc_cfu_count,
                           wanted_experiment = 'CDD02',
                           wanted_vendor = 'charles_river',
                           x_axis = 'mouse_id',
                           y_axis = 'location',
                           alpha_value = 'status_microbePres',
                           color_value = 'diet',
                           shape_value = 'status_microbePres',
                           point_size = 6,
                           y_labels = location_y_labs,
                           color_legend_title = 'Diet',
                           color_labels = diet_fill_labs,
                           alpha_legend_title = surv_presAbs_title,
                           alpha_labels = surv_presAbs_labs,
                           shape_legend_title = surv_presAbs_title,
                           shape_labels = surv_presAbs_labs,
                           x_name = 'Mouse',
                           y_name = 'Location',
                           wanted_title = 'CDD02 Results',
                           wanted_subtitle = 'Charles River Mice')

cdd02_tc_plot <- make_plot(input_table = proc_cfu_count,
                           wanted_experiment = 'CDD02',
                           wanted_vendor = 'taconic',
                           x_axis = 'mouse_id',
                           y_axis = 'location',
                           alpha_value = 'status_microbePres',
                           color_value = 'diet',
                           shape_value = 'status_microbePres',
                           point_size = 6,
                           y_labels = location_y_labs,
                           color_legend_title = 'Diet',
                           color_labels = diet_fill_labs,
                           alpha_legend_title = surv_presAbs_title,
                           alpha_labels = surv_presAbs_labs,
                           shape_legend_title = surv_presAbs_title,
                           shape_labels = surv_presAbs_labs,
                           x_name = 'Mouse',
                           y_name = 'Location',
                           wanted_title = 'CDD02 Results',
                           wanted_subtitle = 'Taconic Mice')

cdd02_cr_plot
cdd02_tc_plot
```

**CDD03 Plots**
```{r, fig.width=10, fig.height=5}
cdd03_cr_plot <- make_plot(input_table = proc_cfu_count,
                           wanted_experiment = 'CDD03',
                           wanted_vendor = 'charles_river',
                           x_axis = 'mouse_id',
                           y_axis = 'location',
                           alpha_value = 'status_microbePres',
                           color_value = 'diet',
                           shape_value = 'status_microbePres',
                           point_size = 6,
                           y_labels = location_y_labs,
                           color_legend_title = 'Diet',
                           color_labels = diet_fill_labs,
                           alpha_legend_title = surv_presAbs_title,
                           alpha_labels = surv_presAbs_labs,
                           shape_legend_title = surv_presAbs_title,
                           shape_labels = surv_presAbs_labs,
                           x_name = 'Mouse',
                           y_name = 'Location',
                           wanted_title = 'CDD03 Results',
                           wanted_subtitle = 'Charles River Mice')

cdd03_tc_plot <- make_plot(input_table = proc_cfu_count,
                           wanted_experiment = 'CDD03',
                           wanted_vendor = 'taconic',
                           x_axis = 'mouse_id',
                           y_axis = 'location',
                           alpha_value = 'status_microbePres',
                           color_value = 'diet',
                           shape_value = 'status_microbePres',
                           point_size = 6,
                           y_labels = location_y_labs,
                           color_legend_title = 'Diet',
                           color_labels = diet_fill_labs,
                           alpha_legend_title = surv_presAbs_title,
                           alpha_labels = surv_presAbs_labs,
                           shape_legend_title = surv_presAbs_title,
                           shape_labels = surv_presAbs_labs,
                           x_name = 'Mouse',
                           y_name = 'Location',
                           wanted_title = 'CDD03 Results',
                           wanted_subtitle = 'Taconic Mice')

cdd03_cr_plot
cdd03_tc_plot
```

**Putting all experiments together in one plot**
```{r, fig.width=15, fig.height=5}
## ordering the diet column so all samples of the same diet are next to each other
ordered_proc_cfuCount <- proc_cfu_count[order(proc_cfu_count$diet),]
ordered_proc_cfuCount <- ordered_proc_cfuCount %>% 
  mutate(ord_mouse_id = fct_reorder(mouse_id, diet))

## plots
allExp_cr_plot <- make_plot(input_table = ordered_proc_cfuCount,
                           wanted_experiment = NULL,
                           wanted_vendor = 'charles_river',
                           x_axis = 'ord_mouse_id',
                           y_axis = 'location',
                           alpha_value = 'status_microbePres',
                           color_value = 'diet',
                           shape_value = 'status_microbePres',
                           point_size = 6,
                           y_labels = location_y_labs,
                           color_legend_title = 'Diet',
                           color_labels = diet_fill_labs,
                           alpha_legend_title = surv_presAbs_title,
                           alpha_labels = surv_presAbs_labs,
                           shape_legend_title = surv_presAbs_title,
                           shape_labels = surv_presAbs_labs,
                           x_name = 'Mouse',
                           y_name = 'Location',
                           wanted_title = 'All Experiment Results',
                           wanted_subtitle = 'Charles River Mice')

allExp_tc_plot <- make_plot(input_table = ordered_proc_cfuCount,
                           wanted_experiment = NULL,
                           wanted_vendor = 'taconic',
                           x_axis = 'ord_mouse_id',
                           y_axis = 'location',
                           alpha_value = 'status_microbePres',
                           color_value = 'diet',
                           shape_value = 'status_microbePres',
                           point_size = 6,
                           y_labels = location_y_labs,
                           color_legend_title = 'Diet',
                           color_labels = diet_fill_labs,
                           alpha_legend_title = surv_presAbs_title,
                           alpha_labels = surv_presAbs_labs,
                           shape_legend_title = surv_presAbs_title,
                           shape_labels = surv_presAbs_labs,
                           x_name = 'Mouse',
                           y_name = 'Location',
                           wanted_title = 'All Experiment Results',
                           wanted_subtitle = 'Taconic Mice')

allExp_cr_plot
allExp_tc_plot
```

**CDD01 Plots Together**
```{r, fig.width=10, fig.height=5.5}
cdd01_together <- put_plots_together(plot1 = cdd01_cr_plot,
                                     plot2 = cdd01_tc_plot,
                                     y_axis_label = 'Location')

cdd01_together
```

**CDD02 Plots Together**
```{r, fig.width=11, fig.height=6}
cdd02_together <- put_plots_together(plot1 = cdd02_cr_plot,
                                     plot2 = cdd02_tc_plot,
                                     y_axis_label = 'Location')

cdd02_together
```

**CDD03 Plots Together**
```{r, fig.width=11, fig.height=6}
cdd03_together <- put_plots_together(plot1 = cdd03_cr_plot,
                                     plot2 = cdd03_tc_plot,
                                     y_axis_label = 'Location')

cdd03_together
```

**All experiments plots together**
```{r, fig.width=20, fig.height=8}
allExp_together <- put_plots_together(plot1 = allExp_cr_plot,
                                      plot2 = allExp_tc_plot,
                                      y_axis_label = 'Location')

allExp_together
```

**Numbers of each to go with plots**
Wanted just the number of each condition (survival, pos/neg culture, diet, vendor) depicted on the plots for easy comparison. 
```{r}
wide_cfu_count <- proc_cfu_count %>% 
                    spread(location, microbe_presence) %>% 
                    mutate(pos_culture = ifelse(blood == 1 | liver == 1 | spleen == 1, TRUE, FALSE)) %>% 
                    filter(experiment != 'CDD01')

variable_list <- c('diet', 'vendor', 'experiment')

cfuSurvival_counts <- tibble()

for(i in unique(unlist(variable_list))) {
  tmp_output <- make_count_table(input_table = wide_cfu_count,
                                 variable_col = i)
  cfuSurvival_counts <- bind_rows(cfuSurvival_counts, tmp_output)
}

cfuSurvival_counts
```

**Putting together a table of whether there was a positive culture result (regardless of location) or not**
```{r}
pos_culture_table <- cfu_count_table %>% 
                      select(mouse_id, presence_spleen, presence_liver, presence_blood) %>% 
                      mutate(pos_culture = ifelse(presence_spleen == TRUE | presence_liver == TRUE | presence_blood == TRUE,
                                                  1, 0)) %>% 
                      select(mouse_id, pos_culture) %>% 
                      na.omit()

pos_culture_table
```

**to save for later!**
```{r}
proc_cfuCount_topTax <- proc_cfu_count %>% 
  left_join(proc_cult_genusAbun, by = c("mouse_id", "location")) %>% 
  filter(experiment != 'CDD01')  %>% 
  mutate(top_tax = ifelse(is.na(top_tax), 'None', top_tax)) %>% 
  select(!c('Genus', 'sanity')) %>%
  rename(cult_rel_abund = rel_abund) %>% 
  mutate(cult_rel_abund = ifelse(is.na(cult_rel_abund), 0, cult_rel_abund))

proc_cfuCount_topTax  
```


**Saving my Outputs**
```{r}
ggsave('../plots/cdd01_culture_results.pdf',
       plot = cdd01_together,
       width = 10,
       height = 5.5)
ggsave('../plots/cdd02_culture_results.pdf',
       plot = cdd02_together,
       width = 11,
       height = 6)
ggsave('../plots/cdd03_culture_results.pdf',
       plot = cdd03_together,
       width = 11,
       height = 6)
ggsave('../plots/allExp_culture_results.pdf',
       plot = allExp_together,
       width = 20,
       height = 8)
ggsave('../plots/allExp_culture_withMicrobes.pdf',
       plot = everything_plot,
       width = 22,
       height = 12)

## saving processed cfu count table as a .tsv for later analysis
write_tsv(proc_cfuCount_topTax,
          '../data/misc/proc_combined_CDD_cfuCounts.tsv')
write_tsv(pos_culture_table,
          '../data/misc/pos_culture_status.tsv')
```

