---
title: "survival"
output: html_document
date: "2024-07-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(survminer)
require(survival)
library(readr)
library(magrittr)
library(viridis)
library(apppleplots)
```

**Functions**
FOR FIT_SURVIVAL:
- column names don't need to be passed in as a string because I'm using *substitute* in the function 
- *substitute* is literally the only way I can pass variables in a function to *Surv()* bc it hates me 
- I use *eval.parent()* instead of *return()* in this function because I'm creating an unevaluated expression with *substitute()*. *eval.parent()* evaluates the expression I create in the function (so in this case, the survival curve fit) and returns it to me. 
```{r}
fit_survival <- function(survival_table,
                         survival_day_col,
                         survival_results_col,
                         test_between_col1,
                         test_between_col2){
  
  params <- list(survival_table = substitute(survival_table),
                 survival_day_col = substitute(survival_day_col),
                 survival_results_col = substitute(survival_results_col),
                 test_between_col1 = substitute(test_between_col1),
                 test_between_col2 = substitute(test_between_col2))
  
  wanted_fit <- substitute(surv_fit(formula = Surv(survival_day_col, survival_results_col) ~ test_between_col1 +
                                      test_between_col2,
                          data = survival_table), params)
  eval.parent(wanted_fit)
}

## to take p-values calculated and put them into a nice table
pretty_pvalue <- function(pairwise_results){
  
  pvalue_table <- as_tibble(pairwise_results$p.value,
                            rownames = 'group1')
  pvalue_table %>% 
    gather(-group1, key = group2, value = p.value) %>% 
    na.omit() -> pvalue_table
  
  pvalue_table['signif'] <- symnum(pvalue_table$p.value,
                                   cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1),
                                   symbols = c("****", "***", "**", "*", "ns"),
                                   abbr.colnames = FALSE,
                                   na = "")
  return(pvalue_table)
}

## making the survival heat map plot
surv_vis <- function(pvalue_table,
                     x_axis,
                     y_axis,
                     facet_by = NULL,
                     x_labs = NULL,
                     y_labs = NULL){
  surv_stat_plot <- pvalue_table %>%
        na.omit() %>% 
        ggplot(aes(x = .data[[x_axis]], y = .data[[y_axis]])) +
        geom_tile(aes(fill = p.value), alpha = 0.6, color = 'black') +
        geom_text(aes(label = signif), size = 5) +
        scale_fill_distiller(limits = c(0, 0.05), palette = 'Reds', name = "P-Value") +
        theme_bw(base_size = 20) +
        theme(axis.title.x = element_blank(),
              axis.title.y = element_blank(),
              legend.text = element_text(size = 16),
              legend.title = element_text(size = 18),
              legend.position = 'none')
        # facet_wrap(~.data[[facet_by]])
        # scale_x_discrete(labels = x_labs) +
        # scale_y_discrete(labels = y_labs)
  return(surv_stat_plot)
}
```

**File Paths**
```{r}
surv_fp <- '../data/misc/survival_data.tsv'

diet_labels <- c('Chow',
                 'High Fat / High Fiber', 
                 'High Fat / Low Fiber',
                 'Low Fat / High Fiber',
                 'Low Fat / Low Fiber')
names(diet_labels) <- c('Chow',
                        'HF/HF',
                        'HF/LF',
                        'LF/HF',
                        'LF/LF')

vendor_labels <- c('Charles River',
                   'Taconic')
names(vendor_labels) <- c('charles_river',
                          'taconic')
```

**Reading in File**
```{r}
surv_table <- read_tsv(surv_fp,
                       col_types = cols(vendor = col_character(),
                                        mouse_id = col_character(),
                                        experiment = col_character(),
                                        mouse_sex = col_character(),
                                        diet = col_character(),
                                        sac_exptDay = col_integer(),
                                        day_post_inf = col_integer(),
                                        survival = col_integer()))

## filtering out the first set of experiments since it could skew the results
surv_table <- surv_table %>% 
                filter(experiment != 'CDD01')
```

**Investigating Differences in Date of Mouse Death per Diet**
Putting together a summary table for how many days post CDI the mice started/stopped dying per diet and infection type.
```{r}
## day the mice started dying 
start_death_df <- surv_table %>% 
  filter(actual_surv == 1) %>% 
  group_by(diet, vendor) %>%
  summarise(start_death = min(day_post_inf))

## how many mice of each died died during the death range 
mice_lost_during <- surv_table %>% 
  filter(actual_surv == 1) %>% 
  count(diet, vendor) %>% 
  rename(num_mice_lost = n)

## day the mice stopped dying and combining it with the counts table
## and the day the mice started dying to create one summary table 
death_range_df <- surv_table %>% 
  filter(actual_surv == 1) %>% 
  group_by(diet, vendor) %>% 
  summarise(stop_death = max(day_post_inf)) %>% 
  left_join(start_death_df, by = c('diet', 'vendor')) %>% 
  mutate(death_range = paste(start_death, stop_death, sep = "_")) %>% 
  left_join(mice_lost_during, by = c('diet', 'vendor'))

death_range_df
```

Running a non-parametric ANOVA (Kruskal-Wallis followed by Dunn's Post Hoc) on the days the mice from different diets died on binned by vendor (charles river vs taconic). 
```{r}
## pulling out the results from just the mice that died 
just_mice_died <- surv_table %>% 
  filter(actual_surv == 1)

death_date_stats <- apppleplots::kruskal_dunn_stats(input_table = just_mice_died,
                                grouped_by = 'vendor',
                                adjust_method = 'BH',
                                formula_left = 'day_post_inf',
                                formula_right = 'diet')

death_date_kruskal <- death_date_stats$KruskalTest
death_date_dunn <- death_date_stats$DunnTest

death_date_dunn
```

**IN THESE PLOTS:**
- Day 0 on the x-axis is the day that the mice are infected with C. diff, so the day that the mice start dying is how soon after CDI that they succumb

**Diet and Vendor Survival Curve**
```{r, fig.width=17, fig.height=7}
## fitting data
dietVendor_surv_fit <- fit_survival(survival_table = surv_table,
                                    survival_day_col = day_post_inf,
                                    survival_results_col = actual_surv,
                                    test_between_col1 = diet,
                                    test_between_col2 = vendor)

## plot
dietVendor_base_plot <- ggsurvplot(dietVendor_surv_fit,
                                   data = surv_table,
                                   color = "black",
                                   ggtheme = theme_bw(base_size = 20),
                                   legend = "none",
                                   conf.int = TRUE,
                                   xlab = 'Days Post Infection',
                                   ylab = 'Survival Probability',
                                   title = 'Mouse Survival After Infection by Diet and Vendor',
                                   subtitle = 'New Experiments')

dietVendor_base_plot$plot +
    facet_grid(cols = vars(diet),
               rows = vars(vendor),
               labeller = labeller(.cols = diet_labels,
                                   .rows = vendor_labels)) +
  theme(strip.text.y.right = element_text(angle = 0)) -> dietVendor_plot

dietVendor_plot
```

**Diet and Experiment Survival Curve**
```{r, fig.width=17, fig.height=7}
## fitting data
dietExp_surv_fit <- fit_survival(survival_table = surv_table,
                                 survival_day_col = day_post_inf,
                                 survival_results_col = actual_surv,
                                 test_between_col1 = diet,
                                 test_between_col2 = experiment)

## plot
dietExp_base_plot <- ggsurvplot(dietExp_surv_fit,
                                data = surv_table,
                                color = "black",
                                ggtheme = theme_bw(base_size = 20),
                                legend = "none",
                                conf.int = TRUE,
                                xlab = 'Days Post Infection',
                                ylab = 'Survival Probability',
                                title = 'Survival by Diet and Experiment')

dietExp_base_plot$plot +
    facet_grid(cols = vars(diet),
               rows = vars(experiment),
               labeller = labeller(.cols = diet_labels)) +
  theme(strip.text.y.right = element_text(angle = 0)) -> dietExp_plot

dietExp_plot
```

**Experiment and Vendor Survial Curve**
```{r, fig.width=10, fig.height=6}
## fitting data
expVendor_surv_fit <- fit_survival(survival_table = surv_table,
                                   survival_day_col = day_post_inf,
                                   survival_results_col = actual_surv,
                                   test_between_col1 = experiment,
                                   test_between_col2 = vendor)

## plot
expVendor_base_plot <- ggsurvplot(expVendor_surv_fit,
                                  data = surv_table,
                                  color = "black",
                                  ggtheme = theme_bw(base_size = 20),
                                  legend = "none",
                                  conf.int = TRUE,
                                  xlab = 'Days Post Infection',
                                  ylab = 'Survival Probability',
                                  title = 'Survival by Experiment and Vendor')

expVendor_base_plot$plot +
    facet_grid(cols = vars(experiment),
               rows = vars(vendor),
               labeller = labeller(.rows = vendor_labels)) +
    theme(strip.text.y.right = element_text(angle = 0)) -> expVendor_plot

expVendor_plot
```

**Stats**
```{r, fig.width=10, fig.height=10}
## diet and vendor stats
dietVendor_pairwise <- pairwise_survdiff(formula = Surv(day_post_inf, actual_surv) ~ vendor + diet,
                                         data = surv_table,
                                         p.adjust.method = "BH")

dietVendor_pvalue <- pretty_pvalue(pairwise_results = dietVendor_pairwise)

dietVendor_stat_vis <- surv_vis(pvalue_table = dietVendor_pvalue,
                                x_axis = 'group2',
                                y_axis = 'group1')

dietVendor_stat_vis
```

running difference between survival by vendor individually for each diet and then smushing them together 
```{r}
test_out <- tibble()
for(i in unique(unlist(surv_table['diet']))){
  survdiff_res <- surv_table %>% 
    filter(diet == i) %>% 
     pairwise_survdiff(formula = Surv(day_post_inf, actual_surv) ~ vendor,
                       data = .,
                       p.adjust.method = "BH")
  pretty_res <- pretty_pvalue(pairwise_results = survdiff_res) %>% 
    mutate(diet = paste(i))
  
  test_out <- rbind(test_out, pretty_res)
}
```

the stat vis for the stats above (it looks like a lot of work for not a lot of stats..its almost unnecessary)
```{r, fig.width=5.5, fig.height=2.5}
surv_vis(pvalue_table = test_out,
         x_axis = 'diet',
         y_axis = 'group2',
         facet_by = 'group1')
```

idk why i didn't just run normal stats on this data bc I don't need the surminer package to do them 
```{r}
dietVendor_stat_list <- kruskal_dunn_stats(input_table = surv_table,
                                           grouped_by = c("diet"),
                                           adjust_method = "BH",
                                           filter_adj_p_value = FALSE,
                                           formula_left = "actual_surv",
                                           formula_right = "vendor")

dietVendor_kruskal <- dietVendor_stat_list$KruskalTest
dietVendor_dunn <- dietVendor_stat_list$DunnTest

dietVendor_dunn
```

```{r}
## just overall diet stats
diet_pairwise <- pairwise_survdiff(formula = Surv(day_post_inf, actual_surv) ~ diet,
                                   data = surv_table,
                                   p.adjust.method = "BH")

diet_pvalue <- pretty_pvalue(pairwise_results = diet_pairwise)
```

```{r}
## diet and experiment stats
dietExp_pairwise <- pairwise_survdiff(formula = Surv(day_post_inf, actual_surv) ~ diet + experiment,
                                      data = surv_table,
                                      p.adjust.method = "BH")

dietExp_pvalue <- pretty_pvalue(pairwise_results = dietExp_pairwise)
```

```{r}
## experiment and vendor stats
expVendor_pairwise <- pairwise_survdiff(formula = Surv(day_post_inf, actual_surv) ~ experiment + vendor,
                                         data = surv_table,
                                         p.adjust.method = "BH")

expVendor_pvalue <- pretty_pvalue(pairwise_results = expVendor_pairwise)
```

**Cox PH Hazard Ratios**
only doing this for the plot that is in the manuscript which is the `dietVendor_plot`

Notes:
- this calculates the hazard ratios for each diet compared to chow
- the hazard ratio for vendor is compared to charles river which is why only taconic shows up in the results 
- if you call summary(coxph_object), the exp(coef) column contains the actual hazard ratio for the variable and the Pr(>|z|) column contains the p-value 
```{r}
dietVendor_surv_obj <- Surv(time = surv_table$day_post_inf, event = surv_table$actual_surv)
surv_table$diet_f = factor(surv_table$diet,
                           levels = c("Chow",
                                      "HF/HF",
                                      "HF/LF",
                                      "LF/HF",
                                      "LF/LF"))
diet.vendor.fit.cox <- coxph(dietVendor_surv_obj ~ diet_f + vendor, data = surv_table)

dietVendor_summary_coxph <- summary(diet.vendor.fit.cox)

dietVendor_summary_coxph
```

```{r}
## pulling the hazard ratio from coxph summary 
dietVendor_haz_df <- as.data.frame(dietVendor_summary_coxph$coefficients)
dietVendor_haz_df["diet_vendor"] <- rownames(dietVendor_haz_df)

dietVendor_haz_table <- as_tibble(dietVendor_haz_df) %>% 
  mutate(diet_vendor = case_when(
    diet_vendor == 'diet_fHF/HF' ~ 'HF/HF',
    diet_vendor == 'diet_fHF/LF' ~ 'HF/LF',
    diet_vendor == 'diet_fLF/HF' ~ 'LF/HF',
    diet_vendor == 'diet_fLF/LF' ~ 'LF/LF',
    diet_vendor == 'vendortaconic' ~ 'taconic'
  )) %>% 
  rename(haz_ratio = `exp(coef)`,
         p_value = `Pr(>|z|)`)

dietVendor_haz_table
```
```{r}
## pulling confidence intervals from the coxph summary
dietVendor_ci_df <- as.data.frame(dietVendor_summary_coxph$conf.int)
dietVendor_ci_df["diet_vendor"] <- rownames(dietVendor_ci_df)

dietVendor_ci_table <- as_tibble(dietVendor_ci_df) %>% 
  mutate(diet_vendor = case_when(
    diet_vendor == 'diet_fHF/HF' ~ 'HF/HF',
    diet_vendor == 'diet_fHF/LF' ~ 'HF/LF',
    diet_vendor == 'diet_fLF/HF' ~ 'LF/HF',
    diet_vendor == 'diet_fLF/LF' ~ 'LF/LF',
    diet_vendor == 'vendortaconic' ~ 'taconic'
  )) %>% 
  rename(lower_95_ci = `lower .95`,
         upper_95_ci = `upper .95`) %>% 
  select(diet_vendor, lower_95_ci, upper_95_ci)

## combining the hazard ratio and confidence interval tables 
final_dietVendor_haz_table <- dietVendor_haz_table %>% 
  left_join(dietVendor_ci_table, by = 'diet_vendor') %>% 
  select(diet_vendor, coef, haz_ratio, p_value, lower_95_ci, upper_95_ci)

final_dietVendor_haz_table
```

**Saving my Outputs**
```{r}
## plots
ggsave('../plots/dietVendor_survival.pdf',
       plot = dietVendor_plot,
       width = 17,
       height = 7)
ggsave('../plots/dietExp_survival.pdf',
       plot = dietExp_plot,
       width = 17,
       height = 7)
ggsave('../plots/experimentVendor_survival.pdf',
       plot = expVendor_plot,
       width = 10,
       height = 6)

## stats
write_tsv(dietVendor_pvalue,
          '../stats/surv_dietVendor.tsv')
write_tsv(dietVendor_dunn,
          '../stats/surv_dietVendor_dunn.tsv')
write_tsv(diet_pvalue,
          '../stats/surv_diet.tsv')
write_tsv(dietExp_pvalue,
          '../stats/surv_dietExp.tsv')
write_tsv(expVendor_pvalue,
          '../stats/surv_expVendor.tsv')
write_tsv(final_dietVendor_haz_table,
          '../stats/surv_hazardRatio_results.tsv')
```

