---
title: "tax_posCult_surv"
output: html_document
date: "2024-11-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggpubr)
library(ggplot2)
library(qiime2R)
library(tidyverse)
library(ggh4x)
library(viridis)
library(ggrepel)
library(cowplot)
```

MAIN QUESTION:
- out of the mice who died and had positive cultures, did we culture the same microbe that was blooming in their gut microbiome at day 3? 

..i'm still trying to figure out how to show this 

**Functions**
```{r}
abun_file_prep <- function(metadata_file,
                           tax_fp,
                           otu_table_fp){
  ## metadata
  metadata <- metadata_file %>% 
            rename(sampleid = `#SampleID`)
  ## taxonomy
  taxonomy <- read_qza(tax_fp)$data %>% 
    parse_taxonomy() %>% 
    rownames_to_column('asv')
  ## otu table 
  otu_table <- read_qza(otu_table_fp)$data
  otu_table %>% 
    as_tibble(rownames = 'asv') %>% 
    gather(-asv, key = sampleid, value = abun) %>% 
    group_by(sampleid) %>% 
    mutate(rel_abun = abun/sum(abun)) %>% 
    mutate(rel_abun = rel_abun + 0.000001) %>% 
    filter(sampleid %in% metadata$sampleid) -> otu_table
  ## joining all tables together 
  otu_table %>% 
    left_join(metadata, by = 'sampleid') %>% 
    left_join(taxonomy, by = 'asv') -> abun_table
  abun_table %>% 
    group_by(sampleid, day_post_inf, diet, mouse_id, vendor,
             experiment, Genus) %>% 
    summarise(rel_abund = sum(rel_abun)) %>% 
    ungroup() -> abun_filt
  ## creating a list for my outputs
  my_list <- list(Taxonomy = taxonomy,
                  OTUTable = otu_table,
                  AbundanceTable = abun_filt)
  return(my_list)
}
```

**File paths**
```{r}
pos_cult_surv_fp <- '../data/misc/proc_combined_CDD_cfuCounts.tsv'
tax_FP <- '../data/qiime/taxonomy.qza'
otu_table_FP <- '../data/qiime/total_sum_otu_table.qza'
metadata_fp <- '../data/misc/proc_newExp_d15-d3_metadata.tsv'

## main microbes cultured out of blood, spleen, and liver 
top_microbes <- c('Enterococcus',
                  'Escherichia-Shigella',
                  'Proteus',
                  'Lactobacillus',
                  'Micrococcus',
                  'Oceanobacillus',
                  'Dietzia',
                  'Gemella',
                  'Corynebacterium',
                  'Bacillus',
                  'Streptococcus',
                  'Staphylococcus',
                  'Cutibacterium',
                  'Kocuria')

other_list <- c('Unknown',
                'Micrococcus',
                'Oceanobacillus',
                'Dietzia',
                'Gemella',
                'Corynebacterium',
                'Bacillus',
                'Streptococcus',
                'Staphylococcus',
                'Cutibacterium',
                'Kocuria')
```

**Reading in files**
```{r}
posCult_surv_table <- read_tsv(pos_cult_surv_fp)
metadata <- read_tsv(metadata_fp)
```

**Data wrangled**
Info:
- 25 mice between experiments CDD02/CDD03 had positive cultures and died 
- filtered out CDD01 as not to skew results 
```{r}
## pulling mouse ids for mice that had a positive culture result and didnt survive
pre_target_mouse_ids <- posCult_surv_table %>% 
  spread(key = "location", value = "microbe_presence", fill = 0) %>% 
  filter(experiment != "CDD01",
         status_microbePres == "didnt_survive,1") %>% 
  select(mouse_id)

target_mouse_ids <- pre_target_mouse_ids$mouse_id

## filtering the metadata mouse ids by the target mouse ids 
wanted_mouse_meta <- metadata %>% 
  filter(mouse_id %in% target_mouse_ids)
```

pulling top microbe cultured out of the mice that died to compare to microbiome results
```{r}
posCult_surv_table %>% 
  filter(experiment != "CDD01",
         status_microbePres == "didnt_survive,1") %>% 
  mutate(location_microbe = paste(location, top_tax, sep = "_")) %>% 
  select(mouse_id, location, location_microbe) %>% 
  spread(key = "location", value = "location_microbe", fill = "None") %>% 
  mutate(blood = ifelse(blood == "None", "blood_None", blood),
         liver = ifelse(liver == "None", "liver_None", liver),
         spleen = ifelse(spleen == "None", "spleen_None", spleen),
         cultured_microbe = paste(blood, liver, spleen, sep = ";")) %>% 
  select(mouse_id, cultured_microbe) -> comb_microbes

comb_microbes
```

```{r}
wide_cult_table <- posCult_surv_table %>% 
  filter(experiment != "CDD01",
         status_microbePres == "didnt_survive,1") %>% 
  select(mouse_id, location, top_tax) %>% 
  spread(key = "location", value = "top_tax", fill = "None") %>% 
  rename(cult_blood = blood,
         cult_spleen = spleen,
         cult_liver = liver)

wide_cult_table
```


microbiome taxonomic information for the mice who had positive cultures and died 
```{r}
abun_files <- abun_file_prep(metadata_file = wanted_mouse_meta,
                             tax_fp = tax_FP,
                             otu_table_fp = otu_table_FP)

wanted_mouse_abun <- abun_files$AbundanceTable

wanted_mouse_abun <- wanted_mouse_abun %>%
  filter(Genus %in% top_microbes)

test_mouse_abun <- wanted_mouse_abun %>% 
  select(-sampleid) %>% 
  spread(key = "day_post_inf", value = "rel_abund") %>% 
  rename(day_neg15 = `-15`,
         day_3 = `3`) %>% 
  mutate(diff_relAbun_days = (day_neg15 - day_3)) %>% 
  select(!c('day_neg15', 'day_3')) %>% 
  filter(diff_relAbun_days != 0,
         diff_relAbun_days >= 0.1 | diff_relAbun_days <= -0.1,
         Genus != 'Lactobacillus') %>%
  left_join(wide_cult_table, by = "mouse_id")

test_mouse_abun %>% 
  group_by(mouse_id) %>% 
  mutate(cult_pres_microbiome = ifelse(Genus == cult_blood | Genus == cult_liver | Genus == cult_spleen, 1, 0),
         cult_pres_microbiome = case_when(
                    Genus == 'Enterococcus' & cult_blood == 'enterococcus_escherichiaShigella_proteus' | 
                    cult_liver == 'enterococcus_escherichiaShigella_proteus' |
                    cult_spleen == 'enterococcus_escherichiaShigella_proteus' ~ 1,
                  .default = cult_pres_microbiome
          )) -> test_mouse_abun

test_mouse_abun %>% 
  filter(cult_pres_microbiome != 0)

```

**Attempt at a plot**
this is not it 
```{r}
test_mouse_abun %>% 
  na.omit() %>% 
  ggplot(aes(x = mouse_id, y = Genus)) +
  geom_tile(aes(fill = diff_relAbun_days), color = 'black') 
```

