---
title: "bile_acid_comparisons"
output: html_document
date: "2024-01-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(broom)
library(cowplot)
library(magrittr)
library(qiime2R)
library(tidyverse)
library(naniar)
library(ggpubr)
library(rstatix)
library(viridis)
library(apppleplots)
```
**Functions**
```{r}
## runs linear modeling on the table, diet, location, and toxin/metabolite wanted and returns full results 
## along with a table containing the r2 and p-value 
## this goes into the following function with a for loop 
r2_value <- function(biom_table,
                     wanted_diet,
                     wanted_comp_col,
                     wanted_comp_value,
                     stat_value,
                     tox_metab_col,
                     wanted_tox_metab,
                     wanted_conc_value){
  ## to get r squared value and p value
  biom_table %>% 
    filter(diet == wanted_diet,
           .data[[wanted_comp_col]] == wanted_comp_value,
           .data[[tox_metab_col]] == wanted_tox_metab) %>%
    do(glance(lm(.data[[stat_value]] ~ .data[[wanted_conc_value]],
                 data = .))) -> lm_glance
  output <- list(r_squared = lm_glance$r.squared,
                 p_value = lm_glance$p.value)
  ## to get estimate and place it with the overall output table
  biom_table %>% 
    filter(diet == wanted_diet,
           .data[[wanted_comp_col]] == wanted_comp_value,
           .data[[tox_metab_col]] == wanted_tox_metab) %>%
    do(tidy(lm(.data[[stat_value]] ~ .data[[wanted_conc_value]],
                 data = .))) %>% 
    filter(term != '(Intercept)') -> lm_full
  output <- as_tibble(output) %>%
            mutate(diet = paste(wanted_diet),
                   comp_col = paste(wanted_comp_value),
                   tox_metab = paste(wanted_tox_metab),
                   just_r = sqrt(r_squared),
                   estimate = lm_full$estimate,
                   r_direct = if_else(estimate < 0, just_r * -1, just_r))
  my_list <- list(LinearModel = lm_full,
                  WantedStats = output)
  return(my_list)
}

## creating a nested for loop since running this function individually is a nightmare
r2_value_for_loop <- function(biom_table,
                              comp_col,
                              wanted_stat_value,
                              wanted_col,
                              wanted_conc) {
  ## nested for loop for the r squared and p values 
  output <- tibble()
  for(i in unique(unlist(biom_table$diet))) {
    for(j in unique(unlist(biom_table[comp_col]))) {
      for(k in unique(unlist(biom_table[wanted_col]))) {
        tmp_output <- r2_value(biom_table,
                               i,
                               comp_col,
                               j,
                               wanted_stat_value,
                               wanted_col,
                               k,
                               wanted_conc)$WantedStats
        output <- bind_rows(output, tmp_output)
      }
    }
  }
  return(output)
}
```

**Bile Acid Ratio Stats Functions**
I'm too lazy to make my r2 value function and for loop for for everything so I have two separate sets of functions lol
```{r}
ratio_r2_value <- function(biom_table,
                     wanted_diet,
                     category_col,
                     col_value,
                     wanted_category,
                     wanted_conc_value){
  ## to get r squared value and p value
  biom_table %>% 
    filter(diet == wanted_diet,
           .data[[category_col]] == wanted_category) %>%
    do(glance(lm(.data[[col_value]] ~ .data[[wanted_conc_value]],
                 data = .))) -> lm_glance
  output <- list(r_squared = lm_glance$r.squared,
                 p_value = lm_glance$p.value)
  ## to get estimate and place it with the overall output table
  biom_table %>% 
    filter(diet == wanted_diet,
           .data[[category_col]] == wanted_category) %>%
    do(tidy(lm(.data[[col_value]] ~ .data[[wanted_conc_value]],
                 data = .))) %>% 
    filter(term != '(Intercept)') -> lm_full
  output <- as_tibble(output) %>%
            mutate(diet = paste(wanted_diet),
                   category = paste(wanted_category),
                   just_r = sqrt(r_squared),
                   estimate = lm_full$estimate,
                   r_direct = if_else(estimate < 0, just_r * -1, just_r))
  my_list <- list(LinearModel = lm_full,
                  WantedStats = output)
  return(my_list)
}

## creating a nested for loop since running this function individually is a nightmare
ratio_r2_value_for_loop <- function(biom_table,
                              wanted_col,
                              col_value,
                              wanted_conc) {
  ## nested for loop for the r squared and p values 
  output <- tibble()
  for(i in unique(unlist(biom_table$diet))) {
    for(j in unique(unlist(biom_table[wanted_col]))) {
        tmp_output <- ratio_r2_value(biom_table,
                               i,
                               wanted_col,
                               col_value,
                               j,
                               wanted_conc)$WantedStats
        output <- bind_rows(output, tmp_output)
      }
    }
  return(output)
}

## putting stats directly on correlation plots 
stat_on_plot <- function(stat_table,
                         wanted_col_list,
                         x_col_name,
                         x_col_value,
                         y_col_name,
                         y_col_value,
                         plot,
                         plot_text_size){
  ## making table conducive to labeling plot 
  proc_stat_table <- stat_table %>% 
    filter(signif != 'ns') %>% 
    select(all_of(wanted_col_list)) %>% 
    ## created columns need to match up with the column names called for the x and y axis in the plot
    ## numeric values added are the x and y coordinates for where the label will sit in the facet block 
    mutate("{x_col_name}" := paste(x_col_value),
           "{x_col_name}" := as.numeric(.data[[x_col_name]]),
           "{y_col_name}" := paste(y_col_value),
           "{y_col_name}" := as.numeric(.data[[y_col_name]]),
           p_value = signif(p_value, digits = 3),
           r_squared = signif(r_squared, digits = 3),
           pVal_rSquare = paste0('p-value: ', p_value, "\n", 
                                 'r2: ', r_squared))
  
  ## putting labels on plot 
  plot_wStats <- plot +
    geom_label(data = proc_stat_table, 
               label = proc_stat_table$pVal_rSquare,
               color = 'black',
               fontface = 'bold',
               size = plot_text_size)
  
  ## creating list of my outputs
  my_list <- list(ProcStatTable = proc_stat_table,
                  PlotwStats = plot_wStats)
  
  return(my_list)
}
```


**File Paths**
```{r}
metadata_FP <- '../data/misc/filt_cecal_processed_metadata.tsv'
histo_FP <- '../data/misc/processed_histopathology.tsv'
metab_FP <- '../data/misc/processed_metabolomics.tsv'
neat_tox_FP <- '../data/misc/processed_neatToxin.tsv'
dil_tox_FP <- '../data/misc/processed_dilutedToxin.tsv'
bile_acid_FP <- '../data/misc/processed_bile_acid.tsv'
bile_acidRatio_FP <- '../data/misc/processed_ratio_bileAcid.tsv'

## lists to redo the diet names on the facet labels of the ggplot created below 
diet_labs <- 
    c('Chow', 
      'High Fat / High Fiber', 
      'High Fat / Low Fiber', 
      'Low Fat / High Fiber', 
      'Low Fat / Low Fiber')

names(diet_labs) <- c('Chow', 'HF/HF', 'HF/LF', 'LF/HF', 'LF/LF')

stat_diet_y_labs <- c("Chow",
                       "HFt/HFb",
                       "HFt/LFb",
                       "LFt/HFb",
                       "LFt/LFb")

tissue_labs <- c('Cecum',
                 'Colon')
names(tissue_labs) <- c('cecum',
                        'colon')

metab_labs <- c('Acetic Acid',
                'Propanoic Acid',
                'n-Butanoic Acid')
names(metab_labs) <- c('Acetic Acid (ug/g)',
                       'Propanoic Acid (ug/g)',
                       'n-Butanoic Acid (ug/g)')
metab_x_labs <- c('Acetic\nAcid',
                  'n-Butanoic\nAcid',
                  'Propanoic\nAcid')

neat_labs <- c('TcdA', 'TcdB')
names(neat_labs) <- c('Total TcA Neat', 'Total TcB Neat')

dil_labs <- c('TcdA', 'TcdB')
names(dil_labs) <- c('Total TcA 1:10', 'Total TcB 1:10')

bile_facet_labs <- c('Promoter',
                     'Inhibitor')
names(bile_facet_labs) <- c('promoter',
                            'inhibitor')
```

**File Prep**
```{r}
metadata <- read_tsv(metadata_FP)
histo <- read_tsv(histo_FP)

metab <- read_tsv(metab_FP)
names(metab)[names(metab) == 'concentration'] <- 'metab_concentration'
metab %>% 
  mutate(metab_conc_norm = (metab_concentration + 2)) -> metab

neat_tox <- read_tsv(neat_tox_FP)
neat_tox %>% 
  mutate(neat_conc_norm = (neat_conc + 2)) -> neat_tox

dil_tox <- read_tsv(dil_tox_FP)
bile_acid <- read_tsv(bile_acid_FP)

ratio_bile_acid <- read_tsv(bile_acidRatio_FP)
ratio_bile_acid %>% 
  select(!c(inhibitor, promoter)) -> ratio_bile_acid
```
# **Histopathology Comparisons**

**Plots**
```{r, warning=FALSE, fig.width=10, fig.height=4}
## cecum
histo %>% 
  select(tissue, score, mouse_id) %>% 
  left_join(bile_acid, by = 'mouse_id') %>% 
  filter(tissue == 'cecum') %>% 
  na.omit() -> histo_bileAcid_cecum

histo_bileCecum_plot <- apppleplots::correlation_plots(input_table = histo_bileAcid_cecum,
                                                        x_value = 'conc_normalized',
                                                        y_value = 'score',
                                                        x_transform = TRUE,
                                                        y_transform = FALSE,
                                                        point_alpha = 0.4,
                                                        regression_method = 'lm',
                                                        facet_rows = 'c_diff_effect',
                                                        facet_cols = 'diet',
                                                        row_labs = bile_facet_labs,
                                                        col_labs = diet_labs,
                                                        x_name = 'Concentration (log10)',
                                                        y_name = 'Score',
                                                        title_content = 'Cecum Bile Acid by Histopathology')
  
  
  
## colon
histo %>% 
  select(tissue, score, mouse_id) %>% 
  left_join(bile_acid, by = 'mouse_id') %>% 
  filter(tissue == 'colon') %>% 
  na.omit() -> histo_bileAcid_colon

histo_bileColon_plot <- apppleplots::correlation_plots(input_table = histo_bileAcid_colon,
                                                        x_value = 'conc_normalized',
                                                        y_value = 'score',
                                                        x_transform = TRUE,
                                                        y_transform = FALSE,
                                                        point_alpha = 0.4,
                                                        regression_method = 'lm',
                                                        facet_rows = 'c_diff_effect',
                                                        facet_cols = 'diet',
                                                        row_labs = bile_facet_labs,
                                                        col_labs = diet_labs,
                                                        x_name = 'Concentration (log10)',
                                                        y_name = 'Score',
                                                        title_content = 'Colon Bile Acid by Histopathology')

histo_bileCecum_plot
histo_bileColon_plot
```

**Stats**
```{r}
histo %>% 
  select(mouse_id, tissue, score) %>% 
  left_join(bile_acid, by = 'mouse_id') %>% 
  na.omit() -> histo_bileAcid

histo_bile_stats <- r2_value_for_loop(histo_bileAcid,
                                      'tissue',
                                      'score',
                                      'c_diff_effect',
                                      'conc_normalized')

## assigning p value significance
histo_bile_stats['signif'] <- symnum(histo_bile_stats$p_value,
                                     cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1),
                                     symbols = c("****", "***", "**", "*", "ns"),
                                     abbr.colnames = FALSE,
                                     na = "")
```
**Stats on Plot**
```{r, warning=FALSE, fig.width=10, fig.height=4}
names(histo_bile_stats)[names(histo_bile_stats) == 'comp_col'] <- 'tissue'
names(histo_bile_stats)[names(histo_bile_stats) == 'tox_metab'] <- 'c_diff_effect'
col_list <- c('diet', 'c_diff_effect', 'signif', 'p_value', 'r_squared')

## cecum
histo_bile_stats %>% 
  filter(tissue == 'cecum') %>% 
  select(-tissue) -> histo_bileCecum_stats

histo_bileCecum_list <- stat_on_plot(stat_table = histo_bileCecum_stats,
                                    wanted_col_list = col_list,
                                    x_col_name = 'conc_normalized',
                                    x_col_value = '1500',
                                    y_col_name = 'score',
                                    y_col_value = '5.5',
                                    plot = histo_bileCecum_plot,
                                    plot_text_size = 3)

histo_bileCecum_stats_plot <- histo_bileCecum_list$PlotwStats

histo_bileCecum_stats_plot
```


**Statistical Visualization**
```{r, warning=FALSE, fig.height=4, fig.width=6.5}
histo_bile_stat_plot <- apppleplots::stat_heat_plot(input_table = histo_bile_stats,
                                                    x_value = 'c_diff_effect',
                                                    y_value = 'diet',
                                                    x_value_type = 'discrete',
                                                    y_value_type = 'discrete',
                                                    pairwise = FALSE,
                                                    correlation = TRUE,
                                                    tile_fill = 'r_direct',
                                                    tile_outline_color = 'black',
                                                    tile_alpha = 0.6,
                                                    text_value = 'signif',
                                                    low_color = 'blue',
                                                    high_color = 'green',
                                                    legend_name = 'Correlation\nCoefficient',
                                                    x_labels = bile_facet_labs,
                                                    y_labels = stat_diet_y_labs,
                                                    facet_by = 'tissue',
                                                    facet_labels = tissue_labs,
                                                    x_name = 'Bile Acid',
                                                    y_name = 'Diet',
                                                    title_content = 'Histopathology Score and Bile Acid\nConcentration Comparisons')

histo_bile_stat_plot
```


**Plots Together**
```{r, warning=FALSE, fig.height=8, fig.width=16.5}
plot_grid(histo_bileCecum_stats_plot,
          histo_bile_stat_plot,
          histo_bileColon_plot,
          nrow = 2,
          rel_widths = c(1, 0.6),
          labels = c('a)', 'c)', 'b)'),
          axis = c('tblr'),
          align = c('hv')) -> histo_bile_plots_together

histo_bile_plots_together
```

**Ratio Plots**
```{r, warning=FALSE, fig.width=8, fig.height=3.5}
histo %>% 
  select(mouse_id, tissue, score) %>% 
  left_join(ratio_bile_acid, by = 'mouse_id') %>% 
  na.omit() -> histo_bileRatio

histo_ratio_plot <- apppleplots::correlation_plots(input_table = histo_bileRatio,
                                                   x_value = 'ratio',
                                                   y_value = 'score',
                                                   x_transform = TRUE,
                                                   y_transform = FALSE,
                                                   point_alpha = 0.4,
                                                   regression_method = 'lm',
                                                   facet_rows = 'tissue',
                                                   facet_cols = 'diet',
                                                   row_labs = tissue_labs,
                                                   col_labs = diet_labs,
                                                   x_name = 'Ratio',
                                                   y_name = 'Score',
                                                   title_content = 'Histopathology Score Against Bile Acid Growth Ratio')

histo_ratio_plot
```

**Ratio Stats**
```{r, warning=FALSE, fig.height=3, fig.width=5}
histo_ratio_stats <- ratio_r2_value_for_loop(histo_bileRatio,
                                              'tissue',
                                             'score',
                                              'ratio')

## assigning p value significance
histo_ratio_stats['signif'] <- symnum(histo_ratio_stats$p_value,
                                     cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1),
                                     symbols = c("****", "***", "**", "*", "ns"),
                                     abbr.colnames = FALSE,
                                     na = "")
```

**Stats on Plot**
```{r, warning=FALSE, fig.width=8, fig.height=3.5}
names(histo_ratio_stats)[names(histo_ratio_stats) == 'category'] <- 'tissue'
col_list <- c('diet', 'tissue', 'signif', 'p_value', 'r_squared')


histo_bileRatio_list <- stat_on_plot(stat_table = histo_ratio_stats,
                                    wanted_col_list = col_list,
                                    x_col_name = 'ratio',
                                    x_col_value = '0.5',
                                    y_col_name = 'score',
                                    y_col_value = '5.5',
                                    plot = histo_ratio_plot,
                                    plot_text_size = 2)

histo_ratioStats_plot <- histo_bileRatio_list$PlotwStats

histo_ratioStats_plot
```


**Statistical Visualization**
```{r, warning=FALSE, fig.height=3, fig.width=5}
histo_ratio_stats %>% 
  mutate(ratio_label = paste('Promoter:Inhibitor')) -> histo_ratio_stats

histo_ratio_stat_plot <- apppleplots::stat_heat_plot(input_table = histo_ratio_stats,
                                                    x_value = 'category',
                                                    y_value = 'diet',
                                                    x_value_type = 'discrete',
                                                    y_value_type = 'discrete',
                                                    pairwise = FALSE,
                                                    correlation = TRUE,
                                                    tile_fill = 'r_direct',
                                                    tile_outline_color = 'black',
                                                    tile_alpha = 0.6,
                                                    text_value = 'signif',
                                                    low_color = 'blue',
                                                    high_color = 'green',
                                                    legend_name = 'Correlation\nCoefficient',
                                                    x_labels = tissue_labs,
                                                    y_labels = stat_diet_y_labs,
                                                    facet_by = 'ratio_label',
                                                    facet_labels = NULL,
                                                    x_name = 'Tissue',
                                                    y_name = 'Diet',
                                                    title_content = 'Histopathology Score and Bile Acid\nRatio Comparisons')

histo_ratio_stat_plot
```


**Ratio Plots Together**
```{r, warning=FALSE, fig.height=3.5, fig.width=15}
plot_grid(histo_ratioStats_plot,
          histo_ratio_stat_plot,
          rel_widths = c(1, 0.5),
          labels = c('a)', 'b)'),
          axis = c('tblr'),
          align = c('hv')) -> histo_ratio_plots_together

histo_ratio_plots_together
```


# **Metabolite Comparisons**

**Plots**
```{r, warning=FALSE, fig.width=10, fig.height=4}
## secondary bile acid cdiff growth/pathogenicity inhibitors 
metab %>% 
  select(metabolite, metab_conc_norm, mouse_id) %>% 
  left_join(bile_acid, by = 'mouse_id') %>% 
  filter(c_diff_effect == 'inhibitor') %>% 
  na.omit() -> metab_bileAcid_inhib

metab_inhibitors_plot <- apppleplots::correlation_plots(input_table = metab_bileAcid_inhib,
                                                        x_value = 'conc_normalized',
                                                        y_value = 'metab_conc_norm',
                                                        x_transform = TRUE,
                                                        y_transform = TRUE,
                                                        point_alpha = 0.4,
                                                        regression_method = 'lm',
                                                        facet_rows = 'metabolite',
                                                        facet_cols = 'diet',
                                                        row_labs = metab_labs,
                                                        col_labs = diet_labs,
                                                        x_name = 'Bile Acid Concentration (log10)',
                                                        y_name = 'Metabolite Concentration (log10)',
                                                        title_content = 'Metabolites Against Bile Acid C. difficile Growth Inhibitors')
  
## secondary bile acid cdiff growth/pathogenicity inhibitors 
metab %>% 
  select(metabolite, metab_conc_norm, mouse_id) %>% 
  left_join(bile_acid, by = 'mouse_id') %>% 
  filter(c_diff_effect == 'promoter') %>% 
  na.omit() -> metab_bileAcid_promot

metab_promoters_plot <- apppleplots::correlation_plots(input_table = metab_bileAcid_promot,
                                                        x_value = 'conc_normalized',
                                                        y_value = 'metab_conc_norm',
                                                        x_transform = TRUE,
                                                        y_transform = TRUE,
                                                        point_alpha = 0.4,
                                                        regression_method = 'lm',
                                                        facet_rows = 'metabolite',
                                                        facet_cols = 'diet',
                                                        row_labs = metab_labs,
                                                        col_labs = diet_labs,
                                                        x_name = 'Bile Acid Concentration (log10)',
                                                        y_name = 'Metabolite Concentration (log10)',
                                                        title_content = 'Metabolites Against Bile Acid C. difficile Growth Promoters')
  

metab_inhibitors_plot
metab_promoters_plot
```

**Stats**
```{r}
metab %>% 
  select(mouse_id, metabolite, metab_conc_norm) %>% 
  left_join(bile_acid, by = 'mouse_id') %>% 
  na.omit() -> metab_bileAcid

metab_bile_stats <- r2_value_for_loop(metab_bileAcid,
                                      'metabolite',
                                      'metab_conc_norm',
                                      'c_diff_effect',
                                      'conc_normalized')

## assigning p value significance
metab_bile_stats['signif'] <- symnum(metab_bile_stats$p_value,
                                     cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1),
                                     symbols = c("****", "***", "**", "*", "ns"),
                                     abbr.colnames = FALSE,
                                     na = "")
```

**Stats on Plot**
```{r, warning=FALSE, fig.width=10, fig.height=4.5}
names(metab_bile_stats)[names(metab_bile_stats) == 'tox_metab'] <- 'c_diff_effect'
names(metab_bile_stats)[names(metab_bile_stats) == 'comp_col'] <- 'metabolite'
col_list <- c('diet', 'metabolite', 'signif', 'p_value', 'r_squared')

## inhibitors
metab_bile_stats %>% 
  filter(c_diff_effect == 'inhibitor') %>% 
  select(-c_diff_effect) -> metab_bileInhib_stats

metab_bileInhib_list <- stat_on_plot(stat_table = metab_bileInhib_stats,
                                    wanted_col_list = col_list,
                                    x_col_name = 'conc_normalized',
                                    x_col_value = '300',
                                    y_col_name = 'metab_conc_norm',
                                    y_col_value = '35',
                                    plot = metab_inhibitors_plot,
                                    plot_text_size = 2)

metab_inhibitorsStats_plot <- metab_bileInhib_list$PlotwStats

## promoters
metab_bile_stats %>% 
  filter(c_diff_effect == 'promoter') %>% 
  select(-c_diff_effect) -> metab_bilePromot_stats

metab_bilePromot_list <- stat_on_plot(stat_table = metab_bilePromot_stats,
                                    wanted_col_list = col_list,
                                    x_col_name = 'conc_normalized',
                                    x_col_value = '10',
                                    y_col_name = 'metab_conc_norm',
                                    y_col_value = '400',
                                    plot = metab_promoters_plot,
                                    plot_text_size = 2)

metab_promotersStats_plot <- metab_bilePromot_list$PlotwStats

metab_inhibitorsStats_plot
metab_promotersStats_plot
```

**Statistical Visualization**
```{r, warning=FALSE, fig.height=4, fig.width=8}
metab_bile_stat_plot <- apppleplots::stat_heat_plot(input_table = metab_bile_stats,
                                                    x_value = 'metabolite',
                                                    y_value = 'diet',
                                                    x_value_type = 'discrete',
                                                    y_value_type = 'discrete',
                                                    pairwise = FALSE,
                                                    correlation = TRUE,
                                                    tile_fill = 'r_direct',
                                                    tile_outline_color = 'black',
                                                    tile_alpha = 0.6,
                                                    text_value = 'signif',
                                                    low_color = 'blue',
                                                    high_color = 'green',
                                                    legend_name = 'Correlation\nCoefficient',
                                                    x_labels = metab_x_labs,
                                                    y_labels = stat_diet_y_labs,
                                                    facet_by = 'c_diff_effect',
                                                    facet_labels = bile_facet_labs,
                                                    x_name = 'Metabolite',
                                                    y_name = 'Diet',
                                                    title_content = 'Metabolite and Bile Acid\nConcentration Comparisons')

metab_bile_stat_plot
```

**Plots Together**
```{r, warning=FALSE, fig.height=8, fig.width=16.5}
plot_grid(metab_inhibitorsStats_plot,
          metab_bile_stat_plot,
          metab_promotersStats_plot,
          nrow = 2,
          rel_widths = c(1, 0.6),
          labels = c('a)', 'c)', 'b)'),
          axis = c('tblr'),
          align = c('hv')) -> metab_bile_plots_together

metab_bile_plots_together
```


# **Toxin Comparisons**

**Plots**
```{r, fig.width=9, fig.height=4, warning=FALSE}
## secondary bile acid cdiff growth/pathogenicity inhibitors 
neat_tox %>% 
  select(neat_toxin, neat_conc_norm, mouse_id) %>% 
  left_join(bile_acid, by = 'mouse_id') %>% 
  filter(c_diff_effect == 'inhibitor') %>% 
  na.omit() -> tox_bileAcid_inhib

toxin_inhibitors_plot <- apppleplots::correlation_plots(input_table = tox_bileAcid_inhib,
                                                        x_value = 'conc_normalized',
                                                        y_value = 'neat_conc_norm',
                                                        x_transform = TRUE,
                                                        y_transform = TRUE,
                                                        point_alpha = 0.4,
                                                        regression_method = 'lm',
                                                        facet_rows = 'neat_toxin',
                                                        facet_cols = 'diet',
                                                        row_labs = neat_labs,
                                                        col_labs = diet_labs,
                                                        x_name = 'Bile Acid Concentration (log10)',
                                                        y_name = 'Toxin Concentration (log10)',
                                                        title_content = 'Toxin Against Bile Acid C. difficile Growth Inhibitors')

## secondary bile acid cdiff growth/pathogenicity inhibitors 
neat_tox %>% 
  select(neat_toxin, neat_conc_norm, mouse_id) %>% 
  left_join(bile_acid, by = 'mouse_id') %>% 
  filter(c_diff_effect == 'promoter') %>% 
  na.omit() -> tox_bileAcid_promot

toxin_promoters_plot <- apppleplots::correlation_plots(input_table = tox_bileAcid_promot,
                                                        x_value = 'conc_normalized',
                                                        y_value = 'neat_conc_norm',
                                                        x_transform = TRUE,
                                                        y_transform = TRUE,
                                                        point_alpha = 0.4,
                                                        regression_method = 'lm',
                                                        facet_rows = 'neat_toxin',
                                                        facet_cols = 'diet',
                                                        row_labs = neat_labs,
                                                        col_labs = diet_labs,
                                                        x_name = 'Bile Acid Concentration (log10)',
                                                        y_name = 'Toxin Concentration (log10)',
                                                        title_content = 'Toxin Against Bile Acid C. difficile Growth Promoters')

toxin_inhibitors_plot
toxin_promoters_plot
```

**Stats**
```{r}
neat_tox %>% 
  select(mouse_id, neat_toxin, neat_conc_norm) %>% 
  left_join(bile_acid, by = 'mouse_id') %>% 
  na.omit() -> tox_bileAcid

tox_bile_stats <- r2_value_for_loop(tox_bileAcid,
                                    'neat_toxin',
                                    'neat_conc_norm',
                                    'c_diff_effect',
                                    'conc_normalized')

## assigning p value significance
tox_bile_stats['signif'] <- symnum(tox_bile_stats$p_value,
                                   cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1),
                                   symbols = c("****", "***", "**", "*", "ns"),
                                   abbr.colnames = FALSE,
                                   na = "")
```

**Stats on the Plot**
```{r, fig.width=9, fig.height=4, warning=FALSE}
names(tox_bile_stats)[names(tox_bile_stats) == 'comp_col'] <- 'neat_toxin'
names(tox_bile_stats)[names(tox_bile_stats) == 'tox_metab'] <- 'c_diff_effect'
col_list <- c('diet', 'neat_toxin', 'signif', 'p_value', 'r_squared')

## inhibitors
tox_bile_stats %>% 
  filter(c_diff_effect == 'inhibitor') %>% 
  select(-c_diff_effect) -> tox_bileInhib_stats

tox_bileInhib_list <- stat_on_plot(stat_table = tox_bileInhib_stats,
                                    wanted_col_list = col_list,
                                    x_col_name = 'conc_normalized',
                                    x_col_value = '10',
                                    y_col_name = 'neat_conc_norm',
                                    y_col_value = '25',
                                    plot = toxin_inhibitors_plot,
                                    plot_text_size = 2.5)

toxin_inhibitors_stats_plot <- tox_bileInhib_list$PlotwStats

## promoters
tox_bile_stats %>% 
  filter(c_diff_effect == 'promoter') %>% 
  select(-c_diff_effect) -> tox_bilePromot_stats

tox_bilePromot_list <- stat_on_plot(stat_table = tox_bilePromot_stats,
                                    wanted_col_list = col_list,
                                    x_col_name = 'conc_normalized',
                                    x_col_value = '3000',
                                    y_col_name = 'neat_conc_norm',
                                    y_col_value = '160',
                                    plot = toxin_promoters_plot,
                                    plot_text_size = 2)

toxin_promoters_stats_plot <- tox_bilePromot_list$PlotwStats


toxin_inhibitors_stats_plot
toxin_promoters_stats_plot
```

**Statistical Visualization**
```{r, warning=FALSE, fig.height=4, fig.width=6.5}
tox_bile_stat_plot <- apppleplots::stat_heat_plot(input_table = tox_bile_stats,
                                                    x_value = 'neat_toxin',
                                                    y_value = 'diet',
                                                    x_value_type = 'discrete',
                                                    y_value_type = 'discrete',
                                                    pairwise = FALSE,
                                                    correlation = TRUE,
                                                    tile_fill = 'r_direct',
                                                    tile_outline_color = 'black',
                                                    tile_alpha = 0.6,
                                                    text_value = 'signif',
                                                    low_color = 'blue',
                                                    high_color = 'green',
                                                    legend_name = 'Correlation\nCoefficient',
                                                    x_labels = neat_labs,
                                                    y_labels = stat_diet_y_labs,
                                                    facet_by = 'c_diff_effect',
                                                    facet_labels = bile_facet_labs,
                                                    x_name = 'Toxin',
                                                    y_name = 'Diet',
                                                    title_content = 'C. difficile Toxin and Bile Acid\nConcentration Comparisons')

tox_bile_stat_plot
```


**Plots Together**
```{r, warning=FALSE, fig.height=7.5, fig.width=16.5}
plot_grid(toxin_inhibitors_stats_plot,
          tox_bile_stat_plot,
          toxin_promoters_stats_plot,
          nrow = 2,
          rel_widths = c(1, 0.6),
          labels = c('a)', 'c)', 'b)'),
          axis = c('tblr'),
          align = c('hv')) -> tox_bile_plots_together

tox_bile_plots_together
```

**Ratio/Toxin Comparisons**
```{r, fig.width=9, fig.height=4, warning=FALSE}
neat_tox %>% 
  select(mouse_id, neat_toxin, neat_conc_norm) %>% 
  left_join(ratio_bile_acid, by = 'mouse_id') %>% 
  na.omit() -> tox_bileRatio

tox_ratio_plot <- apppleplots::correlation_plots(input_table = tox_bileRatio,
                                                   x_value = 'ratio',
                                                   y_value = 'neat_conc_norm',
                                                   x_transform = TRUE,
                                                   y_transform = TRUE,
                                                   point_alpha = 0.4,
                                                   regression_method = 'lm',
                                                   facet_rows = 'neat_toxin',
                                                   facet_cols = 'diet',
                                                   row_labs = neat_labs,
                                                   col_labs = diet_labs,
                                                   x_name = 'Ratio',
                                                   y_name = 'Toxin Concentration',
                                                   title_content = 'Toxin Against Bile Acid Growth Ratio')

tox_ratio_plot
```

**Ratio/Toxin Stats**
```{r}
neat_tox %>% 
  select(mouse_id, neat_toxin, neat_conc_norm) %>% 
  left_join(ratio_bile_acid, by = 'mouse_id') %>% 
  na.omit() -> tox_ratio

tox_ratio_stats <- ratio_r2_value_for_loop(tox_ratio,
                                           'neat_toxin',
                                           'neat_conc_norm',
                                           'ratio')

## assigning p value significance
tox_ratio_stats['signif'] <- symnum(tox_ratio_stats$p_value,
                                    cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1),
                                    symbols = c("****", "***", "**", "*", "ns"),
                                    abbr.colnames = FALSE,
                                    na = "")
```

**Stats on the Plot**
```{r, fig.width=10, fig.height=4, warning=FALSE}
names(tox_ratio_stats)[names(tox_ratio_stats) == 'category'] <- 'neat_toxin'
col_list <- c('diet', 'neat_toxin', 'signif', 'p_value', 'r_squared')


tox_bileRatio_list <- stat_on_plot(stat_table = tox_ratio_stats,
                                    wanted_col_list = col_list,
                                    x_col_name = 'ratio',
                                    x_col_value = '1',
                                    y_col_name = 'neat_conc_norm',
                                    y_col_value = '5.5',
                                    plot = tox_ratio_plot,
                                    plot_text_size = 3)

tox_ratio_stats_plot <- tox_bileRatio_list$PlotwStats

tox_ratio_stats_plot
```

**Statistical Visualization**
```{r, warning=FALSE, fig.height=3, fig.width=5}
tox_ratio_stats %>% 
  mutate(ratio_label = paste('Promoter:Inhibitor')) -> tox_ratio_stats

tox_ratio_stat_plot <- apppleplots::stat_heat_plot(input_table = tox_ratio_stats,
                                                    x_value = 'neat_toxin',
                                                    y_value = 'diet',
                                                    x_value_type = 'discrete',
                                                    y_value_type = 'discrete',
                                                    pairwise = FALSE,
                                                    correlation = TRUE,
                                                    tile_fill = 'r_direct',
                                                    tile_outline_color = 'black',
                                                    tile_alpha = 0.6,
                                                    text_value = 'signif',
                                                    low_color = 'blue',
                                                    high_color = 'green',
                                                    legend_name = 'Correlation\nCoefficient',
                                                    x_labels = neat_labs,
                                                    y_labels = stat_diet_y_labs,
                                                    facet_by = 'ratio_label',
                                                    facet_labels = NULL,
                                                    x_name = 'Toxin',
                                                    y_name = 'Diet',
                                                    title_content = 'Toxin Concentration and Bile Acid\nRatio Comparisons')

tox_ratio_stat_plot
```


**Ratio/Toxin Plots Together**
```{r, fig.width=15, fig.height=3.5}
plot_grid(tox_ratio_stats_plot,
          tox_ratio_stat_plot,
          rel_widths = c(1, 0.5),
          labels = c('a)', 'b)'),
          axis = c('tblr'),
          align = c('hv')) -> tox_ratio_plots_together

tox_ratio_plots_together
```


# **Saving Results**

**Saving my Outputs**
```{r}
## histo comparisons
ggsave('../plots/bileAcid_histo_comp.pdf',
       plot = histo_bile_plots_together,
       width = 16.5,
       height = 8)
ggsave('../plots/bileAcid_histoCecum_comp.pdf',
       plot = histo_bileCecum_stats_plot,
       width = 10,
       height = 4)
ggsave('../plots/bileAcid_histoRatio_comp.pdf',
       plot = histo_ratioStats_plot,
       width = 8,
       height = 3.5)
ggsave('../plots/bileRatio_histo_comp.pdf',
       plot = histo_ratio_plots_together,
       width = 15,
       height = 3.5)

## metab comparisons
ggsave('../plots/bileAcid_metab_comp.pdf',
       plot = metab_bile_plots_together,
       width = 16.5,
       height = 8)
ggsave('../plots/bileAcid_metabInhib_comp.pdf',
       plot = metab_inhibitorsStats_plot,
       width = 10,
       height = 4.5)
ggsave('../plots/bileAcid_metabPromot_comp.pdf',
       plot = metab_promotersStats_plot,
       width = 10.5,
       height = 4.5)

## toxin comparisons
ggsave('../plots/bileAcid_toxin_comp.pdf',
       plot = tox_bile_plots_together,
       width = 16.5,
       height = 8)
ggsave('../plots/bileRatio_toxin_comp.pdf',
       plot = tox_ratio_plots_together,
       width = 15,
       height = 3.5)
ggsave('../plots/bileAcid_toxinInhib_comp.pdf',
       plot = toxin_inhibitors_stats_plot,
       width = 9, 
       height = 4)
ggsave('../plots/bileAcid_toxinPromot_comp.pdf',
       plot = toxin_promoters_stats_plot,
       width = 9, 
       height = 4)
ggsave('../plots/bileRatio_toxin_comp_wStats.pdf',
       plot = tox_ratio_stats_plot,
       width = 10,
       height = 4)

## as rdata objects for manuscript figures
save(histo_inhibitors_plot,
     file = '../../figures/bile_inhibit_histo.rdata')
save(histo_promoters_plot,
     file = '../../figures/bile_promote_histo.rdata')
save(histo_bile_stat_plot,
     file = '../../figures/bile_histo_stats.rdata')
save(histo_ratio_plot,
     file = '../../figures/bile_ratio_histo.rdata')
save(metab_inhibitors_plot,
     file = '../../figures/bile_inhibit_metab.rdata')
save(metab_promoters_plot,
     file = '../../figures/bile_promote_metab.rdata')
save(metab_bile_stat_plot,
     file = '../../figures/bile_metab_stats.rdata')
save(toxin_inhibitors_stats_plot,
     file = '../../figures/bile_inhibit_toxin.rdata')
save(toxin_promoters_plot,
     file = '../../figures/bile_promote_toxin.rdata')
save(tox_bile_stat_plot,
     file = '../../figures/bile_toxin_stats.rdata')
save(histo_ratio_plot,
     file = '../../figures/bile_ratio_histo.rdata')
save(histo_ratio_stat_plot,
     file = '../../figures/bile_ratio_stats.rdata')
```

