---
title: "hypoxia_comparisons"
output: html_document
date: "2023-12-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(broom)
library(cowplot)
library(magrittr)
library(qiime2R)
library(tidyverse)
library(naniar)
library(ggpubr)
library(rstatix)
library(viridis)
library(apppleplots)
```

**Functions**
```{r}
## runs linear modeling on the table, diet, location, and toxin/metabolite wanted and returns full results
## along with a table containing the r2 and p-value 
## this goes into the following function with a for loop 
r2_value <- function(biom_table,
                     wanted_diet,
                     wanted_location,
                     tox_metab_col,
                     wanted_tox_metab,
                     wanted_conc_value){
  ## to get r squared value and p value
  biom_table %>% 
    filter(diet == wanted_diet,
           location == wanted_location,
           .data[[tox_metab_col]] == wanted_tox_metab) %>%
    do(glance(lm(fluorescence ~ .data[[wanted_conc_value]],
                 data = .))) -> lm_glance
  output <- list(r_squared = lm_glance$r.squared,
                 p_value = lm_glance$p.value)
  ## to get estimate and place it with the overall output table
  biom_table %>% 
    filter(diet == wanted_diet,
           location == wanted_location,
           .data[[tox_metab_col]] == wanted_tox_metab) %>%
    do(tidy(lm(fluorescence ~ .data[[wanted_conc_value]],
                 data = .))) %>% 
    filter(term != '(Intercept)') -> lm_full
  output <- as_tibble(output) %>%
            mutate(diet = paste(wanted_diet),
                   location = paste(wanted_location),
                   tox_metab = paste(wanted_tox_metab),
                   just_r = sqrt(r_squared),
                   estimate = lm_full$estimate,
                   r_direct = if_else(estimate < 0, just_r * -1, just_r))
  my_list <- list(LinearModel = lm_full,
                  WantedStats = output)
  return(my_list)
}

## creating a nested for loop since running this function individually is a nightmare
r2_value_for_loop <- function(biom_table,
                              wanted_col,
                              wanted_conc) {
  ## nested for loop for the r squared and p values 
  output <- tibble()
  for(i in unique(unlist(biom_table$diet))) {
    for(j in unique(unlist(biom_table$location))) {
      for(k in unique(unlist(biom_table[wanted_col]))) {
        tmp_output <- r2_value(biom_table,
                               i,
                               j,
                               wanted_col,
                               k,
                               wanted_conc)$WantedStats
        output <- bind_rows(output, tmp_output)
      }
    }
  }
  return(output)
}
```

**File Paths**
```{r}
metadata_FP <- '../data/misc/filt_cecal_processed_metadata.tsv'
histo_FP <- '../data/misc/processed_histopathology.tsv'
metab_FP <- '../data/misc/processed_metabolomics.tsv'
toxin_FP <- '../data/misc/processed_neatToxin.tsv'
hypoxia_FP <- '../data/misc/pimid_fluor.csv'

## lists to redo the diet names on the facet labels of the ggplot created below 
diet_labs <- 
    c('Chow', 
      'High Fat / High Fiber', 
      'High Fat / Low Fiber', 
      'Low Fat / High Fiber', 
      'Low Fat / Low Fiber')

names(diet_labs) <- c('Chow', 'HF/HF', 'HF/LF', 'LF/HF', 'LF/LF')

tissue_labs <- c('Cecum',
                 'Colon')
names(tissue_labs) <- c('cecum',
                        'colon')
diet_y_labs <- c('Chow',
                  'HFt/HFb',
                  'HFt/LFb',
                  'LFt/HFb',
                  'LFt/LFb')

metab_labs <- c('Acetic Acid',
                'Propanoic Acid',
                'n-Butanoic Acid')
names(metab_labs) <- c('Acetic Acid (ug/g)',
                       'Propanoic Acid (ug/g)',
                       'n-Butanoic Acid (ug/g)')

neat_labs <- c('TcdA', 'TcdB')
names(neat_labs) <- c('Total TcA Neat', 'Total TcB Neat')

dil_labs <- c('TcdA', 'TcdB')
names(dil_labs) <- c('Total TcA 1:10', 'Total TcB 1:10')

location_labs <- c('Cecum',
                   'Proximal Colon',
                   'Distal Colon')
names(location_labs) <- c(1,
                          2,
                          3)
```

**Data Wrangling**
```{r}
metadata <- read_tsv(metadata_FP)

hypoxia <- read_csv(hypoxia_FP) %>% 
  mutate(order = ifelse(location == 'Cecum', 1, location),
         order = ifelse(order == 'Prox_colon', 2, order),
         order = ifelse(order == 'Dist_colon', 3, order)) %>% 
  arrange(order)

metab <- read_tsv(metab_FP) %>% 
          select(metabolite, concentration, mouse_id)

neat_toxin <- read_tsv(toxin_FP) %>% 
              select(neat_toxin, neat_conc, mouse_id)

histo <- read_tsv(histo_FP) %>% 
         select(mouse_id, tissue, score)

## filtered tables for plots
## histopathology
hypox_histo_table <- hypoxia %>% 
  left_join(histo, by = 'mouse_id') %>% 
  na.omit()

hypox_histo_table %>% 
  filter(location == 'Cecum') -> hypox_cecumOnly

hypox_histo_table %>% 
  filter(location != 'Cecum') -> hypox_colonOnly

## toxin
hypox_toxin_table <- hypoxia %>% 
  left_join(neat_toxin, by = 'mouse_id') %>% 
  na.omit()

hypox_toxin_table %>% 
  filter(neat_toxin == 'Total TcA Neat') -> hypox_tcda_table

hypox_toxin_table %>% 
  filter(neat_toxin == 'Total TcB Neat') -> hypox_tcdb_table

## metabolomics
hypox_metab_table <- hypoxia %>% 
  left_join(metab, by = 'mouse_id') %>% 
  na.omit()

hypox_metab_table %>% 
  filter(metabolite == 'Acetic Acid (ug/g)') -> hypox_acetic_table

hypox_metab_table %>% 
  filter(metabolite == 'n-Butanoic Acid (ug/g)') -> hypox_butanoic_table

hypox_metab_table %>% 
  filter(metabolite == 'Propanoic Acid (ug/g)') -> hypox_propanoic_table
```

#**Hypoxia and Histopathology Score Comparisons**
**Plot**
```{r, warning=FALSE, fig.width=10, fig.height=3}
## cecum
hypox_cecumHisto_plot <- apppleplots::correlation_plots(input_table = hypox_cecumOnly,
                                                   x_value = 'score',
                                                   y_value = 'fluorescence',
                                                   x_transform = FALSE,
                                                   y_transform = TRUE,
                                                   point_alpha = 0.4,
                                                   regression_method = 'lm',
                                                   facet_rows = 'order',
                                                   facet_cols = 'diet',
                                                   row_labs = location_labs,
                                                   col_labs = diet_labs,
                                                   x_name = 'Histopathology Score',
                                                   y_name = 'Fluorescence (log10)',
                                                   title_content = 'Cecum Hypoxia by Inflammation')


## colon
hypox_colon_plot <- apppleplots::correlation_plots(input_table = hypox_colonOnly,
                                                   x_value = 'score',
                                                   y_value = 'fluorescence',
                                                   x_transform = FALSE,
                                                   y_transform = TRUE,
                                                   point_alpha = 0.4,
                                                   regression_method = 'lm',
                                                   facet_rows = 'order',
                                                   facet_cols = 'diet',
                                                   row_labs = location_labs,
                                                   col_labs = diet_labs,
                                                   x_name = 'Histopathology Score',
                                                   y_name = 'Fluorescence (log10)',
                                                   title_content = 'Colon Hypoxia by Inflammation')


hypox_cecumHisto_plot
hypox_colon_plot
```

**Stats**
```{r}
hypox_histo_stat_table <- r2_value_for_loop(hypox_histo_table,
                                            "tissue",
                                            "score")

hypox_histo_stat_table['signif'] <- symnum(hypox_histo_stat_table$p_value,
                                           cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
                                           symbols = c("****", "***", "**", "*", "+", "ns"),
                                           abbr.colnames = FALSE,
                                           na = "")

hypox_histo_stat_table %>% 
  mutate(order = ifelse(location == 'Cecum', 1, location),
           order = ifelse(order == 'Prox_colon', 2, order),
           order = ifelse(order == 'Dist_colon', 3, order)) %>% 
  arrange(order) -> hypox_histo_stat_table


hypox_histo_stat_plot <- apppleplots::stat_heat_plot(input_table = hypox_histo_stat_table,
                                                     x_value = 'order',
                                                     y_value = 'diet',
                                                     x_value_type = 'discrete',
                                                     y_value_type = 'discrete',
                                                     pairwise = FALSE,
                                                     correlation = TRUE,
                                                     tile_fill = 'r_direct',
                                                     tile_outline_color = 'black',
                                                     tile_alpha = 0.6,
                                                     text_value = 'signif',
                                                     low_color = 'blue',
                                                     high_color = 'green',
                                                     legend_name = 'Correlation\nCoefficient',
                                                     x_labels = location_labs,
                                                     y_labels = diet_y_labs,
                                                     facet_by = 'tox_metab',
                                                     facet_labels = tissue_labs,
                                                     x_name = 'Location',
                                                     y_name = 'Diet',
                                                     title_content = "Hypoxia and Histopathology Score Comparisons")


hypox_histo_stat_plot
```
**Putting the Stats on the Actual Plot**
```{r, fig.width=17, fig.height=4}
hypox_histo_stat_table %>% 
    filter(tox_metab == 'colon',
           location != 'Cecum',
           signif != 'ns') %>% 
    select(diet, location, signif, p_value) %>% 
    mutate(score = paste(8),
           score = as.numeric(score),
           fluorescence = paste(100),
           fluorescence = as.numeric(fluorescence),
           order = ifelse(location == "Prox_colon", 2, 3),
           p_value = signif(p_value, digits = 3),
           p_value = paste('p-value', p_value, sep = ": ")) -> proc_stat_table
  
# names(proc_stat_table)[names(proc_stat_table) == 'location'] <- 'order'
  
hypox_colon_plot +
  geom_label(data = proc_stat_table,
            label = proc_stat_table$p_value,
            color = 'black',
            fontface = 'bold',
            size = 6) -> hypox_colonStats_plot

hypox_colonStats_plot
```

**Everything Together**
```{r, warning=FALSE, fig.height=5, fig.width=17}
hypoxia_histo_plots <- plot_grid(hypox_cecumHisto_plot,
                                 hypox_colon_plot,
                                 hypox_histo_stat_plot,
                                 nrow = 2,
                                 labels = c('a)', 'b)', 'c)'),
                                 label_size = 18,
                                 align = c('hv'),
                                 axis = c('tblr'),
                                 rel_heights = c(1, 0.8))

hypoxia_histo_plots
```

#**Hypoxia and Neat C. difficile Toxin Concentration Comparisons**
**Plots**
```{r, warning=FALSE, fig.width=10, fig.height=5}
hypox_tcda_plot <- apppleplots::correlation_plots(input_table = hypox_tcda_table,
                                                   x_value = 'neat_conc',
                                                   y_value = 'fluorescence',
                                                   x_transform = TRUE,
                                                   y_transform = TRUE,
                                                   point_alpha = 0.4,
                                                   regression_method = 'lm',
                                                   facet_rows = 'order',
                                                   facet_cols = 'diet',
                                                   row_labs = location_labs,
                                                   col_labs = diet_labs,
                                                   x_name = 'Concentration (ng/uL) (log10)',
                                                   y_name = 'Fluorescence (log10)',
                                                   title_content = 'Hypoxia Compared to TcdA Toxin')
  


hypox_tcdb_plot <- apppleplots::correlation_plots(input_table = hypox_tcdb_table,
                                                   x_value = 'neat_conc',
                                                   y_value = 'fluorescence',
                                                   x_transform = TRUE,
                                                   y_transform = TRUE,
                                                   point_alpha = 0.4,
                                                   regression_method = 'lm',
                                                   facet_rows = 'order',
                                                   facet_cols = 'diet',
                                                   row_labs = location_labs,
                                                   col_labs = diet_labs,
                                                   x_name = 'Concentration (ng/uL) (log10)',
                                                   y_name = 'Fluorescence (log10)',
                                                   title_content = 'Hypoxia Compared to TcdB Toxin')

hypox_tcda_plot
hypox_tcdb_plot
```

**Stats**
I think there are too few points for the high-fat/high-fiber diet so there isn't a calculated p-value for that diet. I turned the NAs into 1 so the result would show as non-significant on the plot. 
```{r}
hypox_toxin_stat_table <- r2_value_for_loop(hypox_toxin_table,
                                            "neat_toxin",
                                            "neat_conc")
hypox_toxin_stat_table %>% 
  mutate(p_value = ifelse(p_value == 'NaN', 1, p_value)) -> hypox_toxin_stat_table

hypox_toxin_stat_table['signif'] <- symnum(hypox_toxin_stat_table$p_value,
                                           cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
                                           symbols = c("****", "***", "**", "*", "+", "ns"),
                                           abbr.colnames = FALSE,
                                           na = "")

hypox_toxin_stat_table %>% 
  mutate(order = ifelse(location == 'Cecum', 1, location),
           order = ifelse(order == 'Prox_colon', 2, order),
           order = ifelse(order == 'Dist_colon', 3, order)) %>% 
  arrange(order) -> hypox_toxin_stat_table


hypox_toxin_stat_plot <- apppleplots::stat_heat_plot(input_table = hypox_toxin_stat_table,
                                                     x_value = 'order',
                                                     y_value = 'diet',
                                                     x_value_type = 'discrete',
                                                     y_value_type = 'discrete',
                                                     pairwise = FALSE,
                                                     correlation = TRUE,
                                                     tile_fill = 'r_direct',
                                                     tile_outline_color = 'black',
                                                     tile_alpha = 0.6,
                                                     text_value = 'signif',
                                                     low_color = 'blue',
                                                     high_color = 'green',
                                                     legend_name = 'Correlation\nCoefficient',
                                                     x_labels = location_labs,
                                                     y_labels = diet_y_labs,
                                                     facet_by = 'tox_metab',
                                                     facet_labels = neat_labs,
                                                     x_name = 'Location',
                                                     y_name = 'Diet',
                                                     title_content = "Hypoxia and Toxin Comparisons")

hypox_toxin_stat_plot
```

**Everything Together**
```{r, warning=FALSE, fig.height=7, fig.width=17}
hypoxia_toxin_plots <- plot_grid(hypox_tcda_plot,
                                 hypox_tcdb_plot,
                                 hypox_toxin_stat_plot,
                                 nrow = 2,
                                 labels = c('a)', 'b)', 'c)'),
                                 label_size = 18,
                                 align = c('hv'),
                                 axis = c('tblr'),
                                 rel_heights = c(1, 0.6))

hypoxia_toxin_plots
```

#**Hypoxia and Metabolite Concentration Comparisons**
**Plots**
```{r, warning=FALSE, fig.width=10, fig.height=5}
hypox_acetic_plot <- apppleplots::correlation_plots(input_table = hypox_acetic_table,
                                                    x_value = 'concentration',
                                                    y_value = 'fluorescence',
                                                    x_transform = TRUE,
                                                    y_transform = TRUE,
                                                    point_alpha = 0.4,
                                                    regression_method = 'lm',
                                                    facet_rows = 'order',
                                                    facet_cols = 'diet',
                                                    row_labs = location_labs,
                                                    col_labs = diet_labs,
                                                    x_name = 'Concentration (ug/g) (log10)',
                                                    y_name = 'Fluorescence (log10)',
                                                    title_content = 'Hypoxia Compared to Acetate Concentration')
  
  


hypox_butanoic_plot <- apppleplots::correlation_plots(input_table = hypox_butanoic_table,
                                                      x_value = 'concentration',
                                                      y_value = 'fluorescence',
                                                      x_transform = TRUE,
                                                      y_transform = TRUE,
                                                      point_alpha = 0.4,
                                                      regression_method = 'lm',
                                                      facet_rows = 'order',
                                                      facet_cols = 'diet',
                                                      row_labs = location_labs,
                                                      col_labs = diet_labs,
                                                      x_name = 'Concentration (ug/g) (log10)',
                                                      y_name = 'Fluorescence (log10)',
                                                      title_content = 'Hypoxia Compared to Butyrate Concentration')

hypox_propanoic_plot <- apppleplots::correlation_plots(input_table = hypox_propanoic_table,
                                                       x_value = 'concentration',
                                                       y_value = 'fluorescence',
                                                       x_transform = TRUE,
                                                       y_transform = TRUE,
                                                       point_alpha = 0.4,
                                                       regression_method = 'lm',
                                                       facet_rows = 'order',
                                                       facet_cols = 'diet',
                                                       row_labs = location_labs,
                                                       col_labs = diet_labs,
                                                       x_name = 'Concentration (ug/g) (log10)',
                                                       y_name = 'Fluorescence (log10)',
                                                       title_content = 'Hypoxia Compared to Propionate Concentration')

hypox_acetic_plot
hypox_butanoic_plot
hypox_propanoic_plot
```

**Stats**
```{r, warning=FALSE, fig.height=4, fig.width=10}
hypox_metab_stat_table <- r2_value_for_loop(hypox_metab_table,
                                            "metabolite",
                                            "concentration")
hypox_metab_stat_table %>% 
  mutate(p_value = ifelse(p_value == 'NaN', 1, p_value)) -> hypox_metab_stat_table

hypox_metab_stat_table['signif'] <- symnum(hypox_metab_stat_table$p_value,
                                           cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
                                           symbols = c("****", "***", "**", "*", "+", "ns"),
                                           abbr.colnames = FALSE,
                                           na = "")

hypox_metab_stat_table %>% 
  mutate(order = ifelse(location == 'Cecum', 1, location),
           order = ifelse(order == 'Prox_colon', 2, order),
           order = ifelse(order == 'Dist_colon', 3, order)) %>% 
  arrange(order) -> hypox_metab_stat_table


hypox_metab_stat_plot <- apppleplots::stat_heat_plot(input_table = hypox_metab_stat_table,
                                                     x_value = 'order',
                                                     y_value = 'diet',
                                                     x_value_type = 'discrete',
                                                     y_value_type = 'discrete',
                                                     pairwise = FALSE,
                                                     correlation = TRUE,
                                                     tile_fill = 'r_direct',
                                                     tile_outline_color = 'black',
                                                     tile_alpha = 0.6,
                                                     text_value = 'signif',
                                                     low_color = 'blue',
                                                     high_color = 'green',
                                                     legend_name = 'Correlation\nCoefficient',
                                                     x_labels = location_labs,
                                                     y_labels = diet_y_labs,
                                                     facet_by = 'tox_metab',
                                                     facet_labels = metab_labs,
                                                     x_name = 'Location',
                                                     y_name = 'Diet',
                                                     title_content = "Hypoxia and SCFA Comparisons")

hypox_metab_stat_plot
```

**Everything Together**
```{r, warning=FALSE, fig.height=10, fig.width=20}
hypoxia_metab_plots <- plot_grid(hypox_acetic_plot,
                                 hypox_butanoic_plot,
                                 hypox_propanoic_plot,
                                 hypox_metab_stat_plot,
                                 nrow = 2,
                                 labels = c('a)', 'b)', 'c)', 'd)'),
                                 label_size = 18,
                                 align = c('hv'),
                                 axis = c('tblr'))

hypoxia_metab_plots
```

**Saving my Outputs**
```{r}
ggsave('../plots/hypoxia_histo_comp_plots.pdf',
       plot = hypoxia_histo_plots,
       width = 17, 
       height = 7)
ggsave('../plots/hypoxia_toxin_comp_plots.pdf',
       plot = hypoxia_toxin_plots,
       width = 17, 
       height = 7)
ggsave('../plots/hypoxia_metab_comp_plots.pdf',
       plot = hypoxia_metab_plots,
       width = 20, 
       height = 10)

## as an rdata object 
save(hypox_tcda_plot,
     file = '../../figures/hypoxia_tcda_comp.rdata')
save(hypox_tcdb_plot,
     file = '../../figures/hypoxia_tcdb_comp.rdata')
save(hypox_toxin_stat_plot,
     file = '../../figures/hypoxia_toxin_stats.rdata')
save(hypox_cecumHisto_plot,
     file = '../../figures/hypox_histo_cecum_comp.rdata')
save(hypox_colonStats_plot,
     file = '../../figures/hypox_histo_colon_comp.rdata')
```

