---
title: "toxinHisto_ratio_plot"
output: html_document
date: "2025-04-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(broom)
library(cowplot)
library(magrittr)
library(qiime2R)
library(tidyverse)
library(naniar)
library(ggpubr)
library(rstatix)
library(viridis)
library(apppleplots)
```

**Functions**
```{r}
## will give the r2 and p values for each value in the specified column 
lm_r2_p_value <- function(input_table,
                          wanted_col1,
                          wanted_col2 = NULL,
                          formula_left,
                          formula_right){
  if (is.character(wanted_col1) & is.character(wanted_col2)) {
    output <- tibble()
    for(i in unique(unlist(input_table[wanted_col1]))) {
      for(j in unique(unlist(input_table[wanted_col2]))) {
        funky_formula <- paste(formula_left, formula_right, sep = "~")
        lm <- input_table %>% 
                filter(.data[[wanted_col1]] == i,
                       .data[[wanted_col2]] == j) %>%
                do(glance(lm(as.formula(funky_formula),
                             data = .))) %>% 
              mutate("{wanted_col1}" := paste(i),
                     "{wanted_col2}" := paste(j))
        output <- bind_rows(output, lm)
        
        stat_table <- output %>% 
          select(r.squared, p.value, .data[[wanted_col1]], .data[[wanted_col2]]) %>% 
          rename(r_squared = r.squared,
                 p_value = p.value) %>% 
          mutate(p_value = if_else(is.na(p_value), 1, p_value))
      }
    }
    
  } else {
    output <- tibble()
    for(i in unique(unlist(input_table[wanted_col1]))){
        funky_formula <- paste(formula_left, formula_right, sep = "~")
        lm <- input_table %>% 
                filter(.data[[wanted_col1]] == i) %>%
                do(glance(lm(as.formula(funky_formula),
                             data = .))) %>% 
              mutate("{wanted_col1}" := paste(i))
        output <- bind_rows(output, lm)
        
        ## putting together actual stat table 
        stat_table <- output %>% 
          select(r.squared, p.value, .data[[wanted_col1]]) %>% 
          rename(r_squared = r.squared,
                 p_value = p.value) %>% 
          mutate(p_value = if_else(is.na(p_value), 1, p_value))
    }
  }
  
  stat_table['signif'] <- symnum(stat_table$p_value,
                                 cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1),
                                 symbols = c("****", "***", "**", "*", "ns"),
                                 abbr.colnames = FALSE,
                                 na = "")
  return(stat_table)
}

## putting stats directly on correlation plots 
stat_on_plot <- function(stat_table,
                         wanted_col_list,
                         x_col_name,
                         x_col_value,
                         y_col_name,
                         y_col_value,
                         plot,
                         plot_text_size){
  ## making table conducive to labeling plot 
  proc_stat_table <- stat_table %>% 
    filter(signif != 'ns') %>% 
    select(all_of(wanted_col_list)) %>% 
    ## created columns need to match up with the column names called for the x and y axis in the plot
    ## numeric values added are the x and y coordinates for where the label will sit in the facet block 
    mutate("{x_col_name}" := paste(x_col_value),
           "{x_col_name}" := as.numeric(.data[[x_col_name]]),
           "{y_col_name}" := paste(y_col_value),
           "{y_col_name}" := as.numeric(.data[[y_col_name]]),
           p_value = signif(p_value, digits = 3),
           r_squared = signif(r_squared, digits = 3),
           pVal_rSquare = paste0('p-value: ', p_value, "\n", 
                                 'r2: ', r_squared))
  
  ## putting labels on plot 
  plot_wStats <- plot +
    geom_label(data = proc_stat_table, 
               label = proc_stat_table$pVal_rSquare,
               color = 'black',
               fontface = 'bold',
               size = plot_text_size)
  
  ## creating list of my outputs
  my_list <- list(ProcStatTable = proc_stat_table,
                  PlotwStats = plot_wStats)
  
  return(my_list)
}
```

**File Paths**
```{r}
tox_histo_table_FP <- '../data/misc/histoMetabToxin_results_comb.tsv'
histo_cat_FP <- '../data/misc/histo_categories.txt'

## lists to redo the diet names on the facet labels of the ggplot created below 
diet_x_labs <- c('Chow', 
                  'HFt/\nHFb', 
                  'HFt/\nLFb', 
                  'LFt/\nHFb', 
                  'LFt/\nLFb')

neat_labs <- c('TcdA', 'TcdB')
names(neat_labs) <- c('Total TcA Neat', 'Total TcB Neat')

cecum_labels <- c('Submucosal Edema',
                  'Neutrophil Infiltration',
                  'Number of Goblet Cells',
                  'Epithelial Integrity')
names(cecum_labels) <- c('cecum_submuc_edema',
                        'cecum_PNM_infil',
                        'cecum_num_gobletCells',
                        'cecum_epithelial_integrity')

colon_labels <- c('Severity of Inflammation',
                  'Extent of Injury',
                  'Epithelial Regeneration',
                  'Crypt Damage')
names(colon_labels) <- c('colon_inflamm_sev',
                        'colon_injury_extent',
                        'colon_epithelial_regen',
                        'colon_crypt_damage')
```

**Reading in Files**
```{r}
tox_histo_table <- read_tsv(tox_histo_table_FP)

## histo categories tables 
cecum_histoCat_table <- read_tsv(histo_cat_FP) %>% 
  gather(contains('cecum'), key = 'cecum_category', value = 'cecum_score') %>% 
  select(mouse_id, cecum_category, cecum_score)

colon_histoCat_table <- read_tsv(histo_cat_FP) %>% 
  gather(contains('colon'), key = 'colon_category', value = 'colon_score') %>% 
  select(mouse_id, colon_category, colon_score)
```

**Data Wrangled**

*for toxin/histo score ratio*
calculating the ratio of toxin concentration to histopathology score per mouse
- should I be normalizing the score/concentration before I calculate the ratio bc they have very different units??
- note: normalized both score/toxin concentration using min/max normalization (or scaling)
```{r}
(toxHisto_ratio_table <- tox_histo_table %>% 
  filter(microbe_product == 'Total TcA Neat' | microbe_product == 'Total TcB Neat',
         tissue == 'cecum') %>% 
  mutate(score_norm = ((score - min(score)) / (max(score) - min(score))),
         tox_norm = ((concentration - min(concentration)) / (max(concentration) - min(concentration)))) %>% 
  group_by(mouse_id, microbe_product) %>% 
  mutate(tox_histo_ratio = (tox_norm/score_norm),
         tox_histo_ratio = ifelse(tox_histo_ratio == 'Inf', 0, tox_histo_ratio),
         tox_histo_ratio = ifelse(tox_histo_ratio == 0, tox_histo_ratio + 0.001, tox_histo_ratio)))
```

*for toxin/histo score categories correlation plot*
```{r}
## cecum
(cecum_toxHisto_cat_table <- tox_histo_table %>% 
   filter(tissue == 'cecum') %>% 
   select(mouse_id, diet, sampleid, microbe_product, concentration, conc_normalized) %>% 
   filter(microbe_product == 'Total TcA Neat' | microbe_product == 'Total TcB Neat') %>% 
   left_join(cecum_histoCat_table, by = 'mouse_id', relationship = "many-to-many"))

## colon
(colon_toxHisto_cat_table <- tox_histo_table %>% 
   filter(tissue == 'colon') %>% 
   select(mouse_id, diet, sampleid, microbe_product, concentration, conc_normalized) %>% 
   filter(microbe_product == 'Total TcA Neat' | microbe_product == 'Total TcB Neat') %>% 
   left_join(colon_histoCat_table, by = 'mouse_id', relationship = "many-to-many"))
```


**Plots**

*for toxin/histo score ratio*
scales are not log10 transformed
```{r, fig.width=10, fig.height=5}
(toxHisto_ratio_plot <- toxHisto_ratio_table %>% 
  ggplot(aes(x = diet, y = tox_histo_ratio)) +
  geom_boxplot(aes(group = diet), outlier.shape = NA) +
  geom_jitter(alpha = 0.5, size = 3, width = 0.1, height = 0) +
  theme_bw(base_size = 20) +
  facet_wrap(~microbe_product,
             labeller = labeller(microbe_product = neat_labs)) +
  scale_x_discrete(labels = diet_x_labs) +
  labs(x = 'Diet',
       y = 'Ratio (conc./score)',
       title = 'Cecum Toxin:Histopathology Ratio'))
```
scales are log10 transformed
```{r, fig.width=10, fig.height=5}
(log10_toxHisto_ratio_plot <- toxHisto_ratio_plot +
   scale_y_continuous(trans = 'log10',
                      labels = scales::comma))
```

*for toxin/histo score categories correlation plot*
cecum
```{r, fig.width=15, fig.height=10}
cecum_toxHisto_cat_plot <- apppleplots::correlation_plots(input_table = cecum_toxHisto_cat_table,
                                                           x_value = 'conc_normalized',
                                                           y_value = 'cecum_score',
                                                           x_transform = TRUE,
                                                           y_transform = FALSE,
                                                           point_alpha = 0,
                                                           regression_method = "lm",
                                                           facet_rows = 'cecum_category',
                                                           facet_cols = 'microbe_product',
                                                           row_labs = cecum_labels,
                                                           col_labs = neat_labs,
                                                           x_name = 'Concentration (ng/uL)(log10)',
                                                           y_name = 'Score',
                                                           title_content = 
                                                             'Cecum Toxin vs Cecum Histopathology Scoring Category') +
                    geom_jitter(aes(fill = diet), pch = 21, alpha = 0.6, size = 4) +
                    scale_fill_viridis(option = 'C',
                                       discrete = TRUE,
                                       name = 'Mouse Diet',
                                       labels = c('Chow',
                                                 'HFt / HFb',
                                                 'HFt / LFb',
                                                 'LFt / HFb',
                                                 'LFt / LFb')) +
                    theme_bw(base_size = 20) +
                    theme(strip.text.y.right = element_text(angle = 0))


## stats
cecum_toxHisto_stats <- lm_r2_p_value(input_table = cecum_toxHisto_cat_table,
                                       wanted_col1 = 'microbe_product',
                                       wanted_col2 = 'cecum_category',
                                       formula_left = 'cecum_score',
                                       formula_right = 'conc_normalized * diet')


## stats on plot
cecum_toxHisto_cat_list <- stat_on_plot(stat_table = cecum_toxHisto_stats,
                                        wanted_col_list = colnames(cecum_toxHisto_stats),
                                        x_col_name = 'conc_normalized',
                                        x_col_value = '10',
                                        y_col_name = 'cecum_score',
                                        y_col_value = '1.75',
                                        plot = cecum_toxHisto_cat_plot,
                                        plot_text_size = 5)

(cecum_toxHisto_catStats_plot <- cecum_toxHisto_cat_list$PlotwStats)
```

colon
```{r, fig.width=15, fig.height=10}
colon_toxHisto_cat_plot <- apppleplots::correlation_plots(input_table = colon_toxHisto_cat_table,
                                                           x_value = 'conc_normalized',
                                                           y_value = 'colon_score',
                                                           x_transform = TRUE,
                                                           y_transform = FALSE,
                                                           point_alpha = 0,
                                                           regression_method = "lm",
                                                           facet_rows = 'colon_category',
                                                           facet_cols = 'microbe_product',
                                                           row_labs = colon_labels,
                                                           col_labs = neat_labs,
                                                           x_name = 'Concentration (ng/uL)(log10)',
                                                           y_name = 'Score',
                                                           title_content = 
                                                             'Cecum Toxin vs Colon Histopathology Scoring Category')+
                    geom_jitter(aes(fill = diet), pch = 21, alpha = 0.6, size = 4) +
                    scale_fill_viridis(option = 'C',
                                       discrete = TRUE,
                                       name = 'Mouse Diet',
                                       labels = c('Chow',
                                                 'HFt / HFb',
                                                 'HFt / LFb',
                                                 'LFt / HFb',
                                                 'LFt / LFb')) +
                    theme_bw(base_size = 20) +
                    theme(strip.text.y.right = element_text(angle = 0))

## stats
colon_toxHisto_stats <- lm_r2_p_value(input_table = colon_toxHisto_cat_table,
                                       wanted_col1 = 'microbe_product',
                                       wanted_col2 = 'colon_category',
                                       formula_left = 'colon_score',
                                       formula_right = 'conc_normalized * diet')


## stats on plot
colon_toxHisto_cat_list <- stat_on_plot(stat_table = colon_toxHisto_stats,
                                        wanted_col_list = colnames(colon_toxHisto_stats),
                                        x_col_name = 'conc_normalized',
                                        x_col_value = '10',
                                        y_col_name = 'colon_score',
                                        y_col_value = '5',
                                        plot = colon_toxHisto_cat_plot,
                                        plot_text_size = 5)

(colon_toxHisto_catStats_plot <- colon_toxHisto_cat_list$PlotwStats)
```

**Saving my Outputs**
```{r}
## toxin:cecum histo score ratio plots
ggsave('../plots/cecum_toxHisto_ratio_plot.pdf',
       plot = toxHisto_ratio_plot,
       width = 10, 
       height = 5)
ggsave('../plots/cecum_toxHisto_ratioLog10_plot.pdf',
       plot = log10_toxHisto_ratio_plot,
       width = 10, 
       height = 5)

## toxin to cecum/colon histo scores per category plots
ggsave('../plots/cecum_toxHisto_cat_plot.pdf',
       plot = cecum_toxHisto_catStats_plot,
       width = 15, 
       height = 10)
ggsave('../plots/colon_toxHisto_cat_plot.pdf',
       plot = colon_toxHisto_catStats_plot,
       width = 15, 
       height = 10)
```

