---
title: "az_newExp_diet_deathDate"
output: html_document
date: "2025-01-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(apppleplots)
library(magrittr)
library(readr)
```

**Functions**
```{r}
## puts together a summary table for the death range of the given group(s) in the survival data 
## assumes that you have your survival data set up with day of death = 1 and survival = 0
death_range_table <- function(input_table,
                              surv_col_name,
                              wanted_group,
                              timeline_day_col){
  ## day the mice stopped dying 
  stop_death_df <- input_table %>% 
    filter(.data[[surv_col_name]] == 1) %>% 
    ## need across(all_of(group)) because im passing the groups into mutate/summarise and it gets mad
    group_by(across(all_of(wanted_group))) %>%
    summarise(stop_death = max(.data[[timeline_day_col]]))
  
  ## how many mice of each died died during the death range 
  mice_lost_during <- input_table %>% 
    filter(.data[[surv_col_name]] == 1) %>% 
    count(across(all_of(wanted_group))) %>% 
    rename(num_mice_lost = n)
  
  ## day the mice started dying and combining it with the counts table
  ## and the day the mice stopped dying to create one summary table 
  death_range_df <- input_table %>% 
    filter(.data[[surv_col_name]] == 1) %>% 
    group_by(across(all_of(wanted_group))) %>% 
    summarise(start_death = min(.data[[timeline_day_col]])) %>% 
    left_join(stop_death_df, by = wanted_group) %>% 
    mutate(death_range = paste(start_death, stop_death, sep = "_")) %>% 
    left_join(mice_lost_during, by = wanted_group)
  
  return(death_range_df)
}


## editing my dunn's post hoc test to include the difference in means between groups 
## this version of the function is for when you don't group by any variable for your stats 
edit_dunn_test_v2 <- function(input_table,
                              wanted_group,
                              mean_value,
                              dunn_test){
  input_table %>% 
    group_by(across(all_of(wanted_group))) %>% 
    summarise(mean = mean(.data[[mean_value]])) -> mean_table
  
  ## doing some funky stuff to put the merge by columns together accurately and make my life easier
  ## looks for the column in the dunn test df that matches the wanted_group param and adds it to the 
  ## group1 and group2 lists for merging the dunn test df and mean_table together 
  pre_group_list <- lapply(wanted_group, function(x) x[x %in% colnames(dunn_test)])
  proc_group_list <- sapply(pre_group_list, function(x) paste(sort(x[x %in% wanted_group]), collapse=""))
  wanted_col <- scan(text = proc_group_list, sep = "", what = character())
  
  group1_list <- append('group1', wanted_col)
  group2_list <- append('group2', wanted_col)
  
  dunn_test %>% 
    merge(mean_table, 
          by.x = group1_list,
          by.y = wanted_group) %>% 
    rename_with(~paste0('group1_', mean_value, recycle0 = TRUE), contains('mean')) %>% 
    merge(mean_table,
          by.x = group2_list,
          by.y = wanted_group) %>% 
    rename_with(~paste0('group2_', mean_value, recycle0 = TRUE), contains('mean')) -> int_dunn
  
  group1_col <- paste0('group1_', mean_value)
  group2_col <- paste0('group2_', mean_value)
  
  int_dunn %>% 
    mutate(diff_means = (.data[[group1_col]] - .data[[group2_col]]),
           stat_diff_means = if_else(p.adj > 0.05, 0, diff_means)) -> new_dunn
  
  return(new_dunn)
}
```

**File paths**
```{r}
newExp_surv_fp <- '../../new_experiments/data/misc/survival_data.tsv'
az_surv_fp <- '../../stool/data/misc/aim1a_survival.csv'

## global renaming of axis/facet labels
diet_labels <- c('Chow',
                 'High Fat / High Fiber', 
                 'High Fat / Low Fiber',
                 'Low Fat / High Fiber',
                 'Low Fat / Low Fiber')
names(diet_labels) <- c('Chow',
                        'HF/HF',
                        'HF/LF',
                        'LF/HF',
                        'LF/LF')

experiment_labs <- c('New Anschutz\n(2024)',
                     'U of Arizona')
names(experiment_labs) <- c('new_exp_anschutz',
                            'second_set_arizona')

cult_x_labs <- c('Live\nCultures',
                 'Spores')

diet_y_labs <- c('HFt / HFb',
                 'HFt / LFb',
                 'LFt / HFb',
                 'LFt / LFb')
```

**Reading in files**
```{r}
newExp_surv_res <- read_tsv(newExp_surv_fp)

az_surv_res <- read_csv(az_surv_fp)
```

**Data wrangled**
so i can put the two survival tables together and do stats on them 
```{r}
## new experiments 
proc_newExp_surv_res <- newExp_surv_res %>% 
  filter(experiment != 'CDD01') %>% 
  mutate(inf_stock = paste('spores'),
         experiment_set = paste('new_exp_anschutz')) %>% 
  select(mouse_id, vendor, diet, day_post_inf, actual_surv, inf_stock, experiment_set) %>% 
  rename(status = actual_surv)

## az experiments
proc_az_surv_res <- az_surv_res %>% 
  mutate(vendor = paste('charles_river'),
         experiment_set = paste('second_set_arizona'),
         diet = case_when(
           diet == 'chow' ~ 'Chow',
           diet == 'LF/LF' ~ 'LF/LF',
           diet == 'LF+fiber' ~ 'LF/HF',
           diet == 'HF/LF' ~ 'HF/LF',
           diet == 'HF+fiber' ~ 'HF/HF'
         )) %>% 
  rename(day_post_inf = Day,
         mouse_id = group) %>% 
  select(-barrier)

## smushing them together via rbind 
## i dont think i actually need to combine these 
all_surv_res <- rbind(proc_newExp_surv_res,
                      proc_az_surv_res)
```

**Investigating Differences in Date of Mouse Death per Diet**
Putting together a summary table for how many days post CDI the mice started/stopped dying per diet and infection type.
```{r}
## new experiments
(newExp_death_range_df <- death_range_table(input_table = proc_newExp_surv_res,
                                           surv_col_name = 'status',
                                           wanted_group = 'diet',
                                           timeline_day_col = 'day_post_inf'))

## az experiments
(az_death_range_df <- death_range_table(input_table = proc_az_surv_res,
                                       surv_col_name = 'status',
                                       wanted_group = c('inf_stock', 'diet'),
                                       timeline_day_col = 'day_post_inf'))
```

**Stats**
Running a non-parametric ANOVA (Kruskal-Wallis followed by Dunn's Post Hoc) on the days the mice from different diets died on NOT binned by vendor (charles river vs taconic). 
```{r}
## new experiments
## pulling out the results from just the mice that died 
newExp_just_mice_died <- proc_newExp_surv_res %>% 
  filter(status == 1)

newExp_deathDate_stats <- apppleplots::kruskal_dunn_stats(input_table = newExp_just_mice_died,
                                                          grouped_by = 'inf_stock',
                                                          adjust_method = 'BH',
                                                          formula_left = 'day_post_inf',
                                                          formula_right = 'diet')

newExp_deathDate_kruskal <- newExp_deathDate_stats$KruskalTest
newExp_deathDate_dunn <- newExp_deathDate_stats$DunnTest %>% 
  select(-inf_stock)

(proc_newExp_deathDate_dunn <- edit_dunn_test_v2(input_table = newExp_just_mice_died,
                                                wanted_group = 'diet',
                                                mean_value = 'day_post_inf',
                                                dunn_test = newExp_deathDate_dunn))
```
same stats on arizona experiments but binned by c. diff culture method bc keith did two separate experiments 
```{r}
## arizona experiments
## pulling out the results from just the mice that died 
az_just_mice_died <- proc_az_surv_res %>% 
  filter(status == 1)

az_deathDate_stats <- apppleplots::kruskal_dunn_stats(input_table = az_just_mice_died,
                                                      grouped_by = 'inf_stock',
                                                      adjust_method = 'BH',
                                                      formula_left = 'day_post_inf',
                                                      formula_right = 'diet')

az_deathDate_kruskal <- az_deathDate_stats$KruskalTest
az_deathDate_dunn <- az_deathDate_stats$DunnTest

## this function is still acting funky with multiple groups (depends on the order you write them 
## in the character vector)
(proc_az_deathDate_dunn <- edit_dunn_test_v2(input_table = az_just_mice_died,
                                             wanted_group = c('diet', 'inf_stock'),
                                             mean_value = 'day_post_inf',
                                             dunn_test = az_deathDate_dunn))
```

putting new and arizona experiments together for overall stats
```{r}
## pulling out data from just the mice that died 
all_surv_just_mice_died <- all_surv_res %>% 
  filter(status == 1)

allSurv_deathDate_stats <- apppleplots::kruskal_dunn_stats(input_table = all_surv_just_mice_died,
                                                      grouped_by = c('inf_stock', 'experiment_set'),
                                                      adjust_method = 'BH',
                                                      formula_left = 'day_post_inf',
                                                      formula_right = 'diet')

allSurv_deathDate_kruskal <- allSurv_deathDate_stats$KruskalTest
allSurv_deathDate_dunn <- allSurv_deathDate_stats$DunnTest

## this function is still acting funky with multiple groups (depends on the order you write them 
## in the character vector)
(proc_allSurv_deathDate_dunn <- edit_dunn_test_v2(input_table = all_surv_just_mice_died,
                                             wanted_group = c('diet', 'inf_stock', 'experiment_set'),
                                             mean_value = 'day_post_inf',
                                             dunn_test = allSurv_deathDate_dunn))
```
**Plots**
not sure how I want to show this yet
```{r, fig.width=15, fig.height=7}
(deathDate_stat_plot <- apppleplots::stat_heat_plot(input_table = proc_allSurv_deathDate_dunn,
                            x_value = 'inf_stock',
                            y_value = 'group2',
                            x_value_type = 'discrete',
                            y_value_type = 'discrete',
                            pairwise = TRUE, 
                            tile_fill = 'stat_diff_means',
                            text_value = 'p.adj.signif',
                            legend_name = 'Group 1 -\nGroup 2',
                            x_labels = cult_x_labs,
                            y_labels = diet_y_labs,
                            facet_by = NULL,
                            facet_labels = NULL,
                            x_name = 'C. difficile Infection Method',
                            y_name = 'Group 2',
                            title_content = 'Onset of Mortality Post-CDI Comparisons') +
  facet_grid(rows = vars(experiment_set),
             cols = vars(group1),
             labeller = labeller(.rows = experiment_labs,
                                 .cols = diet_labels)) +
  theme_bw(base_size = 20) +
  theme(strip.text.y.right = element_text(angle = 0),
        plot.subtitle = element_text(hjust = 0.5)))
```

**Saving my Outputs**
```{r}
ggsave('../baseline_day3/plots/deathDate_stat_plot.pdf',
       plot = deathDate_stat_plot,
       width = 15,
       height = 7)

write_tsv(proc_allSurv_deathDate_dunn,
          file = '../baseline_day3/stats/allSurv_deathDate_dunn.tsv')
```

