---
title: "extra_relAbun_stats"
output: html_document
date: "2024-10-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(magrittr)
library(broom)
library(apppleplots)
library(rstatix)
library(qiime2R)
```

**Functions**
```{r}
family_abun_file_prep <- function(metadata_fp,
                                  tax_fp,
                                  otu_table_fp,
                                  tax_level,
                                  wanted_tax){
  ## metadata
  metadata <- read_tsv(metadata_fp)
  names(metadata)[names(metadata) == '#SampleID'] <- 'sampleid'
  ## taxonomy
  taxonomy <- read_qza(tax_fp)$data %>% 
    parse_taxonomy() %>% 
    rownames_to_column('asv')
  ## otu table 
  otu_table <- read_qza(otu_table_fp)$data
  otu_table %>% 
    as_tibble(rownames = 'asv') %>% 
    gather(-asv, key = sampleid, value = abun) %>% 
    group_by(sampleid) %>% 
    mutate(rel_abun = abun/sum(abun)) %>% 
    mutate(rel_abun = rel_abun + 0.000001) -> otu_table
  ## joining all tables together 
  otu_table %>% 
    left_join(metadata, by = 'sampleid') %>% 
    left_join(taxonomy, by = 'asv') -> abun_table
  abun_table %>% 
    group_by(sampleid, diet, mouse_id,
             experiment_set, vendor, mouse_sex, day_post_inf,
             .data[[tax_level]]) %>% 
    summarise(rel_abund = sum(rel_abun)) %>% 
    filter(.data[[tax_level]] %in% wanted_tax) %>% 
    mutate(mouse_fact = as.factor(mouse_id)) -> abun_filt
  ## creating a list for my outputs
  my_list <- list(Metadata = metadata,
                  Taxonomy = taxonomy,
                  OTUTable = otu_table,
                  AbundanceTable = abun_filt)
  return(my_list)
}

## preps dunns post hoc results for statistical visualization
stat_plot_prep <- function(filtered_table,
                           first_group,
                           second_group,
                           mean_value,
                           dunn_test){
  filtered_table %>% 
    group_by(filtered_table[first_group], filtered_table[second_group]) %>%
    summarise(mean = mean(.data[[mean_value]])) -> mean_table
  
  dunn_test %>% 
    merge(mean_table, 
          by.x = c('group1',
                   second_group),
          by.y = c(first_group,
                   second_group)) %>%
    rename_with(~paste0('group1_', mean_value, recycle0 = TRUE), contains('mean')) %>% 
    merge(mean_table,
          by.x = c('group2',
                   second_group),
          by.y = c(first_group,
                   second_group)) %>%
    rename_with(~paste0('group2_', mean_value, recycle0 = TRUE), contains('mean')) -> int_dunn
  
  group1_col <- paste0('group1_', mean_value)
  group2_col <- paste0('group2_', mean_value)
  
  int_dunn %>% 
    mutate(diff_means = (.data[[group1_col]] - .data[[group2_col]]),
           stat_diff_means = if_else(p.adj > 0.05, 0, diff_means)) -> new_dunn
  
  return(new_dunn)
}

## statistical visualization 
stat_plot <- function(new_dunn,
                      facet_by,
                      title,
                      wanted_subtitle){
  new_dunn %>% 
    ggplot(aes(x = group1, y = group2)) +
    geom_tile(aes(fill = stat_diff_means), alpha = 1, color = 'black') +
    scale_fill_gradient2(low = 'blue', high = 'green', name = 'Group 1 -\nGroup 2') +
    geom_text(aes(label = p.adj.signif)) +
    facet_wrap(~.data[[facet_by]],
               nrow = 2) +
    theme_bw(base_size = 20) +
    ggtitle(label = title,
            subtitle = wanted_subtitle) +
    theme(strip.text.y = element_text(angle = 0),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x = element_text(angle = 45, hjust = 1)) +
    xlab('Group 1') +
    ylab('Group 2') -> stat_vis
  return(stat_vis)
}
```

**File paths**
```{r}
## baseline microbiome only
baseline_metadata_FP <- '../data/misc/oldNew_comp_d15_metadata.tsv'
baseline_otu_table_FP <- '../data/baseline_qiime/total_sum_otu_table.qza'
baseline_tax_FP <- '../data/baseline_qiime/taxonomy.qza'

## baseline and day 3 microbiome
d3_metadata_FP <- '../data/misc/oldNew_comp_d15-d3_metadata.tsv'
d3_otu_table_FP <- '../data/baseline_day3_qiime/total_sum_otu_table.qza'
d3_tax_FP <- '../data/baseline_day3_qiime/taxonomy.qza'

## global lists
wanted_family <- c('Enterobacteriaceae', 'Lactobacillaceae', 'Lachnospiraceae', 'Enterococcaceae',
                   'Staphylococcaceae', 'Bacteroidaceae', 'Tannerellaceae', 'Morganellaceae')
wanted_genus <- c('Akkermansia', 'Enterococcus', 'Escherichia-Shigella', 'Proteus', 
                  'Bacteroides', 'Lactobacillus', 'Staphylococcus', 'Muribaculaceae')
```

## **Family level**
**Abundance table**
```{r}
## baseline microbiome 
baseline_family_abun_files <- family_abun_file_prep(metadata_fp = baseline_metadata_FP,
                                          tax_fp = baseline_tax_FP,
                                          otu_table_fp = baseline_otu_table_FP,
                                          tax_level = 'Family',
                                          wanted_tax = wanted_family)
baseline_family_abun_table <- baseline_family_abun_files$AbundanceTable

## adding a separate column the dataframe so I can do stats on it 
proc_baseline_familyAbun_table <- baseline_family_abun_table %>% 
  mutate(exp_vendor = paste(experiment_set, vendor, sep = ":")) %>% 
  ungroup()

## baseline and day 3 microbiome 
d3_family_abun_files <- family_abun_file_prep(metadata_fp = d3_metadata_FP,
                                          tax_fp = d3_tax_FP,
                                          otu_table_fp = d3_otu_table_FP,
                                          tax_level = 'Family',
                                          wanted_tax = wanted_family)
d3_family_abun_table <- d3_family_abun_files$AbundanceTable

proc_d3_familyAbun_table <- d3_family_abun_table %>% 
  mutate(exp_vendor = paste(experiment_set, vendor, sep = ":")) %>% 
  ungroup()
```

**Linear modeling**
```{r}
## baseline microbiome
baseline_family_relAbun_lm <- apppleplots::linear_model(input_table = proc_baseline_familyAbun_table,
                                               grouped_by = 'Family',
                                               adjust_method = 'BH',
                                               filter_adj_p_value = FALSE,
                                               formula_left = 'rel_abund',
                                               formula_right = 'exp_vendor')

## baseline and day 3 microbiome 
d3_family_relAbun_lm <- apppleplots::linear_model(input_table = proc_d3_familyAbun_table,
                                               grouped_by = c('Family', 'day_post_inf'),
                                               adjust_method = 'BH',
                                               filter_adj_p_value = FALSE,
                                               formula_left = 'rel_abund',
                                               formula_right = 'exp_vendor')
```

**Kruskal-Wallis and Dunn's Post Hoc test**
i think i want to use the results from the dunn's post hoc test for the stat visualization
```{r}
## baseline microbiome 
baseline_family_relAbun_krDunn <- apppleplots::kruskal_dunn_stats(input_table = proc_baseline_familyAbun_table,
                                                         grouped_by = 'Family',
                                                         adjust_method = 'BH',
                                                         filter_adj_p_value = FALSE,
                                                         formula_left = 'rel_abund',
                                                         formula_right = 'exp_vendor')

baseline_family_relAbun_kruskal <- baseline_family_relAbun_krDunn$KruskalTest
baseline_family_relAbun_dunn <- baseline_family_relAbun_krDunn$DunnTest

## baseline and day 3 microbiome 
d3_family_relAbun_krDunn <- apppleplots::kruskal_dunn_stats(input_table = proc_d3_familyAbun_table,
                                                         grouped_by = c('Family', 'day_post_inf'),
                                                         adjust_method = 'BH',
                                                         filter_adj_p_value = FALSE,
                                                         formula_left = 'rel_abund',
                                                         formula_right = 'exp_vendor')

d3_family_relAbun_kruskal <- d3_family_relAbun_krDunn$KruskalTest
d3_family_relAbun_dunn <- d3_family_relAbun_krDunn$DunnTest
```

**Wrangling Dunn test for stat visualization**
```{r}
## baseline microbiome
proc_baseline_family_relAbun_dunn <- stat_plot_prep(filtered_table = proc_baseline_familyAbun_table,
                                           first_group = 'exp_vendor',
                                           second_group = 'Family',
                                           mean_value = 'rel_abund',
                                           dunn_test = baseline_family_relAbun_dunn)

## baseline and day 3 microbiome 
proc_d3_family_relAbun_dunn <- stat_plot_prep(filtered_table = proc_d3_familyAbun_table,
                                           first_group = 'exp_vendor',
                                           second_group = c('Family', 'day_post_inf'),
                                           mean_value = 'rel_abund',
                                           dunn_test = d3_family_relAbun_dunn)
```

**Stat visualization**
```{r, fig.width=15, fig.height=9}
## baseline microbiome 
baseline_family_relAbun_stat_plot <- stat_plot(new_dunn = proc_baseline_family_relAbun_dunn,
                                      facet_by = 'Family',
                                      title = 'All Exp Microbe Relative Abundance at Baseline',
                                      wanted_subtitle = 'Microbe Family')

baseline_family_relAbun_stat_plot
```
```{r, fig.width=23, fig.height=9}
## baseline and day 3 microbiome 
d3_family_relAbun_stat_plot <- stat_plot(new_dunn = proc_d3_family_relAbun_dunn,
                                      facet_by = 'Family',
                                      title = 'All Exp Microbe Relative Abundance',
                                      wanted_subtitle = 'Microbe Family')

d3_family_relAbun_stat_plot <- d3_family_relAbun_stat_plot +
  facet_grid(day_post_inf~Family)

d3_family_relAbun_stat_plot
```


## **Genus level**
**Abundance table**
```{r}
## baseline microbiome
baseline_genus_abun_files <- family_abun_file_prep(metadata_fp = baseline_metadata_FP,
                                          tax_fp = baseline_tax_FP,
                                          otu_table_fp = baseline_otu_table_FP,
                                          tax_level = 'Genus',
                                          wanted_tax = wanted_genus)
baseline_genus_abun_table <- baseline_genus_abun_files$AbundanceTable

## adding a separate column the dataframe so I can do stats on it
proc_baseline_genusAbun_table <- baseline_genus_abun_table %>% 
  mutate(exp_vendor = paste(experiment_set, vendor, sep = ":")) %>% 
  ungroup()

## baseline and day 3 microbiome
d3_genus_abun_files <- family_abun_file_prep(metadata_fp = d3_metadata_FP,
                                          tax_fp = d3_tax_FP,
                                          otu_table_fp = d3_otu_table_FP,
                                          tax_level = 'Genus',
                                          wanted_tax = wanted_genus)
d3_genus_abun_table <- d3_genus_abun_files$AbundanceTable

proc_d3_genusAbun_table <- d3_genus_abun_table %>% 
  mutate(exp_vendor = paste(experiment_set, vendor, sep = ":")) %>% 
  ungroup()
```

**Linear modeling**
```{r}
## baseline microbiome
baseline_genus_relAbun_lm <- apppleplots::linear_model(input_table = proc_baseline_genusAbun_table,
                                              grouped_by = 'Genus',
                                              adjust_method = 'BH',
                                              filter_adj_p_value = FALSE,
                                              formula_left = 'rel_abund',
                                              formula_right = 'exp_vendor')

## baseline and day 3 microbiome
d3_genus_relAbun_lm <- apppleplots::linear_model(input_table = proc_d3_genusAbun_table,
                                              grouped_by = c('Genus', 'day_post_inf'),
                                              adjust_method = 'BH',
                                              filter_adj_p_value = FALSE,
                                              formula_left = 'rel_abund',
                                              formula_right = 'exp_vendor')
```

**Kruskal-Wallis and Dunn's Post Hoc test**
```{r}
## baseline microbiome
baseline_genus_relAbun_krDunn <- apppleplots::kruskal_dunn_stats(input_table = proc_baseline_genusAbun_table,
                                                                 grouped_by = 'Genus',
                                                                 adjust_method = 'BH',
                                                                 filter_adj_p_value = FALSE,
                                                                 formula_left = 'rel_abund',
                                                                 formula_right = 'exp_vendor')

baseline_genus_relAbun_kruskal <- baseline_genus_relAbun_krDunn$KruskalTest
baseline_genus_relAbun_dunn <- baseline_genus_relAbun_krDunn$DunnTest

## baseline and day 3 microbiome 
d3_genus_relAbun_krDunn <- apppleplots::kruskal_dunn_stats(input_table = proc_d3_genusAbun_table,
                                                           grouped_by = c('Genus', 'day_post_inf'),
                                                           adjust_method = 'BH',
                                                           filter_adj_p_value = FALSE,
                                                           formula_left = 'rel_abund',
                                                           formula_right = 'exp_vendor')

d3_genus_relAbun_kruskal <- d3_genus_relAbun_krDunn$KruskalTest
d3_genus_relAbun_dunn <- d3_genus_relAbun_krDunn$DunnTest
```

**Wrangling Dunn test for stat visualization**
```{r}
## baseline microbiome
proc_baseline_genus_relAbun_dunn <- stat_plot_prep(filtered_table = proc_baseline_genusAbun_table,
                                                   first_group = 'exp_vendor',
                                                   second_group = 'Genus',
                                                   mean_value = 'rel_abund',
                                                   dunn_test = baseline_genus_relAbun_dunn)

## baseline and day 3 microbiome
proc_d3_genus_relAbun_dunn <- stat_plot_prep(filtered_table = proc_d3_genusAbun_table,
                                             first_group = 'exp_vendor',
                                             second_group = c('Genus', 'day_post_inf'),
                                             mean_value = 'rel_abund',
                                             dunn_test = d3_genus_relAbun_dunn)
```

**Stat visualization**
```{r, fig.width=15, fig.height=9}
## baseline microbiome
baseline_genus_relAbun_stat_plot <- stat_plot(new_dunn = proc_baseline_genus_relAbun_dunn,
                                              facet_by = 'Genus',
                                              title = 'All Exp Microbe Relative Abundance at Baseline',
                                              wanted_subtitle = 'Microbe Genus')

baseline_genus_relAbun_stat_plot
```

```{r, fig.width=24.5, fig.height=9}
## baseline and day 3 microbiome
d3_genus_relAbun_stat_plot <- stat_plot(new_dunn = proc_d3_genus_relAbun_dunn,
                                        facet_by = 'Genus',
                                        title = 'All Exp Microbe Relative Abundance',
                                        wanted_subtitle = 'Microbe Genus')

d3_genus_relAbun_stat_plot <- d3_genus_relAbun_stat_plot +
  facet_grid(day_post_inf~Genus)

d3_genus_relAbun_stat_plot
```


## **Saving my outputs**
```{r}
## baseline microbiome only 
ggsave('../baseline/plots/famAbun_stat_vis.pdf',
       plot = baseline_family_relAbun_stat_plot,
       width = 15, 
       height = 9)
ggsave('../baseline/plots/genusAbun_stat_vis.pdf',
       plot = baseline_genus_relAbun_stat_plot,
       width = 15,
       height = 9)

## baseline and day 3 microbiome 
ggsave('../baseline_day3/plots/famAbun_stat_vis.pdf',
       plot = d3_family_relAbun_stat_plot,
       width = 23, 
       height = 9)
ggsave('../baseline_day3/plots/genusAbun_stat_vis.pdf',
       plot = d3_genus_relAbun_stat_plot,
       width = 24.5,
       height = 9)
```

