---
title: "hypoxia_comparisons"
output: html_document
date: "2023-12-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(broom)
library(cowplot)
library(magrittr)
library(qiime2R)
library(tidyverse)
library(naniar)
library(ggpubr)
library(rstatix)
library(viridis)
```

**Functions**
```{r}
file_prep <- function(metadata_fp,
                      hypoxia_fp,
                      metab_fp,
                      metab_col_filter,
                      metab_filter,
                      histo_fp,
                      toxin_fp){
  ## metadata file
  metadata <- read_tsv(metadata_fp) %>% 
              select(!c(tube_num, date, corr_sample_num))
  ## hypoxia file 
  hypoxia <- read_csv(hypoxia_fp)
  hypoxia %>% 
    mutate(order = ifelse(location == 'Cecum', 1, location),
           order = ifelse(order == 'Prox_colon', 2, order),
           order = ifelse(order == 'Dist_colon', 3, order)) %>% 
    arrange(order) -> hypoxia
  ## metabolomics file
  metab <- read_csv(metab_fp)
  metadata %>% 
    merge(metab, by = 'mouse_id') %>% 
    select(-(all_of(unwanted_columns))) %>% 
    gather(wanted_metabs, key = metabolite, value = concentration) %>% 
    filter(!is.na(mouse_id)) -> pre_metab
  ## changes all 'ND' values in the concentration column to 0 
  pre_metab$concentration[pre_metab$concentration == 'ND'] <- 0
  pre_metab %>% 
    filter(!is.na(concentration)) %>% 
    mutate(concentration = as.numeric(concentration)) -> big_metab
  ## histopathology file
  histo <- read_csv(histo_fp) %>% 
           filter(!is.na(mouse_id))
  metadata %>% 
    merge(histo, by = 'mouse_id') %>% 
    gather(cecum, colon, key = tissue, value = score) -> big_histo
  ## toxin file
  toxin <- read_tsv(toxin_fp)
  ## toxin neat concentrations (non_diluted)
  toxin %>% 
    merge(metadata, by = 'mouse_id') %>% 
    select(!c('Total TcA 1:10', 'Total TcB 1:10')) %>% 
    gather('Total TcA Neat', 'Total TcB Neat', 
           key = neat_toxin, value = neat_conc) -> pre_neat_toxin
  pre_neat_toxin$neat_conc[pre_neat_toxin$neat_conc == 'BDL'] <- '0'
  pre_neat_toxin %>% 
    mutate(neat_conc = as.numeric(neat_conc)) %>% 
    select(!c('Tube_Label',
              'Collection Date',
              'Sample_Type',
              'Extra_Sample')) -> neat_toxin
  ## toxin diluted concentrations 
  toxin %>% 
    merge(metadata, by = 'mouse_id') %>% 
    select(!c('Total TcA Neat', 'Total TcB Neat')) %>% 
    gather('Total TcA 1:10', 'Total TcB 1:10',
           key = dil_toxin, value = dil_conc) -> pre_dil_toxin
  pre_dil_toxin$dil_conc[pre_dil_toxin$dil_conc == 'BDL'] <- '0'
  pre_dil_toxin$dil_conc[pre_dil_toxin$dil_conc == 'Chow'] <- '0'
  pre_dil_toxin %>% 
    mutate(dil_conc = as.numeric(dil_conc)) %>% 
    filter(diet != 'Chow') %>% 
    select(!c('Tube_Label',
              'Collection Date',
              'Sample_Type',
              'Extra_Sample')) -> dil_toxin
  ## creating a list of my outputs
  my_list <- list(Metadata = metadata,
                  Hypoxia = hypoxia, 
                  Metabolomics = big_metab,
                  Histopathology = big_histo,
                  NeatToxin = neat_toxin,
                  DilToxin = dil_toxin)
  return(my_list)
}

## hypoxia plot comparison function
comp_plot <- function(hypoxia_table,
                       wanted_table,
                       ## this is what you want your x axis to be (neat, diluted, or concentration)? 
                       wanted_x,
                       ## this is toxin/metabolite column you want to filter by (neat or diluted)?
                       wanted_y,
                       ## what is the toxin/metabolite you want to filter by called in the table?
                       wanted_y_name,
                       diet_labels,
                       location_labels,
                       x_label,
                       title){
  ## putting tables together
  hypoxia_table %>% 
    left_join(wanted_table, by = 'mouse_id') %>% 
    na.omit() -> compiled_table
  ## constructing plot
  compiled_table %>% 
    filter(.data[[wanted_y]] == wanted_y_name) %>% 
    na.omit() %>% 
    ggplot(aes(x = .data[[wanted_x]], y = fluorescence)) +
    scale_x_continuous(trans = 'log10') +
    scale_y_continuous(trans = 'log10') +
    geom_jitter(alpha = 0.4, width = 0.1, height = 0) +
    geom_smooth(method = 'lm', se = FALSE) +
    facet_grid(order~diet,
               labeller = labeller(order = location_labels,
                                   diet = diet_labels)) +
    theme_bw() +
    theme(strip.text.y = element_text(angle = 0)) +
    xlab(x_label) +
    ylab('Fluorescence') +
    ggtitle(title) -> plot
  return(plot)
}
```

**Statisitical Analysis Functions**
```{r}
## runs linear modeling on the table, diet, location, and toxin/metabolite wanted and returns full results 
## along with a table containing the r2 and p-value 
## this goes into the following function with a for loop 
r2_value <- function(biom_table,
                     wanted_diet,
                     wanted_location,
                     tox_metab_col,
                     wanted_tox_metab,
                     wanted_conc_value){
  ## to get r squared value and p value
  biom_table %>% 
    filter(diet == wanted_diet,
           location == wanted_location,
           .data[[tox_metab_col]] == wanted_tox_metab) %>%
    do(glance(lm(fluorescence ~ .data[[wanted_conc_value]],
                 data = .))) -> lm_glance
  output <- list(r_squared = lm_glance$r.squared,
                 p_value = lm_glance$p.value)
  ## to get estimate and place it with the overall output table
  biom_table %>% 
    filter(diet == wanted_diet,
           location == wanted_location,
           .data[[tox_metab_col]] == wanted_tox_metab) %>%
    do(tidy(lm(fluorescence ~ .data[[wanted_conc_value]],
                 data = .))) %>% 
    filter(term != '(Intercept)') -> lm_full
  output <- as_tibble(output) %>%
            mutate(diet = paste(wanted_diet),
                   location = paste(wanted_location),
                   tox_metab = paste(wanted_tox_metab),
                   just_r = sqrt(r_squared),
                   estimate = lm_full$estimate,
                   r_direct = if_else(estimate < 0, just_r * -1, just_r))
  my_list <- list(LinearModel = lm_full,
                  WantedStats = output)
  return(my_list)
}

## creating a nested for loop since running this function individually is a nightmare
r2_value_for_loop <- function(biom_table,
                              wanted_col,
                              wanted_conc) {
  ## nested for loop for the r squared and p values 
  output <- tibble()
  for(i in unique(unlist(biom_table$diet))) {
    for(j in unique(unlist(biom_table$location))) {
      for(k in unique(unlist(biom_table[wanted_col]))) {
        tmp_output <- r2_value(biom_table,
                               i,
                               j,
                               wanted_col,
                               k,
                               wanted_conc)$WantedStats
        output <- bind_rows(output, tmp_output)
      }
    }
  }
  return(output)
}

## statistical visualization plot function
stat_plot <- function(stat_table,
                      x_lab_list,
                      y_lab_list,
                      facet_labels,
                      title){
  stat_table %>% 
    ggplot(aes(x = order, y = diet)) +
    geom_tile(aes(fill = r_direct), color = 'black', alpha = 0.6) +
    geom_text(aes(label = signif)) +
    scale_fill_gradient2(low = 'blue', high = 'green', name = "Correlation\nCoefficient") +
    scale_x_discrete(labels = x_lab_list) +
    scale_y_discrete(labels = y_lab_list) +
    facet_wrap(~tox_metab,
               labeller = labeller(tox_metab = facet_labels)) +
    theme_bw() +
    xlab("Location") +
    ylab("Diet") +
    ggtitle(title) -> plot
  return(plot)
  }
```

**File Paths**
```{r}
metadata_FP <- '../../cecum/data/misc/filt_cecal_processed_metadata.tsv'
histo_FP <- '../../colon/data/misc/histo_data.csv'
metab_FP <- '../../colon/data/misc/metabolomics.csv'
toxin_FP <- '../../colon/data/misc/toxin_final_data.tsv'
hypoxia_FP <- '../../colon/data/misc/pimid_fluor.csv'

## lists to redo the diet names on the facet labels of the ggplot created below 
diet_labs <- 
    c('Chow', 
      'High Fat / High Fiber', 
      'High Fat / Low Fiber', 
      'Low Fat / High Fiber', 
      'Low Fat / Low Fiber')

names(diet_labs) <- c('Chow', 'HF/HF', 'HF/LF', 'LF/HF', 'LF/LF')

tissue_labs <- c('Cecum',
                 'Colon')
names(tissue_labs) <- c('cecum',
                        'colon')
diet_y_labs <- c('Chow',
                  'HFt/HFb',
                  'HFt/LFb',
                  'LFt/HFb',
                  'LFt/LFb')

wanted_metabs <- c('Acetic Acid (ug/g)',
                   'Propanoic Acid (ug/g)',
                   'n-Butanoic Acid (ug/g)')
unwanted_columns <- c('2-methyl-propanoic acid (ug/g)',
                     'Isopentanoic Acid (ug/g)',
                     '2-methyl-Butanoic Acid (ug/g)',
                     'Pentanoic Acid (ug/g)',
                     'Notes',
                     'Sample Group',
                     'SCFA Data File',
                     'Acq. Date-Time',
                     'Tube_Label',
                     'Sample_Type',
                     'Collection Date',
                     'Dil.')
metab_labs <- c('Acetic Acid',
                'Propanoic Acid',
                'n-Butanoic Acid')
names(metab_labs) <- wanted_metabs

neat_labs <- c('TcdA', 'TcdB')
names(neat_labs) <- c('Total TcA Neat', 'Total TcB Neat')

dil_labs <- c('TcdA', 'TcdB')
names(dil_labs) <- c('Total TcA 1:10', 'Total TcB 1:10')

location_labs <- c('Cecum',
                   'Proximal Colon',
                   'Distal Colon')
names(location_labs) <- c(1,
                          2,
                          3)
```

**Data Wrangling**
```{r}
hypoxia_files <- file_prep(metadata_FP,
                           hypoxia_FP,
                           metab_FP,
                           unwanted_columns,
                           wanted_metabs,
                           histo_FP,
                           toxin_FP)

metadata <- hypoxia_files$Metadata
hypoxia <- hypoxia_files$Hypoxia
metab <- hypoxia_files$Metabolomics
metab %>% 
  select(metabolite, concentration, mouse_id) -> mini_metab

dil_toxin <- hypoxia_files$DilToxin
neat_toxin <- hypoxia_files$NeatToxin
neat_toxin %>% 
  select(neat_toxin, neat_conc, mouse_id) -> mini_neat_toxin

histo <- hypoxia_files$Histopathology
histo %>% 
  select(mouse_id, tissue, score) -> mini_histo
```


**Hypoxia and Histopathology Score Comparisons**

**Plot**
```{r, warning=FALSE, fig.width=10, fig.height=5}
hypox_cecum_plot <- comp_plot(hypoxia,
                              mini_histo,
                              'score',
                              'tissue',
                              'cecum',
                              diet_labs,
                              location_labs,
                              'Histopathology Score',
                              'Hypoxia and Cecal Inflammation')

hypox_colon_plot <- comp_plot(hypoxia,
                              mini_histo,
                              'score',
                              'tissue',
                              'colon',
                              diet_labs,
                              location_labs,
                              'Histopathology Score',
                              'Hypoxia and Colon Inflammation')


hypox_cecum_plot
hypox_colon_plot
```

**Stats**
```{r}
hypox_histo_table <- hypoxia %>% 
  left_join(mini_histo, by = 'mouse_id') %>% 
  na.omit()

hypox_histo_stat_table <- r2_value_for_loop(hypox_histo_table,
                                            "tissue",
                                            "score")

hypox_histo_stat_table['signif'] <- symnum(hypox_histo_stat_table$p_value,
                                           cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
                                           symbols = c("****", "***", "**", "*", "+", "ns"),
                                           abbr.colnames = FALSE,
                                           na = "")

hypox_histo_stat_table %>% 
  mutate(order = ifelse(location == 'Cecum', 1, location),
           order = ifelse(order == 'Prox_colon', 2, order),
           order = ifelse(order == 'Dist_colon', 3, order)) %>% 
  arrange(order) -> hypox_histo_stat_table


hypox_histo_stat_plot <- stat_plot(hypox_histo_stat_table,
                                   location_labs,
                                   histo_y_labs,
                                   tissue_labs,
                                   'Hypoxia and Histopathology Score Comparisons')

hypox_histo_stat_plot
```

**Everything Together**
```{r, warning=FALSE, fig.height=7, fig.width=17}
hypoxia_histo_plots <- plot_grid(hypox_cecum_plot,
                                 hypox_colon_plot,
                                 hypox_histo_stat_plot,
                                 nrow = 2,
                                 labels = c('a)', 'b)', 'c)'),
                                 label_size = 18,
                                 align = c('hv'),
                                 axis = c('tblr'),
                                 rel_heights = c(1, 0.6))

hypoxia_histo_plots
```


**Hypoxia and Neat C. difficile Toxin Concentration Comparisons**

**Plots**
```{r, warning=FALSE, fig.width=10, fig.height=5}
hypox_tcda_plot <- comp_plot(hypoxia,
                             mini_neat_toxin,
                             'neat_conc',
                             'neat_toxin',
                             'Total TcA Neat',
                             diet_labs,
                             location_labs,
                             'Concentration',
                             'Hypoxia Compared to TcdA Toxin')

hypox_tcdb_plot <- comp_plot(hypoxia,
                             mini_neat_toxin,
                             'neat_conc',
                             'neat_toxin',
                             'Total TcB Neat',
                             diet_labs,
                             location_labs,
                             'Concentration',
                             'Hypoxia Compared to TcdB Toxin')

hypox_tcda_plot
hypox_tcdb_plot
```

**Stats**
I think there are too few points for the high-fat/high-fiber diet so there isn't a calculated p-value for that diet. I turned the NAs into 1 so the result would show as non-significant on the plot. 
```{r}
hypox_toxin_table <- hypoxia %>% 
  left_join(mini_neat_toxin, by = 'mouse_id') %>% 
  na.omit()

hypox_toxin_stat_table <- r2_value_for_loop(hypox_toxin_table,
                                            "neat_toxin",
                                            "neat_conc")
hypox_toxin_stat_table %>% 
  mutate(p_value = ifelse(p_value == 'NaN', 1, p_value)) -> hypox_toxin_stat_table

hypox_toxin_stat_table['signif'] <- symnum(hypox_toxin_stat_table$p_value,
                                           cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
                                           symbols = c("****", "***", "**", "*", "+", "ns"),
                                           abbr.colnames = FALSE,
                                           na = "")

hypox_toxin_stat_table %>% 
  mutate(order = ifelse(location == 'Cecum', 1, location),
           order = ifelse(order == 'Prox_colon', 2, order),
           order = ifelse(order == 'Dist_colon', 3, order)) %>% 
  arrange(order) -> hypox_toxin_stat_table


hypox_toxin_stat_plot <- stat_plot(hypox_toxin_stat_table,
                                   location_labs,
                                   diet_y_labs,
                                   neat_labs,
                                   'Hypoxia and Toxin Comparisons')

hypox_toxin_stat_plot
```

**Everything Together**
```{r, warning=FALSE, fig.height=7, fig.width=17}
hypoxia_toxin_plots <- plot_grid(hypox_tcda_plot,
                                 hypox_tcdb_plot,
                                 hypox_toxin_stat_plot,
                                 nrow = 2,
                                 labels = c('a)', 'b)', 'c)'),
                                 label_size = 18,
                                 align = c('hv'),
                                 axis = c('tblr'),
                                 rel_heights = c(1, 0.6))

hypoxia_toxin_plots
```


**Hypoxia and Metabolite Concentration Comparisons**

**Plots**
```{r, warning=FALSE, fig.width=10, fig.height=5}
hypox_acetic_plot <- comp_plot(hypoxia,
                               mini_metab,
                               'concentration',
                               'metabolite',
                               'Acetic Acid (ug/g)',
                               diet_labs,
                               location_labs,
                               'Concentration',
                               'Hypoxia Compared to Acetic Acid Concentration')

hypox_butanoic_plot <- comp_plot(hypoxia,
                                 mini_metab,
                                 'concentration',
                                 'metabolite',
                                 'n-Butanoic Acid (ug/g)',
                                 diet_labs,
                                 location_labs,
                                 'Concentration',
                                 'Hypoxia Compared to n-Butanoic Acid Concentration')

hypox_propanoic_plot <- comp_plot(hypoxia,
                                   mini_metab,
                                   'concentration',
                                   'metabolite',
                                   'Propanoic Acid (ug/g)',
                                   diet_labs,
                                   location_labs,
                                   'Concentration',
                                   'Hypoxia Compared to Propanoic Acid Concentration')

hypox_acetic_plot
hypox_butanoic_plot
hypox_propanoic_plot
```

**Stats**
```{r, warning=FALSE, fig.height=4, fig.width=10}
hypox_metab_table <- hypoxia %>% 
  left_join(mini_metab, by = 'mouse_id') %>% 
  na.omit()

hypox_metab_stat_table <- r2_value_for_loop(hypox_metab_table,
                                            "metabolite",
                                            "concentration")
hypox_metab_stat_table %>% 
  mutate(p_value = ifelse(p_value == 'NaN', 1, p_value)) -> hypox_metab_stat_table

hypox_metab_stat_table['signif'] <- symnum(hypox_metab_stat_table$p_value,
                                           cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
                                           symbols = c("****", "***", "**", "*", "+", "ns"),
                                           abbr.colnames = FALSE,
                                           na = "")

hypox_metab_stat_table %>% 
  mutate(order = ifelse(location == 'Cecum', 1, location),
           order = ifelse(order == 'Prox_colon', 2, order),
           order = ifelse(order == 'Dist_colon', 3, order)) %>% 
  arrange(order) -> hypox_metab_stat_table


hypox_metab_stat_plot <- stat_plot(hypox_metab_stat_table,
                                   location_labs,
                                   diet_y_labs,
                                   metab_labs,
                                   'Hypoxia and Metabolite Comparisons')

hypox_metab_stat_plot
```

**Everything Together**
```{r, warning=FALSE, fig.height=10, fig.width=20}
hypoxia_metab_plots <- plot_grid(hypox_acetic_plot,
                                 hypox_butanoic_plot,
                                 hypox_propanoic_plot,
                                 hypox_metab_stat_plot,
                                 nrow = 2,
                                 labels = c('a)', 'b)', 'c)', 'd)'),
                                 label_size = 18,
                                 align = c('hv'),
                                 axis = c('tblr'))

hypoxia_metab_plots
```


**Saving my Outputs**
```{r}
ggsave('../plots/hypoxia_histo_comp_plots.pdf',
       plot = hypoxia_histo_plots,
       width = 17, 
       height = 7)
ggsave('../plots/hypoxia_toxin_comp_plots.pdf',
       plot = hypoxia_toxin_plots,
       width = 17, 
       height = 7)
ggsave('../plots/hypoxia_metab_comp_plots.pdf',
       plot = hypoxia_metab_plots,
       width = 20, 
       height = 10)
```

