---
title: "combined_cfu_plot"
output: html_document
date: "2024-11-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggpubr)
library(ggplot2)
library(qiime2R)
library(tidyverse)
library(ggh4x)
library(viridis)
library(ggrepel)
library(cowplot)
library(grid)
library(gridExtra)
library(apppleplots)
```

**Functions**
```{r}
## editing my dunn's post hoc test to include the difference in means between groups 
edit_dunn_test <- function(filtered_table,
                           first_group,
                           second_group,
                           mean_value,
                           dunn_test){
  filtered_table %>% 
    group_by(.data[[first_group]], .data[[second_group]]) %>% 
    summarise(mean = mean(.data[[mean_value]])) -> mean_table
  
  dunn_test %>% 
    merge(mean_table, 
          by.x = c('group1',
                   second_group),
          by.y = c(first_group,
                   second_group)) %>% 
    rename_with(~paste0('group1_', mean_value, recycle0 = TRUE), contains('mean')) %>% 
    merge(mean_table,
          by.x = c('group2',
                   second_group),
          by.y = c(first_group,
                   second_group)) %>% 
    rename_with(~paste0('group2_', mean_value, recycle0 = TRUE), contains('mean')) -> int_dunn
  
  group1_col <- paste0('group1_', mean_value)
  group2_col <- paste0('group2_', mean_value)
  
  int_dunn %>% 
    mutate(diff_means = (.data[[group1_col]] - .data[[group2_col]]),
           stat_diff_means = if_else(p.adj > 0.05, 0, diff_means)) -> new_dunn
  
  return(new_dunn)
}
```

**File Paths**
```{r}
cult_genusAbun_fp <- '../data/misc/culture_genusAbun_table.tsv'
cfu_count_fp <- '../data/misc/proc_combined_CDD_cfuCounts.tsv'

location_y_labs <- c('Blood',
                     'Liver',
                     'Spleen')
diet_fill_labs <- c('Chow',
                    'HFt/HFb',
                    'HFt/LFb',
                    'LFt/HFb',
                    'LFt/LFb')
short_diet_fill_labs <- c('Chow',
                          'LFt/LFb')
presAbs_labs <- c('Absent',
                  'Present')
surv_labs <- c("Didn't Survive",
               'Survived')
surv_presAbs_title <- 'Survival and\nMicrobe Status'
surv_presAbs_labs <- c("Didn't Survive, Absent",
                       "Didn't Survive, Present",
                       "Survived, Absent",
                       "Survived, Present")
diet_labs <- c('Chow',
               'High Fat / High Fiber',
               'High Fat / Low Fiber',
               'Low Fat / High Fiber',
               'Low Fat / Low Fiber')
names(diet_labs) <- c('Chow',
                      'HF/HF',
                      'HF/LF',
                      'LF/HF',
                      'LF/LF')

diet_order_labs <- c('High Fat / Low Fiber',
                     'Low Fat / Low Fiber',
                     'High Fat / High Fiber',
                     'Low Fat / High Fiber',
                     'Chow')
names(diet_order_labs) <- c(1,
                            2, 
                            3,
                            4,
                            5)

vendor_labs <- c('Charles River',
                 'Taconic')
names(vendor_labs) <- c('charles_river',
                        'taconic')
```

**Reading in Files**
```{r}
cult_genusAbun_table <- read_tsv(cult_genusAbun_fp)
proc_cfu_count <- read_tsv(cfu_count_fp)
```

**Wrangling the taxonomy of positive cultures**
info on culture results:
- samples taken from 40 different mice
- 64 samples total from different locations 
```{r}
## pulling out mice/locations that have culture results that include enterococcus, escherichia-shigella,
## and proteus 
cult_genusAbun_table %>% 
  select(location, mouse_id, Genus, rel_abund) %>% 
  filter(Genus == 'Enterococcus' | Genus == 'Escherichia-Shigella' | Genus == 'Proteus',
         rel_abund < 0.9) %>% 
  spread(key = 'Genus', value = 'rel_abund', fill = 0) %>% 
  mutate(total_abun = (Enterococcus + `Escherichia-Shigella` + Proteus)) %>% 
  filter(total_abun >= 0.8) %>% 
  mutate(Genus = paste("enterococcus", "escherichiaShigella", "proteus", sep = "_"),
         sanity = paste(mouse_id, location, sep = "_")) %>% 
  rename(rel_abund = total_abun) %>% 
  select(mouse_id, location, Genus, sanity, rel_abund) -> big3_polymicro_table

## pulling out mice/locations with culture results that are 90% or above one microbe
cult_genusAbun_table %>%
    select(mouse_id, Genus, rel_abund, location) %>%
    group_by(mouse_id, Genus) %>%
    filter(rel_abund > 0.9) %>% 
  mutate(sanity = paste(mouse_id, location, sep = "_")) -> top_tax_table

append(top_tax_table$sanity,
      big3_polymicro_table$sanity) -> sanity_list

## pulling out the last mice/locations that are not included in either of the tables above 
cult_genusAbun_table %>% 
  group_by(mouse_id, location) %>% 
  filter(rel_abund == max(rel_abund)) %>% 
  mutate(sanity = paste(mouse_id, location, sep = "_")) %>% 
  filter(!(sanity %in% sanity_list)) %>% 
  select(mouse_id, Genus, rel_abund, location, sanity) -> missing_mice

## putting all tables together and more data wrangling 
proc_cult_genusAbun <- rbind(top_tax_table,
                             big3_polymicro_table,
                             missing_mice)

proc_cult_genusAbun <- proc_cult_genusAbun %>% 
  mutate(top_tax = ifelse(is.na(Genus), 'Unknown', Genus))
```

**Combining all tables together for the plot**
```{r}
## added staph to the other list bc there's only one sample with it 
other_list <- c('Unknown',
                'Micrococcus',
                'Oceanobacillus',
                'Dietzia',
                'Gemella',
                'Corynebacterium',
                'Bacillus',
                'Streptococcus',
                'Staphylococcus',
                'Cutibacterium',
                'Kocuria')


comb_cfu_genusAbun <- proc_cfu_count %>% 
  left_join(proc_cult_genusAbun, by = c("mouse_id", "location")) %>% 
  filter(experiment != 'CDD01')  %>% 
  mutate(top_tax = ifelse(is.na(top_tax), 'None', top_tax),
         top_tax = ifelse(top_tax %in% other_list, 'Other', top_tax),
         diet_order = case_when(
           diet == 'HF/LF' ~ 1,
           diet == 'LF/LF' ~ 2,
           diet == 'HF/HF' ~ 3,
           diet == 'LF/HF' ~ 4,
           diet == 'Chow' ~ 5
         ))
```

**Putting together color palettes**
color palette?
took the hexcodes from scale_color_brewers 'Paired' palette and put gray in the middle for the "None" category. 
```{r}
##RColorBrewer::brewer.pal(n = 8, name = 'Spectral')
my_spectral_palette <- c("#D53E4F",
                         "#F46D43",
                         "#FDAE61",
                         "#FEE08B",
                         "#E6F598",
                         "#ABDDA4",
                         "#66C2A5",
                         "#3288BD",
                         "#999999")

my_paired_palette <- c("#A6CEE3", 
                       "#1F78B4",
                       "#B2DF8A",
                       "#33A02C",
                       # "#FB9A99",
                       # "#E31A1C",
                       # "#FDBF6F",
                       # "#FF7F00",
                       "#CAB2D6",
                       "#6A3D9A",
                       "#999999")


## get hexcodes of viridis color palette option C
# viridis_pal(option = 'C')(6)

my_viridis_c_palette <- c("#0D0887FF",
                          "#6A00A8FF",
                          "#B12A90FF",
                          "#E16462FF",
                          "#FCA636FF",
                          "#F0F921FF",
                          "#999999")
```

**Combined CFU/Survival Plot**
```{r, fig.width=22, fig.height=12}
comb_cfu_genusAbun %>% 
  ggplot(aes(x = mouse_id, y = location)) +
  theme_bw(base_size = 20) +
  theme(axis.text.x = element_blank()) +
  geom_point(aes(color = top_tax, shape = status_microbePres, alpha = status_microbePres), size = 6, 
             stroke = 2) +
  scale_y_discrete(labels = location_y_labs) +
  facet_nested_wrap(~diet_order + vendor,
                    scales = "free_x",
                    labeller = labeller(diet_order = diet_order_labs,
                                        vendor = vendor_labs),
                    nrow = 3) +
  scale_color_manual(breaks = c('Enterococcus',
                                'Escherichia-Shigella',
                                'Proteus',
                                'enterococcus_escherichiaShigella_proteus',
                                'Lactobacillus',
                                'Other',
                                'None'),
                     labels = c('Enterococcus',
                                'Escherichia-Shigella',
                                'Proteus',
                                'Enterococcus, Escherichia-Shigella,\nand Proteus',
                                'Lactobacillus',
                                'Other',
                                'None'),
                     values = my_viridis_c_palette,
                     name = 'Genus') +
  scale_shape_manual(values = c(4, 4, 16, 16),
                     name = surv_presAbs_title,
                     labels = surv_presAbs_labs) +
  scale_alpha_manual(values = c(0.15, 1, 0.15, 1),
                     name = surv_presAbs_title,
                     labels = surv_presAbs_labs) +
  labs(x = 'Mouse',
       y = 'Location',
       title = 'New Experiment Culture and Survival Results') -> everything_plot

everything_plot
```

**Combined CFU/Survival Stats**
status column:
- survived = 1
- didnt_survive = 0
status_microbePres column:
- didnt_survive,0 = 1
- didnt_survive,1 = 2
- survived,0 = 3
- survived,1 = 4
```{r}
num_cfu_table <- proc_cfu_count %>% 
  mutate(status = ifelse(status == 'survived', 1, 0),
         status_microbePres = case_when(
           status_microbePres == 'didnt_survive,0' ~ 1,
           status_microbePres == 'didnt_survive,1' ~ 2,
           status_microbePres == 'survived,0' ~ 3,
           status_microbePres == 'survived,1' ~ 4
         ))

num_cfu_table
```

```{r}
cfu_stats <- apppleplots::kruskal_dunn_stats(input_table = num_cfu_table,
                                              grouped_by = c('vendor'),
                                              adjust_method = 'BH',
                                              filter_adj_p_value = FALSE,
                                              formula_left = 'microbe_presence',
                                              formula_right = 'diet')

cfu_kruskal <- cfu_stats$KruskalTest
cfu_dunn <- cfu_stats$DunnTest

new_cfu_dunn <- edit_dunn_test(filtered_table = num_cfu_table,
                               first_group = 'diet',
                               second_group = 'vendor',
                               mean_value = 'microbe_presence',
                               dunn_test = cfu_dunn)
```

**Stat Plot**
```{r, fig.height=5, fig.width=10}
diet_stat_x_labs <- c('Chow',
                      'HFt/\nHFb',
                      'HFt/\nLFb',
                      'LFt/\nHFb')
diet_stat_y_labs <- c('HFt / HFb',
                      'HFt / LFb',
                      'LFt / HFb',
                      'LFt / LFb')

everything_stats <- apppleplots::stat_heat_plot(input_table = new_cfu_dunn,
                                                x_value = 'group1',
                                                y_value = 'group2',
                                                x_value_type = 'discrete',
                                                y_value_type = 'discrete',
                                                pairwise = TRUE,
                                                correlation = FALSE,
                                                tile_fill = 'stat_diff_means',
                                                text_value = 'p.adj.signif',
                                                legend_name = 'Group 1 -\nGroup2',
                                                x_labels = diet_stat_x_labs,
                                                y_labels = diet_stat_y_labs,
                                                facet_by = 'vendor',
                                                facet_labels = vendor_labs,
                                                x_name = 'Group 1',
                                                y_name = 'Group 2',
                                                title_content = 'New Experiment Culture Comparisons')

everything_stats <- everything_stats +
  labs(subtitle = 'Vendor') +
  theme_bw(base_size = 20) +
  theme(plot.subtitle = element_text(hjust = 0.5))

everything_stats
```

**To Save for Later!**
```{r}
proc_cfuCount_topTax <- proc_cfu_count %>% 
  left_join(proc_cult_genusAbun, by = c("mouse_id", "location")) %>% 
  filter(experiment != 'CDD01')  %>% 
  mutate(top_tax = ifelse(is.na(top_tax), 'None', top_tax)) %>% 
  select(!c('Genus', 'sanity')) %>%
  rename(cult_rel_abund = rel_abund) %>% 
  mutate(cult_rel_abund = ifelse(is.na(cult_rel_abund), 0, cult_rel_abund))

proc_cfuCount_topTax  
```

**Saving my Outputs**
```{r}
## as a pdf
ggsave('../plots/allExp_culture_withMicrobes.pdf',
       plot = everything_plot,
       width = 22,
       height = 12)
ggsave('../plots/allExp_culture_withMicrobes_stats.pdf',
       plot = everything_stats,
       width = 10,
       height = 5)

## as an .rdat object
save(everything_plot,
     file = '../../figures/allExp_culture_withMicrobes.rdat')
save(everything_stats,
     file = '../../figures/allExp_culture_withMicrobes_stats.rdat')


write_tsv(proc_cfuCount_topTax,
          '../data/misc/proc_tax_CDD_cfuCounts.tsv')
```

