---
title: "kidney_histology"
output: html_document
date: "2025-01-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggplot2)
library(magrittr)
library(apppleplots)
```

**File paths**
```{r}
raw_kidney_histo_fp <- '../data/misc/histologySubtotalsUnblinded_wKidney.csv'


## global variables for renaming plot stuff
kidney_labs <- c('Cortex',
                 'Medulla',
                 'Other',
                 'Renal Corpus\nInjury',
                 'Total Injury\nScore')
names(kidney_labs) <- c('Cortex',
                        'Medulla',
                        'Other',
                        'RenalCorpusInjury',
                        'TotInjuryScore')

vendor_x_labs <- c('Charles\nRiver',
                   'Taconic')
diet_labs <- c('Chow',
               'HFt / HFb',
               'HFt / LFb',
               'LFt / HFb',
               'LFt / LFb')
```

**Reading in file**
```{r}
raw_kidney_histo <- read_csv(raw_kidney_histo_fp)
```

**Data wrangled**
only pulling the kidney histo scores and necessary metadata out (does not have starting/current weights, temp, or other histology measures)
```{r}
kidney_histo_only <- raw_kidney_histo %>% 
  select(SampleID, MouseID, BlindingID, Diet, Vendor, ExperimentID, 
         RenalCorpusInjury, Cortex, Medulla, Other, TotInjuryScore) %>% 
  na.omit() %>% 
  gather(c('RenalCorpusInjury', 'Cortex', 'Medulla', 'Other', 'TotInjuryScore'), 
         key = 'kidney_histo', value = 'score')

kidney_histo_only
```

**Stats**
- can't do stats between diets bc I don't think all diets are represented evenly across kidney histo scoring categories
- there is no statistical difference in kidney histo scores between vendor for any categories 
```{r}
kidney_histo_krDunn <- apppleplots::kruskal_dunn_stats(input_table = kidney_histo_only,
                                                       grouped_by = c('kidney_histo', 'ExperimentID'),
                                                       adjust_method = 'BH',
                                                       filter_adj_p_value = FALSE,
                                                       formula_left = 'score',
                                                       formula_right = 'Vendor')

kidney_histo_kruskal <- kidney_histo_krDunn$KruskalTest
kidney_histo_dunn <- kidney_histo_krDunn$DunnTest

kidney_histo_dunn
```

**Plot**
i feel like i just put the plot together opposite of how elena did it and while there may be some small differences in kidney histo between vendors, since they're not statistically significant, I'm not sure there's anything here 
```{r, fig.width=15, fig.height=5}
kidney_histo_plot <- kidney_histo_only %>% 
  ggplot(aes(x = Vendor, y = score)) +
  geom_boxplot(aes(group = Vendor), outlier.shape = NA) +
  geom_jitter(aes(fill = Diet), pch = 21, width = 0.1, height = 0, alpha = 0.8, size = 3.5) +
  scale_fill_brewer(palette = 'Spectral',
                    labels = diet_labs) +
  scale_x_discrete(labels = vendor_x_labs) +
  theme_bw(base_size = 20) +
  facet_grid(~kidney_histo,
             labeller = labeller(kidney_histo = kidney_labs)) +
  labs(x = 'Vendor',
       y = 'Score',
       title = 'New Experiment Kidney Histology Scores')

kidney_histo_plot
```
faceting the above plot by batch/experiment to see if there are any differences between them 
```{r, fig.width=15, fig.height=8}
(kidneyHisto_byExp_plot <- kidney_histo_plot +
  facet_grid(ExperimentID~kidney_histo,
             labeller = labeller(kidney_histo = kidney_labs)) +
  theme(strip.text.y.right = element_text(angle = 0)) +
  ggtitle('Kidney Histology Scores by Experiment'))

# kidneyHisto_byExp_plot
```
**Saving my outputs**
```{r}
ggsave('../plots/kidney_histo_plot.pdf',
       plot = kidney_histo_plot,
       width = 15,
       height = 5)
ggsave('../plots/kidneyHisto_byExp_plot.pdf',
       plot = kidneyHisto_byExp_plot,
       width = 15,
       height = 8)
```


