---
title: "mouse_weight"
output: html_document
date: "2025-01-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(apppleplots)
library(magrittr)
library(readr)
library(viridis)
library(survival)
library(survminer)
```
**Functions**
```{r}
## weight loss hazard stats via coxph 
## p-values corrected via benjamini and hochberg
calculate_weight_loss_stats <- function(input_table){
  ## putting together weight loss table for stats
  weight_loss_df <- input_table %>% 
    ## any percent weight change value less than 0 is 1
    mutate(weight_loss = ifelse(perc_change_weight < 0, 1, 0),
           ## creating levels for every 3 days after infection 
           time_levels = as.factor(case_when(
             between(day_post_inf, 0, 3) ~ 'days 0 to 3',
             between(day_post_inf, 4, 6) ~ 'days 4 to 6',
             between(day_post_inf, 7, 9) ~ 'days 7 to 9',
             between(day_post_inf, 10, 12) ~ 'days 10 to 12',
             between(day_post_inf, 13, 15) ~ 'days 13 to 15',
           ))) %>% 
    select(sampleid, day_post_inf, diet, time_levels, weight_loss) 
  
  ## running coxph stratified by diet and time level
  coxph_out <- tibble()
  
  for (baseline_diet in unique(unlist(weight_loss_df$diet))){
    for (time_period in unique(unlist(weight_loss_df$time_levels))){
        
        ## filtering the df to only include one time period 
        filt_df <- weight_loss_df %>% 
          filter(time_levels == time_period)
        
        surv_obj <- Surv(time = filt_df$day_post_inf, event = filt_df$weight_loss)
        filt_df$diet <- as.factor(filt_df$diet)
        ## releveling diets so that different baseline diet to compare against each time 
        ## or else it defaults to chow as the baseline every time 
        filt_df$diet <- relevel(filt_df$diet, baseline_diet)
        
        fit.cox <- coxph(surv_obj ~ diet, data = filt_df)
        
        coxph_res <- summary(fit.cox)
        
        ## pulling the hazard ratio from coxph summary 
        haz_df <- as.data.frame(coxph_res$coefficients)
        haz_df["diet"] <- rownames(haz_df)
        
        ## making the output table pretty 
        haz_table <- as_tibble(haz_df) %>% 
          rename(haz_ratio = `exp(coef)`,
                 p_value = `Pr(>|z|)`) %>% 
          mutate(p_adj = p.adjust(p_value,
                                  method = 'BH'),
                 signif = as.character(symnum(p_adj,
                                       cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1),
                                       symbols = c("****", "***", "**", "*", "ns"),
                                       abbr.colnames = FALSE,
                                       na = "")),
                 time_frame = paste(time_period),
                 baseline = paste(baseline_diet),
                 haz_ratio = round(haz_ratio, 3),
                 p_adj = round(p_adj, 3),
                 ## will tell you if hazard of weight loss is greater or less than baseline diet
                 weightLoss_haz_toBaseline = ifelse(haz_ratio >= 1, 'greater', 'less'),
                 diet = gsub(diet, pattern = "diet", replacement = "")) %>% 
          select(baseline, diet, time_frame, everything())
        
        coxph_out <- rbind(coxph_out, haz_table)
    
    }
  }
  return(coxph_out)
}
```

**File paths**
```{r}
weight_fp <- '../data/misc/newExp_mouse_weightData.csv'

## global renaming variables
diet_legend_labs <- c('Chow',
                      'HFt / HFb',
                      'HFt / LFb',
                      'LFt / HFb',
                      'LFt / LFb')

vendor_labs <- c('Charles River',
                 'Taconic')
names(vendor_labs) <- c('CharlesRiver',
                        'Taconic')
```

**Reading in files**
```{r}
pre_weight_df <- read_csv(weight_fp) %>% 
  select(SampleID, MouseSex, Diet, Vendor, 
         ExperimentID, ExperimentDay, BaseWeight, DayWeight, 
         PercChange, Temperature)
```

**Data wrangled**
only use days 0-13 since we have sampling problems at day 14 for one of the experiments
```{r}
proc_weight_df <- pre_weight_df %>%
   rename(sampleid = SampleID,
          mouse_sex = MouseSex,
          diet = Diet,
          vendor = Vendor,
          experiment = ExperimentID,
          experiment_day = ExperimentDay,
          base_weight = BaseWeight,
          day_weight = DayWeight,
          perc_change_weight = PercChange,
          temp = Temperature) %>% 
   mutate(day_post_inf = (experiment_day - 15),
          mouse_id = gsub(sampleid, pattern = ".[0-9]{2}$", replacement = "")) %>% 
   filter(experiment != 'CDD01') %>% 
   na.omit()

## taking the average percent change in weights per day and diet for plotting
avWeight_per_dayDiet <- proc_weight_df %>% 
   group_by(diet, day_post_inf, vendor) %>% 
   summarise(av_perc_change = mean(perc_change_weight),
             sd = sd(perc_change_weight)) %>% 
   filter(day_post_inf <= 13)
```

**Plot**
```{r, fig.width=12, fig.height=10}
(newExp_avWeight_plot <- avWeight_per_dayDiet %>% 
  ggplot(aes(x = day_post_inf, y = av_perc_change, group = diet)) +
  geom_line(aes(color = diet), linewidth = 2) +
  geom_point(aes(color = diet), size = 3.5, alpha = 0.5) +
  geom_pointrange(aes(ymin = av_perc_change-sd, ymax = av_perc_change+sd, color = diet), 
                  alpha = 0.5) +
  scale_color_viridis(option = 'C',
                      discrete = TRUE,
                      name = 'Diet',
                      labels = diet_legend_labs) +
  scale_x_continuous(breaks = c(0, 3, 6, 9, 12)) +
  theme_bw(base_size = 20) +
  labs(x = 'Days Relative to Infection',
       y = 'Average % Weight Change',
       title = 'Average Weight Change After Infection by Diet') +
  expand_limits(y = c(-20, 20)) +
  facet_wrap(~vendor,
             nrow=2,
             labeller = labeller(vendor = vendor_labs)))
```

**Stats**
- stats are being done on the full percent weight change data (per mouse) and not the average percent weight change per day, diet, and vendor ! (i.e. what's plotted)

method: creating binary column (0/1) for whether there was weight loss (1) at a given time point or not (0), then calculating the hazard ratio per diet/day based on that

coxph hazard ratio for loop
- FDR p-value correction?? - yes!, FDR via benjamini and hochberg 
- account for vendor in this!! - really hard to do since all LF/LF mice in charles river died after day 7...
- hazard ratio is really large for diets compared to LF/LF days 10-12 and days 13-15 bc all the LF/LF mice in charles river died before then (non-significant though)
```{r}
(newExp_coxph_out <- calculate_weight_loss_stats(input_table = proc_weight_df))
```

**Saving my Outputs**
```{r}
ggsave('../plots/avWeight_change_plot.pdf',
       plot = newExp_avWeight_plot,
       width = 12,
       height = 10)

write_tsv(newExp_coxph_out,
          '../stats/newExp_weightLoss_hazard_coxph_res.tsv')

## as an .rdat object to put in r package 
save(newExp_avWeight_plot,
     file = '../../figures/newExp_avWeight_change.rdat')
```

