---
title: "mouse_survival"
format: html
editor: visual
---

```{r, include=FALSE}
library(tidyverse)
library(survminer)
require(survival)
library(readr)
library(viridis)
library(cowplot)
library(apppleplots)
```

**Input File Path**

```{r}
survival_fp <- '../data/misc/aim1a_survival.csv'
diet_labs <- 
  c('Chow', 
    'High Fat / High Fiber', 
    'High Fat / Low Fiber', 
    'Low Fat / High Fiber', 
    'Low Fat / Low Fiber')
names(diet_labs) <- c('chow',
                      'HF+fiber',
                      'HF/LF',
                      'LF+fiber',
                      'LF/LF')

fib_labs <- c('High Fiber',
              'Low Fiber')
names(fib_labs) <- c('high_fiber',
                     'low_fiber')

fat_labs <- c('High Fat',
              'Low Fat')
names(fat_labs) <- c('high_fat',
                     'low_fat')
```

**Reading in Mouse Survival Data**

col_types tells R what to classify the data in each column as

```{r}
survival <- read_csv(survival_fp,
                     col_types = cols(Day = col_integer(),
                                      status = col_integer(), 
                                      diet = col_character(),
                                      group = col_integer()))

spore_survival <- survival %>% 
  filter(inf_stock == 'spores')

liveCult_survival <- survival %>% 
  filter(inf_stock == 'culture')
```

## Spore Infection Survival Data

**Creating Survival Plot by Diet**

Legit can't figure out how to take color by diet off of this plot so rip.

```{r, fig.height=10, fig.width=15}
diet_fit <- surv_fit(Surv(Day, status) ~ diet, data = spore_survival)
ggsurvplot(diet_fit, 
           data = spore_survival,
           color = 'black',
           ggtheme = theme_bw(base_size = 20),
           legend = "none",
           conf.int = TRUE, 
           xlab = 'Days Post Infection',
           ylab = 'Survival Probability',
           title = 'Mouse Survival After Infection by Diet') -> diet_plot

diet_plot$plot +
    facet_wrap(~factor(diet, levels = c('chow',
                                        'HF+fiber',
                                        'HF/LF',
                                        'LF+fiber',
                                        'LF/LF')),
               labeller = labeller(.cols = diet_labs),
               nrow = 2) +
    scale_fill_grey() -> diet_plot_final

diet_plot_final
```

**Pairwise Comparisons**

```{r}
pairwise_survdiff(Surv(Day, status) ~ diet,
                  data = spore_survival, 
                  p.adjust.method = "BH") -> diet_pairwise

diet_pvalue <- as_tibble(diet_pairwise$p.value,
                         rownames = 'diet')

## changing format from wide to long so I can work with it 
diet_pvalue %>% 
  gather(-diet, key = diet2, value = p.value) -> diet_pvalue

names(diet_pvalue)[names(diet_pvalue) == 'diet'] <- 'diet1'


diet_pvalue['signif'] <- symnum(diet_pvalue$p.value,
                                cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
                                symbols = c("****", "***", "**", "*", "+", "ns"),
                                abbr.colnames = FALSE,
                                na = "")

```

**Pairwise Comparison Visualization**

```{r, fig.height=3.5, fig.width=6.5}
diet_pvalue %>%
  na.omit() %>% 
  ggplot(aes(x = diet2, y = diet1)) +
  geom_tile(aes(fill = p.value), alpha = 0.6, color = 'black') +
  geom_text(aes(label = signif), size = 5) +
  scale_fill_viridis(option = 'C', limits = c(0, 0.05), direction = 1,  name = 'p-value') +
  # scale_fill_distiller(limits = c(0, 0.05), palette = 'Reds', name = "P-Value") +
  theme_bw(base_size = 20) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 18),
        legend.position = 'right') +
  scale_x_discrete(labels = c("Chow",
                              "HFt/\nLFb",
                              "HFt/\nHFb",
                              "LFt/\nLFb")) +
  scale_y_discrete(labels = c("HFt / LFb",
                               "HFt / HFb",
                               "LFt / LFb",
                               "LFt / HFb")) -> surv_stat_vis

surv_stat_vis
```

**Cox PH for Diet**

this calculates the hazard ratios for each diet compared to chow

could assign each diet a numeric key and run that way so that there isn't a reference diet but idk if that will replace pairwise comparisons in general

if you call *summary(coxph_object)*, the *exp(coef)* column contains the actual hazard ratio for the variable and the *Pr(\>\|z\|)* column contains the p-value

```{r}
spore_surv_obj <- Surv(time = spore_survival$Day, event = spore_survival$status)
spore_survival$diet_f = factor(spore_survival$diet, 
                               levels = c("chow",
                                          "HF+fiber",
                                          "HF/LF", 
                                          "LF+fiber",
                                          "LF/LF"))
diet.fit.cox.spore <- coxph(spore_surv_obj ~ diet_f, data = spore_survival)

spore_summary_coxph <- summary(diet.fit.cox.spore)
```

```{r}
## pulling the hazard ratio from coxph summary 
spore_haz_df <- as.data.frame(spore_summary_coxph$coefficients)
spore_haz_df["diet"] <- rownames(spore_haz_df)

spore_haz_table <- as_tibble(spore_haz_df) %>% 
  mutate(diet = case_when(
    diet == 'diet_fHF+fiber' ~ 'HF/HF',
    diet == 'diet_fHF/LF' ~ 'HF/LF',
    diet == 'diet_fLF+fiber' ~ 'LF/HF',
    diet == 'diet_fLF/LF' ~ 'LF/LF'
  )) %>% 
  rename(haz_ratio = `exp(coef)`,
         p_value = `Pr(>|z|)`)

spore_haz_table
```

```{r}
## pulling confidence intervals from the coxph summary
spore_ci_df <- as.data.frame(spore_summary_coxph$conf.int)
spore_ci_df["diet"] <- rownames(spore_ci_df)

spore_ci_table <- as_tibble(spore_ci_df) %>% 
  mutate(diet = case_when(
    diet == 'diet_fHF+fiber' ~ 'HF/HF',
    diet == 'diet_fHF/LF' ~ 'HF/LF',
    diet == 'diet_fLF+fiber' ~ 'LF/HF',
    diet == 'diet_fLF/LF' ~ 'LF/LF'
  )) %>% 
  rename(lower_95_ci = `lower .95`,
         upper_95_ci = `upper .95`) %>% 
  select(diet, lower_95_ci, upper_95_ci)

## combining the hazard ratio and confidence interval tables 
final_spore_haz_table <- spore_haz_table %>% 
  left_join(spore_ci_table, by = 'diet') %>% 
  select(diet, coef, haz_ratio, p_value, lower_95_ci, upper_95_ci) %>% 
  mutate(infection_type = paste('spores'))

final_spore_haz_table
```

forest plot for the hazard ratio because it's nice to look at

```{r}
ggforest(diet.fit.cox.spore)
```

**Putting Survival Plot by Diet with Statistical Visualization**

```{r, fig.width=20, fig.height=8}
plot_grid(diet_plot_final, surv_stat_vis,
          ncol = 2, 
          labels = c('a)', 'b)'),
          label_size = 22,
          rel_widths = c(1, 0.5))
```

## Live Culture Survival Data

**Survival Plot**

```{r, fig.height=7, fig.width=10}
liveCult_diet_fit <- surv_fit(Surv(Day, status) ~ diet, data = liveCult_survival)
ggsurvplot(liveCult_diet_fit, 
           data = liveCult_survival,
           color = 'black',
           ggtheme = theme_bw(base_size = 20),
           legend = "none",
           conf.int = TRUE, 
           xlab = 'Days Post Infection',
           ylab = 'Survival Probability',
           title = 'Mouse Survival After Infection by Diet (Live Cultures)') -> liveCult_diet_plot

liveCult_diet_plot$plot +
    facet_wrap(~factor(diet, levels = c('chow',
                                        'HF/LF',
                                        'LF/LF')),
               labeller = labeller(.cols = diet_labs),
               nrow = 2) +
    scale_fill_grey() -> liveCult_diet_plot_final

liveCult_diet_plot_final
```

**Pairwise Comparisons**

```{r}
pairwise_survdiff(Surv(Day, status) ~ diet,
                  data = liveCult_survival, 
                  p.adjust.method = "BH") -> liveCult_diet_pairwise

liveCult_diet_pvalue <- as_tibble(liveCult_diet_pairwise$p.value,
                         rownames = 'diet')

## changing format from wide to long so I can work with it 
liveCult_diet_pvalue %>% 
  gather(-diet, key = diet2, value = p.value) -> liveCult_diet_pvalue

names(liveCult_diet_pvalue)[names(liveCult_diet_pvalue) == 'diet'] <- 'diet1'


liveCult_diet_pvalue['signif'] <- symnum(liveCult_diet_pvalue$p.value,
                                        cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
                                        symbols = c("****", "***", "**", "*", "+", "ns"),
                                        abbr.colnames = FALSE,
                                        na = "")
```

**Pairwise Comparison Visualization**

```{r, fig.height=3.5, fig.width=5.5}
liveCult_diet_pvalue %>%
  na.omit() %>% 
  ggplot(aes(x = diet2, y = diet1)) +
  geom_tile(aes(fill = p.value), alpha = 0.6, color = 'black') +
  geom_text(aes(label = signif), size = 5) +
  scale_fill_viridis(option = 'C', limits = c(0, 0.05), direction = 1, name = 'p-value') +
  ##scale_fill_distiller(limits = c(0, 0.05), palette = 'Reds', name = "P-Value") +
  theme_bw(base_size = 20) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 18),
        legend.position = 'right') +
  scale_x_discrete(labels = c("Chow",
                              "HFt/\nLFb")) +
  scale_y_discrete(labels = c("HFt / LFb",
                               "LFt / LFb")) -> liveCult_surv_stat_vis

liveCult_surv_stat_vis
```

**Cox PH for Diet**

this calculates the hazard ratios for each diet compared to chow

```{r}
liveCult_surv_obj <- Surv(time = liveCult_survival$Day, event = liveCult_survival$status)
liveCult_survival$diet_f = factor(liveCult_survival$diet, 
                                  levels = c("chow",
                                             "HF/LF",
                                             "LF/LF"))
diet.fit.cox.liveCult <- coxph(liveCult_surv_obj ~ diet_f, data = liveCult_survival)

liveCult_summary_coxph <- summary(diet.fit.cox.liveCult)
```

```{r}
## pulling the hazard ratio from coxph summary 
liveCult_haz_df <- as.data.frame(liveCult_summary_coxph$coefficients)
liveCult_haz_df["diet"] <- rownames(liveCult_haz_df)

liveCult_haz_table <- as_tibble(liveCult_haz_df) %>% 
  mutate(diet = case_when(
    diet == 'diet_fHF/LF' ~ 'HF/LF',
    diet == 'diet_fLF/LF' ~ 'LF/LF'
  )) %>% 
  rename(haz_ratio = `exp(coef)`,
         p_value = `Pr(>|z|)`)

liveCult_haz_table
```

```{r}
## pulling confidence intervals from the coxph summary
liveCult_ci_df <- as.data.frame(liveCult_summary_coxph$conf.int)
liveCult_ci_df["diet"] <- rownames(liveCult_ci_df)

liveCult_ci_table <- as_tibble(liveCult_ci_df) %>% 
  mutate(diet = case_when(
    diet == 'diet_fHF/LF' ~ 'HF/LF',
    diet == 'diet_fLF/LF' ~ 'LF/LF'
  )) %>% 
  rename(lower_95_ci = `lower .95`,
         upper_95_ci = `upper .95`) %>% 
  select(diet, lower_95_ci, upper_95_ci)

## combining the hazard ratio and confidence interval tables 
final_liveCult_haz_table <- liveCult_haz_table %>% 
  left_join(liveCult_ci_table, by = 'diet') %>% 
  select(diet, coef, haz_ratio, p_value, lower_95_ci, upper_95_ci) %>% 
  mutate(infection_type = paste('live_culture'))

final_liveCult_haz_table
```

```{r}
ggforest(diet.fit.cox.liveCult)
```

**Putting Survival Plot by Diet with Statistical Visualization**

also combining this with the weight data plot for the supplemental figure

```{r, fig.height=11, fig.width=22}
liveCult_bottom_row <- plot_grid(liveCult_surv_stat_vis, NULL,
                                 rel_widths = c(1, 0.8),
                                 labels = c('c)', ' '),
                                 label_size = 20)

plot_grid(liveCult_diet_plot_final +
            theme_bw(base_size = 20), 
          az_liveCult_weights_plot +
            labs(x = 'Days Post Infection',
                 y = 'Average % Weight Change',
                 title = 'Weight Change After Infection by Diet'),
          liveCult_bottom_row,
          rel_heights = c(1, 0.4),
          nrow = 2,
          labels = c('a)', 'b) ', ' '),
          align = 'hv',
          label_size = 20) -> liveCult_surv_res

liveCult_surv_res
```

## Combining the Hazard Ratio Tables

for the spore and live culture c. diff infections

```{r}
combined_haz_ratio_table <- rbind(final_spore_haz_table,
                                  final_liveCult_haz_table)
combined_haz_ratio_table
```

## Saving my Outputs

```{r}
## as a pdf
ggsave('../plots/survival_curve_all.pdf',
       plot = diet_plot_final,
       width = 15,
       height = 10)
ggsave('../plots/survival_curve_stats.pdf',
       plot = surv_stat_vis,
       width = 4.5,
       height = 3)
ggsave('../plots/liveCult_survival.pdf',
       plot = liveCult_surv_res,
       width = 22,
       height = 11)
## saving supplemental figure with all other figures 
ggsave('../../../figures/figS6.pdf',
       plot = liveCult_surv_res,
       width = 22,
       height = 11)

## as an rdata object
save(diet_plot_final,
     file = '../../../figures/survival.rdata')
save(surv_stat_vis,
     file = '../../../figures/survival_stats.rdata')

## saving the hazard ratio table results 
write_tsv(combined_haz_ratio_table,
          '../stats/survival_hazardRatio_results.tsv')
```
