---
title: "survCohort_mouse_weights"
output: html_document
date: "2025-05-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(apppleplots)
library(magrittr)
library(readr)
library(viridis)
```
**Functions**
```{r}
## calculates percent change between the first value in the timeline and each subsequent one
## then calculates the average percent change (based on timeline day and another variable) and the 
## standard deviation
## outputs the processed table 
calculate_percent_change <- function(input_table,
                                     ## id per subject (i.e. mouse_id)
                                     ident_col,
                                     ## which column denotes the timeline day?
                                     time_col,
                                     ## the column you actually want the percent change calculated for
                                     calc_wanted_col,
                                     ## any other variables you want the average and sd calc grouped by 
                                     ## (besides timeline day)
                                     other_var_col){
  proc_table <- input_table %>% 
    group_by(.data[[ident_col]]) %>% 
    arrange(.data[[time_col]], .by_group = TRUE) %>% 
    mutate(perc_change = ((.data[[calc_wanted_col]] - 
                             first(.data[[calc_wanted_col]]))/first(.data[[calc_wanted_col]])) * 100) %>% 
    group_by(.data[[time_col]], .data[[other_var_col]]) %>% 
    mutate(av_perc_change = mean(perc_change),
           sd_perc_change = sd(perc_change)) %>% 
    ungroup()
  
  return(proc_table)
}

## puts together ggplot for average percent change over time of some variable 
## to be used on output table from function above 
make_weights_plot <- function(input_table,
                              x_axis,
                              y_axis,
                              color_by,
                              sd_col,
                              viridis_pal_option,
                              legend_name,
                              legend_labels,
                              x_name,
                              y_name,
                              wanted_title){
  plot <- input_table %>% 
    ggplot(aes(x = .data[[x_axis]], y = .data[[y_axis]], group = .data[[color_by]])) +
    geom_line(aes(color = .data[[color_by]]), linewidth = 2) +
    geom_point(aes(color = .data[[color_by]]), size = 3.5, alpha = 0.5) +
    geom_pointrange(aes(ymin = .data[[y_axis]]-.data[[sd_col]], ymax = .data[[y_axis]]+.data[[sd_col]], 
                        color = .data[[color_by]]),
                    alpha = 0.5) +
    scale_color_viridis(option = viridis_pal_option,
                        discrete = TRUE,
                        name = legend_name,
                        labels = legend_labels) +
    theme_bw(base_size = 20) +
    labs(x = x_name,
         y = y_name,
         title = wanted_title)
  
  return(plot)
}
```

**File paths**
```{r}
spore_weights_fp <- '../data/misc/survivalCohort_weight_data.csv'
liveCult_weights_fp <- '../data/misc/liveCult_weight_data.csv'

## global renaming variables
diet_legend_labs <- c('Chow',
                      'HFt / HFb',
                      'HFt / LFb',
                      'LFt / HFb',
                      'LFt / LFb')

liveCult_diet_labs <- c('Chow',
                        'HFt / LFb',
                        'LFt / LFb')
```

**Reading in files**
idk if i can do an average percent weight change calculation since i dont have a base weight for these mice
```{r}
spore_weights_table <- read_csv(spore_weights_fp) %>% 
  na.omit() %>% 
  mutate(day_post_inf = (day - 15)) %>% 
  filter(day_post_inf >= 0)


liveCult_weights_table <- read_csv(liveCult_weights_fp) %>% 
  na.omit() %>% 
  mutate(day_post_inf = (day - 15)) %>% 
  filter(day_post_inf >= 0)
```

**Data wrangled**
calculating the percent weight change from day 0 instead of day -15 to match the amc follow up experiments weight change plot 
```{r}
weight_dfs_list <- list(spore = spore_weights_table,
                        liveCult = liveCult_weights_table)

for (weight_table in unique(unlist(names(weight_dfs_list)))){
  var_name <- paste0("proc_", names(weight_dfs_list[weight_table]), "_weights_table")
  
  int_table <- calculate_percent_change(input_table = weight_dfs_list[[weight_table]],
                                        ident_col = 'mouse_id',
                                        time_col = 'day_post_inf',
                                        calc_wanted_col = 'weight',
                                        other_var_col = 'diet')
  assign(var_name,
         int_table)
}
```

**Plots**
spore infections (survival cohort in manuscript)
```{r, fig.width=14, fig.height=7}
(az_spore_weights_plot <- make_weights_plot(input_table = proc_spore_weights_table,
                                        x_axis = 'day_post_inf',
                                        y_axis = 'av_perc_change',
                                        color_by = 'diet',
                                        sd_col = 'sd_perc_change',
                                        viridis_pal_option = 'C',
                                        legend_name = 'Diet',
                                        legend_labels = diet_legend_labs,
                                        x_name = 'Days Post Infection',
                                        y_name = 'Average % Weight Change',
                                        wanted_title = 'Average Weight Change After Infection by Diet') +
  scale_x_continuous(breaks = c(0, 3, 6, 9, 12)))
```

live culture infections (in supplemental)
```{r, fig.width=14, fig.height=7}
(az_liveCult_weights_plot <- make_weights_plot(input_table = proc_liveCult_weights_table,
                                        x_axis = 'day_post_inf',
                                        y_axis = 'av_perc_change',
                                        color_by = 'diet',
                                        sd_col = 'sd_perc_change',
                                        viridis_pal_option = 'C',
                                        legend_name = 'Diet',
                                        legend_labels = liveCult_diet_labs,
                                        x_name = 'Days Post Infection',
                                        y_name = 'Average % Weight Change',
                                        wanted_title = 'Average Weight Change After Infection by Diet') +
  scale_x_continuous(breaks = c(0, 3, 6, 9, 12)) +
  expand_limits(y = c(-20, 20)))
```

**Stats**
spore infections 
```{r}
## the zeroes at day 0 were causing problems in dunn test (creating NAs) so im filtering them out
noZero_spore_weights_table <- proc_spore_weights_table %>% 
  filter(day_post_inf != 0)

spore_avWeight_stats <- apppleplots::kruskal_dunn_stats(input_table = noZero_spore_weights_table,
                                                        grouped_by = c('day_post_inf'),
                                                        adjust_method = 'BH',
                                                        filter_adj_p_value = FALSE,
                                                        formula_left = 'av_perc_change',
                                                        formula_right = 'diet')

spore_avWeight_kruskal <- spore_avWeight_stats$KruskalTest
(spore_avWeight_dunn <- spore_avWeight_stats$DunnTest %>% 
    ## only including significant results bc hundreds of rows if not
    filter(p.adj.signif != 'ns'))
```

live culture infections 
```{r}
## the zeroes at day 0 were causing problems in dunn test (creating NAs) so im filtering them out
noZero_liveCult_weights_table <- proc_liveCult_weights_table %>% 
  filter(day_post_inf != 0)

liveCult_avWeight_stats <- apppleplots::kruskal_dunn_stats(input_table = noZero_liveCult_weights_table,
                                                        grouped_by = c('day_post_inf'),
                                                        adjust_method = 'BH',
                                                        filter_adj_p_value = FALSE,
                                                        formula_left = 'av_perc_change',
                                                        formula_right = 'diet')

liveCult_avWeight_kruskal <- liveCult_avWeight_stats$KruskalTest
(liveCult_avWeight_dunn <- liveCult_avWeight_stats$DunnTest %>% 
    ## only including significant results bc hundreds of rows if not
    filter(p.adj.signif != 'ns'))
```

**Saving my outputs**
```{r}
ggsave('../plots/spore_avWeight_change_plot.pdf',
       plot = az_spore_weights_plot,
       width = 14,
       height = 7)
ggsave('../plots/liveCult_avWeight_change_plot.pdf',
       plot = az_liveCult_weights_plot,
       width = 14,
       height = 7)

## stats
write_tsv(spore_avWeight_dunn,
          '../stats/spore_avWeight_change_dunnTest.tsv')
write_tsv(liveCult_avWeight_dunn,
          '../stats/liveCult_avWeight_change_dunnTest.tsv')

## as an .rdat object
save(az_spore_weights_plot,
     file = "../../../figures/az_spore_avWeight_change.rdat")
save(az_liveCult_weights_plot,
     file = "../../../figures/az_liveCult_avWeight_change.rdat")
```

