---
title: "bile_acid"
output: html_document
date: "2023-12-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(rstatix)
library(ggplot2)
library(ggpubr)
library(ggh4x)
library(cowplot)
library(magrittr)
library(glue)
```

**Functions**
```{r}
## data wrangling
bile_file_prep <- function(bile_acid_fp,
                           inhibitors,
                           promoters){
  ## reading in bile acid file
  bile_acid <- read_tsv(bile_acid_fp)
  ## putting together inhibitor table
  bile_acid %>% 
    select(diet, tube_numb, tube_label, date, 
           mouse_id, facility, study, contains(unlist(inhibitors))) %>% 
    gather(contains('acid'), key = bile_acid, value = concentration) %>% 
    mutate(c_diff_effect = paste('inhibitor')) %>% 
    filter(concentration != '#VALUE!') -> inhibit_bile
  ## putting together promoter table
  bile_acid %>% 
    select(diet, tube_numb, tube_label, date, 
           mouse_id, facility, study, contains(unlist(promoters))) %>% 
    gather(contains('acid'), key = bile_acid, value = concentration) %>% 
    mutate(c_diff_effect = paste('promoter')) -> promote_bile
  ## putting tables back together
  big_bile_acid <- rbind(inhibit_bile,
                         promote_bile)
  ## changing all undetectable values to 0 
  big_bile_acid$concentration[big_bile_acid$concentration == '<LOD' | big_bile_acid$concentration == '<LOQ'] <- 0
  ## adding 2 to all values so that all points will be on the plot when transformed to log10
  big_bile_acid %>% 
    na.omit() %>% 
    mutate(concentration = as.numeric(concentration),
           conc_normalized = concentration + 2) %>% 
    filter(study != 1) -> big_bile_acid
  ## creating list of outputs
  my_list <- list(NonProcBile = bile_acid,
                  ProcBile = big_bile_acid)
  return(my_list)
}

## statistical analysis function
bile_stats <- function(big_bile_table,
                       group_cols,
                       conc_value){
  ## kruskal test
  big_bile_table %>% 
    group_by(big_bile_table[unlist(group_cols)]) %>% 
    do(tidy(kruskal.test(.data[[conc_value]] ~ diet,
               data = .))) %>% 
    ungroup() %>% 
    arrange(p.value) %>% 
    mutate(p.adj = p.adjust(p.value,
                            method = "BH")) -> kruskal_test
  ## dunns post hoc test
  diet_name <- 'diet'
  test <- reformulate(glue("{diet_name}"),glue("{conc_value}"))
  
  big_bile_table %>% 
    group_by(big_bile_table[unlist(group_cols)]) %>% 
    dunn_test(test,
              p.adjust.method = 'BH',
              data = .) %>% 
    add_y_position(scales = 'free',
                   fun = 'max',
                   y.trans = log10) -> dunn_test
  # ## creating a list of outputs
  my_list <- list(KruskalTest = kruskal_test,
                  DunnPostHoc = dunn_test)
  return(my_list)
}

## creating plots
bile_plots <- function(big_bile_table,
                       x_labs,
                       non_bile_dunn,
                       bile_dunn){
  ## plot not faceted by bile acid
  big_bile_table %>% 
    ggplot(aes(x = diet, y = conc_normalized)) +
      geom_boxplot(aes(group = diet), outlier.shape = NA) +
      geom_jitter(alpha = 0.4, width = 0.1, height = 0) +
      scale_x_discrete(labels = x_labs) +
      theme_bw() +
      facet_wrap(~c_diff_effect,
                 labeller = labeller(c_diff_effect = cdiff_effect_labs)) +
      scale_y_log10() +
      stat_pvalue_manual(non_bile_dunn,
                         tip.length = 0.01,
                         step.increase = 0.05,
                         label = 'p.adj.signif',
                         hide.ns = TRUE) +
      labs(x = 'Diet',
           y = 'log10(Concentration (pmol/mg))',
           title = 'Cecal Bile Acid Concentration') -> overall_bile_plot
  ## plot faceted by bile acid
  big_bile_acid %>% 
    ggplot(aes(x = diet, y = conc_normalized)) +
      geom_boxplot(aes(group = diet), outlier.shape = NA) +
      geom_jitter(alpha = 0.4, width = 0.1, height = 0) +
      scale_x_discrete(labels = x_labs) +
      facet_nested_wrap(~c_diff_effect + bile_acid,
                        labeller = labeller(c_diff_effect = cdiff_effect_labs,
                                            bile_acid = bile_labs),
                        nrow = 3) +
      theme_bw() +
      scale_y_log10() +
      stat_pvalue_manual(dunn_bile,
                         tip.length = 0.01,
                         step.increase = 0.05,
                         label = 'p.adj.signif',
                         hide.ns = TRUE) +
    labs(x = 'Diet',
         y = 'log10(Concentration (pmol/mg))',
         title = 'Cecal Bile Acid Concentration') -> bile_include_plot
  ## creating list of outputs
  my_list <- list(NoFacetBilePlot = overall_bile_plot,
                  FacetBilePlot = bile_include_plot)
  return(my_list)
}
```

**File Paths**
```{r}
bile_acid_FP <- '../data/misc/corrected_bile_acid.tsv'

inhibitor_list <- c('_a-MCA',
                    '_b-MCA',
                    '_LCA',
                    '_DCA')
                    # '_CDCA',
                    # '_UDCA')

promoter_list <- c('_T-CA',
                   '_CA')

cdiff_effect_labs <- c('C. difficile Inhibitor',
                       'C. difficile Promoter')

names(cdiff_effect_labs) <- c('inhibitor',
                              'promoter')

bile_labs <- c('a-MCA',
               'b-MCA',
               'LCA',
               'DCA',
               # 'CDCA',
               # 'UDCA',
               'TCA',
               'CA')
  
names(bile_labs) <- c("Alpha_muricholic_acid_a-MCA",
                     "Beta_muricholic_acid_b-MCA",
                     "Lithocholic_acid_LCA", 
                     "Deoxycholic_acid_DCA", 
                     # "Chenodeoxycholic_acid_CDCA",
                     # "Ursodeoxycholic_acid_UDCA", 
                     "Taurocholic_acid_T-CA",
                     "Cholic_acid_CA")

bile_groups <- c('c_diff_effect',
                 'bile_acid')
non_bile_groups <- c('c_diff_effect')

diet_x_labs <- c('Chow',
                 'HFt/\nHFb',
                 'HFt/\nLFb',
                 'LFt/\nHFb',
                 'LFt/\nLFb')
```

Use prescence/abscence tests or rank-based test instead of a non-parametric statistical test!!
- Look at whether you'd like to include additional bile acids in the promoters/inhibitors categories
- join with metadata and filter out replicate one 

**Bile Acid File Prep**
```{r}
bile_files <- bile_file_prep(bile_acid_FP,
                             inhibitor_list,
                             promoter_list)

big_bile_acid <- bile_files$ProcBile
```
**Stats**
Had to filter out CDCA and UDCA from the 'inhibitors' category because the dunn_test wouldn't run - I think it's because they're all zeroes so there aren't any pairwise comparisons to do between diets for them. 
```{r}
## not faceted by bile acid
no_bile_stat_files <- bile_stats(big_bile_acid,
                                 non_bile_groups,
                                 'conc_normalized')

kruskal <- no_bile_stat_files$KruskalTest
dunn <- no_bile_stat_files$DunnPostHoc

## faceted by bile acid 
bile_stat_files <- bile_stats(big_bile_acid,
                              bile_groups,
                              'conc_normalized')

kruskal_bile <- bile_stat_files$KruskalTest
dunn_bile <- bile_stat_files$DunnPostHoc
```

**Plots**
```{r, fig.width=6, fig.height=3.5}
all_bile_plots <- bile_plots(big_bile_acid,
                             diet_x_labs,
                             dunn,
                             dunn_bile)

overall_plot <- all_bile_plots$NoFacetBilePlot
bile_plot <- all_bile_plots$FacetBilePlot

overall_plot
```

```{r, fig.width=8, fig.height=8}
bile_plot
```

**Saving my Outputs**
```{r}
## plots
ggsave('../plots/overall_bile_acid.pdf',
       plot = overall_plot,
       width = 6,
       height = 3.5)
ggsave('../plots/bile_acid.pdf',
       plot = bile_plot,
       width = 8,
       height = 8)

## rdata object 
save(overall_plot,
     file = '../../figures/overall_bile_acid.rdata')
save(bile_plot,
     file = '../../figures/bile_acid.rdata')

## stats
write_tsv(dunn,
          '../stats/overall_bile_dunn.tsv')
write_tsv(dunn,
          '../stats/bile_acid_dunn.tsv')
```



