---
title: "bile_acid_comparisons"
output: html_document
date: "2024-01-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(broom)
library(cowplot)
library(magrittr)
library(qiime2R)
library(tidyverse)
library(naniar)
library(ggpubr)
library(rstatix)
library(viridis)
```

**Functions**
```{r}
## histo plot comparison function
comp_plot <- function(base_table,
                      ## names associated with the score/concentration value on the y-axis
                      base_table_col_select1,
                      ## score/concentration value that you want on the y-axis
                      base_table_col_select2,
                      ## toxin or metabolomics?
                       wanted_table,
                       ## this is what you want your x axis to be (neat, diluted, or concentration)? 
                       wanted_x,
                       ## this is toxin/metabolite column you want to filter by (neat or diluted)?
                       wanted_y=NULL,
                       ## what is the toxin/metabolite you want to filter by called in the table?
                       wanted_y_name=NULL,
                       diet_labels,
                       row_labels,
                       y_label,
                       x_label,
                       title){
  ## putting tables together
  base_table %>% 
    select(mouse_id, .data[[base_table_col_select1]], .data[[base_table_col_select2]]) %>% 
    left_join(wanted_table, by = 'mouse_id') %>% 
    na.omit() -> compiled_table
  ## constructing plot
  compiled_table %>% 
    # filter(.data[[wanted_y]] == wanted_y_name) %>%
    na.omit() %>% 
    ggplot(aes(x = .data[[wanted_x]], y = .data[[base_table_col_select2]])) +
    scale_x_continuous(trans = 'log10') +
    # scale_y_continuous(trans = 'log10') +
    geom_jitter(alpha = 0.4, width = 0.1, height = 0) +
    geom_smooth(method = 'lm', se = FALSE) +
    facet_grid(.data[[base_table_col_select1]]~diet,
               labeller = labeller(.rows = row_labels,
                                   diet = diet_labels),
               scales = 'free_y') +
    theme_bw() +
    theme(strip.text.y = element_text(angle = 0)) +
    xlab(x_label) +
    ylab(y_label) +
    ggtitle(title) -> plot
  return(plot)
}

## runs linear modeling on the table, diet, location, and toxin/metabolite wanted and returns full results 
## along with a table containing the r2 and p-value 
## this goes into the following function with a for loop 
r2_value <- function(biom_table,
                     wanted_diet,
                     wanted_comp_col,
                     wanted_comp_value,
                     stat_value,
                     tox_metab_col,
                     wanted_tox_metab,
                     wanted_conc_value){
  ## to get r squared value and p value
  biom_table %>% 
    filter(diet == wanted_diet,
           .data[[wanted_comp_col]] == wanted_comp_value,
           .data[[tox_metab_col]] == wanted_tox_metab) %>%
    do(glance(lm(.data[[stat_value]] ~ .data[[wanted_conc_value]],
                 data = .))) -> lm_glance
  output <- list(r_squared = lm_glance$r.squared,
                 p_value = lm_glance$p.value)
  ## to get estimate and place it with the overall output table
  biom_table %>% 
    filter(diet == wanted_diet,
           .data[[wanted_comp_col]] == wanted_comp_value,
           .data[[tox_metab_col]] == wanted_tox_metab) %>%
    do(tidy(lm(.data[[stat_value]] ~ .data[[wanted_conc_value]],
                 data = .))) %>% 
    filter(term != '(Intercept)') -> lm_full
  output <- as_tibble(output) %>%
            mutate(diet = paste(wanted_diet),
                   comp_col = paste(wanted_comp_value),
                   tox_metab = paste(wanted_tox_metab),
                   just_r = sqrt(r_squared),
                   estimate = lm_full$estimate,
                   r_direct = if_else(estimate < 0, just_r * -1, just_r))
  my_list <- list(LinearModel = lm_full,
                  WantedStats = output)
  return(my_list)
}

## creating a nested for loop since running this function individually is a nightmare
r2_value_for_loop <- function(biom_table,
                              comp_col,
                              wanted_stat_value,
                              wanted_col,
                              wanted_conc) {
  ## nested for loop for the r squared and p values 
  output <- tibble()
  for(i in unique(unlist(biom_table$diet))) {
    for(j in unique(unlist(biom_table[comp_col]))) {
      for(k in unique(unlist(biom_table[wanted_col]))) {
        tmp_output <- r2_value(biom_table,
                               i,
                               comp_col,
                               j,
                               wanted_stat_value,
                               wanted_col,
                               k,
                               wanted_conc)$WantedStats
        output <- bind_rows(output, tmp_output)
      }
    }
  }
  return(output)
}

## statistical visualization plot function
stat_plot <- function(stat_table,
                      wanted_labels,
                      x_lab_list,
                      y_lab_list,
                      x_axis_label,
                      title){
  stat_table %>% 
    ggplot(aes(x = comp_col, y = diet)) +
    geom_tile(aes(fill = r_direct), color = 'black', alpha = 0.6) +
    geom_text(aes(label = signif)) +
    scale_fill_gradient2(low = 'blue', high = 'green', name = "Correlation\nCoefficient") +
    scale_x_discrete(labels = x_lab_list) +
    scale_y_discrete(labels = y_lab_list) +
    facet_wrap(~tox_metab,
               labeller = labeller(tox_metab = wanted_labels)) +
    theme_bw() +
    xlab(x_axis_label) +
    ylab("Diet") +
    ggtitle(title) -> plot
  return(plot)
}
```

**Bile Acid Ratio Stats Functions**
```{r}
ratio_r2_value <- function(biom_table,
                     wanted_diet,
                     category_col,
                     col_value,
                     wanted_category,
                     wanted_conc_value){
  ## to get r squared value and p value
  biom_table %>% 
    filter(diet == wanted_diet,
           .data[[category_col]] == wanted_category) %>%
    do(glance(lm(.data[[col_value]] ~ .data[[wanted_conc_value]],
                 data = .))) -> lm_glance
  output <- list(r_squared = lm_glance$r.squared,
                 p_value = lm_glance$p.value)
  ## to get estimate and place it with the overall output table
  biom_table %>% 
    filter(diet == wanted_diet,
           .data[[category_col]] == wanted_category) %>%
    do(tidy(lm(.data[[col_value]] ~ .data[[wanted_conc_value]],
                 data = .))) %>% 
    filter(term != '(Intercept)') -> lm_full
  output <- as_tibble(output) %>%
            mutate(diet = paste(wanted_diet),
                   category = paste(wanted_category),
                   just_r = sqrt(r_squared),
                   estimate = lm_full$estimate,
                   r_direct = if_else(estimate < 0, just_r * -1, just_r))
  my_list <- list(LinearModel = lm_full,
                  WantedStats = output)
  return(my_list)
}

## creating a nested for loop since running this function individually is a nightmare
ratio_r2_value_for_loop <- function(biom_table,
                              wanted_col,
                              col_value,
                              wanted_conc) {
  ## nested for loop for the r squared and p values 
  output <- tibble()
  for(i in unique(unlist(biom_table$diet))) {
    for(j in unique(unlist(biom_table[wanted_col]))) {
        tmp_output <- ratio_r2_value(biom_table,
                               i,
                               wanted_col,
                               col_value,
                               j,
                               wanted_conc)$WantedStats
        output <- bind_rows(output, tmp_output)
      }
    }
  return(output)
}

## statistical visualization plot function
ratio_stat_plot <- function(stat_table,
                      y_lab_list,
                      x_lab_list,
                      x_label,
                      title){
  stat_table %>% 
    ggplot(aes(x = category, y = diet)) +
    geom_tile(aes(fill = r_direct), color = 'black', alpha = 0.6) +
    geom_text(aes(label = signif)) +
    scale_fill_gradient2(low = 'blue', high = 'green', name = "Correlation\nCoefficient") +
    scale_y_discrete(labels = y_lab_list) +
    scale_x_discrete(labels = x_lab_list) +
    facet_wrap(~ratio_label) +
    theme_bw() +
    xlab(x_label) +
    ylab("Diet") +
    ggtitle(title) -> plot
  return(plot)
}
```


**File Paths**
```{r}
metadata_FP <- '../data/misc/filt_cecal_processed_metadata.tsv'
histo_FP <- '../data/misc/processed_histopathology.tsv'
metab_FP <- '../data/misc/processed_metabolomics.tsv'
neat_tox_FP <- '../data/misc/processed_neatToxin.tsv'
dil_tox_FP <- '../data/misc/processed_dilutedToxin.tsv'
bile_acid_FP <- '../data/misc/processed_bile_acid.tsv'
bile_acidRatio_FP <- '../data/misc/processed_ratio_bileAcid.tsv'

## lists to redo the diet names on the facet labels of the ggplot created below 
diet_labs <- 
    c('Chow', 
      'High Fat / High Fiber', 
      'High Fat / Low Fiber', 
      'Low Fat / High Fiber', 
      'Low Fat / Low Fiber')

names(diet_labs) <- c('Chow', 'HF/HF', 'HF/LF', 'LF/HF', 'LF/LF')

stat_diet_y_labs <- c("Chow",
                       "HFt/HFb",
                       "HFt/LFb",
                       "LFt/HFb",
                       "LFt/LFb")

tissue_labs <- c('Cecum',
                 'Colon')
names(tissue_labs) <- c('cecum',
                        'colon')

metab_labs <- c('Acetic Acid',
                'Propanoic Acid',
                'n-Butanoic Acid')
names(metab_labs) <- c('Acetic Acid (ug/g)',
                       'Propanoic Acid (ug/g)',
                       'n-Butanoic Acid (ug/g)')

neat_labs <- c('TcdA', 'TcdB')
names(neat_labs) <- c('Total TcA Neat', 'Total TcB Neat')

dil_labs <- c('TcdA', 'TcdB')
names(dil_labs) <- c('Total TcA 1:10', 'Total TcB 1:10')

bile_facet_labs <- c('Promoter',
                     'Inhibitor')
names(bile_facet_labs) <- c('promoter',
                            'inhibitor')
```

**File Prep**
```{r}
metadata <- read_tsv(metadata_FP)
histo <- read_tsv(histo_FP)

metab <- read_tsv(metab_FP)
names(metab)[names(metab) == 'concentration'] <- 'metab_concentration'
metab %>% 
  mutate(metab_conc_norm = (metab_concentration + 2)) -> metab

neat_tox <- read_tsv(neat_tox_FP)
neat_tox %>% 
  mutate(neat_conc_norm = (neat_conc + 2)) -> neat_tox

dil_tox <- read_tsv(dil_tox_FP)
bile_acid <- read_tsv(bile_acid_FP)

ratio_bile_acid <- read_tsv(bile_acidRatio_FP)
ratio_bile_acid %>% 
  select(!c(inhibitor, promoter)) -> ratio_bile_acid
```

**Histopathology/Bile Acid Comparison Plot**
```{r, warning=FALSE, fig.width=9, fig.height=4}
histo_inhibitors_plot <- comp_plot(histo,
                                   'tissue',
                                   'score',
                                   bile_acid,
                                   'conc_normalized',
                                   'c_diff_effect',
                                   'inhibitor',
                                   diet_labs,
                                   tissue_labs,
                                   'Score',
                                   'Concentration (log10)',
                                   'Histopathology Score Against Bile Acid Inhibitors')

histo_promoters_plot <- comp_plot(histo,
                                  'tissue',
                                  'score',
                                   bile_acid,
                                   'conc_normalized',
                                   'c_diff_effect',
                                   'promoter',
                                   diet_labs,
                                   tissue_labs,
                                   'Score',
                                   'Concentration (log10)',
                                   'Histopathology Score Against Bile Acid Promoters')

histo_inhibitors_plot
histo_promoters_plot
```

**Stats**
```{r, warning=FALSE, fig.height=4, fig.width=6.5}
histo %>% 
  select(mouse_id, tissue, score) %>% 
  left_join(bile_acid, by = 'mouse_id') %>% 
  na.omit() -> histo_bile_acid

histo_bile_stats <- r2_value_for_loop(histo_bile_acid,
                                      'tissue',
                                      'score',
                                      'c_diff_effect',
                                      'conc_normalized')

## assigning p value significance
histo_bile_stats['signif'] <- symnum(histo_bile_stats$p_value,
                                     cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1),
                                     symbols = c("****", "***", "**", "*", "ns"),
                                     abbr.colnames = FALSE,
                                     na = "")

histo_bile_stat_plot <- stat_plot(histo_bile_stats,
                                  bile_facet_labs,
                                  tissue_labs,
                                  stat_diet_y_labs,
                                  'Tissue',
                                  'Histopathology Score and Bile Acid\nConcentration Comparisons')

histo_bile_stat_plot
```
**Plots Together**
```{r, warning=FALSE, fig.height=8, fig.width=16.5}
plot_grid(histo_inhibitors_plot,
          histo_bile_stat_plot,
          histo_promoters_plot,
          nrow = 2,
          rel_widths = c(1, 0.6),
          labels = c('a)', 'c)', 'b)'),
          axis = c('tblr'),
          align = c('hv')) -> histo_bile_plots_together

histo_bile_plots_together
```

**Histopathology/Bile Acid Ratio Comparison Plots**
```{r, warning=FALSE, fig.width=8, fig.height=3.5}
histo_ratio_plot <- comp_plot(histo,
                              'tissue',
                              'score',
                              ratio_bile_acid,
                              'ratio',
                              NULL,
                              NULL,
                              diet_labs,
                              tissue_labs,
                              'Score',
                              'Ratio',
                              'Histopathology Score Against Bile Acid Growth Ratio')

histo_ratio_plot
```

**Stats**
```{r, warning=FALSE, fig.height=3, fig.width=5}
histo %>% 
  select(mouse_id, tissue, score) %>% 
  left_join(ratio_bile_acid, by = 'mouse_id') %>% 
  na.omit() -> histo_ratio

histo_ratio_stats <- ratio_r2_value_for_loop(histo_ratio,
                                              'tissue',
                                             'score',
                                              'ratio')

## assigning p value significance
histo_ratio_stats['signif'] <- symnum(histo_ratio_stats$p_value,
                                     cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1),
                                     symbols = c("****", "***", "**", "*", "ns"),
                                     abbr.colnames = FALSE,
                                     na = "")

histo_ratio_stats %>% 
  mutate(ratio_label = paste('Promoter:Inhibitor')) -> histo_ratio_stats

histo_ratio_stat_plot <- ratio_stat_plot(histo_ratio_stats,
                                         stat_diet_y_labs,
                                        tissue_labs,
                                        'Tissue',
                                         'Histopathology Score and Bile Acid\nRatio Comparisons')

histo_ratio_stat_plot
```

**Plots Together**
```{r, warning=FALSE, fig.height=3.5, fig.width=15}
plot_grid(histo_ratio_plot,
          histo_ratio_stat_plot,
          rel_widths = c(1, 0.5),
          labels = c('a)', 'b)'),
          axis = c('tblr'),
          align = c('hv')) -> histo_ratio_plots_together

histo_ratio_plots_together
```


**Metabolomics/Bile Acid Comparison Plot**
```{r, warning=FALSE, fig.width=10, fig.height=4}
metab_inhibitors_plot <- comp_plot(metab,
                                   'metabolite',
                                   'metab_conc_norm',
                                   bile_acid,
                                   'conc_normalized',
                                   'c_diff_effect',
                                   'inhibitor',
                                   diet_labs,
                                   metab_labs,
                                   'Metabolite Concentration',
                                   'Bile Acid Concentration',
                                   'Metabolites Against Bile Acid C. difficile Growth Inhibitors')

metab_promoters_plot <- comp_plot(metab,
                                  'metabolite',
                                  'metab_conc_norm',
                                   bile_acid,
                                   'conc_normalized',
                                   'c_diff_effect',
                                   'promoter',
                                   diet_labs,
                                   metab_labs,
                                   'Metabolite Concentration',
                                   'Bile Acid Concentration',
                                   'Metabolites Against Bile Acid C. difficile Growth Promoters')

metab_inhibitors_plot
metab_promoters_plot
```

**Stats**
```{r, warning=FALSE, fig.height=4, fig.width=8}
metab_x_labs <- c('Acetic\nAcid',
                  'n-Butanoic\nAcid',
                  'Propanoic\nAcid')
metab %>% 
  select(mouse_id, metabolite, metab_conc_norm) %>% 
  left_join(bile_acid, by = 'mouse_id') %>% 
  na.omit() -> metab_bile_acid

metab_bile_stats <- r2_value_for_loop(metab_bile_acid,
                                      'metabolite',
                                      'metab_conc_norm',
                                      'c_diff_effect',
                                      'conc_normalized')

## assigning p value significance
metab_bile_stats['signif'] <- symnum(metab_bile_stats$p_value,
                                     cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1),
                                     symbols = c("****", "***", "**", "*", "ns"),
                                     abbr.colnames = FALSE,
                                     na = "")

metab_bile_stat_plot <- stat_plot(metab_bile_stats,
                                  bile_facet_labs,
                                  metab_x_labs,
                                  stat_diet_y_labs,
                                  'Metabolite',
                                  'Metabolite and Bile Acid\nConcentration Comparisons')

metab_bile_stat_plot
```

**Plots Together**
```{r, warning=FALSE, fig.height=8, fig.width=16.5}
plot_grid(metab_inhibitors_plot,
          metab_bile_stat_plot,
          metab_promoters_plot,
          nrow = 2,
          rel_widths = c(1, 0.6),
          labels = c('a)', 'c)', 'b)'),
          axis = c('tblr'),
          align = c('hv')) -> metab_bile_plots_together

metab_bile_plots_together
```


**Neat Toxin/Bile Acid Comparison Plots**
```{r, fig.width=9, fig.height=4, warning=FALSE}
toxin_inhibitors_plot <- comp_plot(neat_tox,
                                   'neat_toxin',
                                   'neat_conc_norm',
                                   bile_acid,
                                   'conc_normalized',
                                   'c_diff_effect',
                                   'inhibitor',
                                   diet_labs,
                                   neat_labs,
                                   'Toxin Concentration',
                                   'Bile Acid Concentration',
                                   'Toxin Against Bile Acid C. difficile Growth Inhibitors')

toxin_promoters_plot <- comp_plot(neat_tox,
                                  'neat_toxin',
                                  'neat_conc_norm',
                                   bile_acid,
                                   'conc_normalized',
                                   'c_diff_effect',
                                   'promoter',
                                   diet_labs,
                                   neat_labs,
                                   'Toxin Concentration',
                                   'Bile Acid Concentration',
                                   'Toxin Against Bile Acid C. difficile Growth Promoters')

toxin_inhibitors_plot
toxin_promoters_plot
```

**Stats**
```{r, warning=FALSE, fig.height=4, fig.width=6.5}
neat_tox %>% 
  select(mouse_id, neat_toxin, neat_conc_norm) %>% 
  left_join(bile_acid, by = 'mouse_id') %>% 
  na.omit() -> tox_bile_acid

tox_bile_stats <- r2_value_for_loop(tox_bile_acid,
                                      'neat_toxin',
                                      'neat_conc_norm',
                                      'c_diff_effect',
                                      'conc_normalized')

## assigning p value significance
tox_bile_stats['signif'] <- symnum(tox_bile_stats$p_value,
                                     cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1),
                                     symbols = c("****", "***", "**", "*", "ns"),
                                     abbr.colnames = FALSE,
                                     na = "")

tox_bile_stat_plot <- stat_plot(tox_bile_stats,
                                  bile_facet_labs,
                                  neat_labs,
                                  stat_diet_y_labs,
                                  'Toxin',
                                  'C. difficile Toxin and Bile Acid\nConcentration Comparisons')

tox_bile_stat_plot
```

**Stats on the Plot**
```{r, fig.width=17, fig.height=5, warning=FALSE}
tox_bile_stats %>% 
  filter(tox_metab == 'inhibitor',
         signif != 'ns') %>% 
  select(diet, tox_metab, comp_col, signif, p_value) %>% 
  mutate(neat_conc_norm = paste(160),
         neat_conc_norm = as.numeric(neat_conc_norm),
         conc_normalized = paste(3000),
         conc_normalized = as.numeric(conc_normalized),
         p_value = signif(p_value, digits = 3),
         p_value = paste('p-value', p_value, sep = ": ")) -> tox_bile_stats2

names(tox_bile_stats2)[names(tox_bile_stats2) == 'comp_col'] <- 'neat_toxin'
names(tox_bile_stats2)[names(tox_bile_stats2) == 'tox_metab'] <- 'c_diff_effect'

toxin_inhibitors_plot +
  geom_label(data = tox_bile_stats2, 
            label = tox_bile_stats2$p_value,
            color = 'black',
            fontface = 'bold',
            size = 6) -> toxin_inhibitors_stats_plot

toxin_inhibitors_stats_plot
```

**Plots Together**
```{r, warning=FALSE, fig.height=7.5, fig.width=16.5}
plot_grid(toxin_inhibitors_plot,
          tox_bile_stat_plot,
          toxin_promoters_plot,
          nrow = 2,
          rel_widths = c(1, 0.6),
          labels = c('a)', 'c)', 'b)'),
          axis = c('tblr'),
          align = c('hv')) -> tox_bile_plots_together

tox_bile_plots_together
```

**Ratio/Toxin Comparisons**
```{r, fig.width=9, fig.height=4, warning=FALSE}
tox_ratio_plot <- comp_plot(neat_tox,
                              'neat_toxin',
                              'neat_conc_norm',
                              ratio_bile_acid,
                              'ratio',
                              NULL,
                              NULL,
                              diet_labs,
                              neat_labs,
                              'Toxin Concentration',
                              'Ratio',
                              'Toxin Against Bile Acid Growth Ratio')

tox_ratio_plot
```

**Ratio/Toxin Stats**
```{r, warning=FALSE, fig.height=3, fig.width=5}
neat_tox %>% 
  select(mouse_id, neat_toxin, neat_conc_norm) %>% 
  left_join(ratio_bile_acid, by = 'mouse_id') %>% 
  na.omit() -> tox_ratio

tox_ratio_stats <- ratio_r2_value_for_loop(tox_ratio,
                                              'neat_toxin',
                                           'neat_conc_norm',
                                              'ratio')

## assigning p value significance
tox_ratio_stats['signif'] <- symnum(tox_ratio_stats$p_value,
                                     cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1),
                                     symbols = c("****", "***", "**", "*", "ns"),
                                     abbr.colnames = FALSE,
                                     na = "")

tox_ratio_stats %>% 
  mutate(ratio_label = paste('Promoter:Inhibitor')) -> tox_ratio_stats

tox_ratio_stat_plot <- ratio_stat_plot(tox_ratio_stats,
                                         stat_diet_y_labs,
                                        neat_labs,
                                       'Toxin',
                                         'Toxin Concentration and Bile Acid\nRatio Comparisons')

tox_ratio_stat_plot
```

**Stats on the Plot**
```{r, fig.width=10, fig.height=4, warning=FALSE}
tox_ratio_stats %>% 
  filter(signif != 'ns') %>% 
  select(diet, category, signif) %>% 
  mutate(neat_conc_norm = paste(500),
         neat_conc_norm = as.numeric(neat_conc_norm),
         ratio = paste(0.1),
         ratio = as.numeric(ratio)) -> tox_ratio_stats2

names(tox_ratio_stats2)[names(tox_ratio_stats2) == 'category'] <- 'neat_toxin'

tox_ratio_plot +
  geom_text(data = tox_ratio_stats2, 
            label = tox_ratio_stats2$signif,
            color = 'red',
            size = 6) -> tox_ratio_stats_plot

tox_ratio_stats_plot
```

**Ratio/Toxin Plots Together**
```{r, fig.width=15, fig.height=3.5}
plot_grid(tox_ratio_plot,
          tox_ratio_stat_plot,
          rel_widths = c(1, 0.5),
          labels = c('a)', 'b)'),
          axis = c('tblr'),
          align = c('hv')) -> tox_ratio_plots_together

tox_ratio_plots_together
```



**Saving my Outputs**
```{r}
ggsave('../plots/bileAcid_histo_comp.pdf',
       plot = histo_bile_plots_together,
       width = 16.5,
       height = 8)
ggsave('../plots/bileAcid_metab_comp.pdf',
       plot = metab_bile_plots_together,
       width = 16.5,
       height = 8)
ggsave('../plots/bileAcid_toxin_comp.pdf',
       plot = tox_bile_plots_together,
       width = 16.5,
       height = 8)
ggsave('../plots/bileRatio_histo_comp.pdf',
       plot = histo_ratio_plots_together,
       width = 15,
       height = 3.5)
ggsave('../plots/bileRatio_toxin_comp.pdf',
       plot = tox_ratio_plots_together,
       width = 15,
       height = 3.5)
ggsave('../plots/bileRatio_toxin_comp_wStats.pdf',
       plot = tox_ratio_stats_plot,
       width = 10,
       height = 4)

## as rdata objects for manuscript figures
save(histo_inhibitors_plot,
     file = '../../figures/bile_inhibit_histo.rdata')
save(histo_promoters_plot,
     file = '../../figures/bile_promote_histo.rdata')
save(histo_bile_stat_plot,
     file = '../../figures/bile_histo_stats.rdata')
save(histo_ratio_plot,
     file = '../../figures/bile_ratio_histo.rdata')
save(metab_inhibitors_plot,
     file = '../../figures/bile_inhibit_metab.rdata')
save(metab_promoters_plot,
     file = '../../figures/bile_promote_metab.rdata')
save(metab_bile_stat_plot,
     file = '../../figures/bile_metab_stats.rdata')
save(toxin_inhibitors_stats_plot,
     file = '../../figures/bile_inhibit_toxin.rdata')
save(toxin_promoters_plot,
     file = '../../figures/bile_promote_toxin.rdata')
save(tox_bile_stat_plot,
     file = '../../figures/bile_toxin_stats.rdata')
save(histo_ratio_plot,
     file = '../../figures/bile_ratio_histo.rdata')
save(histo_ratio_stat_plot,
     file = '../../figures/bile_ratio_stats.rdata')
```

