---
title: "bile_acid"
output: html_document
date: "2023-12-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(rstatix)
library(ggplot2)
library(ggpubr)
library(ggh4x)
library(cowplot)
library(magrittr)
library(glue)
library(apppleplots)
```

**Functions**
```{r}
## statistical analysis function
bile_stats <- function(big_bile_table,
                       group_cols,
                       conc_value){
  ## kruskal test
  big_bile_table %>% 
    group_by(big_bile_table[unlist(group_cols)]) %>% 
    do(tidy(kruskal.test(.data[[conc_value]] ~ diet,
               data = .))) %>% 
    ungroup() %>% 
    arrange(p.value) %>% 
    mutate(p.adj = p.adjust(p.value,
                            method = "BH")) -> kruskal_test
  ## dunns post hoc test
  diet_name <- 'diet'
  test <- reformulate(glue("{diet_name}"),glue("{conc_value}"))
  
  big_bile_table %>% 
    group_by(big_bile_table[unlist(group_cols)]) %>% 
    dunn_test(test,
              p.adjust.method = 'BH',
              data = .) %>% 
    add_y_position(scales = 'free',
                   fun = 'max',
                   y.trans = log10) -> dunn_test
  # ## creating a list of outputs
  my_list <- list(KruskalTest = kruskal_test,
                  DunnPostHoc = dunn_test)
  return(my_list)
}

## creating plots
bile_plots <- function(big_bile_table,
                       x_labs,
                       non_bile_dunn,
                       bile_dunn){
  ## plot not faceted by bile acid
  big_bile_table %>% 
    ggplot(aes(x = diet, y = conc_normalized)) +
      geom_boxplot(aes(group = diet), outlier.shape = NA) +
      geom_jitter(alpha = 0.4, width = 0.1, height = 0) +
      scale_x_discrete(labels = x_labs) +
      theme_bw() +
      facet_wrap(~c_diff_effect,
                 labeller = labeller(c_diff_effect = cdiff_effect_labs)) +
      scale_y_log10() +
      stat_pvalue_manual(non_bile_dunn,
                         tip.length = 0.01,
                         step.increase = 0.05,
                         label = 'p.adj.signif',
                         hide.ns = TRUE) +
      labs(x = 'Diet',
           y = 'Concentration (pmol/mg) (log10)',
           title = 'Cecal Bile Acid Concentration') -> overall_bile_plot
  ## plot faceted by bile acid
  big_bile_acid %>% 
    ggplot(aes(x = diet, y = conc_normalized)) +
      geom_boxplot(aes(group = diet), outlier.shape = NA) +
      geom_jitter(alpha = 0.4, width = 0.1, height = 0) +
      scale_x_discrete(labels = x_labs) +
      facet_nested_wrap(~c_diff_effect + bile_acid,
                        labeller = labeller(c_diff_effect = cdiff_effect_labs,
                                            bile_acid = bile_labs),
                        nrow = 2) +
      theme_bw() +
      scale_y_log10() +
      stat_pvalue_manual(dunn_bile,
                         tip.length = 0.01,
                         step.increase = 0.05,
                         label = 'p.adj.signif',
                         hide.ns = TRUE) +
    labs(x = 'Diet',
         y = 'Concentration (pmol/mg) (log10)',
         title = 'Cecal Bile Acid Concentration') -> bile_include_plot
  ## creating list of outputs
  my_list <- list(NoFacetBilePlot = overall_bile_plot,
                  FacetBilePlot = bile_include_plot)
  return(my_list)
}

## promoter to inhibitor ratio file prep, stats, and plot function
bile_ratio_func <- function(ratio_table){
  ## stats
  ## kruskal test
  ratio_table %>% 
    do(tidy(kruskal.test(ratio ~ diet,
                         data = .))) %>% 
    arrange(p.value) %>% 
    mutate(p.adj = p.adjust(p.value,
                            method = 'BH')) -> ratio_kruskal
  ## dunns test
  ratio_table %>% 
    dunn_test(ratio ~ diet,
              p.adjust.method = 'BH',
              data = .) %>% 
    add_y_position(scales = 'free_y',
                   fun = 'max') -> ratio_dunn
  ## plot
  ratio_table %>% 
    ggplot(aes(x = diet, y = ratio)) +
      geom_boxplot(aes(group = diet), outlier.shape = NA) +
      geom_jitter(width = 0.1, height = 0, alpha = 0.4) +
      theme_bw() +
      scale_x_discrete(labels = diet_x_labs) +
      facet_wrap(~ratio_label) +
      stat_pvalue_manual(ratio_dunn,
                         tip.length = 0.01,
                         step.increase = 0.05,
                         label = 'p.adj.signif',
                         hide.ns = TRUE) +
      labs(x = 'Diet',
           y = 'Ratio',
           title = 'Cecal Bile Acid Ratio') -> ratio_bile_plot
  ## creating a list of my outputs
  my_list <- list(RatioTable = ratio_table,
                  RatioKruskal = ratio_kruskal,
                  RatioDunn = ratio_dunn,
                  RatioPlot = ratio_bile_plot)
  return(my_list)
}
```

**File Paths**
```{r}
bile_acid_FP <- '../data/misc/processed_bile_acid.tsv'
bile_ratio_FP <- '../data/misc/processed_ratio_bileAcid.tsv'

cdiff_effect_labs <- c('C. difficile Inhibitor',
                       'C. difficile Promoter')
names(cdiff_effect_labs) <- c('inhibitor',
                              'promoter')

bile_labs <- c('a-MCA',
               'b-MCA',
               'LCA',
               'DCA',
               'TCA',
               'CA')
names(bile_labs) <- c("Alpha_muricholic_acid_a-MCA",
                     "Beta_muricholic_acid_b-MCA",
                     "Lithocholic_acid_LCA", 
                     "Deoxycholic_acid_DCA",
                     "Taurocholic_acid_T-CA",
                     "Cholic_acid_CA")

bile_groups <- c('c_diff_effect',
                 'bile_acid')
non_bile_groups <- c('c_diff_effect')

diet_x_labs <- c('Chow',
                 'HFt/\nHFb',
                 'HFt/\nLFb',
                 'LFt/\nHFb',
                 'LFt/\nLFb')
```

Use prescence/abscence tests or rank-based test instead of a non-parametric statistical test!!
- Look at whether you'd like to include additional bile acids in the promoters/inhibitors categories

**Reading in Needed Files**
```{r}
big_bile_acid <- read_tsv(bile_acid_FP)
bile_ratio <- read_tsv(bile_ratio_FP)
```
**Stats**
Had to filter out CDCA and UDCA from the 'inhibitors' category because the dunn_test wouldn't run - I think it's because they're all zeroes so there aren't any pairwise comparisons to do between diets for them. Maybe include T-CDCA and T-DCA as inhibitors instead?
```{r}
## not faceted by bile acid
no_bile_stat_files <- bile_stats(big_bile_acid,
                                 non_bile_groups,
                                 'conc_normalized')

kruskal <- no_bile_stat_files$KruskalTest
dunn <- no_bile_stat_files$DunnPostHoc

## faceted by bile acid 
bile_stat_files <- bile_stats(big_bile_acid,
                              bile_groups,
                              'conc_normalized')

kruskal_bile <- bile_stat_files$KruskalTest
dunn_bile <- bile_stat_files$DunnPostHoc
```

**Plots**
```{r, fig.width=6, fig.height=3.5}
all_bile_plots <- bile_plots(big_bile_acid,
                             diet_x_labs,
                             dunn,
                             dunn_bile)

overall_plot <- all_bile_plots$NoFacetBilePlot
bile_plot <- all_bile_plots$FacetBilePlot

overall_plot
```

```{r, fig.width=8, fig.height=6}
bile_plot
```

**Ratio of C. difficile Promoter:Inhibitor**
```{r, fig.width=5, fig.height=3.5}
ratio_files <- bile_ratio_func(bile_ratio)

bile_ratio <- ratio_files$RatioTable
ratio_kruskal <- ratio_files$RatioKruskal
ratio_dunn <- ratio_files$RatioDunn
ratio_bile_plot <- ratio_files$RatioPlot

ratio_bile_plot
```

**Saving my Outputs**
```{r}
## plots
ggsave('../plots/overall_bile_acid.pdf',
       plot = overall_plot,
       width = 6,
       height = 3.5)
ggsave('../plots/bile_acid.pdf',
       plot = bile_plot,
       width = 8,
       height = 6)
ggsave('../plots/bile_acid_ratio.pdf',
       plot = ratio_bile_plot,
       width = 5,
       height = 3.5)

## rdata object 
save(overall_plot,
     file = '../../figures/overall_bile_acid.rdata')
save(bile_plot,
     file = '../../figures/bile_acid.rdata')
save(ratio_bile_plot,
     file = '../../figures/bile_acid_ratio.rdata')

## stats
write_tsv(dunn,
          '../stats/overall_bile_dunn.tsv')
write_tsv(dunn_bile,
          '../stats/bile_acid_dunn.tsv')
write_tsv(ratio_dunn,
          '../stats/ratio_bile_acid.tsv')

## saving processed bile acid files so I don't have to keep reprocessing them
write_tsv(big_bile_acid,
          '../data/misc/processed_bile_acid.tsv')
write_tsv(bile_ratio,
          '../data/misc/processed_ratio_bileAcid.tsv')
```



