---
title: "hypoxia_ancombc"
output: html_document
date: "2023-12-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(qiime2R)
library(cowplot)
library(magrittr)
library(rstatix)
library(vegan)
library(broom)
library(ANCOMBC)
library(microViz)
library(phyloseq)
library(cowplot)
```

**Functions**
```{r}
## ancombc file prep and running it function
run_ancombc <- function(otu_fp,
                        tax_fp,
                        metadata_table,
                        wanted_tax_level,
                        formula,
                        adj_method){
  ## metadata table 
  metadata <- as.data.frame(metadata_table)
  rownames(metadata) <- metadata$sampleid
  ## otu table
  otu_table <- read_qza(otu_fp)$data
  otu_table %>% 
    as_tibble(rownames = 'taxon') %>% 
    gather(-taxon, key = sampleid, value = abun) %>% 
    filter(sampleid %in% metadata$sampleid) %>% 
    spread(sampleid, abun) -> otu_table

  otu_table <- as.data.frame(otu_table)
  rownames(otu_table) <- otu_table$taxon
  proc_otu <- subset(otu_table,
                     select = -taxon)
  metadata %>% 
    filter(metadata$sampleid %in% colnames(proc_otu)) -> metadata
  ## taxonomy table
  tax_table <- read_qza(tax_fp)$data %>% 
    parse_taxonomy()
  tax_table %>% 
    filter(rownames(tax_table) %in% rownames(proc_otu)) -> tax_table
  ## creating phyloseq object
  pseq <- phyloseq::phyloseq(otu_table(proc_otu, taxa_are_rows = TRUE),
                             tax_table(as.matrix(tax_table)),
                             sample_data(metadata))
  ## running ancombc
  ## everything but data (pseq) can be a string
  ancombc_res <- ancombc2(data = pseq,
                          tax_level = wanted_tax_level,
                          fix_formula = formula,
                          p_adj_method = adj_method)
  ## pulling wanted results out
  results <- ancombc_res$res
  ## creating a list of outputs
  my_list <- list(ANCOMBCResults = results,
                  Metadata = metadata,
                  OTUTable = proc_otu,
                  Taxonomy = tax_table,
                  Pseq = pseq)
  return(my_list)
}

## mini metadata filtering function
meta_filter <- function(metadata_table,
                        wanted_filter){
  metadata_table %>% 
    # select(sampleid, mouse_id, location, diet, fiber, fluorescence, batch, .data[[wanted_filter]]) -> filt_meta
    filter(location == wanted_filter) -> filt_meta
  return(filt_meta)
}

## creating a for loop to run ancombc on each day_post_inf and then put all the tables together into one giant table
ancom_diet_for_loop <- function(otu_fp,
                                tax_fp,
                                metadata_table,
                                wanted_tax_level,
                                formula,
                                adj_method,
                                strat_column){
  output <- list()
  for (i in unique(unlist(metadata_table[strat_column]))) {
    new_metadata <- meta_filter(metadata_table,
                                i)
    ancombc_results <- run_ancombc(otu_fp,
                                    tax_fp,
                                    new_metadata,
                                    wanted_tax_level,
                                    formula,
                                    adj_method)
    tmp_output <- ancombc_results$ANCOMBCResults
    tmp_output[strat_column] <- i
    output <- append(output, list(tmp_output))
  }
  output <- bind_rows(output)
  return(output)
}

## function to put together plots of lfc by day for each diet baseline 
baseline_plot <- function(baseline_table,
                          facet_labs=NULL,
                          names_of_labs=NULL,
                          x_labels,
                          title){
  names(facet_labs) <- names_of_labs
  baseline_table %>% 
    ggplot(aes(x = component, y = taxon)) +
    geom_tile(aes(fill = lfc), color = 'black') +
    scale_fill_gradient2(low = 'blue', high = 'red') +
    geom_text(aes(label = lfc), size = 3) +
    theme_bw(base_size = 16) +
    scale_x_discrete(labels = x_labels) +
    # facet_grid(~order,
    #            labeller = labeller(order = facet_labs)) +
    theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          legend.title = element_blank(),
          plot.title = element_text(face = "bold")) +
    ggtitle(title) -> plot
  return(plot)
}
```

**File Paths**
```{r}
metadata_FP <- '../data/misc/filt_cecal_processed_metadata.tsv'
hypoxia_FP <- '../data/misc/pimid_fluor.csv'
otu_table_FP <- '../data/cecal_qiime/tax_filt_actual.qza'
tax_FP <- '../data/cecal_qiime/taxonomy.qza'

col_names <- c('taxon',
               'location',
               'high_fat',
               'high_fiber',
               'purified_diet')

strat_value <- 'location'

hypoxia_facet_labs <- c('Cecum',
                        'Proximal Colon',
                        'Distal Colon')
hypoxia_names_labs <- c(1,
                        2,
                        3)

diet_comp_x_labs <- c('High\nFat',
                      'High\nFiber',
                      'Purified\nDiet')
location_x_labs <- c('Cecum',
                     'Proximal Colon',
                     'Distal Colon')
```

**Combining Metadata Table with Hypoxia Data**
Run it only on the cecum by fluorescence!! 'high_fat + high_fiber + purified_diet + fluorescence'
- histo epithelium involvement and neutrophil invasion correlated with hypoxia? 

We don't have hypoxia data for all mice that we have cecal data for (not sure what's going on here). 
Mouse ID 9743 was duplicated in the hypoxia data but the rest of the values were different (i.e. fluorescence). I think the mouse ids are supposed to go 9740, 9741, 9743, 9742 but I will need to ask Keith. I'm temporarily altering the mouse ids to reflect that order (was 9740, 9742, 9743, 9743). 

Mouse ID 9941 was also duplicated with separate values for each. Need to ask about this. Changed the first duplicate to 9939 (now 9939, 9941, 9943... was 9941, 9941, 9943). 
```{r}
## metadata
hypox_metadata <- read_tsv(metadata_FP)

## hypoxia 
hypoxia <- read_csv(hypoxia_FP)

hypoxia %>% 
  # group_by(mouse_id, diet, fiber, batch) %>%
  # pivot_wider(names_from = location,
  #             values_from = fluorescence) %>%
  # ungroup() %>%
  mutate(fiber = ifelse(diet == 'Chow', 'chow', fiber)) -> hypoxia

hypoxia[[1]][[4]] <- '9939'
hypoxia[[1]][[37]] <- '9939'
hypoxia[[1]][[70]] <- '9939'

hypox_metadata %>% 
  select(sampleid, mouse_id, high_fat, high_fiber, purified_diet) %>% 
  mutate(mouse_id = as.character(mouse_id)) %>% 
  left_join(hypoxia, by = 'mouse_id') %>% 
  na.omit() -> hypox_metadata

hypox_metadata %>% 
  group_by(location) %>% 
  mutate(sum = sum(fluorescence),
         ratio = (fluorescence/sum),
         high_fat_scaled = (high_fat * ratio),
         high_fiber_scaled = (high_fiber * ratio),
         purified_diet_scaled = (purified_diet * ratio)) %>% 
  ungroup() -> hypox_metadata
```

```{r, warning=FALSE}
hypox_metadata %>% 
  filter(location == 'Cecum') -> hypox_meta_cecum

cecum_res <- run_ancombc(otu_table_FP,
                        tax_FP,
                        hypox_meta_cecum,
                        'Family',
                        'fluorescence',
                        'BH')

cecum_ancom <- cecum_res$ANCOMBCResults
```

```{r}
cecum_ancom %>% 
  filter(diff_fluorescence == TRUE) %>% 
  mutate(lfc_fluorescence_corr = (lfc_fluorescence * diff_fluorescence)) %>% 
  select(taxon, contains('corr')) -> cecum_lfc

colnames(cecum_lfc) <- c('taxon',
                         'fluorescence')

cecum_lfc %>% 
  gather(-taxon, key = component, value = lfc) %>% 
  mutate(lfc = as.numeric(lfc),
         lfc = signif(lfc, digits = 5),
         lfc = ifelse(is.na(lfc), 0, lfc)) -> cecum_lfc
```

```{r, fig.width=10, fig.height=7}
cecum_lfc %>% 
    ggplot(aes(x = component, y = taxon)) +
    geom_tile(aes(fill = lfc), color = 'black') +
    scale_fill_gradient2(low = 'blue', high = 'red') +
    geom_text(aes(label = lfc), size = 3) +
    theme_bw(base_size = 16) +
    # scale_x_discrete(labels = x_labels) +
    # facet_grid(~order,
    #            labeller = labeller(order = facet_labs)) +
    theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          legend.title = element_blank(),
          plot.title = element_text(face = "bold")) 
```


# **ANCOMBC Stratified by Location between Dietary Components**
This is scaled to the ratio of fluorescence seen per location. I have no idea if it was okay that I did this but it allowed me to do the analysis this exact way. Otherwise, it would give me the exact same values for everything. 


**Attempting ANCOMBC**
```{r, warning=FALSE}
hypox_res <- ancom_diet_for_loop(otu_table_FP,
                                tax_FP,
                                hypox_metadata,
                                'Family',
                                'high_fat_scaled + high_fiber_scaled + purified_diet_scaled',
                                'BH',
                                strat_value)
```

**Data Wrangling for ANCOMBC Plot**
```{r}
hypox_res %>% 
  filter(diff_high_fat_scaled == TRUE | diff_high_fiber_scaled == TRUE | diff_purified_diet_scaled == TRUE) %>% 
  mutate(lfc_high_fat_corr = (lfc_high_fat_scaled * diff_high_fat_scaled),
         lfc_high_fiber_corr = (lfc_high_fiber_scaled * diff_high_fiber_scaled),
         lfc_purified_diet_corr = (lfc_purified_diet_scaled * diff_purified_diet_scaled)) %>% 
  select(taxon, location, contains('corr')) -> hypox_lfc

colnames(hypox_lfc) <- col_names

hypox_lfc %>% 
  gather(-taxon, -location, key = component, value = lfc) %>% 
  mutate(lfc = as.numeric(lfc),
         lfc = signif(lfc, digits = 5)) %>% 
  mutate(order = ifelse(location == 'Cecum', 1, location),
         order = ifelse(order == 'Prox_colon', 2, order),
         order = ifelse(order == 'Dist_colon', 3, order),
         lfc = ifelse(is.na(lfc), 0, lfc)) %>% 
    arrange(order) -> hypox_lfc
```

**Plot**
I think I want to go through and run the ANCOMBC analysis based on location between diets and then smush them together. Or I can run the analysis between locations for each diet and then smush it together for a plot. I like my first suggestion better but we'll see.

Another update: When I try to run this analysis filtered by location between dietary components it gives me the exact same log fold changes for all three locations because each location has the exact same dietary component distribution (so it's not coded any differently and ANCOMBC is running the same analysis three times, regardless of which location you have specified)

Update: I tried to do this by diet but there were too few data points and ANCOMBC got upset with me so that's why it's sectioned out by dietary fiber content instead. 
```{r, fig.width=12, fig.height=5}
hypox_plot <- baseline_plot(hypox_lfc,
                           hypoxia_facet_labs,
                           hypoxia_names_labs,
                           diet_comp_x_labs,
                           'Log Fold Change of Hypoxia\nby Location and Dietary Component')

hypox_plot
```

# **ANCOMBC Between Gut Locations ONLY**

**ANCOMBC**
```{r}
location_res <- run_ancombc(otu_table_FP,
                            tax_FP,
                            hypox_metadata_long,
                            'Family',
                            'Cecum + Prox_colon + Dist_colon',
                            'BH')

location_ancom <- location_res$ANCOMBCResults
```

**Data Wrangling**
```{r}
location_ancom %>% 
  filter(diff_Cecum == TRUE | diff_Prox_colon == TRUE | diff_Dist_colon == TRUE) %>% 
  mutate(lfc_cecum_corr = (lfc_Cecum * diff_Cecum),
         lfc_prox_colon_corr = (lfc_Prox_colon * diff_Prox_colon),
         lfc_dist_colon_corr = (lfc_Dist_colon * diff_Dist_colon)) %>% 
  select(taxon, contains('corr')) -> location_lfc

colnames(location_lfc) <- c('taxon',
                            'Cecum',
                            'Prox_colon',
                            'Dist_colon')

location_lfc %>% 
  gather(-taxon, key = location, value = lfc) %>% 
  mutate(lfc = as.numeric(lfc),
         lfc = signif(lfc, digits = 5)) %>% 
  mutate(component = ifelse(location == 'Cecum', 1, location),
         component = ifelse(component == 'Prox_colon', 2, component),
         component = ifelse(component == 'Dist_colon', 3, component),
         lfc = ifelse(is.na(lfc), 0, lfc)) %>% 
    arrange(component) -> location_lfc
```

**Plot**
```{r, fig.width=10, fig.height=7}
location_plot <- baseline_plot(location_lfc,
                               NULL,
                               NULL,
                               location_x_labs,
                               'Log Fold Change of Hypoxia by\nLocation (Family Level)')

location_plot
```


**Saving my Outputs**
```{r}
ggsave('../plots/hypoxia_ancombc.pdf',
       plot = test_plot,
       width = 12, 
       height = 7)
ggsave('../plots/hypoxia_dietCompFam_ancombc.pdf',
       plot = diet_comp_plot,
       width = 10, 
       height = 7)
ggsave('../plots/hypoxia_locationFam_ancombc.pdf',
       plot = location_plot,
       width = 10, 
       height = 7)
```

