---
title: "hypoxia"
output: html_document
date: "2023-11-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(cowplot)
library(broom)
library(rstatix)
library(ggpubr)
```

**File Path**
```{r}
hypoxia_fp <- '../data/misc/pimid_fluor.csv'

diet_labs <- c('Chow',
               'High Fat / High Fiber',
               'High Fat / Low Fiber',
               'Low Fat / High Fiber',
               'Low Fat / Low Fiber')
names(diet_labs) <- c('Chow',
                      'HF/HF',
                      'HF/LF',
                      'LF/HF',
                      'LF/LF')

fiber_labs <- c('High Fiber',
                'Low Fiber')
names(fiber_labs) <- c('high_fiber',
                       'low_fiber')

location_labs <- c('Cecum',
                   'Proximal Colon',
                   'Distal Colon')
names(location_labs) <- c(1,
                          2,
                          3)
```

**Reading in Table**
```{r}
hypoxia <- read_csv(hypoxia_fp)

hypoxia %>% 
  mutate(order = ifelse(location == 'Cecum', 1, location),
         order = ifelse(order == 'Prox_colon', 2, order),
         order = ifelse(order == 'Dist_colon', 3, order)) %>% 
  arrange(order) -> hypoxia
```

## **Stats (by diet/fiber content)**

**Mouse Diet**
```{r}
## kruskal test
hypoxia %>% 
  group_by(diet) %>% 
  do(tidy(kruskal.test(fluorescence ~ order,
             data = .))) %>% 
  ungroup() %>% 
  arrange(p.value) %>% 
  mutate(p.adj = p.adjust(p.value,
                          method = "BH"),
        test_id = paste(diet)) -> kruskal

## dunns post hoc test
hypoxia %>% 
  group_by(diet) %>% 
  mutate(test_id = paste(diet)) %>% 
  filter(test_id %in% kruskal$test_id) %>% 
  dunn_test(fluorescence ~ order,
            p.adjust.method = 'BH',
            data = .) %>% 
  add_y_position(scales = 'free_y') -> dunn
```


**Fiber Content**
```{r}
## kruskal test
hypoxia %>% 
  group_by(fiber) %>% 
  do(tidy(kruskal.test(fluorescence ~ order,
             data = .))) %>% 
  ungroup() %>% 
  arrange(p.value) %>% 
  mutate(p.adj = p.adjust(p.value,
                          method = "BH"),
        test_id = paste(fiber)) -> kruskal_fiber

## dunns post hoc test
hypoxia %>% 
  group_by(fiber) %>% 
  mutate(test_id = paste(fiber)) %>% 
  filter(test_id %in% kruskal_fiber$test_id) %>% 
  dunn_test(fluorescence ~ order,
            p.adjust.method = 'BH',
            data = .) %>% 
  add_y_position(scales = 'free_y') -> dunn_fiber
```


## **Plots (faceted by diet/fiber content)**

**Hypoxia Plot by Mouse Diet**
```{r, fig.height=5, fig.width=15}
hypoxia %>% 
  ggplot(aes(x = order, y = fluorescence)) +
  geom_boxplot(aes(group = order)) +
  geom_jitter(alpha = 0.4, width = 0.1, height = 0) +
  # scale_y_log10() +
  scale_x_discrete(labels = c('Cecum',
                              'Proximal\nColon',
                              'Distal\nColon')) +
  theme_bw(base_size = 16) +
  facet_grid(~diet,
             labeller = labeller(diet = diet_labs)) +
  stat_pvalue_manual(dunn,
                     tip.length = 0.01,
                     label = 'p.adj.signif',
                     hide.ns = TRUE) +
  labs(x = 'Intestinal Location',
       y = 'Fluorescence',
       title = 'Mouse GI Oxygen Concentration by Diet') -> hypox_plot

hypox_plot
```

**Hypoxia Plot by Fiber Content**
```{r, fig.height=5, fig.width=8}
hypoxia %>% 
  ggplot(aes(x = order, y = fluorescence)) +
  geom_boxplot(aes(group = order)) +
  geom_jitter(alpha = 0.4, width = 0.1, height = 0) +
  # scale_y_log10() +
  scale_x_discrete(labels = c('Cecum',
                              'Proximal\nColon',
                              'Distal\nColon')) +
  theme_bw(base_size = 16) +
  facet_grid(~fiber,
             labeller = labeller(fiber = fiber_labs)) +
  stat_pvalue_manual(dunn_fiber,
                     tip.length = 0.01,
                     label = 'p.adj.signif',
                     hide.ns = TRUE) +
  labs(x = 'Intestinal Location',
       y = 'Fluorescence',
       title = 'Mouse GI Oxygen Concentration by Fiber Content') -> hypox_fib_plot

hypox_fib_plot
```

## **Stats (by location)**

**Mouse Diet**
```{r}
## kruskal test
hypoxia %>% 
  group_by(order) %>% 
  do(tidy(kruskal.test(fluorescence ~ diet,
             data = .))) %>% 
  ungroup() %>% 
  arrange(p.value) %>% 
  mutate(p.adj = p.adjust(p.value,
                          method = "BH"),
        test_id = paste(order)) -> kruskal_location

## dunns post hoc test
hypoxia %>% 
  group_by(order) %>% 
  mutate(test_id = paste(order)) %>% 
  filter(test_id %in% kruskal_location$test_id) %>% 
  dunn_test(fluorescence ~ diet,
            p.adjust.method = 'BH',
            data = .) %>% 
  add_y_position(scales = 'free_y') -> dunn_location
```


**Fiber Content**
```{r}
## kruskal test
hypoxia %>% 
  group_by(order) %>% 
  do(tidy(kruskal.test(fluorescence ~ fiber,
             data = .))) %>% 
  ungroup() %>% 
  arrange(p.value) %>% 
  mutate(p.adj = p.adjust(p.value,
                          method = "BH"),
        test_id = paste(order)) -> kruskal_fiber_location

## dunns post hoc test
hypoxia %>% 
  group_by(order) %>% 
  mutate(test_id = paste(order)) %>% 
  filter(test_id %in% kruskal_fiber_location$test_id) %>% 
  dunn_test(fluorescence ~ fiber,
            p.adjust.method = 'BH',
            data = .) %>% 
  add_y_position(scales = 'free_y') -> dunn_fiber_location
```

## **Plots (faceted by location)**

**Hypoxia Plot by Mouse Diet**
Overall
```{r, fig.height=5, fig.width=10}
hypoxia %>% 
  ggplot(aes(x = diet, y = fluorescence)) +
  geom_boxplot(aes(group = diet)) +
  geom_jitter(alpha = 0.4, width = 0.1, height = 0) +
  scale_x_discrete(labels = c('Chow',
                              'HFt/\nHFb',
                              'HFt/\nLFb',
                              'LFt/\nHFb',
                              'LFt/\nLFb')) +
  theme_bw(base_size = 16) +
  facet_grid(~order,
             labeller = labeller(order = location_labs)) +
  stat_pvalue_manual(dunn_location,
                     tip.length = 0.01,
                     label = 'p.adj.signif',
                     hide.ns = TRUE) +
  labs(x = 'Diet',
       y = 'Fluorescence',
       title = 'Cecum Hypoxia') -> hypox_location_plot

hypox_location_plot
```


Cecum!!
```{r, fig.height=4, fig.width=6}
hypoxia %>% 
  filter(order == 1) %>% 
  ggplot(aes(x = diet, y = fluorescence)) +
  geom_boxplot(aes(group = diet)) +
  geom_jitter(alpha = 0.4, width = 0.1, height = 0) +
  scale_x_discrete(labels = c('Chow',
                              'HFt/\nHFb',
                              'HFt/\nLFb',
                              'LFt/\nHFb',
                              'LFt/\nLFb')) +
  theme_bw(base_size = 16) +
  facet_grid(~order,
             labeller = labeller(order = location_labs)) +
  stat_pvalue_manual(dunn_location,
                     tip.length = 0.01,
                     label = 'p.adj.signif',
                     hide.ns = TRUE) +
  labs(x = 'Diet',
       y = 'Fluorescence',
       title = 'Cecum Hypoxia') -> hypox_cecumLocation_plot

hypox_cecumLocation_plot
```

Colon!!
No statistical significance so that's why no stats on plot
```{r, fig.height=4, fig.width=8}
hypoxia %>% 
  filter(order != 1) %>% 
  ggplot(aes(x = diet, y = fluorescence)) +
  geom_boxplot(aes(group = diet)) +
  geom_jitter(alpha = 0.4, width = 0.1, height = 0) +
  scale_x_discrete(labels = c('Chow',
                              'HFt/\nHFb',
                              'HFt/\nLFb',
                              'LFt/\nHFb',
                              'LFt/\nLFb')) +
  theme_bw(base_size = 16) +
  facet_grid(~order,
             labeller = labeller(order = location_labs)) +
  labs(x = 'Diet',
       y = 'Fluorescence',
       title = 'Colon Hypoxia') -> hypox_colonLocation_plot

hypox_colonLocation_plot
```

**Hypoxia Plot by Fiber Content**
```{r, fig.height=4.5, fig.width=8}
hypoxia %>% 
  ggplot(aes(x = fiber, y = fluorescence)) +
  geom_boxplot(aes(group = fiber)) +
  geom_jitter(alpha = 0.4, width = 0.1, height = 0) +
  scale_x_discrete(labels = c('High\nFiber',
                              'Low\nFiber')) +
  theme_bw(base_size = 16) +
  facet_grid(~order,
             labeller = labeller(order = location_labs)) +
  stat_pvalue_manual(dunn_fiber_location,
                     tip.length = 0.01,
                     label = 'p.adj.signif',
                     hide.ns = TRUE) +
  labs(x = 'Fiber Content',
       y = 'Fluorescence',
       title = 'Mouse GI Oxygen Concentration by Location') -> location_fib_plot

location_fib_plot
```

**Saving my Outputs**
```{r}
## as a pdf
ggsave('../plots/hypoxia_by_diet.pdf',
       plot = hypox_location_plot,
       width = 10, 
       height = 5)
ggsave('../plots/hypoxia_by_fiberContent.pdf',
       plot = location_fib_plot,
       width = 8, 
       height = 4.5)
ggsave('../plots/hypoxia_by_location.pdf',
       plot = hypox_plot,
       width = 15, 
       height = 5)

## as an rdata object
save(hypox_location_plot,
     file = '../../figures/hypoxia_by_diet.rdata')
save(location_fib_plot,
     file = '../../figures/hypoxia_by_fiberContent.rdata')
save(hypox_plot,
     file = '../../figures/hypoxia_by_location.rdata')
save(hypox_cecumLocation_plot,
     file = '../../figures/hypoxia_cecumONLY.rdata')
save(hypox_colonLocation_plot,
     file = '../../figures/hypoxia_colonONLY.rdata')

## stats
write_tsv(dunn_location,
          '../stats/hypoxia_diet_dunn.tsv')
write_tsv(dunn_fiber_location,
          '../stats/hypoxia_fiberContent_dunn.tsv')
```

