---
title: "metab_toxin_comp"
output: html_document
date: "2023-09-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(broom)
library(cowplot)
library(magrittr)
library(qiime2R)
library(tidyverse)
library(naniar)
library(ggpubr)
library(rstatix)
library(viridis)
```

**File Prep and Plot Functions**
```{r}
file_prep <- function(metadata_fp,
                      metab_fp,
                      metab_col_filter,
                      metab_filter,
                      toxin_fp){
  ## metadata file
  metadata <- read_tsv(metadata_fp) %>% 
              select(!c(tube_num, date, corr_sample_num))
  # names(metadata)[names(metadata) == '#SampleID'] <- 'sampleid'
  ## metabolomics file
  metab <- read_csv(metab_fp)
  metadata %>% 
    merge(metab, by = 'mouse_id') %>% 
    select(-(all_of(unwanted_columns))) %>% 
    gather(wanted_metabs, key = metabolite, value = concentration) %>% 
    filter(!is.na(mouse_id)) -> pre_metab
  ## changes all 'ND' values in the concentration column to 0 
  pre_metab$concentration[pre_metab$concentration == 'ND'] <- 0
  pre_metab %>% 
    filter(!is.na(concentration)) %>% 
    mutate(concentration = as.numeric(concentration)) -> big_metab
  ## toxin file
  toxin <- read_tsv(toxin_fp)
  ## toxin neat concentrations (non_diluted)
  toxin %>% 
    merge(metadata, by = 'mouse_id') %>% 
    select(!c('Total TcA 1:10', 'Total TcB 1:10')) %>% 
    gather('Total TcA Neat', 'Total TcB Neat', 
           key = neat_toxin, value = neat_conc) -> pre_neat_toxin
  pre_neat_toxin$neat_conc[pre_neat_toxin$neat_conc == 'BDL'] <- '0'
  pre_neat_toxin %>% 
    mutate(neat_conc = as.numeric(neat_conc)) %>% 
    select(!c('Tube_Label',
              'Collection Date',
              'Sample_Type',
              'Extra_Sample')) -> neat_toxin
  ## toxin diluted concentrations 
  toxin %>% 
    merge(metadata, by = 'mouse_id') %>% 
    select(!c('Total TcA Neat', 'Total TcB Neat')) %>% 
    gather('Total TcA 1:10', 'Total TcB 1:10',
           key = dil_toxin, value = dil_conc) -> pre_dil_toxin
  pre_dil_toxin$dil_conc[pre_dil_toxin$dil_conc == 'BDL'] <- '0'
  pre_dil_toxin$dil_conc[pre_dil_toxin$dil_conc == 'Chow'] <- '0'
  pre_dil_toxin %>% 
    mutate(dil_conc = as.numeric(dil_conc)) %>% 
    filter(diet != 'Chow') %>% 
    select(!c('Tube_Label',
              'Collection Date',
              'Sample_Type',
              'Extra_Sample')) -> dil_toxin
  ## creating a list of my outputs
  my_list <- list(Metadata = metadata,
                  Metabolomics = big_metab,
                  NeatToxin = neat_toxin,
                  DilToxin = dil_toxin)
  return(my_list)
}

## histo plot comparison function
comp_plot <- function(metab_table,
                      ## toxin or metabolomics?
                       wanted_table,
                       ## this is what you want your x axis to be (neat, diluted, or concentration)? 
                       wanted_x,
                       ## this is toxin/metabolite column you want to filter by (neat or diluted)?
                       wanted_y,
                       ## what is the toxin/metabolite you want to filter by called in the table?
                       wanted_y_name,
                       diet_labels,
                       tox_labels,
                       title){
  ## putting tables together
  metab_table %>% 
    select(mouse_id, metabolite, log_conc, concentration) %>% 
    left_join(wanted_table, by = 'mouse_id') %>% 
    na.omit() -> compiled_table
  ## constructing plot
  compiled_table %>% 
    filter(.data[[wanted_y]] == wanted_y_name) %>% 
    na.omit() %>% 
    ggplot(aes(x = .data[[wanted_x]], y = concentration)) +
    scale_x_continuous(trans = 'log10') +
    scale_y_continuous(trans = 'log10') +
    geom_jitter(alpha = 0.4, width = 0.1, height = 0) +
    geom_smooth(method = 'lm', se = FALSE) +
    facet_grid(neat_toxin~diet,
               labeller = labeller(neat_toxin = tox_labels,
                                   diet = diet_labels),
               scales = 'free_y') +
    theme_bw() +
    theme(strip.text.y = element_text(angle = 0)) +
    xlab('Toxin Concentration') +
    ylab('Metabolite Concentration') +
    ggtitle(title) -> plot
  return(plot)
}
```

**Statisitical Analysis Functions**
```{r}
## runs linear modeling on the table, diet, location, and toxin/metabolite wanted and returns full results 
## along with a table containing the r2 and p-value 
## this goes into the following function with a for loop 
r2_value <- function(biom_table,
                     wanted_diet,
                     wanted_metab,
                     tox_col,
                     wanted_tox,
                     wanted_conc_value){
  ## to get r squared value and p value
  biom_table %>% 
    filter(diet == wanted_diet,
           metabolite == wanted_metab,
           .data[[tox_col]] == wanted_tox) %>% 
    do(glance(lm(concentration ~ .data[[wanted_conc_value]],
                 data = .))) -> lm_glance
  output <- list(r_squared = lm_glance$r.squared,
                 p_value = lm_glance$p.value)
  ## to get estimate and place it with the overall output table
  biom_table %>% 
    filter(diet == wanted_diet,
           metabolite == wanted_metab,
           .data[[tox_col]] == wanted_tox) %>% 
    do(tidy(lm(concentration ~ .data[[wanted_conc_value]],
                 data = .))) %>% 
    filter(term != '(Intercept)') -> lm_full
  output <- as_tibble(output) %>%
            mutate(diet = paste(wanted_diet),
                   metabolite = paste(wanted_metab),
                   toxin = paste(wanted_tox),
                   just_r = sqrt(r_squared),
                   estimate = lm_full$estimate,
                   r_direct = if_else(estimate < 0, just_r * -1, just_r))
  my_list <- list(LinearModel = lm_full,
                  WantedStats = output)
  return(my_list)
}

## creating a nested for loop since running this function individually is a nightmare
r2_value_for_loop <- function(biom_table,
                              wanted_col,
                              wanted_conc) {
  ## nested for loop for the r squared and p values 
  output <- tibble()
  for(i in unique(unlist(biom_table$diet))) {
    for(j in unique(unlist(biom_table$metabolite))) {
      for(k in unique(unlist(biom_table[wanted_col]))) {
        tmp_output <- r2_value(biom_table,
                               i,
                               j,
                               wanted_col,
                               k,
                               wanted_conc)$WantedStats
        output <- bind_rows(output, tmp_output)
      }
    }
  }
  return(output)
}

## statistical visualization plot function
stat_plot <- function(stat_table,
                      wanted_labels,
                      title){
  stat_table %>% 
    ggplot(aes(x = toxin, y = diet)) +
    geom_tile(aes(fill = r_direct), color = 'black', alpha = 0.6) +
    geom_text(aes(label = signif)) +
    scale_fill_gradient2(low = 'blue', high = 'green', name = "Correlation\nCoefficient") +
    scale_x_discrete(labels = c("TcdA",
                                "TcdB")) +
    scale_y_discrete(labels = c("Chow",
                                "HFt/HFb",
                                "HFt/LFb",
                                "LFt/HFb",
                                "LFt/LFb")) +
    facet_wrap(~metabolite,
               labeller = labeller(metabolite = wanted_labels)) +
    theme_bw() +
    xlab("Toxin") +
    ylab("Diet") +
    ggtitle(title) -> plot
  return(plot)
  }
```

**File Paths**
```{r}
metadata_FP <- '../data/misc/filt_cecal_processed_metadata.tsv'
metab_FP <- '../data/misc/metabolomics.csv'
toxin_FP <- '../data/misc/toxin_final_data.tsv'

## lists to redo the diet names on the facet labels of the ggplot created below 
diet_labs <- 
    c('Chow', 
      'High Fat / High Fiber', 
      'High Fat / Low Fiber', 
      'Low Fat / High Fiber', 
      'Low Fat / Low Fiber')

names(diet_labs) <- c('Chow', 'HF/HF', 'HF/LF', 'LF/HF', 'LF/LF')

wanted_metabs <- c('Acetic Acid (ug/g)',
                   'Propanoic Acid (ug/g)',
                   'n-Butanoic Acid (ug/g)')
unwanted_columns <- c('2-methyl-propanoic acid (ug/g)',
                     'Isopentanoic Acid (ug/g)',
                     '2-methyl-Butanoic Acid (ug/g)',
                     'Pentanoic Acid (ug/g)',
                     'Notes',
                     'Sample Group',
                     'SCFA Data File',
                     'Acq. Date-Time',
                     'Tube_Label',
                     'Sample_Type',
                     'Collection Date',
                     'Dil.')
metab_labs <- c('Acetic Acid',
                'Propanoic Acid',
                'n-Butanoic Acid')
names(metab_labs) <- wanted_metabs
neat_labs <- c('TcdA', 'TcdB')
names(neat_labs) <- c('Total TcA Neat', 'Total TcB Neat')

dil_labs <- c('TcdA', 'TcdB')
names(dil_labs) <- c('Total TcA 1:10', 'Total TcB 1:10')
```

**File Prep**
```{r}
comp_files <- file_prep(metadata_FP,
                        metab_FP,
                        unwanted_columns,
                        wanted_metabs,
                        toxin_FP)

metadata <- comp_files$Metadata
metab <- comp_files$Metabolomics
dil_toxin <- comp_files$DilToxin
neat_toxin <- comp_files$NeatToxin
```

YOURE PLOTTING/DOING STATS ON THE LOG10 TRANSFORMATION IN THE ACTUAL BIOM TABLE RATHER THAN TRANSFORMING THE AXES!!
This is inconsistent with what you've done so far with all of the other plots - you may want to change this back. Need further thought/information into what exactly you're doing and whether you should change things or not. 
```{r}
metab %>% 
  mutate(log_conc = log(concentration + 0.01)) -> metab

neat_toxin %>% 
  mutate(log_neat_conc = log(neat_conc + 0.001)) -> neat_toxin
```


**Comparison Plots**
Acetic Acid
```{r, warning=FALSE, fig.height=4, fig.width=10}
tox_acetic <- comp_plot(metab,
                        neat_toxin,
                        "neat_conc",
                        "metabolite",
                        "Acetic Acid (ug/g)",
                        diet_labs,
                        neat_labs,
                        "Acetic Acid")
tox_acetic
```

n-Butanoic Acid
```{r, warning=FALSE, fig.height=4, fig.width=10}
tox_butanoic <- comp_plot(metab,
                        neat_toxin,
                        "neat_conc",
                        "metabolite",
                        "n-Butanoic Acid (ug/g)",
                        diet_labs,
                        neat_labs,
                        "n-Butanoic Acid")
tox_butanoic
```

Propanoic Acid 
```{r, warning=FALSE, fig.height=4, fig.width=10}
tox_propanoic <- comp_plot(metab,
                            neat_toxin,
                            "neat_conc",
                            "metabolite",
                            "Propanoic Acid (ug/g)",
                            diet_labs,
                            neat_labs,
                            "Propanoic Acid")
tox_propanoic
```

**Statistical Analysis**
```{r}
metab %>% 
    select(mouse_id, metabolite, log_conc, concentration) %>% 
    left_join(neat_toxin, by = 'mouse_id') %>% 
    na.omit() %>% 
    ungroup() -> metab_tox

stat_table <- r2_value_for_loop(metab_tox,
                                "neat_toxin",
                                "neat_conc")
## assigning p value significance
stat_table['signif'] <- symnum(stat_table$p_value,
                               cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
                               symbols = c("****", "***", "**", "*", "+", "ns"),
                               abbr.colnames = FALSE,
                               na = "")
```


**Statistical Visualization**
```{r, fig.height=3, fig.width=5.5}
stat_plot(stat_table,
          metab_labs,
          "Metabolite and Toxin Concentration Comparisons") -> tox_metab_stat_vis

tox_metab_stat_vis
```

**Putting my Plots Together**
```{r, warning=FALSE, fig.height=7, fig.width=16}
plot_grid(tox_acetic, tox_butanoic,
          tox_propanoic, tox_metab_stat_vis,
          nrow = 2,
          labels = c('a)', 'b)', 'c)', 'd)')) -> metab_tox_plots

metab_tox_plots
```

**Saving my Outputs**
```{r}
## as a pdf
ggsave('metab_tox_comp.pdf',
       plot = metab_tox_plots,
       width = 16,
       height = 7,
       path = '../plots')

## as an rdata object
save(tox_acetic,
     file = '../../figures/tox_metab_acetic.rdata')
save(tox_butanoic,
     file = '../../figures/tox_metab_butanoic.rdata')
save(tox_propanoic,
     file = '../../figures/tox_metab_propanoic.rdata')
save(tox_metab_stat_vis,
     file = '../../figures/tox_metab_stats.rdata')
```

