---
title: "microbiome_cecum_comp"
output: html_document
date: "2024-10-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggpubr)
library(ggplot2)
library(magrittr)
library(qiime2R)
library(tidyverse)
library(broom)
library(cowplot)
library(vegan)
library(viridis)
library(apppleplots)
```

**Functions**
```{r}
## rel abun file prep
rel_abun_file_prep <- function(metadata_fp,
                               tax_fp,
                               otu_table_fp,
                               tax_level,
                               wanted_tax){
  ## metadata
  metadata <- read_tsv(metadata_fp)
  ## taxonomy
  taxonomy <- read_qza(tax_fp)$data %>% 
    parse_taxonomy() %>% 
    rownames_to_column('asv')
  ## otu table 
  otu_table <- read_qza(otu_table_fp)$data
  otu_table %>% 
    as_tibble(rownames = 'asv') %>% 
    gather(-asv, key = sampleid, value = abun) %>% 
    group_by(sampleid) %>% 
    mutate(rel_abun = abun/sum(abun)) %>% 
    mutate(rel_abun = rel_abun + 0.000001) -> otu_table
  ## joining all tables together 
  otu_table %>% 
    left_join(metadata, by = 'sampleid') %>% 
    left_join(taxonomy, by = 'asv') -> abun_table
  abun_table %>% 
    group_by(sampleid, day_post_inf, diet, mouse_id, 
             purified_diet, high_fat, high_fiber, 
             seq_depth, .data[[tax_level]], facility, study) %>% 
    summarise(rel_abund = sum(rel_abun)) %>% 
    filter(.data[[tax_level]] %in% wanted_tax) %>% 
    mutate(mouse_fact = as.factor(mouse_id),
           day_fact = as.factor(day_post_inf)) -> abun_filt
  ## creating a list for my outputs
  my_list <- list(Metadata = metadata,
                  Taxonomy = taxonomy,
                  OTUTable = otu_table,
                  AbundanceTable = abun_filt)
  return(my_list)
}

## will give the r2 and p values for each value in the specified column 
lm_r2_p_value <- function(input_table,
                          wanted_col1,
                          wanted_col2 = NULL,
                          formula_left,
                          formula_right){
  if (is.character(wanted_col1) & is.character(wanted_col2)) {
    output <- tibble()
    for(i in unique(unlist(input_table[wanted_col1]))) {
      for(j in unique(unlist(input_table[wanted_col2]))) {
        funky_formula <- paste(formula_left, formula_right, sep = "~")
        lm <- input_table %>% 
                filter(.data[[wanted_col1]] == i,
                       .data[[wanted_col2]] == j) %>%
                do(glance(lm(as.formula(funky_formula),
                             data = .))) %>% 
              mutate("{wanted_col1}" := paste(i),
                     "{wanted_col2}" := paste(j))
        output <- bind_rows(output, lm)
        
        stat_table <- output %>% 
          select(r.squared, p.value, .data[[wanted_col1]], .data[[wanted_col2]]) %>% 
          rename(r_squared = r.squared,
                 p_value = p.value) %>% 
          mutate(p_value = if_else(is.na(p_value), 1, p_value))
      }
    }
    
  } else {
    output <- tibble()
    for(i in unique(unlist(input_table[wanted_col1]))){
        funky_formula <- paste(formula_left, formula_right, sep = "~")
        lm <- input_table %>% 
                filter(.data[[wanted_col1]] == i) %>%
                do(glance(lm(as.formula(funky_formula),
                             data = .))) %>% 
              mutate("{wanted_col1}" := paste(i))
        output <- bind_rows(output, lm)
        
        ## putting together actual stat table 
        stat_table <- output %>% 
          select(r.squared, p.value, .data[[wanted_col1]]) %>% 
          rename(r_squared = r.squared,
                 p_value = p.value) %>% 
          mutate(p_value = if_else(is.na(p_value), 1, p_value))
    }
  }
  
  stat_table['signif'] <- symnum(stat_table$p_value,
                                 cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1),
                                 symbols = c("****", "***", "**", "*", "ns"),
                                 abbr.colnames = FALSE,
                                 na = "")
  return(stat_table)
}



## putting stats directly on correlation plots 
stat_on_plot <- function(stat_table,
                         wanted_col_list,
                         x_col_name,
                         x_col_value,
                         y_col_name,
                         y_col_value,
                         plot,
                         plot_text_size){
  ## making table conducive to labeling plot 
  proc_stat_table <- stat_table %>% 
    filter(signif != 'ns') %>% 
    select(all_of(wanted_col_list)) %>% 
    ## created columns need to match up with the column names called for the x and y axis in the plot
    ## numeric values added are the x and y coordinates for where the label will sit in the facet block 
    mutate("{x_col_name}" := paste(x_col_value),
           "{x_col_name}" := as.numeric(.data[[x_col_name]]),
           "{y_col_name}" := paste(y_col_value),
           "{y_col_name}" := as.numeric(.data[[y_col_name]]),
           p_value = signif(p_value, digits = 3),
           r_squared = signif(r_squared, digits = 3),
           pVal_rSquare = paste0('p-value: ', p_value, "\n", 
                                 'r2: ', r_squared))
  
  ## putting labels on plot 
  plot_wStats <- plot +
    geom_label(data = proc_stat_table, 
               label = proc_stat_table$pVal_rSquare,
               color = 'black',
               fontface = 'bold',
               size = plot_text_size)
  
  ## creating list of my outputs
  my_list <- list(ProcStatTable = proc_stat_table,
                  PlotwStats = plot_wStats)
  
  return(my_list)
}
```

**File paths**
bile acids:
- look at DCA against alpha diversity at day 3 
metabolites:
- look at butyrate against alpha diversity at day 3

- FOR PLOTS:
  - don't facet by diet but facet by high/low fiber diets and chow to give us slightly more power 
```{r}
metadata_FP <- '../data/misc/filt_cecal_processed_metadata.tsv'
faith_pd_FP <- '../data/core_outputs/faith_pd.tsv'
otu_table_FP <- '../data/cecal_qiime/tax_filt_actual.qza'
tax_FP <- '../data/cecal_qiime/taxonomy.qza'
bile_acid_FP <- '../data/misc/processed_bile_acid.tsv'
bile_ratio_FP <- '../data/misc/processed_ratio_bileAcid.tsv'
metab_FP <- '../data/misc/processed_metabolomics.tsv'


diet_labs <- 
  c('Chow', 
    'High Fat / High Fiber', 
    'High Fat / Low Fiber', 
    'Low Fat / High Fiber', 
    'Low Fat / Low Fiber')
names(diet_labs) <- c('Chow', 
                      'HF/HF', 
                      'HF/LF', 
                      'LF/HF', 
                      'LF/LF')

diet_content_labs <- c('Chow',
                       'High Fiber Diets',
                       'Low Fiber Diets')
names(diet_content_labs) <- c('chow',
                              'high_fiber',
                              'low_fiber')

faith_comb_labs <- c('Deoxycholic Acid (pmol/mg)',
                     'Butyrate (ug/g)')
names(faith_comb_labs) <- c('Deoxycholic_acid_DCA',
                            'n-Butanoic Acid (ug/g)')

microbe_labs <- c('Lachnospiraceae',
                  'Ruminococcaceae',
                  'Peptostreptococcaceae',
                  'Lactobacillaceae',
                  'Enterobacteriaceae',
                  'Enterococcaceae',
                  'Staphylococcaceae')
names(microbe_labs) <- c(1,
                         2,
                         3,
                         4,
                         5,
                         6,
                         7)


wanted_level <- 'Family'
wanted_family <- c('Enterobacteriaceae', 
                   'Lactobacillaceae', 
                   'Lachnospiraceae', 
                   'Enterococcaceae',
                   'Staphylococcaceae', 
                   'Tannerellaceae', 
                   'Muribaculaceae', 
                   'Bacteroidaceae', 
                   'Marinifilaceae', 
                   'Ruminococcaceae',
                   'Morganellaceae',
                   'Peptostreptococcaceae')
diet_x_labs <- c("Chow",
                 "HFt/\nHFb",
                 "HFt/\nLFb",
                 "LFt/\nHFb",
                 "LFt/\nLFb")
```

# **Alpha Diversity Comparisons**

**Joining Faith's PD with bile acid and metabolomics data**
```{r}
## reading in files
faith <- read_tsv(faith_pd_FP) %>% 
            rename('sampleid' = '#SampleID')

sample_mouseID_table <- read_tsv(metadata_FP) %>% 
  select(mouse_id, sampleid)

full_metab <- read_tsv(metab_FP) %>% 
  select(mouse_id, diet, sampleid, metabolite, concentration) %>% 
  mutate(metab_conc_norm = (concentration + 2)) %>% 
  mutate(dietary_content = case_when(
    diet == 'HF/HF' | diet == 'LF/HF' ~ 'high_fiber',
    diet == 'Chow' ~ 'chow',
    diet == 'HF/LF' | diet == 'LF/LF' ~ 'low_fiber'
  ))

full_bile_acid <- read_tsv(bile_acid_FP) %>% 
  select(diet, mouse_id, bile_acid, concentration, c_diff_effect, conc_normalized) %>% 
  mutate(dietary_content = case_when(
    diet == 'HF/HF' | diet == 'LF/HF' ~ 'high_fiber',
    diet == 'Chow' ~ 'chow',
    diet == 'HF/LF' | diet == 'LF/LF' ~ 'low_fiber'
  ))

## joining
butyrate_faith_table <- full_metab %>% 
  left_join(faith, by = 'sampleid') %>% 
  filter(metabolite == 'n-Butanoic Acid (ug/g)')

dca_faith_table <- full_bile_acid %>% 
  left_join(sample_mouseID_table, by = 'mouse_id') %>% 
  left_join(faith, by = 'sampleid') %>% 
  na.omit() %>% 
  filter(bile_acid == 'Deoxycholic_acid_DCA',
         mouse_id != 106) ## filtered out a LF/LF mouse that had an outlier concentration of DCA 

## putting the two faith tables together to see if they work in one visualization
miniDCA_faith_table <- dca_faith_table %>% 
  select(-c_diff_effect) %>% 
  rename(microbe_product = bile_acid)

miniButyrate_faith_table <- butyrate_faith_table %>% 
  rename(microbe_product = metabolite,
         conc_normalized = metab_conc_norm)

microbeProd_faith_comb <- rbind(miniDCA_faith_table,
                                miniButyrate_faith_table)
```

**Alpha Diversity Compared to Butyrate and DCA Concentration**
secondary bile acid: 
  - the outlier point in the low fiber facet is a LF/LF mouse that has a DCA concentration of ~65 pmol/mg (all    the others have a DCA concentration of 0) - mouse_id is 106!!
  - i want to filter that point out
```{r, fig.width=9.5, fig.height=4.5}
faith_comb_plot <- apppleplots::correlation_plots(input_table = microbeProd_faith_comb,
                                                 x_value = 'conc_normalized',
                                                 y_value = 'faith_pd',
                                                 x_transform = TRUE,
                                                 y_transform = TRUE,
                                                 point_alpha = 0,
                                                 regression_method = "lm",
                                                 facet_rows = NULL,
                                                 facet_cols = 'microbe_product',
                                                 row_labs = NULL,
                                                 col_labs = faith_comb_labs,
                                                 x_name = 'Microbiome Product Concentration (log10)',
                                                 y_name = "Faith's PD (log10)",
                                                 title_content = 'Cecum Microbiome Products by Alpha Diversity')

faith_comb_plot <- faith_comb_plot +
                    geom_jitter(aes(fill = diet), pch = 21, alpha = 0.7, size = 3) +
                    scale_fill_brewer(palette = 'Spectral',
                                      name = 'Mouse Diet',
                                      labels = c('Chow',
                                                 'HFt / HFb',
                                                 'HFt / LFb',
                                                 'LFt / HFb',
                                                 'LFt / LFb')) +
                    theme_bw(base_size = 20)

## stats
faith_comb_stats <- lm_r2_p_value(input_table = microbeProd_faith_comb,
                                  wanted_col1 = 'microbe_product',
                                  wanted_col2 = NULL,
                                  formula_left = 'faith_pd',
                                  formula_right = 'conc_normalized * diet')


## stats on plot
faith_comb_list <- stat_on_plot(stat_table = faith_comb_stats,
                               wanted_col_list = colnames(faith_comb_stats),
                               x_col_name = 'conc_normalized',
                               x_col_value = '200',
                               y_col_name = 'faith_pd',
                               y_col_value = '4',
                               plot = faith_comb_plot,
                               plot_text_size = 3.5)

faith_combStats_plot <- faith_comb_list$PlotwStats


faith_combStats_plot
```



# **Microbe Relative Abundance Comparisons**

**Data wrangling**
```{r}
rel_abun_files <- rel_abun_file_prep(metadata_fp = metadata_FP,
                                     tax_fp = tax_FP,
                                     otu_table_fp = otu_table_FP,
                                     tax_level = 'Family',
                                     wanted_tax = wanted_family)

rel_abun_table <- rel_abun_files$AbundanceTable %>% 
                    ungroup() %>% 
                    select(sampleid, Family, rel_abund)

relAbun_butyrate_table <- rel_abun_table %>% 
  left_join(full_metab, by = 'sampleid',
            relationship = "many-to-many") %>% 
  filter(metabolite == 'n-Butanoic Acid (ug/g)') %>% 
  mutate(dietary_content = case_when(
    diet == 'HF/HF' | diet == 'LF/HF' ~ 'high_fiber',
    diet == 'Chow' ~ 'chow',
    diet == 'HF/LF' | diet == 'LF/LF' ~ 'low_fiber'
  ))

## microbe product combined table
microbeProd_relAbun_comb <- microbeProd_faith_comb %>% 
                              select(-faith_pd) %>% 
                              left_join(rel_abun_table, by = 'sampleid',
                                        relationship = 'many-to-many')

mini_family <- c('Enterobacteriaceae', 
                 'Lactobacillaceae', 
                 'Lachnospiraceae', 
                 'Enterococcaceae',
                 'Ruminococcaceae',
                 'Staphylococcaceae',
                 'Peptostreptococcaceae')

mini_microbeProd_relAbun_comb <- microbeProd_relAbun_comb %>% 
  filter(Family %in% mini_family) %>% 
  mutate(microbe_order = case_when(
    Family == 'Lachnospiraceae' ~ 1,
    Family == 'Ruminococcaceae' ~ 2,
    Family == 'Peptostreptococcaceae' ~ 3,
    Family == 'Lactobacillaceae' ~ 4,
    Family == 'Enterobacteriaceae' ~ 5,
    Family == 'Enterococcaceae' ~ 6,
    Family == 'Staphylococcaceae' ~ 7
  ))
```


**Correlation comparison plot w stats**
- added five regression lines (one for each diet) instead of just one line per facet - im not sure how I feel about it 
```{r, fig.height=7, fig.width=25}
microbeProd_relAbun_plot <- apppleplots::correlation_plots(input_table = mini_microbeProd_relAbun_comb,
                                                       x_value = 'conc_normalized',
                                                       y_value = 'rel_abund',
                                                       x_transform = TRUE,
                                                       y_transform = TRUE,
                                                       point_alpha = 0,
                                                       regression_method = "none",
                                                       facet_rows = 'microbe_product',
                                                       facet_cols = 'microbe_order',
                                                       row_labs = faith_comb_labs,
                                                       col_labs = microbe_labs,
                                                       x_name = 'Concentration (log10)',
                                                       y_name = "Relative Abundance (log10)",
                                                       title_content = 'Microbe Relative Abundance by Product Concentration')

microbeProd_relAbun_plot <- microbeProd_relAbun_plot +
                    ## this adds the black outline to the regression lines so you can see them better 
                    geom_smooth(aes(group = diet), color = 'black', 
                                linewidth = 1.5, method = 'lm', se = FALSE) +
                    ## this is the actual regression lines 
                    geom_smooth(aes(group = diet, color = diet), method = 'lm', se = FALSE) +
                    scale_color_brewer(palette = 'Spectral',
                                       guide = 'none') +
                    geom_jitter(aes(fill = diet), pch = 21, alpha = 0.6, size = 4) +
                    scale_fill_brewer(palette = 'Spectral',
                                      name = 'Mouse Diet',
                                      labels = c('Chow',
                                                 'HFt / HFb',
                                                 'HFt / LFb',
                                                 'LFt / HFb',
                                                 'LFt / LFb')) +
                    theme_bw(base_size = 20) +
                    theme(strip.text.y.right = element_text(angle = 0)) +
                    ## changed y-axis limits so I could fit the stat label in all facet blocks
                    expand_limits(y = c(0, 1000))


## stats
microbeProd_relAbun_stats <- lm_r2_p_value(input_table = mini_microbeProd_relAbun_comb,
                                       wanted_col1 = 'microbe_product',
                                       wanted_col2 = 'microbe_order',
                                       formula_left = 'rel_abund',
                                       formula_right = 'conc_normalized * diet')

## stats on plot
microbeProd_relAbun_list <- stat_on_plot(stat_table = microbeProd_relAbun_stats,
                                     wanted_col_list = colnames(microbeProd_relAbun_stats),
                                     x_col_name = 'conc_normalized',
                                     x_col_value = '15',
                                     y_col_name = 'rel_abund',
                                     y_col_value = '100',
                                     plot = microbeProd_relAbun_plot,
                                     plot_text_size = 4)

microbeProd_relAbunStats_plot <- microbeProd_relAbun_list$PlotwStats

microbeProd_relAbunStats_plot
```

# **Saving my Outputs**
```{r}
ggsave('../plots/alphaDiv_microbeProd_plot.pdf',
       plot = faith_combStats_plot,
       width = 9.5,
       height = 4.5)
ggsave('../plots/relAbun_microbeProd_plot.pdf',
       plot = microbeProd_relAbunStats_plot,
       width = 25,
       height = 7)

## as an rdat object
save(faith_combStats_plot,
     file = '../../figures/alphaDiv_microbeProd_plot.rdat')
save(microbeProd_relAbunStats_plot,
     file = '../../figures/relAbun_microbeProd_plot.rdat')
```

