---
title: "toxin_metab_histo_comp"
output: html_document
date: "2023-09-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(broom)
library(cowplot)
library(magrittr)
library(qiime2R)
library(tidyverse)
library(naniar)
library(ggpubr)
library(rstatix)
library(viridis)
library(apppleplots)
```

**Functions**
```{r}
## runs linear modeling on the table, diet, location, and toxin/metabolite wanted and returns full results
## along with a table containing the r2 and p-value 
## this goes into the following function with a for loop 
r2_value <- function(biom_table,
                     wanted_diet,
                     wanted_tissue,
                     tox_metab_col,
                     wanted_tox_metab,
                     wanted_conc_value){
  ## to get r squared value and p value
  biom_table %>% 
    filter(diet == wanted_diet,
           tissue == wanted_tissue,
           .data[[tox_metab_col]] == wanted_tox_metab) %>%
    do(glance(lm(score ~ .data[[wanted_conc_value]],
                 data = .))) -> lm_glance
  output <- list(r_squared = lm_glance$r.squared,
                 p_value = lm_glance$p.value)
  ## to get estimate and place it with the overall output table
  biom_table %>% 
    filter(diet == wanted_diet,
           tissue == wanted_tissue,
           .data[[tox_metab_col]] == wanted_tox_metab) %>%
    do(tidy(lm(score ~ .data[[wanted_conc_value]],
                 data = .))) %>% 
    filter(term != '(Intercept)') -> lm_full
  output <- as_tibble(output) %>%
            mutate(diet = paste(wanted_diet),
                   tissue = paste(wanted_tissue),
                   tox_metab = paste(wanted_tox_metab),
                   just_r = sqrt(r_squared),
                   estimate = lm_full$estimate,
                   r_direct = if_else(estimate < 0, just_r * -1, just_r))
  my_list <- list(LinearModel = lm_full,
                  WantedStats = output)
  return(my_list)
}

## creating a nested for loop since running this function individually is a nightmare
r2_value_for_loop <- function(biom_table,
                              wanted_col,
                              wanted_conc) {
  ## nested for loop for the r squared and p values 
  output <- tibble()
  for(i in unique(unlist(biom_table$diet))) {
    for(j in unique(unlist(biom_table$tissue))) {
      for(k in unique(unlist(biom_table[wanted_col]))) {
        tmp_output <- r2_value(biom_table,
                               i,
                               j,
                               wanted_col,
                               k,
                               wanted_conc)$WantedStats
        output <- bind_rows(output, tmp_output)
      }
    }
  }
  return(output)
}

## putting stats directly on correlation plots 
stat_on_plot <- function(stat_table,
                         wanted_col_list,
                         x_col_name,
                         x_col_value,
                         y_col_name,
                         y_col_value,
                         plot,
                         plot_text_size){
  ## making table conducive to labeling plot 
  proc_stat_table <- stat_table %>% 
    filter(signif != 'ns') %>% 
    select(all_of(wanted_col_list)) %>% 
    ## created columns need to match up with the column names called for the x and y axis in the plot
    ## numeric values added are the x and y coordinates for where the label will sit in the facet block 
    mutate("{x_col_name}" := paste(x_col_value),
           "{x_col_name}" := as.numeric(.data[[x_col_name]]),
           "{y_col_name}" := paste(y_col_value),
           "{y_col_name}" := as.numeric(.data[[y_col_name]]),
           p_value = signif(p_value, digits = 3),
           r_squared = signif(r_squared, digits = 3),
           pVal_rSquare = paste0('p-value: ', p_value, "\n", 
                                 'r2: ', r_squared))
  
  ## putting labels on plot 
  plot_wStats <- plot +
    geom_label(data = proc_stat_table, 
               label = proc_stat_table$pVal_rSquare,
               color = 'black',
               fontface = 'bold',
               size = plot_text_size)
  
  ## creating list of my outputs
  my_list <- list(ProcStatTable = proc_stat_table,
                  PlotwStats = plot_wStats)
  
  return(my_list)
}
```

**File Paths**
```{r}
metadata_FP <- '../data/misc/filt_cecal_processed_metadata.tsv'
histo_FP <- '../data/misc/processed_histopathology.tsv'
metab_FP <- '../data/misc/processed_metabolomics.tsv'
toxin_FP <- '../data/misc/processed_neatToxin.tsv'

## lists to redo the diet names on the facet labels of the ggplot created below 
diet_labs <- 
    c('Chow', 
      'High Fat / High Fiber', 
      'High Fat / Low Fiber', 
      'Low Fat / High Fiber', 
      'Low Fat / Low Fiber')

names(diet_labs) <- c('Chow', 
                      'HF/HF', 
                      'HF/LF', 
                      'LF/HF', 
                      'LF/LF')

tissue_labs <- c('Cecum',
                 'Colon')
names(tissue_labs) <- c('cecum',
                        'colon')

metab_labs <- c('Acetic Acid',
                'Propanoic Acid',
                'n-Butanoic Acid')
names(metab_labs) <- c('Acetic Acid (ug/g)',
                       'Propanoic Acid (ug/g)',
                       'n-Butanoic Acid (ug/g)')

neat_labs <- c('TcdA', 'TcdB')
names(neat_labs) <- c('Total TcA Neat', 'Total TcB Neat')


stat_y_labs <- c("Chow",
                 "HFt/HFb",
                 "HFt/LFb",
                 "LFt/HFb",
                 "LFt/LFb")
```

**Reading in Needed Files**
```{r}
metadata <- read_tsv(metadata_FP)
metab <- read_tsv(metab_FP)
histo <- read_tsv(histo_FP)
neat_toxin <- read_tsv(toxin_FP)

## adding two so that I don't have any zeroes 
## zeroes get cut off on the plot since I have the axis in log10
metab %>% 
  select(metabolite, concentration, mouse_id) %>% 
  mutate(metab_conc_norm = (concentration + 2)) -> metab

neat_toxin %>% 
  select(neat_toxin, neat_conc, mouse_id) %>% 
  mutate(neat_conc_norm = (neat_conc + 2)) -> neat_toxin

## putting histo and metab tables together
histo %>% 
  select(mouse_id, tissue, score, diet, sampleid) %>% 
  left_join(metab, by = 'mouse_id') %>% 
  na.omit() -> histo_metab

## putting histo and neat toxin tables together
histo %>% 
  select(mouse_id, tissue, score, diet, sampleid) %>% 
  left_join(neat_toxin, by = 'mouse_id') %>% 
  na.omit() %>% 
  mutate(log_neat_conc = log10(neat_conc_norm)) -> histo_neat_toxin

## creating mini histo_neat_toxin table that only contains low fiber diets 
histo_neat_toxin %>% 
  filter(diet == "HF/LF" | diet == "LF/LF") -> mini_histo_neat_toxin

## making a giant table of everything combined to save for later
## so I don't have to do this data wrangling again 
toComb_histo_metab <- histo_metab %>% 
  rename(microbe_product = metabolite,
         conc_normalized = metab_conc_norm) 

toComb_histo_neatTox <- histo_neat_toxin %>% 
  rename(microbe_product = neat_toxin,
         concentration = neat_conc,
         conc_normalized = neat_conc_norm)

everything_table <- rbind(toComb_histo_metab,
                          toComb_histo_neatTox)

everything_table 
```

#**Toxin to Histopathology Scores Comparisons**
**Neat TcdA and TcdB Plot Construction**
```{r, warning=FALSE, fig.height=4, fig.width=9}
## neat cecum
histo_neat_toxin %>% 
  filter(tissue == 'cecum') -> cecum_neat_toxin

neat_histo_cecum_plot <- apppleplots::correlation_plots(input_table = cecum_neat_toxin,
                                                        x_value = 'neat_conc_norm',
                                                        y_value = 'score',
                                                        x_transform = TRUE,
                                                        y_transform = FALSE,
                                                        point_alpha = 0.4,
                                                        regression_method = 'lm',
                                                        facet_rows = 'neat_toxin',
                                                        facet_cols = 'diet',
                                                        row_labs = neat_labs,
                                                        col_labs = diet_labs,
                                                        x_name = 'Concentration (ng/uL) (log10)',
                                                        y_name = 'Score',
                                                        title_content = 
                                                          'Cecum C. difficile Toxin by Histopathology')


## mini neat cecum
mini_histo_neat_toxin %>% 
  filter(tissue == 'cecum') -> mini_cecum_neat_toxin

mini_neat_histo_cecum_plot <- apppleplots::correlation_plots(input_table = mini_cecum_neat_toxin,
                                                             x_value = 'neat_conc_norm',
                                                             y_value = 'score',
                                                             x_transform = TRUE,
                                                             y_transform = FALSE,
                                                             point_alpha = 0.4,
                                                             regression_method = 'lm',
                                                             facet_rows = 'neat_toxin',
                                                             facet_cols = 'diet',
                                                             row_labs = neat_labs,
                                                             col_labs = diet_labs,
                                                             x_name = 'Concentration (ng/uL) (log10)',
                                                             y_name = 'Score',
                                                             title_content = 'Cecum C. difficile Toxin by Histopathology')

## neat colon
histo_neat_toxin %>% 
  filter(tissue == 'colon') -> colon_neat_toxin

neat_histo_colon_plot <- apppleplots::correlation_plots(input_table = colon_neat_toxin,
                                                        x_value = 'neat_conc_norm',
                                                        y_value = 'score',
                                                        x_transform = TRUE,
                                                        y_transform = FALSE,
                                                        point_alpha = 0.4,
                                                        regression_method = 'lm',
                                                        facet_rows = 'neat_toxin',
                                                        facet_cols = 'diet',
                                                        row_labs = neat_labs,
                                                        col_labs = diet_labs,
                                                        x_name = 'Concentration (ng/uL) (log10)',
                                                        y_name = 'Score',
                                                        title_content = 'Colon C. difficile Toxin by Histopathology')
  
  
  
neat_histo_cecum_plot
neat_histo_colon_plot
mini_neat_histo_cecum_plot
```

**Toxin/Histo Correlations Not Faceted by Diet**
```{r, fig.height=4, fig.width=20}
## base plot
noDiet_histoCecum_tox_plot <- apppleplots::correlation_plots(input_table = cecum_neat_toxin,
                                                          x_value = 'neat_conc_norm',
                                                          y_value = 'score',
                                                          x_transform = TRUE,
                                                          y_transform = FALSE,
                                                          point_alpha = 0,
                                                          regression_method = 'lm',
                                                          facet_rows = NULL,
                                                          facet_cols = 'neat_toxin',
                                                          row_labs = NULL,
                                                          col_labs = neat_labs,
                                                          x_name = 'Concentration (ng/uL) (log10)',
                                                          y_name = 'Score',
                                                          title_content = 
                                                          'Cecum C. difficile Toxin by Histopathology')

noDiet_histoCecum_tox_plot <- noDiet_histoCecum_tox_plot +
                    geom_jitter(aes(fill = diet), pch = 21, alpha = 0.7, size = 3) +
                    scale_fill_brewer(palette = 'Spectral',
                                      name = 'Mouse Diet',
                                      labels = c('Chow',
                                                 'HFt / HFb',
                                                 'HFt / LFb',
                                                 'LFt / HFb',
                                                 'LFt / LFb')) +
                    theme_bw(base_size = 20)
  
## stats
cecum_neat_toxin %>% 
    filter(neat_toxin == "Total TcA Neat") %>%
    do(glance(lm(score ~ log_neat_conc * diet,
                 data = .))) -> lm_tca

cecum_neat_toxin %>% 
    filter(neat_toxin == "Total TcB Neat") %>%
    do(glance(lm(score ~ log_neat_conc * diet,
                 data = .))) -> lm_tcb

noDiet_histoCecum_tox_stats <- data.frame(r_squared = c(lm_tca$r.squared,
                                                        lm_tcb$r.squared),
                                          p_value = c(lm_tca$p.value,
                                                      lm_tcb$p.value),
                                          neat_toxin = c('Total TcA Neat',
                                                         'Total TcB Neat'))

noDiet_histoCecum_tox_stats['signif'] <- symnum(noDiet_histoCecum_tox_stats$p_value,
                                                cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1),
                                                symbols = c("****", "***", "**", "*", "ns"),
                                                abbr.colnames = FALSE,
                                                na = "")

## stats on plot
noDiet_histoCecum_col_list <- c('r_squared', 'p_value', 'neat_toxin', 'signif')


noDiet_histoCecum_list <- stat_on_plot(stat_table = noDiet_histoCecum_tox_stats,
                                       wanted_col_list = noDiet_histoCecum_col_list,
                                       x_col_name = 'neat_conc_norm',
                                       x_col_value = '5',
                                       y_col_name = 'score',
                                       y_col_value = '5.5',
                                       plot = noDiet_histoCecum_tox_plot,
                                       plot_text_size = 4)

noDiet_histoCecum_toxStats_plot <- noDiet_histoCecum_list$PlotwStats

noDiet_histoCecum_tox_stats
noDiet_histoCecum_toxStats_plot
```

**Toxin Stats (to put on plots)**
```{r, fig.height=4, fig.width=10}
## neat toxin stats
neatTox_stat_table <- r2_value_for_loop(histo_neat_toxin,
                                        "neat_toxin",
                                        "neat_conc")
## assigning p value significance
neatTox_stat_table['signif'] <- symnum(neatTox_stat_table$p_value,
                                       cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1),
                                       symbols = c("****", "***", "**", "*", "ns"),
                                       abbr.colnames = FALSE,
                                       na = "")
```

**P-values on Neat Toxin Plots**
Adding p-value to actual facet block on the plot for increased readability. 
```{r, fig.height=4, fig.width=9}
names(neatTox_stat_table)[names(neatTox_stat_table) == 'tox_metab'] <- 'neat_toxin'
col_list <- c('diet', 'neat_toxin', 'signif', 'p_value', 'r_squared')

## cecum 
neatTox_stat_table %>% 
  filter(tissue == 'cecum') %>% 
  select(-tissue) -> cecal_neat_stat_table

neat_cecumHisto_list <- stat_on_plot(stat_table = cecal_neat_stat_table,
                                     wanted_col_list = col_list,
                                     x_col_name = 'neat_conc_norm',
                                     x_col_value = '15',
                                     y_col_name = 'score',
                                     y_col_value = '5.5',
                                     plot = neat_histo_cecum_plot,
                                     plot_text_size = 3)

neat_cecumHisto_stats_plot <- neat_cecumHisto_list$PlotwStats

## mini cecum
mini_cecumHisto_list <- stat_on_plot(stat_table = cecal_neat_stat_table,
                                     wanted_col_list = col_list,
                                     x_col_name = 'neat_conc_norm',
                                     x_col_value = '5',
                                     y_col_name = 'score',
                                     y_col_value = '5.5',
                                     plot = mini_neat_histo_cecum_plot,
                                     plot_text_size = 3)

mini_neat_cecumHisto_stats_plot <- mini_cecumHisto_list$PlotwStats

## colon
neatTox_stat_table %>% 
  filter(tissue == 'colon') %>% 
  select(-tissue) -> colon_neat_stat_table

neat_colonHisto_list <- stat_on_plot(stat_table = colon_neat_stat_table,
                                     wanted_col_list = col_list,
                                     x_col_name = 'neat_conc_norm',
                                     x_col_value = '15',
                                     y_col_name = 'score',
                                     y_col_value = '9',
                                     plot = neat_histo_colon_plot,
                                     plot_text_size = 3)

neat_colonHisto_stats_plot <- neat_colonHisto_list$PlotwStats


neat_cecumHisto_stats_plot
mini_neat_cecumHisto_stats_plot
neat_colonHisto_stats_plot
```


**Heat Map Statistical Visualization**
```{r, warning=FALSE, fig.height=4, fig.width=6.5}
## neat stat plot
neat_tox_stat_vis <- apppleplots::stat_heat_plot(input_table = neatTox_stat_table,
                                                 x_value = 'neat_toxin',
                                                 y_value = 'diet',
                                                 x_value_type = 'discrete',
                                                 y_value_type = 'discrete',
                                                 pairwise = FALSE,
                                                 correlation = TRUE,
                                                 tile_fill = 'r_direct',
                                                 tile_outline_color = 'black',
                                                 tile_alpha = 0.6,
                                                 text_value = 'signif',
                                                 low_color = 'blue',
                                                 high_color = 'green',
                                                 legend_name = 'Correlation\nCoefficient',
                                                 x_labels = neat_labs,
                                                 y_labels = stat_y_labs,
                                                 facet_by = 'tissue',
                                                 facet_labels = tissue_labs,
                                                 x_name = 'Toxin',
                                                 y_name = 'Diet',
                                                 title_content = "Histopathology Score and Toxin Concentration Comparisons (neat)")

neat_tox_stat_vis
```
**Toxin Plots Together!**
```{r, warning=FALSE, fig.height=7, fig.width=15.5}
## neat
plot_grid(neat_histo_cecum_plot, neat_tox_stat_vis,
          neat_histo_colon_plot,
          ncol = 2,
          rel_widths = c(1, 0.7),
          labels = c('a)', 'c)', 'b)')) -> neat_toxin_histo_plots

neat_toxin_histo_plots
```

#**Metabolite Concentration Comparisons to Histopathology Score**
**Metabolomics Plots**
```{r, warning=FALSE, fig.height=5, fig.width=10}
## cecum 
histo_metab %>% 
  filter(tissue == 'cecum') -> cecum_metab

histo_metabCecum_plot <- apppleplots::correlation_plots(input_table = cecum_metab,
                                                        x_value = 'metab_conc_norm',
                                                        y_value = 'score',
                                                        x_transform = TRUE,
                                                        y_transform = FALSE,
                                                        point_alpha = 0.4,
                                                        regression_method = 'lm',
                                                        facet_rows = 'metabolite',
                                                        facet_cols = 'diet',
                                                        row_labs = metab_labs,
                                                        col_labs = diet_labs,
                                                        x_name = 'Concentration (log10)',
                                                        y_name = 'Score',
                                                        title_content = 'Cecum SCFAs by Histopathology')
  
  

## colon
histo_metab %>% 
  filter(tissue == 'colon') -> colon_metab

histo_metabColon_plot <- apppleplots::correlation_plots(input_table = colon_metab,
                                                        x_value = 'metab_conc_norm',
                                                        y_value = 'score',
                                                        x_transform = TRUE,
                                                        y_transform = FALSE,
                                                        point_alpha = 0.4,
                                                        regression_method = 'lm',
                                                        facet_rows = 'metabolite',
                                                        facet_cols = 'diet',
                                                        row_labs = metab_labs,
                                                        col_labs = diet_labs,
                                                        x_name = 'Concentration (log10)',
                                                        y_name = 'Score',
                                                        title_content = 'Colon SCFAs by Histopathology')
  
  
histo_metabCecum_plot
histo_metabColon_plot
```

**Metabolite Stats**
```{r}
## metabolite stats
metab_stat_table <- r2_value_for_loop(histo_metab,
                                      "metabolite",
                                      "metab_conc_norm")
## assigning p value significance 
metab_stat_table['signif'] <- symnum(metab_stat_table$p_value,
                                     cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1),
                                     symbols = c("****", "***", "**", "*", "ns"),
                                     abbr.colnames = FALSE,
                                     na = "")
```

**P-values on Metabolite Plots**
Putting significant p-values on the actual plot. 
```{r, fig.height=5, fig.width=10}
names(metab_stat_table)[names(metab_stat_table) == 'tox_metab'] <- 'metabolite'
col_list <- c('diet', 'metabolite', 'signif', 'p_value', 'r_squared')

## cecum 
metab_stat_table %>% 
  filter(tissue == 'cecum') %>% 
  select(-tissue) -> cecal_metab_stat_table

metab_cecumHisto_list <- stat_on_plot(stat_table = cecal_metab_stat_table,
                                     wanted_col_list = col_list,
                                     x_col_name = 'metab_conc_norm',
                                     x_col_value = '300',
                                     y_col_name = 'score',
                                     y_col_value = '5',
                                     plot = histo_metabCecum_plot,
                                     plot_text_size = 3)

metab_cecumHisto_stats_plot <- metab_cecumHisto_list$PlotwStats

## colon
metab_stat_table %>% 
  filter(tissue == 'colon') %>% 
  select(-tissue) -> colon_metab_stat_table

metab_colonHisto_list <- stat_on_plot(stat_table = colon_metab_stat_table,
                                     wanted_col_list = col_list,
                                     x_col_name = 'metab_conc_norm',
                                     x_col_value = '300',
                                     y_col_name = 'score',
                                     y_col_value = '9',
                                     plot = histo_metabColon_plot,
                                     plot_text_size = 3)

metab_colonHisto_stats_plot <- metab_colonHisto_list$PlotwStats

metab_cecumHisto_stats_plot
metab_colonHisto_stats_plot
```

**Heat Map Statistical Visualization**
```{r, warning=FALSE, fig.height=3.5, fig.width=7.5}
metab_stat_vis <- apppleplots::stat_heat_plot(input_table = metab_stat_table,
                                              x_value = 'metabolite',
                                              y_value = 'diet',
                                              x_value_type = 'discrete',
                                              y_value_type = 'discrete',
                                              pairwise = FALSE,
                                              correlation = TRUE,
                                              tile_fill = 'r_direct',
                                              tile_outline_color = 'black',
                                              tile_alpha = 0.6,
                                              text_value = 'signif',
                                              low_color = 'blue',
                                              high_color = 'green',
                                              legend_name = 'Correlation\nCoefficient',
                                              x_labels = metab_labs,
                                              y_labels = stat_y_labs,
                                              facet_by = 'tissue',
                                              facet_labels = tissue_labs,
                                              x_name = 'SCFA',
                                              y_name = 'Diet',
                                              title_content = "Histopathology Score and SCFA Concentration Comparisons")
  
metab_stat_vis
```

**Metabolite Plots Together!!**
```{r, warning=FALSE, fig.height=7, fig.width=19}
plot_grid(histo_metabCecum_plot, histo_metabColon_plot,
          metab_stat_vis,
          nrow = 2,
          rel_heights = c(1, 0.6),
          align = c('hv'),
          axis = c('tblr'),
          labels = c('a)', 'b)', 'c)')) -> metab_histo_plot

metab_histo_plot
```

#**Saving my Outputs**
```{r}
## as a pdf 

## stats on actual plot
ggsave('../plots/neat_cecumHisto_stats.pdf',
       plot = neat_cecumHisto_stats_plot,
       width = 10,
       height = 4)
ggsave('../plots/neat_colonHisto_stats.pdf',
       plot = neat_colonHisto_stats_plot,
       width = 9,
       height = 4)
ggsave('../plots/metab_cecumHisto_stats.pdf',
       plot = metab_cecumHisto_stats_plot,
       width = 10.5,
       height = 5)
ggsave('../plots/metab_colonHisto_stats.pdf',
       plot = metab_colonHisto_stats_plot,
       width = 10.5,
       height = 5)

## plots put together w their stat heat map via cowplot
ggsave('neat_histo_toxin_comp.pdf',
       plot = neat_toxin_histo_plots,
       width = 15.5,
       height = 7,
       path = '../plots')
ggsave('neat_histo_toxin_stats.pdf',
       plot = neat_tox_stat_vis,
       width = 6,
       height = 4,
       path = '../plots')
ggsave('histo_metab_comp.pdf',
       plot = metab_histo_plot,
       width = 19,
       height = 7,
       path = '../plots')
ggsave('histo_metab_stats.pdf',
       plot = metab_stat_vis,
       width = 8,
       height = 4,
       path = '../plots')

## saving histo, metab, and toxin results combined table as a .tsv
write_tsv(everything_table,
          '../data/misc/histoMetabToxin_results_comb.tsv')

## as an rdata object
save(neat_histo_cecum_plot,
     file = '../../figures/neat_histo_cecum.rdata')
save(neat_cecumHisto_stats_plot,
     file = '../../figures/neat_cecumHisto_stats.rdata')
save(neat_histo_colon_plot,
     file = '../../figures/neat_histo_colon.rdata')
save(neat_tox_stat_vis,
     file = '../../figures/neat_histo_stats.rdata')
save(histo_metabCecum_plot,
     file = '../../figures/histo_metab_cecum.rdata')
save(histo_metabColon_plot,
     file = '../../figures/histo_metab_colon.rdata')
save(metab_stat_vis,
     file = '../../figures/histo_metab_stats.rdata')
save(mini_neat_cecumHisto_stats_plot,
     file = '../../posters/mini_neat_cecumHisto_stats.rdat')
save(noDiet_histoCecum_toxStats_plot,
     file = '../../figures/noDiet_histoCecum_tox_plot.rdat')
```

