---
title: "tss_core_outputs"
author: "Madi"
date: "2023-05-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(qiime2R)
library(tidyverse)
library(cowplot)
library(magrittr)
library(vegan)
library(viridis)
library(microshades)
library(phyloseq)
library(ggh4x)
library(broom)
```

**Basic File-Prep Functions**
```{r}
## this function will pull out the percent variations from a specified column so you can add it to your pcoa plots 
pcoa_ax_lab <- function(unifrac_var, col_name){
  uni_lab <- as.character(round(unifrac_var[col_name] * 100, 2))
  uni_lab <- paste0(col_name, ' - ', uni_lab, '%')
  return(uni_lab)
}

## unweighted/weighted unifrac pcoa result, faith's pd, and shannon entropy file prep 
## going to attempt to return multiple outputs so I can just have one function for file prep
biom_table_prep <- function(unweighted_fp,
                            weighted_fp,
                            faith_fp,
                            shannon_fp,
                            metadata_file){
  ## unweighted pcoa
  unweighted <- read_qza(unweighted_fp)$data
  unweighted_var <- unweighted$ProportionExplained
  unweighted_pcoa <- unweighted$Vectors ##used for pcoa plot
  names(unweighted_pcoa)[names(unweighted_pcoa) == 'SampleID'] <- 'sampleid'
  ## weighted pcoa
  weighted <- read_qza(weighted_fp)$data
  weighted_var <- weighted$ProportionExplained
  weighted_pcoa <- weighted$Vectors
  names(weighted_pcoa)[names(weighted_pcoa) == 'SampleID'] <- 'sampleid'
  ## faith's 
  faith <- read_tsv(faith_fp)
  names(faith)[names(faith) == '#SampleID'] <- 'sampleid'
  ## shannon 
  shannon <- read_tsv(shannon_fp)
  names(shannon)[names(shannon) == '...1'] <- 'sampleid'
  ## unweighted biom 
  unweighted_pcoa %>% 
    left_join(metadata_file, by = 'sampleid') %>% 
    left_join(faith, by = 'sampleid') %>% 
    left_join(shannon, by = 'sampleid') -> unweighted_biom
  ## weighted biom
  weighted_pcoa %>% 
    left_join(metadata_file, by = 'sampleid') %>% 
    left_join(faith, by = 'sampleid') %>% 
    left_join(shannon, by = 'sampleid') -> weighted_biom
  ## creating a list to return multiple outputs 
  my_list <- list(UnweightedVar = unweighted_var, 
                  WeightedVar = weighted_var,
                  UnweightedBiom = unweighted_biom,
                  WeightedBiom = weighted_biom)
  return(my_list)
}

```

**Alpha Diversity Plot Function**
```{r}
## overall alpha diversity plot function
## assumes that the input file path includes a .tsv
alpha_div_plot <- function(alpha_div_fp,
                           rename_me,
                           metadata_file,
                           labels,
                           names_labels,
                           x_axis,
                           y_axis,
                           box_grouped_by,
                           line_grouped_by,
                           facet_by,
                           title,
                           x_label,
                           y_label){
  alpha_table <- read_tsv(alpha_div_fp)
  names(alpha_table)[names(alpha_table) == rename_me] <- 'sampleid'
  metadata_file %>% 
    left_join(alpha_table, by = 'sampleid') %>% 
    filter(!is.na(diet)) -> joint_table
  ## what you want the grid labels to be (list)
  labs <- (labels) 
  ## what grid labels currently are (list)
  names(labs) <- (names_labels) 
  joint_table %>%
    ggplot(aes(x = .data[[x_axis]], y = .data[[y_axis]])) +
    geom_boxplot(aes(group = .data[[box_grouped_by]]), outlier.shape = NA) +
    geom_line(aes(group = .data[[line_grouped_by]]), alpha = 0.1) +
    geom_vline(xintercept = -3, linetype = 'dashed', color = 'red', size = 0.2) +
    geom_vline(xintercept = 0, linetype = 'dashed', color = 'purple', size = 0.2) +
    geom_jitter(width = 0.1, height = 0, alpha = 0.4) +
    scale_x_continuous(breaks = c(-15, -8, -3, 0, 3)) +
    geom_smooth(se = FALSE) +
    theme_bw(base_size = 20) +
    facet_grid(~.data[[facet_by]], 
               labeller = labeller(.cols = labs) ) +
    ggtitle(title) +
    xlab(x_label) +
    ylab(y_label) -> plot
  return(plot)
}
```

**Beta Diversity Plot Function**
```{r}
## pcoa plot function
## xlab and ylab are outputs from pcoa_ax_lab function
pcoa_plot <- function(biom_file,
                      labels,
                      names_labels,
                      xlab,
                      ylab,
                      title){
  ## what you want the grid labels to be (list)
  labs <- (labels) 
  ## what grid labels currently are (list)
  names(labs) <- (names_labels) 
  biom_file %>% 
    filter(!is.na(diet)) %>% 
    ggplot(aes(x = PC1, y = PC2)) +
    geom_point(aes(fill = faith_pd), pch = 21, alpha = 0.7, size = 3) +
    theme_bw(base_size = 20) +
    scale_fill_distiller(palette = 'Spectral', name = "Faith's PD") +
    facet_grid(day_post_inf~diet, 
               labeller = labeller(diet = labs)) +
    theme(legend.text = element_text(size = 16),
          strip.text.y = element_text(angle = 0)) +
    ggtitle(title) +
    labs(x = xlab, y = ylab) -> pcoa
  return(pcoa)
}
```


**File paths for all needed files**
Doing this so that I don't have to worry about re-entering entire file paths all the time, they'll just be at the beginning
```{r}
metadata_FP <- '../data/misc/processed_metadata.tsv'
unweighted_FP <- '../data/qiime/core_outputs/unweighted_unifrac_pcoa_results.qza'
weighted_FP <- '../data/qiime/core_outputs/weighted_unifrac_pcoa_results.qza'
faith_pd_FP <- '../data/qiime/core_outputs/faith_pd.tsv'
shannon_FP <- '../data/qiime/core_outputs/shannon_entropy.tsv'

diet_labs <- 
    c('Chow', 
      'High Fat / High Fiber', 
      'High Fat / Low Fiber', 
      'Low Fat / High Fiber', 
      'Low Fat / Low Fiber')

diet_names_labels <- c('Chow', 
                       'HF/HF', 
                       'HF/LF', 
                       'LF/HF', 
                       'LF/LF')
```

**Core Metrics File-Prep (for plots)**
```{r}
## metadata file prep
metadata <- read_tsv(metadata_FP)

## preparing core beta diversity files for ggplot
core_files <- biom_table_prep(unweighted_FP,
                              weighted_FP,
                              faith_pd_FP,
                              shannon_FP,
                              metadata)
## extracting core beta diversity files from named list 
uw_var <- core_files$UnweightedVar
w_var <- core_files$WeightedVar
unweighted_biom <- core_files$UnweightedBiom
weighted_biom <- core_files$WeightedBiom

```


**Unweighted PcoA plot**
```{r, fig.height=8.5, fig.width=15, warning=FALSE}
uw_uni_xlab <- pcoa_ax_lab(uw_var, 'PC1')
uw_uni_ylab <- pcoa_ax_lab(uw_var, 'PC2')

uw_title <- 'Unweighted UniFrac PCoA Plot'

unweighted_pcoa <- pcoa_plot(unweighted_biom,
                             diet_labs,
                             diet_names_labels,
                             uw_uni_xlab,
                             uw_uni_ylab,
                             uw_title)

unweighted_pcoa
```

**Weighted PcoA plot**
```{r, fig.height=8.5, fig.width=17, warning=FALSE}
w_uni_xlab <- pcoa_ax_lab(w_var, 'PC1')
w_uni_ylab <- pcoa_ax_lab(w_var, 'PC2')

w_title <- 'Weighted UniFrac PCoA Plot'

weighted_pcoa <- pcoa_plot(weighted_biom,
                             diet_labs,
                             diet_names_labels,
                             w_uni_xlab,
                             w_uni_ylab,
                             w_title)

weighted_pcoa
```

**Faith's PD plot**
```{r, warning=FALSE, fig.height=4.5, fig.width=14}
faith_title <- "Faith's Phylogenetic Diversity"
faith_renamed <- "#SampleID"
faith_x_lab <- 'Days Relative to Infection'
faith_y_lab <- "Faith's PD"

faith_plot <- alpha_div_plot(alpha_div_fp = faith_pd_FP,
                             rename_me = faith_renamed,
                             metadata_file = metadata,
                             labels = diet_labs,
                             names_labels = diet_names_labels,
                             x_axis = 'day_post_inf',
                             y_axis = 'faith_pd',
                             box_grouped_by = 'day_post_inf',
                             line_grouped_by = 'mouse_id',
                             facet_by = 'diet',
                             title = faith_title,
                             x_label = faith_x_lab,
                             y_label = faith_y_lab)
faith_plot
```

**Shannon Entropy plot**
```{r, warning=FALSE, fig.height=4.5, fig.width=14}
shannon_title <- "Shannon Entropy"
shannon_renamed <- '...1'
shannon_x_lab <- 'Days Relative to Infection'
shannon_y_lab <- 'Shannon Entropy'

shannon_entropy_plot <- alpha_div_plot(alpha_div_fp = shannon_FP,
                                       rename_me = shannon_renamed,
                                       metadata_file = metadata,
                                       labels = diet_labs,
                                       names_labels = diet_names_labels,
                                       x_axis = 'day_post_inf',
                                       y_axis = 'shannon_entropy',
                                       box_grouped_by = 'day_post_inf',
                                       line_grouped_by = 'mouse_id',
                                       facet_by = 'diet',
                                       title = shannon_title,
                                       x_label = shannon_x_lab,
                                       y_label = shannon_y_lab)
shannon_entropy_plot
```

**Putting my Alpha Diversity Plots together**
NOTE: you did something lazy here and you had the faith_stat_vis plot loaded into your global environment so if you want to replicate this, you'll need to run the code in qiime_core_stats.Rmd before you run this chunk!!
```{r, fig.height=8.5, fig.width=14}
alpha_plots <- plot_grid(faith_plot, shannon_entropy_plot,
                          nrow = 2,
                          labels = c('a)', 'b)'),
                          label_size = 22)
alpha_plots
```

**Putting my Beta Diversity Plots Together**
```{r, fig.height=8, fig.width=30}
beta_plots <- plot_grid(unweighted_pcoa, weighted_pcoa,
                        nrow = 1,
                        labels = c('a)', 'b)'),
                        label_size = 22)
beta_plots
```

**Saving my plot outputs**
These go to the plots directory. 
```{r}
## this is an example of how to save my plot outputs as a pdf
ggsave("faith_pd.pdf",
       plot = faith_plot,
       width = 14,
       height = 4.5,
       path = '../plots')
ggsave("shannon_entropy.pdf",
       plot = shannon_entropy_plot,
       width = 14,
       height = 4.5,
       path = '../plots')
ggsave("tss_weighted_unifrac.pdf",
       plot = weighted_pcoa,
       width = 17,
       height = 8.5,
       path = '../plots')
ggsave("tss_unweighted_unifrac.pdf",
       plot = unweighted_pcoa,
       width = 17,
       height = 8.5,
       path = '../plots')
ggsave("alpha_plots.pdf",
       plot = alpha_plots,
       width = 14,
       height = 8.5,
       path = '../plots')
ggsave("beta_plots.pdf",
       plot = beta_plots,
       width = 30,
       height = 8,
       path = '../plots')
```




