---
title: "alpha_div_stats"
author: "Madi"
date: "2023-05-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(qiime2R)
library(tidyverse)
library(cowplot)
library(magrittr)
library(vegan)
library(viridis)
library(microshades)
library(phyloseq)
library(ggh4x)
library(broom)
library(rstatix)
library(dunn.test)
```

**Alpha Diversity Functions**
```{r}
## alpha diversity file prep 
alpha_div_prep <- function(file_path1,
                           file_path2,
                           metadata_fp){
  ## faith's pd 
  faith_pd <- read_tsv(file_path1)
  names(faith_pd)[names(faith_pd) == '#SampleID'] <- 'sampleid'
  ## metadata file for both
  stat_meta <- read_tsv(metadata_fp)
  ## joining faith's pd and metadata file together into one table
  stat_meta %>% 
    filter(sampleid %in% faith_pd$sampleid) %>% 
    left_join(faith_pd, by = 'sampleid') %>% 
    filter(!is.na(diet)) -> faith_biom
  ## shannon entropy
  shannon <- read_tsv(file_path2)
  names(shannon)[names(shannon) == '...1'] <- 'sampleid'
  ## joining shannon and metadata file together into one table 
  stat_meta %>% 
    filter(sampleid %in% shannon$sampleid) %>% 
    left_join(shannon, by = 'sampleid') %>% 
    filter(!is.na(diet)) -> shannon_biom
  ## creating a list for outputs 
  my_list <- list(FaithPD = faith_biom,
                  Shannon = shannon_biom, 
                  Metadata = stat_meta)
  return(my_list)
}


## stats calculations
## faith's pd 
faith_div_stats <- function(biom_table){
  ## alpha_cat is what the alpha div column is called (faith_pd or shannon_entropy)
  ## sectioned out by diet 
  biom_table %>% 
    filter(day_post_inf != -15) %>%
    group_by(day_post_inf) %>% 
    do(glance(lm(faith_pd ~ (purified_diet * seq_depth) + high_fat * high_fiber + study,
               data = .))) %>% 
    ungroup() %>% 
    na.omit() %>% 
    mutate(adj.p = p.adjust(p.value, 
                            method = "BH"),
          test_id = paste(day_post_inf)) %>% 
    filter(adj.p <= 0.05) -> lm_full
  biom_table %>% 
    group_by(day_post_inf) %>% 
    mutate(test_id = paste(day_post_inf)) %>% 
    filter(test_id %in% lm_full$test_id) %>% 
    do(tidy(lm(faith_pd ~ (purified_diet * seq_depth) + high_fat * high_fiber + study,
               data = .))) %>%
    filter(term != '(Intercept)') -> sectioned_lm
  sectioned_lm['signif'] <- symnum(sectioned_lm$p.value,
                                   cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
                                   symbols = c("****", "***", "**", "*", "+", "ns"),
                                   abbr.colnames = FALSE,
                                   na = "")
  ## not sectioned out by diet 
  ## haven't used these results much so decided not to do anything to this
  biom_table %>%
    group_by(day_post_inf) %>% 
    do(tidy(lm(faith_pd ~ diet * seq_depth,
               data = .))) -> not_sectioned_lm
  not_sectioned_lm %>% 
    filter(day_post_inf != -15) %>% 
    filter(p.value <= 0.05) -> not_sectioned_lm
  ## kruskal wallis and dunns post hoc tests
  biom_table %>% 
    na.omit() %>% 
    group_by(day_post_inf) %>% 
    do(tidy(kruskal.test(faith_pd ~ diet,
             data = .))) %>% 
    ungroup() %>% 
    arrange(p.value) %>% 
    mutate(p.adj = p.adjust(p.value,
                            method = "BH"),
          test_id = paste(day_post_inf)) %>%
    filter(p.adj <= 0.05) -> kruskal
  biom_table %>% 
    na.omit() %>% 
    group_by(day_post_inf) %>% 
    mutate(test_id = paste(day_post_inf)) %>% 
    filter(test_id %in% kruskal$test_id) %>% 
    dunn_test(faith_pd ~ diet,
              p.adjust.method = 'BH',
              data = .) -> dunn
  ## creating a list 
  my_list <- list(DietSpecific = sectioned_lm,
                  OverallDiet = not_sectioned_lm,
                  KruskalTest = kruskal,
                  DunnPostHoc = dunn)
  return(my_list)
}

## shannon entropy 
shannon_div_stats <- function(biom_table){
  ## alpha_cat is what the alpha div column is called (faith_pd or shannon_entropy)
  ## sectioned out by diet 
  biom_table %>% 
    filter(day_post_inf != -15) %>% 
    group_by(day_post_inf) %>% 
    do(glance(lm(shannon_entropy ~ (purified_diet * seq_depth) + high_fat * high_fiber + study,
               data = .))) %>% 
    ungroup() %>% 
    na.omit() %>% 
    mutate(adj.p = p.adjust(p.value, 
                            method = "BH"),
          test_id = paste(day_post_inf)) %>% 
    filter(adj.p <= 0.05) -> lm_full
  biom_table %>% 
    group_by(day_post_inf) %>% 
    mutate(test_id = paste(day_post_inf)) %>% 
    filter(test_id %in% lm_full$test_id) %>% 
    do(tidy(lm(shannon_entropy ~ (purified_diet * seq_depth) + high_fat * high_fiber + study,
               data = .))) %>%
    filter(term != '(Intercept)') -> sectioned_lm
  sectioned_lm['signif'] <- symnum(sectioned_lm$p.value,
                                   cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
                                   symbols = c("****", "***", "**", "*", "+", "ns"),
                                   abbr.colnames = FALSE,
                                   na = "")
  ## not sectioned out by diet 
  biom_table %>%
    group_by(day_post_inf) %>% 
    do(tidy(lm(shannon_entropy ~ diet * seq_depth,
               data = .))) -> not_sectioned_lm
  not_sectioned_lm %>% 
    filter(day_post_inf != -15) %>% 
    filter(p.value <= 0.05) -> not_sectioned_lm
    ## kruskal wallis and dunns post hoc tests
  biom_table %>% 
    na.omit() %>% 
    group_by(day_post_inf) %>% 
    do(tidy(kruskal.test(shannon_entropy ~ diet,
             data = .))) %>% 
    ungroup() %>% 
    arrange(p.value) %>% 
    mutate(p.adj = p.adjust(p.value,
                            method = "BH"),
          test_id = paste(day_post_inf)) %>%
    filter(p.adj <= 0.05) -> kruskal
  biom_table %>% 
    na.omit() %>% 
    group_by(day_post_inf) %>% 
    mutate(test_id = paste(day_post_inf)) %>% 
    filter(test_id %in% kruskal$test_id) %>% 
    dunn_test(shannon_entropy ~ diet,
              p.adjust.method = 'BH',
              data = .) -> dunn
  ## creating a list 
  my_list <- list(DietSpecific = sectioned_lm,
                  OverallDiet = not_sectioned_lm,
                  KruskalTest = kruskal,
                  DunnPostHoc = dunn)
  return(my_list)
}
```

**Statistical Visualization Functions**
Note: this may need to be edited based on the statistical tests used and column names of the dataset, could make these more easiy accessible for other types of data but I was being lazy when I created them. 
```{r}
## creating a function for this so I don't have to keep doing each one by hand 
stat_plot_prep <- function(biom_table,
                           dunn_test,
                           value){
  biom_table %>% 
    group_by(diet, day_post_inf) %>% 
    summarise(means = mean(.data[[value]])) -> mean_table
  dunn_test %>% 
    merge(mean_table, 
          by.x = c('group1',
                   'day_post_inf'),
          by.y = c('diet',
                   'day_post_inf')) %>% 
    rename('group1_means' = 'means') %>% 
    merge(mean_table,
          by.x = c('group2',
                   'day_post_inf'),
          by.y = c('diet',
                   'day_post_inf')) %>% 
    rename('group2_means' = 'means') %>% 
    mutate(diff_means = (group1_means - group2_means),
           stat_diff_means = if_else(p.adj > 0.05, 0, diff_means)) -> new_dunn
  return(new_dunn)
}

stat_plot <- function(new_dunn){
  new_dunn %>% 
    filter(day_post_inf != -15) %>%
    ggplot(aes(x = group1, y = group2)) +
    geom_tile(aes(fill = stat_diff_means), alpha = 0.8, color = 'black') +
    scale_fill_gradient2(low = 'blue', high = 'green', name = 'Group 1 -\nGroup 2') +
    geom_text(aes(label = p.adj.signif)) +
    scale_x_discrete(labels = c('Chow',
                                'HFt/\nHFb',
                                'HFt/\nLFb',
                                'LFt/\nHFb')) +
    scale_y_discrete(labels = c('HFt / HFb',
                                'HFt / LFb',
                                'LFt / HFb',
                                'LFt / LFb')) +
    facet_grid(~day_post_inf,
              scales = 'free_x') +
    theme_bw(base_size = 20) +
    theme(strip.text.y = element_text(angle = 0)) +
    xlab('Group 1') +
    ylab('Group 2') -> stat_vis
  return(stat_vis)
}
```

**File paths for all needed files**
Doing this so that I don't have to worry about re-entering entire file paths all the time, they'll just be at the beginning
```{r}
metadata_FP <- '../data/misc/processed_metadata.tsv'
faith_pd_fp <- '../data/qiime/core_outputs/faith_pd.tsv'
shannon_fp <- '../data/qiime/core_outputs/shannon_entropy.tsv'
```


**Alpha Diversity Statistical Analysis**
```{r}
## alpha diversity analysis  
alpha_files <- alpha_div_prep(faith_pd_fp,
                              shannon_fp,
                              metadata_FP)

faith <- alpha_files$FaithPD
shannon <- alpha_files$Shannon
metadata <- alpha_files$Metadata

## faith's pd stats
faith_stats <- faith_div_stats(faith)
sectioned_faith_lm <- faith_stats$DietSpecific
faith_lm <- faith_stats$OverallDiet
faith_kruskal <- faith_stats$KruskalTest
faith_dunn <- faith_stats$DunnPostHoc
new_faith_dunn <- stat_plot_prep(faith,
                                 faith_dunn,
                                 'faith_pd')
faith_stat_vis <- stat_plot(new_faith_dunn)

## shannon entropy stats
shannon_stats <- shannon_div_stats(shannon)
sectioned_shannon_lm <- shannon_stats$DietSpecific
shannon_lm <- shannon_stats$OverallDiet
shannon_kruskal <- shannon_stats$KruskalTest
shannon_dunn <- shannon_stats$DunnPostHoc
new_shannon_dunn <- stat_plot_prep(shannon,
                                   shannon_dunn,
                                   'shannon_entropy')
shannon_stat_vis <- stat_plot(new_shannon_dunn)
```


**Alpha Diversity Statistical Visualizations**
I'm doing this for my poster presentation but these will be helpful overall in the future.
```{r, fig.height=3.5, fig.width=13.75}
faith_stat_vis
```

```{r, fig.height=3.5, fig.width=13.75}
shannon_stat_vis
```


**Saving my Outputs**
```{r}
## plot as a pdf
ggsave("faith_stats.pdf",
       plot = faith_stat_vis, 
       width = 14, 
       height = 4, 
       path = '../plots')
ggsave("shannon_stats.pdf",
       plot = shannon_stat_vis, 
       width = 14, 
       height = 4, 
       path = '../plots')

## as an rdata object
save(faith_stat_vis,
     file = '../figures/faith_stats.rdata')
save(shannon_stat_vis,
     file = '../figures/shannon_stats.rdata')

## example of how to write results out as a .tsv 
write_tsv(new_faith_dunn, '../stats/faith_dunn.tsv')
write_tsv(sectioned_faith_lm, '../stats/faith_total_results.tsv')
write_tsv(new_shannon_dunn, '../stats/shannon_dunn.tsv')
write_tsv(sectioned_shannon_lm, '../stats/shannon_total_results.tsv')
```


