---
title: "family_abundance"
output: html_document
date: "2023-06-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# change to scripts directory if not there already
# curr_dir <- getwd()
# curr_dir <- str_split(curr_dir, '\\/')
# if (curr_dir[length(curr_dir)] != 'scripts'){
#   setwd('./scripts')
# } 
```

```{r}
library(qiime2R)
library(tidyverse)
library(cowplot)
library(magrittr)
library(vegan)
library(viridis)
library(broom)
```

**Function**
.data[[]] let's you insert a string variable into tidyverse commands which makes your functions more customizable!
```{r}
family_abun_file_prep <- function(metadata_fp,
                                  tax_fp,
                                  otu_table_fp,
                                  tax_level,
                                  wanted_tax){
  ## metadata
  metadata <- read_tsv(metadata_FP)
  ## taxonomy
  taxonomy <- read_qza(tax_FP)$data %>% 
  parse_taxonomy() %>% 
  rownames_to_column('asv')
  ## otu table 
  otu_table <- read_qza(otu_table_FP)$data
  otu_table %>% 
    as_tibble(rownames = 'asv') %>% 
    gather(-asv, key = sampleid, value = abun) %>% 
    group_by(sampleid) %>% 
    mutate(rel_abun = abun/sum(abun)) %>% 
    mutate(rel_abun = rel_abun + 0.000001) -> otu_table
  ## joining all tables together 
  otu_table %>% 
    left_join(metadata, by = 'sampleid') %>% 
    left_join(taxonomy, by = 'asv') -> abun_table
  abun_table %>% 
    group_by(sampleid, day_post_inf, diet, mouse_id, 
             purified_diet, high_fat, high_fiber, 
             seq_depth, .data[[tax_level]]) %>% 
    summarise(rel_abund = sum(rel_abun)) %>% 
    filter(.data[[tax_level]] %in% wanted_tax) %>% 
    mutate(mouse_fact = as.factor(mouse_id),
           day_fact = as.factor(day_post_inf)) -> abun_filt
  ## creating a list for my outputs
  my_list <- list(Metadata = metadata,
                  Taxonomy = taxonomy,
                  OTUTable = otu_table,
                  AbundanceTable = abun_filt)
  return(my_list)
}

abun_plots <- function(abundance_table){
  ## first plot
  abundance_table %>%
    filter(!is.na(diet)) %>% 
    ggplot(aes(x = day_post_inf, y = rel_abund)) +
      scale_y_continuous(trans = 'log10') +
      scale_x_continuous(breaks = c(-15, -8, -3, 0, 3)) +
      geom_boxplot(aes(group = day_post_inf), outlier.shape = NA) +
      geom_vline(xintercept = -3, linetype = 'dashed', color = 'red', size = 0.2) +
      geom_vline(xintercept = 0, linetype = 'dashed', color = 'purple', size = 0.2) +
      geom_line(aes(group = mouse_id), alpha = 0.1) +
      geom_smooth(se = FALSE) +
      geom_jitter(width = 0.1, height = 0, alpha = 0.4) +
      theme_bw(base_size = 16) +
      facet_grid(Family~diet, labeller = labeller(diet = diet_labs)) +
      theme(strip.text.y = element_text(angle = 0)) +
      ggtitle("Microbe Family Relative Abundance") +
      ylab("Relative Abundance") +
      xlab("Days Relative to Infection") -> family_abun1
  ## second plot
  abundance_table %>%
    filter(!is.na(diet)) %>% 
    ggplot(aes(x = mouse_fact, y = day_fact)) +
      geom_tile(aes(fill = rel_abund), alpha = 0.5) +
      scale_fill_viridis(option = "H", name = 'Relative\nAbundance') +
      theme_bw(base_size = 16) +
      facet_grid(Family~diet, scales = 'free',
                 labeller = labeller(diet = diet_labs)) +
      theme(strip.text.y = element_text(angle = 0),
            axis.text.x = element_blank()) + 
      xlab("Mouse ID") +
    ylab("Days Relative to Infection") +
    scale_y_discrete(limits = rev) -> family_abun2
  ## creating a list of my two plots
  my_list <- list(FamilyAbundance1 = family_abun1,
                  FamilyAbundance2 = family_abun2)
  return(my_list)
}
```

**Needed File Paths**
```{r}
otu_table_FP <- '../data/qiime/taxonomy_filtered.qza'
tax_FP <- '../data/qiime/taxonomy.qza'
metadata_FP <- '../data/misc/processed_metadata.tsv'
diet_labs <- 
  c('Chow', 
    'High Fat / High Fiber', 
    'High Fat / Low Fiber', 
    'Low Fat / High Fiber', 
    'Low Fat / Low Fiber')
names(diet_labs) <- c('Chow', 
                      'HF/HF', 
                      'HF/LF', 
                      'LF/HF', 
                      'LF/LF')
wanted_level <- 'Family'
long_wanted_family <- c('Enterobacteriaceae', 'Lactobacillaceae', 'Lachnospiraceae', 'Enterococcaceae',
                        'Staphylococcaceae', 'Tannerellaceae', 'Muribaculaceae', 'Bacteroidaceae', 
                        'Marinifilaceae')
short_wanted_family <- c('Enterobacteriaceae', 
                         'Enterococcaceae', 
                         'Staphylococcaceae',
                         'Bacteroidaceae',
                         'Lachnospiraceae')
```

**File Prep for Plot Construction**
```{r}
## full table 
long_abun_files <- family_abun_file_prep(metadata_FP,
                                         tax_FP,
                                         otu_table_FP,
                                         wanted_level,
                                         long_wanted_family)

long_abun_filt <- long_abun_files$AbundanceTable

## shortened table for poster presentation
short_abun_files <- family_abun_file_prep(metadata_FP,
                                          tax_FP,
                                          otu_table_FP,
                                          wanted_level,
                                          short_wanted_family)

short_abun_filt <- short_abun_files$AbundanceTable
```

**Testing the New Plots Function**
```{r, warning=FALSE, fig.width=15, fig.height=8}
## full table
long_family_abun_plots <- abun_plots(long_abun_filt)

long_abun1 <- long_family_abun_plots$FamilyAbundance1
long_abun2 <- long_family_abun_plots$FamilyAbundance2

## shortened table for poster presentation
short_family_abun_plots <- abun_plots(short_abun_filt)

short_abun1 <- short_family_abun_plots$FamilyAbundance1
short_abun2 <- short_family_abun_plots$FamilyAbundance2

short_abun2
```

**Attempt at Linear Modeling for Wanted Family Abundances**
```{r}
## for overall plot 
long_abun_filt %>%
  na.omit() %>% 
  group_by(Family, day_post_inf) %>% 
  do(tidy(lm(rel_abund ~ (purified_diet * seq_depth) + high_fat + high_fiber,
             data =.))) %>% 
  adjust_pvalue(method = 'BH') %>% 
  na.omit() %>% 
  filter(term != '(Intercept)',
         p.value <= 0.05) -> family_abun_lm

family_abun_lm
```

these stats are for the shortened figure that I am putting on my poster presentation!
```{r}
## for poster presentation plot linear model
short_abun_filt %>%
  na.omit() %>% 
  group_by(Family, day_post_inf) %>% 
  do(tidy(lm(rel_abund ~ (purified_diet * seq_depth) + high_fat + high_fiber,
             data =.))) %>% 
  adjust_pvalue(method = 'BH') %>% 
  na.omit() %>% 
  filter(term != '(Intercept)') -> poster_family_abun_lm

## kruskal-wallis and dunn's post hoc test
short_abun_filt %>% 
  group_by(Family) %>% 
  do(tidy(kruskal.test(rel_abund ~ diet,
             data = .))) -> poster_kruskal

short_abun_filt %>% 
  group_by(Family) %>% 
    dunn_test(rel_abund ~ diet,
            p.adjust.method = 'BH',
            data = .) -> poster_dunn

poster_dunn

```

possible heat map for statistical results for my poster presentation
```{r, fig.height=7, fig.width=10}
## option #1 based on dunn's post hoc test 
poster_dunn %>% 
  ggplot(aes(x = group1, y = group2)) +
  geom_tile(aes(fill = p.adj.signif), alpha = 0.5) +
  scale_fill_viridis(option = "H", discrete = TRUE) +
  facet_wrap(~Family,
             strip.position = "right",
             ncol = 1) +
  theme_bw(base_size = 16) +
  theme(strip.text.y = element_text(angle = 0)) -> stat_plot1

## option #2 based on the linear model
poster_family_abun_lm %>% 
  filter(term != 'seq_depth',
         term != 'purified_diet:seq_depth') %>% 
  mutate(day_fact = as.factor(day_post_inf)) %>% 
  ggplot(aes(x = day_fact, y = term)) +
  geom_tile(aes(fill = p.value.adj), alpha = 0.5) +
  theme_bw(base_size = 16) +
  scale_fill_viridis(option = "H", name = 'Adjusted P-Value') +
  facet_wrap(~Family,
             scales = "free_x",
             ncol = 1,
             strip.position = "right") +
  theme(strip.text.y = element_text(angle = 0)) +
  xlab('Days Relative to Infection') +
  ylab('Diet Composition') -> stat_plot2

stat_plot2
```


**Saving my Outputs**
These go to the plots and stats directories. 
```{r}
## microbe family relative abundance plots
## option #1
ggsave("family_abun1.pdf",
       plot = long_abun1, 
       width = 15, 
       height = 10, 
       path = '../plots')
## option #2
ggsave("family_abun2.pdf",
       plot = long_abun2, 
       width = 25, 
       height = 10, 
       path = '../plots')

## family abundance plots for poster presentation
ggsave("poster_family_abun.pdf",
       plot = short_abun1, 
       width = 15, 
       height = 8, 
       path = '../plots')

ggsave("poster_famAbun_stats1.pdf",
       plot = stat_plot1, 
       width = 10, 
       height = 7, 
       path = '../plots')

ggsave("poster_famAbun_stats2.pdf",
       plot = stat_plot2, 
       width = 10, 
       height = 7, 
       path = '../plots')

## linear modeling results
write_tsv(family_abun_lm,
          '../stats/family_abun_lm.tsv')
```
