---
title: "family_abundance"
output: html_document
date: "2023-06-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(qiime2R)
library(tidyverse)
library(cowplot)
library(magrittr)
library(vegan)
library(viridis)
library(broom)
library(rstatix)
```

**Function**
.data[[]] let's you insert a string variable into tidyverse commands which makes your functions more customizable!
```{r}
## abundance file prep and filtering function
## combines all files into one large table
family_abun_file_prep <- function(metadata_fp,
                                  tax_fp,
                                  otu_table_fp,
                                  tax_level,
                                  wanted_tax){
  ## metadata
  metadata <- read_tsv(metadata_fp)
  ## taxonomy
  taxonomy <- read_qza(tax_fp)$data %>% 
  parse_taxonomy() %>% 
  rownames_to_column('asv')
  ## otu table 
  otu_table <- read_qza(otu_table_fp)$data
  otu_table %>% 
    as_tibble(rownames = 'asv') %>% 
    gather(-asv, key = sampleid, value = abun) %>% 
    group_by(sampleid) %>% 
    mutate(rel_abun = abun/sum(abun)) %>% 
    mutate(rel_abun = rel_abun + 0.000001) -> otu_table
  ## joining all tables together 
  otu_table %>% 
    left_join(metadata, by = 'sampleid') %>% 
    left_join(taxonomy, by = 'asv') -> abun_table
  abun_table %>% 
    group_by(sampleid, day_post_inf, diet, mouse_id, 
             purified_diet, high_fat, high_fiber, 
             seq_depth, .data[[tax_level]]) %>% 
    summarise(rel_abund = sum(rel_abun)) %>% 
    filter(.data[[tax_level]] %in% wanted_tax) %>% 
    mutate(mouse_fact = as.factor(mouse_id),
           day_fact = as.factor(day_post_inf)) -> abun_filt
  ## creating a list for my outputs
  my_list <- list(Metadata = metadata,
                  Taxonomy = taxonomy,
                  OTUTable = otu_table,
                  AbundanceTable = abun_filt)
  return(my_list)
}

## takes large assembled table and creates the quantitative plot 
abun_plots <- function(abundance_table,
                       tax_level){
  ## first plot
  abundance_table %>%
    filter(!is.na(diet)) %>% 
    ggplot(aes(x = day_post_inf, y = rel_abund)) +
      scale_y_continuous(trans = 'log10') +
      scale_x_continuous(breaks = c(-15, -8, -3, 0, 3)) +
      geom_boxplot(aes(group = day_post_inf), outlier.shape = NA) +
      geom_vline(xintercept = -3, linetype = 'dashed', color = 'red', size = 0.2) +
      geom_vline(xintercept = 0, linetype = 'dashed', color = 'purple', size = 0.2) +
      geom_line(aes(group = mouse_id), alpha = 0.1) +
      geom_smooth(se = FALSE) +
      geom_jitter(width = 0.1, height = 0, alpha = 0.4) +
      theme_bw(base_size = 20) +
      facet_grid(.data[[tax_level]]~diet, labeller = labeller(diet = diet_labs)) +
      theme(strip.text.y = element_text(angle = 0)) +
      ggtitle("Microbe Relative Abundance") +
      ylab("Relative Abundance") +
      xlab("Days Relative to Infection") -> family_abun1
  ## second plot
  abundance_table %>%
    filter(!is.na(diet)) %>% 
    ggplot(aes(x = mouse_fact, y = day_fact)) +
      geom_tile(aes(fill = rel_abund), alpha = 0.5) +
      scale_fill_viridis(option = "H", name = 'Relative\nAbundance') +
      theme_bw(base_size = 16) +
      facet_grid(.data[[tax_level]]~diet, scales = 'free',
                 labeller = labeller(diet = diet_labs)) +
      theme(strip.text.y = element_text(angle = 0),
            axis.text.x = element_blank()) + 
      xlab("Mouse ID") +
    ylab("Days Relative to Infection") +
    scale_y_discrete(limits = rev) -> family_abun2
  ## creating a list of my two plots
  my_list <- list(FamilyAbundance1 = family_abun1,
                  FamilyAbundance2 = family_abun2)
  return(my_list)
}

## runs statistics on the large assembled data table for stat visualization
abun_stats <- function(filtered_table,
                       tax_level){
  ## linear modeling 
  filtered_table %>%
    na.omit() %>% 
    filter(day_post_inf > -15) %>% 
    group_by(.data[[tax_level]], day_post_inf) %>% 
    do(glance(lm(rel_abund ~ (purified_diet * seq_depth) + high_fat * high_fiber,
              data =.))) %>% 
    ungroup() %>% 
    na.omit() %>% 
    mutate(adj.p = p.adjust(p.value, 
                            method = "BH"),
          test_id = paste(.data[[tax_level]], day_post_inf, sep = "_")) %>% 
    filter(adj.p <= 0.05) -> lm_full
  filtered_table %>%
    na.omit() %>% 
    group_by(.data[[tax_level]], day_post_inf) %>% 
    mutate(test_id = paste(.data[[tax_level]], day_post_inf, sep = "_")) %>% 
    filter(test_id %in% lm_full$test_id) %>% 
    do(tidy(lm(rel_abund ~ (purified_diet * seq_depth) + high_fat * high_fiber,
             data =.))) %>% 
    na.omit() %>% 
    filter(term != '(Intercept)') -> linear_model
  linear_model['signif'] <- symnum(linear_model$p.value,
                                   cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
                                   symbols = c("****", "***", "**", "*", "+", "ns"),
                                   abbr.colnames = FALSE,
                                   na = "")
  ## kruskal wallis test 
  filtered_table %>% 
    na.omit() %>% 
    group_by(.data[[tax_level]], day_post_inf) %>% 
    do(tidy(kruskal.test(rel_abund ~ diet,
               data = .))) %>% 
    ungroup() %>% 
    arrange(p.value) %>% 
    mutate(p.adj = p.adjust(p.value,
                            method = "BH"),
           test_id = paste(.data[[tax_level]], day_post_inf, sep = "_")) %>% 
    filter(p.adj <= 0.05) -> kruskal
  ## dunn's post hoc test
  filtered_table %>% 
    na.omit() %>% 
    group_by(.data[[tax_level]], day_post_inf) %>%
    mutate(test_id = paste(.data[[tax_level]], day_post_inf, sep = "_")) %>% 
    filter(test_id %in% kruskal$test_id) %>% 
    dunn_test(rel_abund ~ diet,
              p.adjust.method = 'BH',
              data = .) -> dunn
  ## editing my dunn's post hoc test to include the difference in means between groups 
  filtered_table %>% 
    group_by(diet, .data[[tax_level]], day_post_inf) %>% 
    summarise(mean_rel_abund = mean(rel_abund)) -> mean_abun
  dunn %>% 
    merge(mean_abun, 
          by.x = c('group1',
                   'day_post_inf',
                   'Family'),
          by.y = c('diet',
                   'day_post_inf',
                   'Family')) %>% 
    rename('group1_rel_abun' = 'mean_rel_abund') %>% 
    merge(mean_abun,
          by.x = c('group2',
                   'day_post_inf',
                   'Family'),
          by.y = c('diet',
                   'day_post_inf',
                   'Family')) %>% 
    rename('group2_rel_abun' = 'mean_rel_abund') %>% 
    mutate(diff_means = (group1_rel_abun - group2_rel_abun),
           stat_diff_means = if_else(p.adj > 0.05, 0, diff_means)) -> new_dunn
  ## creating a list of my outputs
  my_list <- list(LinearModel = linear_model,
                  KruskalTest = kruskal,
                  DunnPostHoc = new_dunn)
  return(my_list)
}

## can drop the 'new' dunn test results into this function for the 
## stat visualization 
stat_plot <- function(new_dunn,
                      tax_level){
  new_dunn %>% 
    filter(day_post_inf != -15) %>% 
    ggplot(aes(x = group1, y = group2)) +
    geom_tile(aes(fill = stat_diff_means), alpha = 0.8, color = 'black') +
    scale_fill_gradient2(low = 'blue', high = 'green', name = 'Group 1 -\nGroup 2') +
    geom_text(aes(label = p.adj.signif)) +
    scale_x_discrete(labels = c('Chow',
                                'HFt/\nHFb',
                                'HFt/\nLFb',
                                'LFt/\nHFb')) +
    scale_y_discrete(labels = c('HFt / HFb',
                                'HFt / LFb',
                                'LFt / HFb',
                                'LFt / LFb')) +
    facet_grid(.data[[tax_level]]~day_post_inf,
               scales = 'free_x') +
    theme_bw(base_size = 20) +
    theme(strip.text.y = element_text(angle = 0)) +
    xlab('Group 1') +
    ylab('Group 2') -> stat_plot
  return(stat_plot)
}

```

**Needed File Paths**
```{r}
otu_table_FP <- '../data/qiime/taxonomy_filtered.qza'
tax_FP <- '../data/qiime/taxonomy.qza'
metadata_FP <- '../data/misc/processed_metadata.tsv'
diet_labs <- 
  c('Chow', 
    'High Fat / High Fiber', 
    'High Fat / Low Fiber', 
    'Low Fat / High Fiber', 
    'Low Fat / Low Fiber')
names(diet_labs) <- c('Chow', 
                      'HF/HF', 
                      'HF/LF', 
                      'LF/HF', 
                      'LF/LF')
wanted_level <- 'Family'
long_wanted_family <- c('Enterobacteriaceae', 'Lactobacillaceae', 'Lachnospiraceae', 'Enterococcaceae',
                        'Staphylococcaceae', 'Tannerellaceae', 'Muribaculaceae', 'Bacteroidaceae', 
                        'Marinifilaceae', 'Ruminococcaceae', 'Peptostreptococcaceae')
short_wanted_family <- c('Enterobacteriaceae', 
                         'Enterococcaceae', 
                         'Staphylococcaceae',
                         'Lachnospiraceae',
                         'Ruminococcaceae',
                         'Bacteroidaceae',
                         'Marinifilaceae',
                         'Peptostreptococcaceae')
cdiff_level <- 'Genus'
wanted_genus <- 'Clostridioides'
```

**File Prep for Plot Construction**
```{r}
## full table 
long_abun_files <- family_abun_file_prep(metadata_FP,
                                         tax_FP,
                                         otu_table_FP,
                                         wanted_level,
                                         long_wanted_family)

long_abun_filt <- long_abun_files$AbundanceTable
long_otu <- long_abun_files$OTUTable

## shortened table for poster presentation
short_abun_files <- family_abun_file_prep(metadata_FP,
                                          tax_FP,
                                          otu_table_FP,
                                          wanted_level,
                                          short_wanted_family)

short_abun_filt <- short_abun_files$AbundanceTable
short_tax <- short_abun_files$Taxonomy
```

**Family Abundance Plots**
Long = the full family abundance plot. Short = the revised family abundance plot for my poster presentation. 
```{r, warning=FALSE, fig.width=17, fig.height=15}
## full table
long_family_abun_plots <- abun_plots(long_abun_filt,
                                     wanted_level)

long_abun1 <- long_family_abun_plots$FamilyAbundance1
long_abun2 <- long_family_abun_plots$FamilyAbundance2

## shortened table for poster presentation
short_family_abun_plots <- abun_plots(short_abun_filt,
                                      wanted_level)

short_abun1 <- short_family_abun_plots$FamilyAbundance1
short_abun2 <- short_family_abun_plots$FamilyAbundance2

short_abun1
```

**Statistical Analysis for Wanted Family Abundances**
stats run include, a linear model, a kruskal-wallis test, and a dunn's post hoc test. the dunn's post hoc test is used for the statistical visualization shown. 
```{r, fig.height=15, fig.width=17}
long_abun_stats <- abun_stats(long_abun_filt,
                              wanted_level)
long_kruskal <- long_abun_stats$KruskalTest
new_long_dunn <- long_abun_stats$DunnPostHoc
family_abun_lm <- long_abun_stats$LinearModel

long_stat_plot <- stat_plot(new_long_dunn,
                            wanted_level)
long_stat_plot
```

**Putting My (long) Plots Together!**
```{r, fig.height = 16, fig.width = 36}
plot_grid(long_abun1, long_stat_plot,
          nrow = 1,
          labels = c('a)', 'b)'),
          label_size = 22,
          rel_widths = c(1, 0.8)) -> long_plots_together

long_plots_together
```

**Poster Presentation Stats**
these stats are for the shortened figure that I am putting on my poster presentation!
```{r, fig.height=11, fig.width=17}
poster_abun_stats <- abun_stats(short_abun_filt,
                                wanted_level)
poster_kruskal <- poster_abun_stats$KruskalTest
new_poster_dunn <- poster_abun_stats$DunnPostHoc
poster_family_abun_lm <- poster_abun_stats$LinearModel

stat_plot1 <- stat_plot(new_poster_dunn,
                        wanted_level)
stat_plot1
```

**Statistical Figure Visualization**
this is option #2. 
```{r, fig.height=7, fig.width=9}
## option #2 based on the linear model
poster_family_abun_lm %>% 
  filter(term != 'seq_depth',
         term != 'purified_diet:seq_depth') %>% 
  mutate(day_fact = as.factor(day_post_inf)) %>% 
  ggplot(aes(x = day_fact, y = term)) +
  geom_tile(aes(fill = p.value), alpha = 0.5) +
  geom_text(aes(label = signif)) +
  theme_gray(base_size = 16) +
  scale_fill_viridis(limits = c(0, 0.05), option = "H", name = 'P-Value') +
  facet_wrap(~Family,
             scales = "free_x",
             ncol = 1,
             strip.position = "right") +
  theme(strip.text.y = element_text(angle = 0)) +
  xlab('Days Relative to Infection') +
  ylab('Diet Composition') -> stat_plot2

stat_plot2
```

**Putting my Plots Together**
I'm doing this for my poster presentation!
```{r, fig.height = 15, fig.width = 35}
plot_grid(short_abun1, stat_plot1,
          nrow = 1,
          labels = c('a)', 'b)'),
          label_size = 22,
          rel_widths = c(1, 0.8)) -> short_plots_together

short_plots_together
```

**Plotting the Relative Abundance of C. diff Specifically**
could go to the genus level because all species under the genus level were c. diff
```{r, warning=FALSE, fig.height=5, fig.width=15}
cdiff_files <- family_abun_file_prep(metadata_FP,
                                     tax_FP,
                                     otu_table_FP,
                                     cdiff_level,
                                     wanted_genus)

cdiff_abun_filt <- cdiff_files$AbundanceTable

cdiff_plots <- abun_plots(cdiff_abun_filt,
                         cdiff_level)

cdiff_plot1 <- cdiff_plots$FamilyAbundance1

cdiff_stats <- abun_stats(cdiff_abun_filt,
                          cdiff_level)
```


**Saving my Outputs**
These go to the plots and stats directories. 
```{r}
## microbe family relative abundance plots as a .pdf
## option #1
ggsave("family_abun1.pdf",
       plot = long_abun1, 
       width = 17, 
       height = 15, 
       path = '../plots')
ggsave("family_abun1_w_stats.pdf",
       plot = long_plots_together, 
       width = 36, 
       height = 16, 
       path = '../plots')
## option #2
ggsave("family_abun2.pdf",
       plot = long_abun2, 
       width = 25, 
       height = 10, 
       path = '../plots')

## family abundance plots for poster presentation
ggsave("poster_family_abun.pdf",
       plot = short_abun1, 
       width = 17, 
       height = 10, 
       path = '../plots')

ggsave("poster_famAbun_stats1.pdf",
       plot = stat_plot1, 
       width = 14.5, 
       height = 7.5, 
       path = '../plots')

ggsave("poster_famAbun_stats2.pdf",
       plot = stat_plot2, 
       width = 9, 
       height = 5, 
       path = '../plots')

ggsave("poster_famAbun_together.pdf",
       plot = plots_together, 
       width = 35, 
       height = 10, 
       path = '../plots')

ggsave("cdiff_rel_abun.pdf",
       plot = cdiff_plot1, 
       width = 15, 
       height = 5, 
       path = '../plots')

## plots as an rdata object
save(long_abun1,
     file = '../figures/long_fam_abun.rdata')
save(long_stat_plot,
     file = '../figures/long_fam_abun_stats.rdata')
save(short_abun1,
     file = '../figures/short_fam_abun.rdata')
save(stat_plot1,
     file = '../figures/short_fam_abun_stats.rdata')

## linear modeling results
write_tsv(family_abun_lm,
          '../stats/family_abun_lm.tsv')
write_tsv(poster_family_abun_lm,
          '../stats/poster_family_abun_lm.tsv')
write_tsv(new_poster_dunn,
          '../stats/family_abun_dunn.tsv')
```
