---
title: "histology"
format: html
editor: visual
---

```{r, include=FALSE}
library(broom)
library(cowplot)
library(magrittr)
library(qiime2R)
library(tidyverse)
library(rstatix)
library(ggpubr)
```

TO DO:

-   look at histopathology scores between batches - studies (look at each one separately) to see if there's a batch effect, especially in the high-fat/low-fiber diet.

-   combined analyses with data from previous study published

**File Paths**

```{r}
metadata_FP <- '../data/misc/processed_metadata.tsv'
seq_depth_FP <- '../data/misc/tss_seq_depth.tsv'
histo_FP <- '../data/misc/histo_data.csv'

## lists to redo the diet names on the facet labels of the ggplot created below 
diet_labs <- 
    c('Chow', 
      'High Fat / High Fiber', 
      'High Fat / Low Fiber', 
      'Low Fat / High Fiber', 
      'Low Fat / Low Fiber')

names(diet_labs) <- c('Chow', 'HF/HF', 'HF/LF', 'LF/HF', 'LF/LF')

tissue_labs <- c('Cecum',
                 'Colon')
names(tissue_labs) <- c('cecum',
                        'colon')
```

**Reading in Metadata and Histopathology Files**

the bulk of the histopathology scores are from mice that died before day 3

```{r}
metadata <- read_tsv(metadata_FP)

histo <- read_csv(histo_FP) %>% 
  filter(!is.na(mouse_id))

wanted_ids <- histo$mouse_id

## joining them all together for ggplot rendering 
metadata %>% 
  merge(histo, by = 'mouse_id') %>% 
  group_by(mouse_id) %>% 
  filter(day_post_inf == max(day_post_inf)) %>% 
  ungroup() %>% 
  mutate(day_post_inf = as.factor(day_post_inf)) %>% 
  gather(cecum, colon, key = tissue, value = score) -> big_histo
```

**Histopathology Score Kruskal-Wallis and Dunn's Post Hoc Tests by Diet**

All mice and days are included in the below statistical analysis and ggplot.

```{r}
big_histo %>% 
  group_by(tissue) %>% 
  do(tidy(kruskal.test(score ~ diet,
                       data = .))) %>% 
  ungroup() %>%
  arrange(p.value) %>%
  mutate(p.adj = p.adjust(p.value,
                          method = "BH"),
        test_id = paste(tissue)) -> all_day_kruskal

big_histo %>% 
  group_by(tissue) %>% 
  mutate(test_id = paste(tissue)) %>% 
  filter(test_id %in% all_day_kruskal$test_id) %>% 
  dunn_test(score ~ diet,
            p.adjust.method = 'BH',
            data =.) %>% 
  add_y_position(scales = 'free_y', step.increase = 0) -> all_day_dunn

all_day_dunn
```

**Histopathology Score by Diet Linear Modeling**\
running a linear model on the histopathology of all mice, regardless of day of death and cecal removal.

```{r}
big_histo %>% 
  group_by(tissue) %>% 
  do(glance(lm(score ~ (purified_diet * seq_depth) + high_fat * high_fiber + study,
             data =.))) %>% 
  ungroup() %>%
  na.omit() %>%
  mutate(adj.p = p.adjust(p.value,
                          method = "BH"),
        test_id = paste(tissue)) %>% 
  filter(adj.p <= 0.05) -> big_histo_lm_full

big_histo %>% 
  group_by(tissue) %>% 
  mutate(test_id = paste(tissue)) %>% 
  filter(test_id %in% big_histo_lm_full$test_id) %>% 
  do(tidy(lm(score ~ (purified_diet * seq_depth) + high_fat * high_fiber + study,
             data =.))) %>%
  filter(term != '(Intercept)') %>% 
  na.omit() -> big_histo_lm

big_histo_lm['signif'] <- symnum(big_histo_lm$p.value,
                                 cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
                                 symbols = c("****", "***", "**", "*", "+", "ns"),
                                 abbr.colnames = FALSE,
                                 na = "")
big_histo_lm
```

**Histopathology Score by Diet Plot**

```{r, fig.height=6, fig.width=10}
big_histo %>% 
  mutate(day_post_inf = as.numeric(day_post_inf)) %>% 
  ggplot(aes(x = diet, y = score)) +
  geom_violin(aes(group = diet),  draw_quantiles = c(0.5)) +
  geom_jitter(alpha = 0.4, width = 0.1, height = 0) +
  scale_x_discrete(labels = c('Chow', 
                              'HFt/\nHFb', 
                              'HFt/\nLFb',
                              'LFt/\nHFb', 
                              'LFt/\nLFb')) +
  facet_wrap(~tissue, labeller = labeller(tissue = tissue_labs),
             scales = "free_y") +
  stat_pvalue_manual(all_day_dunn,
                     tip.length = 0.01,
                     label = 'p.adj.signif',
                     hide.ns = TRUE,
                     step.increase = 0.1) +
  theme_bw(base_size = 20) +
  theme(strip.text.y = element_text(angle = 0)) +
  xlab('Diet') +
  ylab('Histopathology Score') +
  ggtitle("Histopathology Score by Diet") -> all_day_plot

all_day_plot
```

**Running Kruskal Wallis, Dunn's Post Hoc, and T-Tests on histopathology score by barrier facility**

You only need a t-test for this analysis since you're only doing two comparisons. I'm keeping the Dunn's Post Hoc test on here because I can't get the t-test to show up on the plot below.

```{r}
## kruskal wallis test by barrier facility
big_histo %>% 
  group_by(tissue, diet) %>% 
  do(tidy(kruskal.test(score ~ facility,
                       data = .))) %>% 
  ungroup() %>%
  arrange(p.value) %>%
  mutate(p.adj = p.adjust(p.value,
                          method = "BH"),
        test_id = paste(tissue, diet, sep = "_")) -> barrier_kruskal

## dunn's post hoc test
big_histo %>% 
  group_by(tissue, diet) %>% 
  mutate(test_id = paste(tissue, diet, sep = "_")) %>% 
  filter(test_id %in% barrier_kruskal$test_id) %>% 
  dunn_test(score ~ facility,
            p.adjust.method = 'BH',
            data =.) %>% 
  add_xy_position(scales = 'free', step.increase = 0) -> barrier_dunn

## this is a t test based on histopathology score of tissue site per diet based on the barrier facility the mice came from 
big_histo %>% 
  group_by(tissue, diet) %>% 
  do(tidy(t.test(score ~ facility,
                 data =.))) %>% 
  mutate(adj.p.value = p.adjust(p.value)) %>% 
  filter(adj.p.value <= 0.05) -> barrier_t_test

barrier_t_test['p.adj.signif'] <- symnum(barrier_t_test$adj.p.value,
                                         cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
                                         symbols = c("****", "***", "**", "*", "+", "ns"),
                                         abbr.colnames = FALSE,
                                         na = "")

barrier_t_test
```

**Histopathology Score by Barrier Facility Plot**

```{r, fig.height=6.5, fig.width=12}
big_histo %>% 
  mutate(day_post_inf = as.numeric(day_post_inf)) %>% 
  ggplot(aes(x = facility, y = score)) +
  geom_violin(aes(group = facility),  draw_quantiles = c(0.5)) +
  geom_jitter(alpha = 0.4, width = 0.1, height = 0) +
  scale_x_discrete(labels = c(1, 2)) +
  facet_grid(tissue~diet, labeller = labeller(tissue = tissue_labs,
                                              diet = diet_labs),
             scales = "free_y") +
  stat_pvalue_manual(barrier_dunn,
                     tip.length = 0.01,
                     label = 'p.adj.signif',
                     hide.ns = TRUE,
                     step.increase = 0.5) +
  theme_bw(base_size = 16) +
  theme(strip.text.y = element_text(angle = 0)) +
  xlab('Facility') +
  ylab('Score') +
  ggtitle("Histopathology Score by Diet per Barrier Facility") -> barrier_plot

barrier_plot
```

**Histopathology Score by Batch (aka study in the metadata file)**\
Statistical analysis! I chose to run a Kruskal-Wallis followed by a Dunn's Post Hoc test for this plot since there are multiple comparisons between mouse batches.

```{r}
## kruskal wallis test by barrier facility
big_histo %>% 
  group_by(tissue, diet) %>% 
  do(tidy(kruskal.test(score ~ study,
                       data = .))) %>% 
  ungroup() %>%
  arrange(p.value) %>%
  mutate(p.adj = p.adjust(p.value,
                          method = "BH"),
        test_id = paste(tissue, diet, sep = "_")) %>% 
  filter(p.adj <= 0.05) -> batch_kruskal

## dunn's post hoc test
big_histo %>% 
  group_by(tissue, diet) %>% 
  mutate(test_id = paste(tissue, diet, sep = "_")) %>% 
  filter(test_id %in% batch_kruskal$test_id) %>% 
  dunn_test(score ~ study,
            p.adjust.method = 'BH',
            data =.) %>% 
  add_xy_position(scales = 'free', step.increase = 0) -> batch_dunn

batch_dunn
```

Plot!

Bad news, the mice on each diet are not evenly distributed between batches (or "studies") so each batch is not equally represented between each diet. Some diets have more batches than others but no diet has every batch represented in it. I assume this is because we didn't account for mouse batch in the study design? Each batch also does not have an equal number of mice/samples in it.

```{r, fig.height=6.5, fig.width=12, warning=FALSE}
big_histo %>% 
  mutate(day_post_inf = as.numeric(day_post_inf)) %>% 
  ggplot(aes(x = study, y = score)) +
  geom_violin(aes(group = study),  draw_quantiles = c(0.5)) +
  geom_jitter(alpha = 0.4, width = 0.1, height = 0) +
  scale_x_continuous(breaks = c(1, 2, 3, 4, 5, 6, 7)) +
  facet_grid(tissue~diet, labeller = labeller(tissue = tissue_labs,
                                              diet = diet_labs),
             scales = "free_y") +
  stat_pvalue_manual(batch_dunn,
                     tip.length = 0.01,
                     label = 'p.adj.signif',
                     hide.ns = TRUE,
                     step.increase = 0.1) +
  theme_bw(base_size = 16) +
  theme(strip.text.y = element_text(angle = 0)) +
  xlab('Batch') +
  ylab('Score') +
  ggtitle("Histopathology Score by Diet per Mouse Batch") -> batch_plot

batch_plot
```

```{r}
big_histo %>% 
  count(study)
```

**Saving Plot and Stats**

```{r}
## regular plots
ggsave("histopathology_all.pdf", 
       plot = all_day_plot,
       width = 10, 
       height = 6,
       path = '../plots')
write_tsv(big_histo_lm,
          '../stats/histopathology_all.tsv')

## by barrier facility plots
ggsave("histo_facility_comp.pdf", 
       plot = barrier_plot,
       width = 12, 
       height = 6.5,
       path = '../plots')
write_tsv(barrier_dunn,
          '../stats/histo_facil_dunn.tsv')
write_tsv(barrier_t_test,
          '../stats/histo_facil_tTest.tsv')

## by mouse batch (aka study)
ggsave("histo_batch_comp.pdf", 
       plot = batch_plot,
       width = 12, 
       height = 6.5,
       path = '../plots')
write_tsv(batch_dunn,
          '../stats/histo_batch_dunn.tsv')

```
