---
title: "histology"
format: html
editor: visual
---

```{r, include=FALSE}
library(broom)
library(cowplot)
library(magrittr)
library(qiime2R)
library(tidyverse)
```

**Basic Functions**

```{r}
## lists to redo the diet names on the facet labels of the ggplot created below 
diet_labs <- 
    c('Chow', 
      'High Fat / High Fiber', 
      'High Fat / Low Fiber', 
      'Low Fat / High Fiber', 
      'Low Fat / Low Fiber')

names(diet_labs) <- c('Chow', 'HF/HF', 'HF/LF', 'LF/HF', 'LF/LF')

tissue_labs <- c('Cecum',
                 'Colon')
names(tissue_labs) <- c('cecum',
                        'colon')

## general function to prep the metadata file for further data analyses 
metadata_fixer <- function(metadata_fp) {
  tmpMeta <- read_tsv(metadata_fp, n_max = 2)
  mycols <- colnames(tmpMeta)
  metadata <- read_tsv(metadata_fp, skip = 2, col_names = mycols)
  names(metadata)[names(metadata) == '#SampleID'] <- 'sampleid'
  metadata %>% 
    filter(!is.na(diet)) %>% 
    mutate(day_post_inf = if_else(day_post_inf == 2, 3, day_post_inf)) %>% 
    mutate(diet = as.factor(diet)) -> metadata
  return(metadata)
}

## for editing my metadata file post metadata fixer 
meta_diet_fixer <- function(metadata_file,
                            seq_depth_fp){
  seq_depths <- read_tsv(seq_depth_fp)
  metadata_file %>% 
    select(sampleid, diet, day_post_inf, mouse_id, study) %>% 
    mutate(diet_true = diet,
    diet_true = if_else(day_post_inf == -15, "Chow", diet_true),
    high_fat = case_when(
      diet_true == 'HF/HF' ~ 1,
      diet_true == 'HF/LF' ~ 1,
      .default = 0
      ), 
      high_fiber = case_when(
      diet_true == 'HF/HF' ~ 1,
      diet_true == 'LF/HF' ~ 1,
      diet_true == 'Chow' ~ 1,
      .default = 0
      ), 
      purified_diet = case_when(
      diet_true == 'Chow' ~ 0,
      .default = 1
      )
    ) %>% 
    left_join(seq_depths) -> metadata
  return(metadata)
}
```

**File Paths**

```{r}
metadata_FP <- '../data/misc/processed_metadata.tsv'
seq_depth_FP <- '../data/misc/tss_seq_depth.tsv'
histo_FP <- '../data/misc/histo_data.csv'
```

**Reading in Metadata and Histopathology Files**

the bulk of the histopathology scores are from mice that died before day 3

```{r}
metadata <- read_tsv(metadata_FP)

histo <- read_csv(histo_FP) %>% 
  filter(!is.na(mouse_id))

wanted_ids <- histo$mouse_id

## joining them all together for ggplot rendering 
metadata %>% 
  merge(histo, by = 'mouse_id') %>% 
  group_by(mouse_id) %>% 
  filter(day_post_inf == max(day_post_inf)) %>% 
  ungroup() %>% 
  mutate(day_post_inf = as.factor(day_post_inf)) %>% 
  gather(cecum, colon, key = tissue, value = score) -> big_histo
```

**Histopathology Score Kruskal-Wallis and Dunn's Post Hoc Tests**

All mice and days are included in the below statistical analysis and ggplot.

```{r}
big_histo %>% 
  group_by(tissue, diet) %>% 
  do(tidy(kruskal.test(score ~ facility,
                       data = .))) %>% 
  ungroup() %>%
  arrange(p.value) %>%
  mutate(p.adj = p.adjust(p.value,
                          method = "BH"),
        test_id = paste(tissue, diet)) -> all_day_kruskal

big_histo %>% 
  group_by(tissue, diet) %>% 
  mutate(test_id = paste(tissue, diet)) %>% 
  filter(test_id %in% all_day_kruskal$test_id) %>% 
  dunn_test(score ~ facility,
            p.adjust.method = 'BH',
            data =.) %>% 
  add_y_position(scales = 'free_y', step.increase = 0) -> all_day_dunn

all_day_dunn
```

**Running a t test on histopathology score by barrier facility**

```{r}
## this is a t test
big_histo %>% 
  group_by(tissue, diet) %>% 
  do(tidy(t.test(score ~ facility,
                 data =.))) %>% 
  mutate(adj.p.value = p.adjust(p.value)) %>% 
  filter(adj.p.value <= 0.05) -> t_test

t_test
```

**Histopathology Plot with Significance Levels**

```{r, fig.height=6, fig.width=13}
big_histo %>% 
  mutate(day_post_inf = as.numeric(day_post_inf)) %>% 
  ggplot(aes(x = diet, y = score)) +
  geom_violin(aes(group = diet),  draw_quantiles = c(0.5)) +
  geom_jitter(alpha = 0.4, width = 0.1, height = 0) +
  scale_x_discrete(labels = c('Chow', 'HFt/\nHFb', 'HFt/\nLFb',
                              'LFt/\nHFb', 'LFt/\nLFb')) +
  facet_grid(tissue~facility, labeller = labeller(tissue = tissue_labs),
             scales = "free_y") +
  # stat_pvalue_manual(all_day_dunn,
  #                    tip.length = 0.01,
  #                    label = 'p.adj.signif',
  #                    hide.ns = TRUE,
  #                    step.increase = 0.1) +
  theme_bw(base_size = 20) +
  xlab('Diet') +
  ylab('Histopathology Score') +
  ggtitle("Histopathology Score by Diet") -> all_day_plot

all_day_plot
```

**Histopathology Score Linear Modeling**\
running a linear model on the histopathology of all mice, regardless of day of death and cecal removal.

```{r}
big_histo %>% 
  group_by(tissue) %>% 
  do(glance(lm(score ~ (purified_diet * seq_depth) + high_fat * high_fiber,
             data =.))) %>% 
  ungroup() %>%
  na.omit() %>%
  mutate(adj.p = p.adjust(p.value,
                          method = "BH"),
        test_id = paste(tissue)) %>% 
  filter(adj.p <= 0.05) -> big_histo_lm_full

big_histo %>% 
  group_by(tissue) %>% 
  mutate(test_id = paste(tissue)) %>% 
  filter(test_id %in% big_histo_lm_full$test_id) %>% 
  do(tidy(lm(score ~ (purified_diet * seq_depth) + high_fat * high_fiber,
             data =.))) %>%
  filter(term != '(Intercept)') %>% 
  na.omit() -> big_histo_lm

big_histo_lm['signif'] <- symnum(big_histo_lm$p.value,
                                 cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
                                 symbols = c("****", "***", "**", "*", "+", "ns"),
                                 abbr.colnames = FALSE,
                                 na = "")
big_histo_lm
```

**Plotting the Cecum and Colon Histopathology Scores**

comparing the histopathology scores between the cecum and colon, can see a positive correlation between the two.

```{r}
histo %>% 
  ggplot(aes(x = colon, y = cecum)) +
    geom_point() +
    geom_smooth(method = 'lm') +
    theme_bw(base_size = 14) -> histo_compare

histo_compare
```

**Linear Modeling for Cecum and Colon Histopathology Scores**

```{r}
histo %>% 
  do(tidy(lm(colon ~ cecum,
            data =.))) %>% 
  adjust_pvalue(method = 'BH')
```

**Saving Plot and Stats**

```{r}
ggsave("histopathology_all.pdf", 
       plot = all_day_plot,
       width = 13, 
       height = 6,
       path = '../plots')
write_tsv(big_histo_lm,
          '../stats/histopathology_all.tsv')
```
