---
title: "mouse_survival"
format: html
editor: visual
---

```{r, include=FALSE}
library(tidyverse)
library(survminer)
require(survival)
library(readr)
library(viridis)
```

**Input File Path**

```{r}
survival_fp <- '../data/misc/aim1a_survival.csv'
diet_labs <- 
  c('Chow', 
    'High Fat / High Fiber', 
    'High Fat / Low Fiber', 
    'Low Fat / High Fiber', 
    'Low Fat / Low Fiber')
names(diet_labs) <- c('chow',
                      'HF+fiber',
                      'HF/LF',
                      'LF+fiber',
                      'LF/LF')

fib_labs <- c('High Fiber',
              'Low Fiber')
names(fib_labs) <- c('high_fiber',
                     'low_fiber')

fat_labs <- c('High Fat',
              'Low Fat')
names(fat_labs) <- c('high_fat',
                     'low_fat')
```

**Reading in Mouse Survival Data**

col_types tells R what to classify the data in each column as

```{r}
survival <- read_csv(survival_fp,
                     col_types = cols(Day = col_integer(),
                                      status = col_integer(), 
                                      diet = col_character(),
                                      group = col_integer()))
```

**Creating Survival Plot by Diet**

Legit can't figure out how to take color by diet off of this plot so rip.

```{r, fig.height=10, fig.width=15}
diet_fit <- surv_fit(Surv(Day, status) ~ diet, data = survival)
ggsurvplot(diet_fit, 
           data = survival,
           color = 'black',
           ggtheme = theme_bw(base_size = 20),
           legend = "none",
           conf.int = TRUE, 
           xlab = 'Days Post Infection',
           ylab = 'Survival Probability',
           title = 'Mouse Survival After Infection by Diet') -> diet_plot

diet_plot$plot +
    facet_wrap(~factor(diet, levels = c('chow',
                                        'HF+fiber',
                                        'HF/LF',
                                        'LF+fiber',
                                        'LF/LF')),
               labeller = labeller(.cols = diet_labs),
               nrow = 2) +
    # scale_color_brewer(option = "H", discrete = TRUE) +
    scale_fill_grey() -> diet_plot_final

diet_plot_final
```

**Pairwise Comparisons**

```{r}
pairwise_survdiff(Surv(Day, status) ~ diet,
                  data = survival, 
                  p.adjust.method = "BH") -> diet_pairwise

diet_pvalue <- as_tibble(diet_pairwise$p.value,
                         rownames = 'diet')

## changing format from wide to long so I can work with it 
diet_pvalue %>% 
  gather(-diet, key = diet2, value = p.value) -> diet_pvalue

names(diet_pvalue)[names(diet_pvalue) == 'diet'] <- 'diet1'


diet_pvalue['signif'] <- symnum(diet_pvalue$p.value,
                                cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
                                symbols = c("****", "***", "**", "*", "+", "ns"),
                                abbr.colnames = FALSE,
                                na = "")

```

pairwise comparison visualization

REVERSE THE ORDER OF YOUR Y-AXIS (GROUP 2) ON ALL PLOTS!!

```{r, fig.height=3, fig.width=4.5}
diet_pvalue %>%
  na.omit() %>% 
  ggplot(aes(x = diet2, y = diet1)) +
  geom_tile(aes(fill = p.value), alpha = 0.6, color = 'black') +
  geom_text(aes(label = signif), size = 5) +
  # scale_fill_gradient2(low = 'blue', high = 'green', midpoint = 0.025) +
  scale_fill_distiller(limits = c(0, 0.05), palette = 'Reds', name = "P-Value") +
  theme_bw(base_size = 20) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 18),
        legend.position = 'none') +
  scale_x_discrete(labels = c("Chow",
                              "HFt/\nLFb",
                              "HFt/\nHFb",
                              "LFt/\nLFb")) +
  scale_y_discrete(labels = c("HFt / LFb",
                               "HFt / HFb",
                               "LFt / LFb",
                               "LFt / HFb")) -> surv_stat_vis

surv_stat_vis
```

**Cox PH for Diet**

```{r}
surv_obj <- Surv(time = survival$Day, event = survival$status)
survival$diet_f = factor(survival$diet, 
                         levels = c("HF/LF", 
                                    "LF+fiber",
                                    "LF/LF", 
                                    "chow", 
                                    "HF+fiber"))
diet.fit.cox <- coxph(surv_obj ~ diet_f, data = survival)
summary(diet.fit.cox)
```

**Putting Survival Plot by Diet with Statistical Visualization**\

```{r, fig.width=18, fig.height=12}
plot_grid(diet_plot_final, surv_stat_vis,
          nrow = 2, 
          labels = c('a)', 'b)'),
          label_size = 22,
          rel_heights = c(1, 0.5))
```

**Creating Survival Plot by Fiber Content**

```{r, fig.height=5, fig.width=10}
fiber_fit <- surv_fit(Surv(Day, status) ~ fiber, data = survival)
ggsurvplot(fiber_fit, 
           data = survival,
           ggtheme = theme_bw (base_size = 14),
           break.x.by = 1,
           legend = "none",
           font.x = 14, 
           font.y = 14, 
           pval = TRUE, 
           pval.method =TRUE,
           pval.size = 4,
           conf.int = TRUE, 
           xlab = 'Days Post Infection',
           ylab = 'Survival Probability',
           title = 'Mouse Survival After Infection by Diet Fiber Content') -> fiber_plot

fiber_plot$plot +
  facet_wrap(~fiber,
             labeller = labeller(fiber = fib_labs)) +
  scale_color_viridis(option = "H", discrete = TRUE) +
  scale_fill_viridis(option = "H", discrete = TRUE) -> fiber_plot_final

fiber_plot_final
```

**Cox PH for Fiber Content**

```{r}
surv_obj <- Surv(time = survival$Day, event = survival$status)
survival$fiber_f = factor(survival$fiber, 
                          levels = c("high_fiber", 
                                     "low_fiber"))
fiber.fit.cox <- coxph(surv_obj ~ fiber_f, data = survival)
summary(fiber.fit.cox)
```

**Survival Plot by Fat Content**

```{r, fig.width=10, fig.height=5}
fat_fit <- surv_fit(Surv(Day, status) ~ fat, data = survival)
ggsurvplot(fat_fit, 
           data = survival,
           ggtheme = theme_bw (base_size = 14),
           break.x.by = 1,
           legend = "none",
           font.x = 14, 
           font.y = 14, 
           pval = TRUE, 
           pval.method =TRUE,
           pval.size = 4, 
           conf.int = TRUE, 
           xlab = 'Days Post Infection',
           ylab = 'Survival Probability',
           title = 'Mouse Survival After Infection by Diet Fat Content') -> fat_plot

fat_plot$plot +
  facet_wrap(~fat,
             labeller = labeller(fat = fat_labs)) +
  scale_color_viridis(option = "H", discrete = TRUE) +
  scale_fill_viridis(option = "H", discrete = TRUE) -> fat_plot_final

fat_plot_final
```

**Cox PH by Fat Content**

```{r}
surv_obj <- Surv(time = survival$Day, event = survival$status)
survival$fat_f = factor(survival$fat, 
                        levels = c("low_fat", 
                                   "high_fat"))
fat.fit.cox <- coxph(surv_obj ~ fat_f, data = survival)
summary(fat.fit.cox)
```

**Repeating Above Analysis Without Chow Diet**

```{r}
no_chow <- filter(survival, diet != "chow")
```

Fiber Content Plot

```{r, fig.width=10, fig.height=5}
nochow_fib_fit <- surv_fit(Surv(Day, status) ~ fiber, data = no_chow)
ggsurvplot(nochow_fib_fit, 
           data = survival,
           ggtheme = theme_bw (base_size = 14),
           break.x.by = 1,
           legend = "none",
           font.x = 14, 
           font.y = 14, 
           pval = TRUE, 
           pval.method =TRUE,
           pval.size = 4,
           conf.int = TRUE, 
           xlab = 'Days Post Infection',
           ylab = 'Survival Probability',
           title = 'Mouse Survival by Diet Fiber Content (No Chow)') -> nc_fiber_plot

nc_fiber_plot$plot +
  facet_wrap(~fiber,
             labeller = labeller(fiber = fib_labs)) +
  scale_color_viridis(option = "H", discrete = TRUE) +
  scale_fill_viridis(option = "H", discrete = TRUE) -> nc_fiber_plot_final

nc_fiber_plot_final
```

Fiber Content Cox PH

```{r}
nc_surv_obj <- Surv(time = no_chow$Day, event = no_chow$status)
no_chow$fiber_f = factor(no_chow$fiber, 
                         levels = c("low_fiber", 
                                    "high_fiber"))
nc.fib.fit.cox <- coxph(nc_surv_obj ~ fiber_f, data = no_chow)
summary(nc.fib.fit.cox)
```

Fat Content Plot

```{r, fig.height=5, fig.width=10}
nc_fat_fit <- surv_fit(Surv(Day, status) ~ fat, data = no_chow)
ggsurvplot(nc_fat_fit, 
           data = survival,
           ggtheme = theme_bw(base_size = 14),
           break.x.by = 1,
           legend = "none",
           font.x = 14, 
           font.y = 14, 
           pval = TRUE, 
           pval.method =TRUE,
           pval.size = 4,
           conf.int = TRUE, 
           xlab = 'Days Post Infection',
           ylab = 'Survival Probability',
           title = 'Mouse Survival by Diet Fat Content (No Chow)') -> nc_fat_plot

nc_fat_plot$plot +
  facet_wrap(~fat,
             labeller = labeller(fat = fat_labs)) +
  scale_color_viridis(option = "H", discrete = TRUE) +
  scale_fill_viridis(option = "H", discrete = TRUE) -> nc_fat_plot_final

nc_fat_plot_final
```

Fat Content Cox PH

```{r}
nc_surv_obj <- Surv(time = no_chow$Day, event = no_chow$status)
no_chow$fat_f = factor(no_chow$fat, 
                       levels = c("low_fat", 
                                  "high_fat"))
nc.fat.fit.cox <- coxph(nc_surv_obj ~ fat_f, data = no_chow)
summary(nc.fat.fit.cox)
```

**Saving my Outputs**

need to have plot = print() to save ggsurvplot plots! regular ggsave doesn't work.

```{r}
## have to extract the actual plot from the ggsurvplot object to save it and 
## actually see it, otherwise it won't work
# surv_all <- diet_plot$plot
# fib_surv <- fiber_plot$plot
# fat_surv <- fat_plot$plot

ggsave('survival_curve_all.pdf',
       plot = diet_plot_final,
       width = 15,
       height = 10,
       path = '../plots')
ggsave('fiber_survival_curve.pdf',
       plot = fiber_plot_final,
       width = 10,
       height = 5,
       path = '../plots')
ggsave('fat_survival_curve.pdf',
       plot = fat_plot_final,
       width = 10,
       height = 5,
       path = '../plots')
ggsave('survival_curve_stats.pdf',
       plot = surv_stat_vis,
       width = 4.5,
       height = 3,
       path = '../plots')
```
