---
title: "Resilience and Homogeneity"
format: html
editor: visual
---

```{r include=FALSE}
library(broom)
library(cowplot)
library(magrittr)
library(qiime2R)
library(tidyverse)
library(virdis)
```

```{r}
## general function to prep the metadata file for further data analyses 
metadata_fixer <- function(metadata_fp) {
  tmpMeta <- read_tsv(metadata_fp, n_max = 2)
  mycols <- colnames(tmpMeta)
  metadata <- read_tsv(metadata_fp, skip = 2, col_names = mycols)
  names(metadata)[names(metadata) == '#SampleID'] <- 'sampleid'
  metadata %>% 
    filter(!is.na(diet)) %>% 
    mutate(day_post_inf = if_else(day_post_inf == 2, 3, day_post_inf)) %>% 
    mutate(diet = as.factor(diet)) -> metadata
  return(metadata)
}

## for editing my metadata file post metadata fixer 
meta_diet_fixer <- function(metadata_file,
                            seq_depth_fp){
  seq_depths <- read_tsv(seq_depth_fp)
  metadata_file %>% 
    select(sampleid, diet, day_post_inf, mouse_id, study) %>% 
    mutate(diet_true = diet,
    diet_true = if_else(day_post_inf == -15, "Chow", diet_true),
    high_fat = case_when(
      diet_true == 'HF/HF' ~ 1,
      diet_true == 'HF/LF' ~ 1,
      .default = 0
      ), 
      high_fiber = case_when(
      diet_true == 'HF/HF' ~ 1,
      diet_true == 'LF/HF' ~ 1,
      diet_true == 'Chow' ~ 1,
      .default = 0
      ), 
      purified_diet = case_when(
      diet_true == 'Chow' ~ 0,
      .default = 1
      )
    ) %>% 
    left_join(seq_depths) -> metadata
  return(metadata)
}

## this function combines your metadata file and your distance matrix 
## it also pulls out the wanted diet variable from the metadata file 
## requires the input of a distance matrix (dist_mat), metadata file (meta),
## and the list of unique diets (mydiet)
## used in subsequent resil/homog_diet_assembly functions
dist_filter <- function(dist_mat, meta, mydiet){
  meta %>% 
    filter(diet == mydiet) -> min_meta

  dist_mat <- data.frame(dist_mat)
  rownames(dist_mat) <- dist_mat[,1]
  dist_mat <- dist_mat[,-1]
  
  sample_idx <- rownames(dist_mat) %in% min_meta$sampleid 
  
  min_dist_mat <- dist_mat[sample_idx, sample_idx]
  min_dist_mat <- as_tibble(min_dist_mat, rownames = 'sampleid')
  colnames(min_dist_mat)[2:ncol(min_dist_mat)] <- min_dist_mat$sampleid
  return(min_dist_mat)
}

stat_plot_prep <- function(biom_table,
                           dunn_test){
  biom_table %>% 
    group_by(diet, day_post_inf.y) %>% 
    summarise(mean_dist = mean(dist)) -> mean_dist
  dunn_test %>% 
    merge(mean_dist, 
        by.x = c('group1',
                 'day_post_inf.y'),
        by.y = c('diet',
                 'day_post_inf.y')) %>% 
    rename('group1_dist' = 'mean_dist') %>% 
    merge(mean_dist,
        by.x = c('group2',
                 'day_post_inf.y'),
        by.y = c('diet',
                 'day_post_inf.y')) %>% 
    rename('group2_dist' = 'mean_dist') %>% 
    mutate(diff_means = (group1_dist - group2_dist),
          stat_diff_means = if_else(p.adj > 0.05, 0, diff_means)) -> new_dunn
  return(new_dunn)
}

stat_plot <- function(new_dunn){
  new_dunn %>% 
    filter(day_post_inf.y != -15) %>%
    ggplot(aes(x = group1, y = group2)) +
    geom_tile(aes(fill = stat_diff_means), alpha = 0.8, color = 'black') +
    scale_fill_gradient2(low = 'blue', high = 'green', name = 'Group 1 -\nGroup 2') +
    geom_text(aes(label = p.adj.signif)) +
    scale_x_discrete(labels = c('Chow',
                                'HFt/\nHFb',
                                'HFt/\nLFb',
                                'LFt/\nHFb')) +
    scale_y_discrete(labels = c('LFt / LFb',
                                'LFt / HFb',
                                'HFt / LFb',
                                'HFt / HFb')) +
    facet_grid(~day_post_inf.y) +
    theme_bw(base_size = 20) +
    theme(strip.text.y = element_text(angle = 0)) +
    xlab('Group 1') +
    ylab('Group 2') -> stat_vis
  return(stat_vis)
}
```

Load in datums

Needed file paths for analysis and reading in the needed files for simplicity later on.

```{r, include=FALSE}
metadata_FP <- '../data/misc/merged_metadata1.tsv'
seq_depth_FP <- '../data/misc/tss_seq_depth.tsv'
uu_dist_fp <- '../data/qiime/core_outputs/uw_dist_matrix.tsv'
wu_dist_fp <- '../data/qiime/core_outputs/w_dist_matrix.tsv'
## lists to redo the diet names on the facet labels of the ggplot created below 
diet_labs <- 
    c('Chow', 
      'High Fat / High Fiber', 
      'High Fat / Low Fiber', 
      'Low Fat / High Fiber', 
      'Low Fat / Low Fiber')

names(diet_labs) <- c('Chow', 'HF/HF', 'HF/LF', 'LF/HF', 'LF/LF')
stat_labs <- c('Chow',
               'High Fat / High Fiber', 
               'High Fat / Low Fiber', 
               'Low Fat / High Fiber')
names(stat_labs) <- c('Chow',
                      'HF/HF',
                      'HF/LF',
                      'LF/HF')

prep_meta <- metadata_fixer(metadata_fp = metadata_FP)
meta <- meta_diet_fixer(prep_meta,
                        seq_depth_FP)
uu_dist <- read_tsv(uu_dist_fp)
wu_dist <- read_tsv(wu_dist_fp)
## pulling out unique diets 
diets <- unique(meta$diet)
```

# Functions for Homogeneity and Resilience Calculations

```{r}
## resiliency ggplot prep functions 

## this function takes the output from dist_filter and the metadata file 
## and pulls all unique distance matrix comparisons out for that particular diet 
## and places them in a new column 
## it also esta the baseline that all sample pairs will be compared to 
## requires the distance matrix output from the dist_filter (min_dist_mat),
## and the metadata file (meta)
resil_dist_comp <- function(min_dist_mat, meta){
  meta %>% 
    select(sampleid, diet, day_post_inf, high_fat, high_fiber, purified_diet,
           seq_depth, mouse_id) -> min_meta
  
  min_dist_mat %>% 
    rename(sampleid_x = sampleid) %>% 
    gather(-sampleid_x, key = sampleid_y, value = dist) %>% 
    merge(min_meta, by.x = 'sampleid_x', by.y = 'sampleid') %>% 
    merge(min_meta, by.x = 'sampleid_y', by.y = 'sampleid') -> long_dist_mat
  
  long_dist_mat %>% 
    mutate(key = paste0(pmin(sampleid_x, sampleid_y), 
                        pmax(sampleid_x, sampleid_y), 
                        sep = "_")) %>% 
    distinct(key, .keep_all = TRUE) %>% 
    filter(sampleid_x != sampleid_y,
           ## for homogenicity testing, do day_post_inf.x == day_post_inf.y 
           ## and don't filter out day -15 
           day_post_inf.x == -8,
           day_post_inf.y > -8) %>% 
    select(-diet.y) %>% 
    rename(diet = diet.x) -> long_dist_mat
  
  return(long_dist_mat)
}

## this is a for loop that does the above analysis for each diet variable and places them 
## in a list of dataframes that can be bound together to create one plot for comparison
## requires the input of a distance matrix (dist_mat), metadata file (meta),
## and the list of unique diets (mydiet)
resil_diet_assembly <- function(dist_mat, 
                                meta,
                                mydiet){
  resil <- list()
    for (i in 1:length(mydiet)){
      tmp_dist_mat <- dist_filter(dist_mat, meta, mydiet[i])
      tmp_resil <- resil_dist_comp(tmp_dist_mat, meta)
      resil[[i]] <- tmp_resil
    }
  resil <- bind_rows(resil)
  return(resil)
}


## homogeneity ggplot prep functions 

## this function takes the output from dist_filter and the metadata file 
## and pulls all unique distance matrix comparisons out for that particular diet 
## and places them in a new column 
## it also esta the baseline that all sample pairs will be compared to 
## requires the distance matrix output from the dist_filter (min_dist_mat),
## and the metadata file (meta)
homog_dist_comp <- function(min_dist_mat, meta){
  meta %>% 
    select(sampleid, diet, day_post_inf, high_fat, high_fiber, purified_diet,
           seq_depth, mouse_id) -> min_meta
  
  min_dist_mat %>% 
    rename(sampleid_x = sampleid) %>% 
    gather(-sampleid_x, key = sampleid_y, value = dist) %>% 
    merge(min_meta, by.x = 'sampleid_x', by.y = 'sampleid') %>% 
    merge(min_meta, by.x = 'sampleid_y', by.y = 'sampleid') -> long_dist_mat
  
  long_dist_mat %>% 
    mutate(key = paste0(pmin(sampleid_x, sampleid_y), 
                        pmax(sampleid_x, sampleid_y), 
                        sep = "_")) %>% 
    distinct(key, .keep_all = TRUE) %>% 
    filter(sampleid_x != sampleid_y,
           ## for homogenicity testing, do day_post_inf.x == day_post_inf.y 
           ## and don't filter out day -15 
           day_post_inf.x == day_post_inf.y) %>% 
           select(-diet.y) %>% 
           rename(diet = diet.x) -> long_dist_mat
  
  return(long_dist_mat)
}

## this is a for loop that does the above analysis for each diet variable and places them 
## in a list of dataframes that can be bound together to create one plot for comparison 
## requires the input of a distance matrix (dist_mat), metadata file (meta),
## and the list of unique diets (mydiet)
homog_diet_assembly <- function(dist_mat, 
                                meta,
                                mydiet){
  homog <- list()
    for (i in 1:length(mydiet)){
      tmp_dist_mat <- dist_filter(dist_mat, meta, mydiet[i])
      tmp_homog <- homog_dist_comp(tmp_dist_mat, meta)
      homog[[i]] <- tmp_homog
    }
  homog <- bind_rows(homog)
  return(homog)
}

```

# Resilience and Homogeneity Calculations

```{r}
## unweighted unifrac
uu_resil <- resil_diet_assembly(uu_dist,
                                meta,
                                diets)

## weighted unifrac
wu_resil <- resil_diet_assembly(wu_dist,
                                meta,
                                diets)
## unweighted unifrac
uu_homog <- homog_diet_assembly(uu_dist,
                                meta,
                                diets)

## weighted unifrac
wu_homog <- homog_diet_assembly(wu_dist,
                                meta,
                                diets)
```

**Unweighted UniFrac Resiliency Plot**

This plot looks at microbiome resiliency over time by the five different diets using the unweighted UniFrac distance matrix and metadata file (above functions).

```{r, warning=FALSE, fig.width=14, fig.height=5}
uu_resil %>% 
  ggplot(aes(x = day_post_inf.y, y = dist)) +
    geom_boxplot((aes(group = day_post_inf.y)), outlier.shape = NA) +
    geom_vline(xintercept = -3, linetype = 'dashed', color = 'red', size = 0.2) +
    geom_vline(xintercept = 0, linetype = 'dashed', color = 'purple', size = 0.2) +
    geom_line(aes(group = mouse_id.y), alpha = 0.1) +
    geom_jitter(width = 0.1, alpha = 0.2) +
    geom_smooth(se = FALSE, method = 'loess') + 
    scale_x_continuous(breaks = c(-3, 0, 3)) + 
    scale_y_reverse() +
    facet_wrap(~diet, 
               labeller = labeller(diet = diet_labs),
               nrow = 1) +
    theme_bw(base_size = 20) +
    xlab('Days Relative to Infection') +
    ylab('Unweighted UniFrac Distance\n(Pairwise Distances from Day -8)') +
    ggtitle("Microbiome Resilience Over Time") -> uu_resil_plot

uu_resil_plot
```

**Unweighted UniFrac Resiliency Linear Modeling**

Statistical analysis by diet for the unweighted UniFrac resiliency plot above.

```{r}
uu_resil %>% 
  group_by(day_post_inf.y) %>% 
  do(tidy(lm(dist ~ (purified_diet.y * seq_depth.y) + high_fat.y + high_fiber.y,
             data = .))) %>% 
  filter(p.value <= 0.05) -> uu_resil_results

uu_resil_results
```

**Weighted UniFrac Resiliency Plot**

This plot looks at microbiome resiliency over time by the five different diets using the weighted UniFrac distance matrix and metadata file (above functions).

```{r, warning=FALSE, fig.width=14, fig.height=5}
wu_resil %>% 
  ggplot(aes(x = day_post_inf.y, y = dist)) +
    geom_boxplot((aes(group = day_post_inf.y)), outlier.shape = NA) +
    geom_vline(xintercept = -3, linetype = 'dashed', color = 'red', size = 0.2) +
    geom_vline(xintercept = 0, linetype = 'dashed', color = 'purple', size = 0.2) +
    geom_line(aes(group = mouse_id.y), alpha = 0.1) +
    geom_jitter(width = 0.1, alpha = 0.2) +
    geom_smooth(se = FALSE, method = 'loess') + 
    scale_x_continuous(breaks = c(-3, 0, 3)) + 
    scale_y_reverse() +
    facet_wrap(~diet, 
               labeller = labeller(diet = diet_labs),
               nrow = 1) +
    theme_bw(base_size = 20) +
    xlab('Days Relative to Infection') +
    ylab('Weighted UniFrac Distance\n(Pairwise Distances from Day -8)') +
    ggtitle("Microbiome Resilience Over Time") -> wu_resil_plot

wu_resil_plot
```

**Weighted UniFrac Resiliency Linear Modeling**

```{r}
wu_resil %>% 
  group_by(day_post_inf.y) %>% 
  do(tidy(lm(dist ~ (purified_diet.y * seq_depth.y) + high_fat.y + high_fiber.y,
             data = .))) %>% 
  filter(p.value <= 0.05) -> wu_resil_results


## kruskal wallis and dunns post hoc test
wu_resil %>% 
  na.omit() %>% 
  group_by(day_post_inf.y) %>% 
  do(tidy(kruskal.test(dist ~ diet,
             data = .))) -> w_resil_kruskal

wu_resil %>% 
  na.omit() %>% 
  group_by(day_post_inf.y) %>% 
  dunn_test(dist ~ diet,
            p.adjust.method = 'BH',
            data = .) -> w_resil_dunn

stat_plot_prep(wu_resil,
               w_resil_dunn) -> new_w_resil_dunn

```

Weighted UniFrac Resiliency Dunn's Post Hoc Visualization

```{r, fig.height=4, fig.width=14}
stat_plot(new_w_resil_dunn) -> w_resil_stat_vis

w_resil_stat_vis
```

**Unweighted UniFrac Homogeneity Plot**\

```{r, warning=FALSE, fig.width=14, fig.height=5}
uu_homog %>% 
  ggplot(aes(x = day_post_inf.y, y = dist)) +
    geom_boxplot((aes(group = day_post_inf.y)), outlier.shape = NA) +
    geom_vline(xintercept = -3, linetype = 'dashed', color = 'red', size = 0.2) +
    geom_vline(xintercept = 0, linetype = 'dashed', color = 'purple', size = 0.2) +
    geom_line(aes(group = mouse_id.y), alpha = 0.1) +
    geom_jitter(width = 0.1, alpha = 0.2) +
    geom_smooth(se = FALSE, method = 'loess', color = 'blue') + 
    scale_x_continuous(breaks = c(-15, -8, -3, 0, 3)) + 
    scale_y_reverse() +
    facet_wrap(~diet, 
               labeller = labeller(diet = diet_labs),
               nrow = 1) +
    theme_bw(base_size = 20) +
    xlab('Days Relative to Infection') +
    ylab('Unweighted UniFrac Distance\n(Pairwise Distances within Day)') +
    ggtitle("Microbiome Homogeneity Over Time") -> uu_homog_plot

uu_homog_plot
```

**Unweighted UniFrac Homogeneity Linear Modeling**

```{r}
uu_homog %>% 
  group_by(day_post_inf.y) %>% 
  do(tidy(lm(dist ~ (purified_diet.y * seq_depth.y) + high_fat.y + high_fiber.y,
             data = .))) %>% 
  filter(p.value <= 0.05) -> uu_homog_results

## kruskal wallis and dunns post hoc test
uu_homog %>% 
  na.omit() %>% 
  group_by(day_post_inf.y) %>% 
  do(tidy(kruskal.test(dist ~ diet,
             data = .))) -> uu_homog_kruskal

uu_homog %>% 
  na.omit() %>% 
  group_by(day_post_inf.y) %>% 
  dunn_test(dist ~ diet,
            p.adjust.method = 'BH',
            data = .) -> uu_homog_dunn
```

**Weighted UniFrac Homogeneity Plot**

```{r, warning=FALSE, fig.width=14, fig.height=5}
wu_homog %>% 
  ggplot(aes(x = day_post_inf.y, y = dist)) +
    geom_boxplot((aes(group = day_post_inf.y)), outlier.shape = NA) +
    geom_vline(xintercept = -3, linetype = 'dashed', color = 'red', size = 0.2) +
    geom_vline(xintercept = 0, linetype = 'dashed', color = 'purple', size = 0.2) +
    geom_line(aes(group = mouse_id.y), alpha = 0.1) +
    geom_jitter(width = 0.1, alpha = 0.2) +
    geom_smooth(se = FALSE, method = 'loess', color = 'blue') + 
    scale_x_continuous(breaks = c(-15, -8, -3, 0, 3)) + 
    scale_y_reverse() +
    facet_wrap(~diet, 
               labeller = labeller(diet = diet_labs),
               nrow = 1) +
    theme_bw(base_size = 20) +
    xlab('Days Relative to Infection') +
    ylab('Weighted UniFrac Distance\n(Pairwise Distances within Day)') +
    ggtitle("Microbiome Homogeneity Over Time") -> wu_homog_plot

wu_homog_plot
```

**Weighted UniFrac Homogeneity Linear Modeling**

```{r}
wu_homog %>% 
  group_by(day_post_inf.y) %>% 
  do(tidy(lm(dist ~ (purified_diet.y * seq_depth.y) + high_fat.y + high_fiber.y,
             data = .))) %>% 
  filter(p.value <= 0.05) -> wu_homog_results

## kruskal wallis and dunns post hoc test
wu_homog %>% 
  na.omit() %>% 
  group_by(day_post_inf.y) %>% 
  do(tidy(kruskal.test(dist ~ diet,
             data = .))) -> w_homog_kruskal

wu_homog %>% 
  na.omit() %>% 
  group_by(day_post_inf.y) %>% 
  dunn_test(dist ~ diet,
            p.adjust.method = 'BH',
            data = .) -> w_homog_dunn

stat_plot_prep(wu_homog,
               w_homog_dunn) -> new_w_homog_dunn
```

Weighted UniFrac Homogeneity Dunn's Post Hoc Visualization

```{r, fig.height=4, fig.width=14}
stat_plot(new_w_homog_dunn) -> w_homog_stat_vis

w_homog_stat_vis
```

**Putting Plots Together for my Poster Presentation**\

```{r, fig.width=30, fig.height=9}
plot_grid(wu_resil_plot, wu_homog_plot,
          w_resil_stat_vis, w_homog_stat_vis,
          nrow = 2,
          labels = c('b)', 'c)', 'd)', 'e)'),
          label_size = 22,
          rel_heights = c(1, 0.8), 
          rel_widths = c(0.9, 1)) -> resil_homog_together

resil_homog_together
```

**Writing all Statistical Results to a .tsv**

```{r}
write_tsv(wu_homog_results,
          '../stats/wu_homogeneity.tsv')
write_tsv(uu_homog_results,
          '../stats/uu_homogeneity.tsv')
write_tsv(uu_resil_results,
          '../stats/uu_resiliency.tsv')
write_tsv(wu_resil_results,
          '../stats/wu_resiliency.tsv')
```

**Saving all Plots**

```{r}
ggsave("wu_resiliency.pdf", 
       plot = wu_resil_plot,
       width = 14, 
       height = 5,
       path = '../plots')
ggsave("uu_resiliency.pdf", 
       plot = uu_resil_plot,
       width = 14, 
       height = 5,
       path = '../plots')
ggsave("wu_homogeneity.pdf", 
       plot = wu_homog_plot,
       width = 14, 
       height = 5,
       path = '../plots')
ggsave("uu_homogeneity.pdf", 
       plot = uu_homog_plot,
       width = 14, 
       height = 5,
       path = '../plots')
ggsave("wu_resil_stats.pdf", 
       plot = w_resil_stat_vis,
       width = 14, 
       height = 4,
       path = '../plots')
ggsave("wu_homog_stats.pdf", 
       plot = w_homog_stat_vis,
       width = 14, 
       height = 4,
       path = '../plots')
ggsave("resil_homog_stat_vis.pdf", 
       plot = resil_homog_together,
       width = 30, 
       height = 9,
       path = '../plots')
```
