---
title: "bile_acid"
output: html_document
date: "2023-12-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(rstatix)
library(ggplot2)
library(ggpubr)
library(ggh4x)
library(cowplot)
library(magrittr)
library(glue)
library(apppleplots)
```

**File Paths**
```{r}
bile_acid_FP <- '../data/misc/processed_bile_acid.tsv'
bile_ratio_FP <- '../data/misc/processed_ratio_bileAcid.tsv'

cdiff_effect_labs <- c('C. difficile Inhibitor',
                       'C. difficile Promoter')
names(cdiff_effect_labs) <- c('inhibitor',
                              'promoter')

dca_sep_labs <- c('Deoxycholic Acid',
                 'C. difficile Inhibitor',
                 'C. difficile Promoter')
names(dca_sep_labs) <- c('DCA',
                         'inhibitor',
                         'promoter')

bile_labs <- c('a-MCA',
               'b-MCA',
               'LCA',
               'DCA',
               'TCA',
               'CA')
names(bile_labs) <- c("Alpha_muricholic_acid_a-MCA",
                     "Beta_muricholic_acid_b-MCA",
                     "Lithocholic_acid_LCA", 
                     "Deoxycholic_acid_DCA",
                     "Taurocholic_acid_T-CA",
                     "Cholic_acid_CA")

bile_groups <- c('c_diff_effect',
                 'bile_acid')
non_bile_groups <- c('c_diff_effect')

diet_x_labs <- c('Chow',
                 'HFt/\nHFb',
                 'HFt/\nLFb',
                 'LFt/\nHFb',
                 'LFt/\nLFb')
```

Use prescence/abscence tests or rank-based test instead of a non-parametric statistical test!!
- Look at whether you'd like to include additional bile acids in the promoters/inhibitors categories

**Reading in Needed Files**
```{r}
big_bile_acid <- read_tsv(bile_acid_FP)
bile_ratio <- read_tsv(bile_ratio_FP)

## changing the c.diff effect column to include DCA for a special plot
dca_sep_bileAcid <- big_bile_acid %>% 
  mutate(c_diff_effect = ifelse(bile_acid == "Deoxycholic_acid_DCA", "DCA", c_diff_effect))
```

**Stats**
Had to filter out CDCA and UDCA from the 'inhibitors' category because the dunn_test wouldn't run - I think it's because they're all zeroes so there aren't any pairwise comparisons to do between diets for them. Maybe include T-CDCA and T-DCA as inhibitors instead?
```{r}
## not faceted by bile acid
no_bile_krDunn <- apppleplots::kruskal_dunn_stats(input_table = big_bile_acid,
                                                  grouped_by = non_bile_groups,
                                                  adjust_method = 'BH',
                                                  filter_adj_p_value = FALSE,
                                                  formula_left = 'conc_normalized',
                                                  formula_right = 'diet')
kruskal <- no_bile_krDunn$KruskalTest
dunn <- no_bile_krDunn$DunnTest

dunn <- dunn %>% 
          add_y_position(scales = 'free',
                         fun = 'max',
                         y.trans = log10)

## with dca separated out
dca_sep_krDunn <- apppleplots::kruskal_dunn_stats(input_table = dca_sep_bileAcid,
                                                  grouped_by = non_bile_groups,
                                                  adjust_method = 'BH',
                                                  filter_adj_p_value = FALSE,
                                                  formula_left = 'conc_normalized',
                                                  formula_right = 'diet')
dca_sep_kruskal <- dca_sep_krDunn$KruskalTest
dca_sep_dunn <- dca_sep_krDunn$DunnTest

dca_sep_dunn <- dca_sep_dunn %>% 
          add_y_position(scales = 'free',
                         fun = 'max',
                         y.trans = log10)

## faceted by bile acid 
bile_krDunn <- apppleplots::kruskal_dunn_stats(input_table = big_bile_acid,
                                               grouped_by = bile_groups,
                                               adjust_method = 'BH',
                                               filter_adj_p_value = FALSE,
                                               formula_left = 'conc_normalized',
                                               formula_right = 'diet')

kruskal_bile <- bile_krDunn$KruskalTest
dunn_bile <- bile_krDunn$DunnTest

dunn_bile <- dunn_bile %>% 
              add_y_position(scales = 'free',
                             fun = 'max',
                             y.trans = log10)
```

**Plots**
```{r, fig.width=6, fig.height=3.5}
overall_plot <- apppleplots::facet_once_plots(input_table = big_bile_acid,
                                              stat_on_plot = FALSE,
                                              x_value = 'diet',
                                              y_value = 'conc_normalized',
                                              x_value_type = 'discrete',
                                              grouped_by = 'diet',
                                              point_alpha = 0.4,
                                              x_labels = diet_x_labs,
                                              facet_by = 'c_diff_effect',
                                              facet_labels = cdiff_effect_labs,
                                              stat_table = NA,
                                              x_name = 'Diet',
                                              y_name = 'Concentration (pmol/mg) (log10)',
                                              title_content = 'Cecal Bile Acid Concentration')

overall_plot <- overall_plot +
                  scale_y_continuous(trans = 'log10') +
                  stat_pvalue_manual(dunn,
                                     tip.length = 0.01,
                                     step.increase = 0.05,
                                     label = 'p.adj.signif',
                                     hide.ns = TRUE)

overall_plot
```

same as the plot above but there's a separate facet for dca rather than it being combined with the c. diff inhibitors facet
```{r, fig.height=3.5, fig.width=7}
dca_sep_plot <- apppleplots::facet_once_plots(input_table = dca_sep_bileAcid,
                                              stat_on_plot = FALSE,
                                              x_value = 'diet',
                                              y_value = 'conc_normalized',
                                              x_value_type = 'discrete',
                                              grouped_by = 'diet',
                                              point_alpha = 0.4,
                                              x_labels = diet_x_labs,
                                              facet_by = 'c_diff_effect',
                                              facet_labels = dca_sep_labs,
                                              stat_table = NA,
                                              x_name = 'Diet',
                                              y_name = 'Concentration (pmol/mg) (log10)',
                                              title_content = 'Cecum Bile Acid Concentration')

dca_sep_plot <- dca_sep_plot +
                  scale_y_continuous(trans = 'log10') +
                  stat_pvalue_manual(dca_sep_dunn,
                                     tip.length = 0.01,
                                     step.increase = 0.05,
                                     label = 'p.adj.signif',
                                     hide.ns = TRUE)

dca_sep_plot
```


```{r, fig.width=8, fig.height=6}
bile_plot <- apppleplots::facet_once_plots(input_table = big_bile_acid,
                                              stat_on_plot = FALSE,
                                              x_value = 'diet',
                                              y_value = 'conc_normalized',
                                              x_value_type = 'discrete',
                                              grouped_by = 'diet',
                                              point_alpha = 0.4,
                                              x_labels = diet_x_labs,
                                              facet_by = 'c_diff_effect',
                                              facet_labels = cdiff_effect_labs,
                                              stat_table = NA,
                                              x_name = 'Diet',
                                              y_name = 'Concentration (pmol/mg) (log10)',
                                              title_content = 'Cecal Bile Acid Concentration')

bile_plot <- bile_plot +
              scale_y_continuous(trans = 'log10') +
              facet_nested_wrap(~c_diff_effect + bile_acid,
                        labeller = labeller(c_diff_effect = cdiff_effect_labs,
                                            bile_acid = bile_labs),
                        nrow = 2) +
              stat_pvalue_manual(dunn_bile,
                                 tip.length = 0.01,
                                 step.increase = 0.05,
                                 label = 'p.adj.signif',
                                 hide.ns = TRUE)
bile_plot
```

**Ratio of C. difficile Promoter:Inhibitor**
```{r, fig.width=5, fig.height=3.5}
## stats
ratio_krDunn <- apppleplots::kruskal_dunn_stats(input_table = bile_ratio,
                                                grouped_by = 'ratio_label',
                                                adjust_method = 'BH',
                                                filter_adj_p_value = FALSE,
                                                formula_left = 'ratio',
                                                formula_right = 'diet')
ratio_kruskal <- ratio_krDunn$KruskalTest
ratio_dunn <- ratio_krDunn$DunnTest

ratio_dunn <- ratio_dunn %>% 
                add_y_position(scales = 'free_y',
                               fun = 'max')
## plot
ratio_bile_plot <- apppleplots::facet_once_plots(input_table = bile_ratio,
                                                 stat_on_plot = TRUE,
                                                 x_value = 'diet',
                                                 y_value = 'ratio',
                                                 x_value_type = 'discrete',
                                                 grouped_by = 'diet',
                                                 point_alpha = 0.4,
                                                 x_labels = diet_x_labs,
                                                 facet_by = 'ratio_label',
                                                 facet_labels = NULL,
                                                 stat_table = ratio_dunn,
                                                 x_name = 'Diet',
                                                 y_name = 'Ratio',
                                                 title_content = 'Cecal Bile Acid Ratio')

ratio_bile_plot
```

**Saving my Outputs**
```{r}
## plots
ggsave('../plots/overall_bile_acid.pdf',
       plot = overall_plot,
       width = 6,
       height = 3.5)
ggsave('../plots/bile_acid.pdf',
       plot = bile_plot,
       width = 8,
       height = 6)
ggsave('../plots/bile_acid_ratio.pdf',
       plot = ratio_bile_plot,
       width = 5,
       height = 3.5)
ggsave('../plots/dcaSep_overall_bileAcid.pdf',
       plot = dca_sep_plot,
       width = 7,
       height = 3.5)

## rdata object 
save(overall_plot,
     file = '../../figures/overall_bile_acid.rdata')
save(bile_plot,
     file = '../../figures/bile_acid.rdata')
save(ratio_bile_plot,
     file = '../../figures/bile_acid_ratio.rdata')
save(dca_sep_plot,
     file = '../../figures/dcaSep_overall_bileAcid.rdat')

## stats
write_tsv(dunn,
          '../stats/overall_bile_dunn.tsv')
write_tsv(dunn_bile,
          '../stats/bile_acid_dunn.tsv')
write_tsv(ratio_dunn,
          '../stats/ratio_bile_acid.tsv')

## saving processed bile acid files so I don't have to keep reprocessing them
# write_tsv(big_bile_acid,
#           '../data/misc/processed_bile_acid.tsv')
# write_tsv(bile_ratio,
#           '../data/misc/processed_ratio_bileAcid.tsv')
```



