---
title: "metabolomics"
format: html
editor: visual
---

```{r, include=FALSE}
library(broom)
library(cowplot)
library(magrittr)
library(qiime2R)
library(tidyverse)
library(naniar)
library(ggpubr)
library(rstatix)
```

```{r}
# change to scripts directory if not there already
# curr_dir <- getwd()
# curr_dir <- str_split(curr_dir, '\\/')
# if (curr_dir[length(curr_dir)] != 'scripts'){
#   setwd('./scripts')
# }
```

**Basic Functions**

```{r}
## general function to prep the metadata file for further data analyses 
metadata_fixer <- function(metadata_fp) {
  tmpMeta <- read_tsv(metadata_fp, n_max = 2)
  mycols <- colnames(tmpMeta)
  metadata <- read_tsv(metadata_fp, skip = 2, col_names = mycols)
  names(metadata)[names(metadata) == '#SampleID'] <- 'sampleid'
  metadata %>% 
    filter(!is.na(diet)) %>% 
    mutate(day_post_inf = if_else(day_post_inf == 2, 3, day_post_inf)) %>% 
    mutate(diet = as.factor(diet)) -> metadata
  return(metadata)
}

## for editing my metadata file post metadata fixer 
meta_diet_fixer <- function(metadata_file,
                            seq_depth_fp){
  seq_depths <- read_tsv(seq_depth_fp)
  metadata_file %>% 
    select(sampleid, diet, day_post_inf, mouse_id, study) %>% 
    mutate(diet_true = diet,
    diet_true = if_else(day_post_inf == -15, "Chow", diet_true),
    high_fat = case_when(
      diet_true == 'HF/HF' ~ 1,
      diet_true == 'HF/LF' ~ 1,
      .default = 0
      ), 
      high_fiber = case_when(
      diet_true == 'HF/HF' ~ 1,
      diet_true == 'LF/HF' ~ 1,
      diet_true == 'Chow' ~ 1,
      .default = 0
      ), 
      purified_diet = case_when(
      diet_true == 'Chow' ~ 0,
      .default = 1
      )
    ) %>% 
    left_join(seq_depths) -> metadata
  return(metadata)
}
```

**File Paths**

```{r}
metadata_FP <- '../data/misc/merged_metadata1.tsv'
metab_FP <- '../data/misc/metabolomics.csv'
seq_depth_FP <- '../data/misc/tss_seq_depth.tsv'
```

**Prepping Metadata and Metabolomics Files**

```{r}
## metabolomics file prep 
metab <- read_csv(metab_FP)
wanted_ids <- metab$mouse_id

## metadta file prep 
metadata <- metadata_fixer(metadata_FP)
metadata <- meta_diet_fixer(metadata,
                            seq_depth_FP)
metadata %>% 
  group_by(mouse_id) %>% 
  filter(mouse_id %in% wanted_ids) -> meta_filt

meta_filt %>% 
  left_join(metab, by = 'mouse_id') %>% 
  gather('Acetic Acid (ug/g)', 'Propanoic Acid (ug/g)', '2-methyl-propanoic acid (ug/g)',
         'n-Butanoic Acid (ug/g)', 'Isopentanoic Acid (ug/g)', '2-methyl-Butanoic Acid (ug/g)',
         'Pentanoic Acid (ug/g)', key = metabolite, value = concentration) %>% 
  filter(!is.na(mouse_id)) -> pre_metab

## changes all 'ND' values in the concentration column to 0 
pre_metab$concentration[pre_metab$concentration == 'ND'] <- 0

pre_metab %>% 
  select(-'Notes', -'Sample Group', -'SCFA Data File',
         -'Acq. Date-Time', -'Tube_Label', -'Sample_Type',
         -'Collection Date', -'Dil.') %>%
  filter(!is.na(concentration)) %>% 
  mutate(concentration = as.numeric(concentration)) %>% 
  distinct(mouse_id, concentration, .keep_all = TRUE) -> big_metab
```

**Putting Together Stats for the Ggplot**\

```{r}
## generating stats to place on ggplot for visualization
## running a kruskal-wallis test on the metabolite data 
big_metab %>% 
  group_by(metabolite) %>% 
  do(tidy(kruskal.test(concentration ~ diet,
             data = .))) -> metab_kruskal

## running dunn's post hoc test on the metabolite data 
big_metab %>% 
  group_by(metabolite) %>% 
  dunn_test(concentration ~ diet,
            p.adjust.method = 'bonferroni',
            data = .) %>% 
  add_y_position(scales = 'free_y') -> metab_dunn
```

**Ggplot Construction for Metabolites**

Very rough draft of the plot for the metabolite data - I know that I will probably need to edit this.

```{r, fig.width=35, fig.height=6}
metab_labs <- c('Acetic Acid', 'Propanoic Acid', '2-methyl-propanoic acid',
                'n-Butanoic Acid', 'Isopentanoic Acid', '2-methyl-Butanoic Acid',
                'Pentanoic Acid')
names(metab_labs) <- c('Acetic Acid (ug/g)', 'Propanoic Acid (ug/g)', 
                       '2-methyl-propanoic acid (ug/g)','n-Butanoic Acid (ug/g)', 
                       'Isopentanoic Acid (ug/g)', '2-methyl-Butanoic Acid (ug/g)',
                       'Pentanoic Acid (ug/g)')
big_metab %>% 
  ggplot(aes(x = diet, y = concentration)) +
  geom_violin(aes(group = diet)) +
  geom_jitter(alpha = 0.4, width = 0.1, height = 0) +
  scale_x_discrete(labels = c('Chow', 'High Fat/\nHigh Fiber', 'High Fat/\nLow Fiber',
                              'Low Fat/\nHigh Fiber', 'Low Fat/\nLow Fiber')) +
  facet_wrap(~metabolite, 
             nrow = 1,
             scales = 'free_y',
             labeller = labeller(metabolite = metab_labs)) +
  theme_bw(base_size = 14) +
  stat_pvalue_manual(metab_dunn,
                     tip.length = 0.01,
                     label = 'p.adj.signif',
                     hide.ns = TRUE) +
  xlab('Diet') +
  ylab('Concentration (ug/g)') +
  ggtitle('Metabolite Concentration by Mouse Diet') -> metab_plot

metab_plot
```

**Linear Modeling**

Grouped the stats per metabolite detected by the core and compared against concentration of said metabolite per diet. I also am not sure if this is the best way to run the stats but it's a start.

```{r}
big_metab %>%
  group_by(metabolite) %>% 
  do(tidy(lm(concentration ~ diet,
             data = .))) %>% 
  adjust_pvalue(method = 'bonferroni') -> metab_results

metab_results
```

**Saving my Outputs**

```{r}
## plot
ggsave("metabolomics.pdf",
       plot = metab_plot,
       width = 35,
       height = 6,
       path = '../plots')

## statistical tests 
write_tsv(metab_results,
          '../stats/metab_linear_model.tsv')
write_tsv(metab_dunn,
          '../stats/metab_dunn_test.tsv')
write_tsv(metab_kruskal,
          '../stats/metab_kruskal_test.tsv')
```
