---
title: "metabolomics"
format: html
editor: visual
---

```{r, include=FALSE}
library(broom)
library(cowplot)
library(magrittr)
library(qiime2R)
library(tidyverse)
```

```{r}
# change to scripts directory if not there already
# curr_dir <- getwd()
# curr_dir <- str_split(curr_dir, '\\/')
# if (curr_dir[length(curr_dir)] != 'scripts'){
#   setwd('./scripts')
# }
```

**Basic Functions**

```{r}
## general function to prep the metadata file for further data analyses 
metadata_fixer <- function(metadata_fp) {
  tmpMeta <- read_tsv(metadata_fp, n_max = 2)
  mycols <- colnames(tmpMeta)
  metadata <- read_tsv(metadata_fp, skip = 2, col_names = mycols)
  names(metadata)[names(metadata) == '#SampleID'] <- 'sampleid'
  metadata %>% 
    filter(!is.na(diet)) %>% 
    mutate(day_post_inf = if_else(day_post_inf == 2, 3, day_post_inf)) %>% 
    mutate(diet = as.factor(diet)) -> metadata
  return(metadata)
}

## for editing my metadata file post metadata fixer 
meta_diet_fixer <- function(metadata_file,
                            seq_depth_fp){
  seq_depths <- read_tsv(seq_depth_fp)
  metadata_file %>% 
    select(sampleid, diet, day_post_inf, mouse_id, study) %>% 
    mutate(diet_true = diet,
    diet_true = if_else(day_post_inf == -15, "Chow", diet_true),
    high_fat = case_when(
      diet_true == 'HF/HF' ~ 1,
      diet_true == 'HF/LF' ~ 1,
      .default = 0
      ), 
      high_fiber = case_when(
      diet_true == 'HF/HF' ~ 1,
      diet_true == 'LF/HF' ~ 1,
      diet_true == 'Chow' ~ 1,
      .default = 0
      ), 
      purified_diet = case_when(
      diet_true == 'Chow' ~ 0,
      .default = 1
      )
    ) %>% 
    left_join(seq_depths) -> metadata
  return(metadata)
}
```

**File Paths**

```{r}
metadata_FP <- '../data/misc/merged_metadata1.tsv'
metab_FP <- '../data/misc/metabolomics.csv'
seq_depth_FP <- '../data/misc/tss_seq_depth.tsv'
```

**Prepping Metadata and Metabolomics Files**

```{r}
## metabolomics file prep 
metab <- read_csv(metab_FP)
wanted_ids <- metab$mouse_id

## metadta file prep 
metadata <- metadata_fixer(metadata_FP)
metadata <- meta_diet_fixer(metadata,
                            seq_depth_FP)
metadata %>% 
  group_by(mouse_id) %>% 
  filter(mouse_id %in% wanted_ids) -> meta_filt

meta_filt %>% 
  left_join(metab, by = 'mouse_id') %>% 
  gather('Acetic Acid (ug/g)', 'Propanoic Acid (ug/g)', '2-methyl-propanoic acid (ug/g)',
         'n-Butanoic Acid (ug/g)', 'Isopentanoic Acid (ug/g)', '2-methyl-Butanoic Acid (ug/g)',
         'Pentanoic Acid (ug/g)', key = metabolite, value = concentration) %>% 
  filter(!is.na(mouse_id)) %>% 
  filter(concentration != 'ND') %>% 
  filter(!is.na(concentration)) %>% 
  mutate(conc_num = as.numeric(concentration)) -> big_metab
```

**Ggplot Construction for Metabolites**

Very rough draft of the plot for the metabolite data - I know that I will probably need to edit this.

```{r, fig.width=25, fig.height=8}
big_metab %>% 
  ggplot(aes(x = diet, y = conc_num)) +
  geom_boxplot(aes(group = diet)) +
  scale_y_continuous(trans = 'log10') +
  geom_jitter(alpha = 0.4, width = 0.1, height = 0) +
  facet_wrap(~metabolite, 
             nrow = 1) +
  theme_bw(base_size = 14) +
  xlab('Diet') +
  ylab('Concentration') +
  ggtitle('Metabolite Concentration by Mouse Diet') -> metab_plot

metab_plot
```

**Linear Modeling**

Grouped the stats per metabolite detected by the core and compared against concentration of said metabolite per diet. I also am not sure if this is the best way to run the stats but it's a start.

```{r}
big_metab %>%
  group_by(metabolite) %>% 
  do(tidy(lm(conc_num ~ purified_diet + high_fat + high_fiber,
             data = .))) -> metab_results

metab_results
```

**Saving my Outputs**

```{r}
ggsave("metabolomics.pdf",
       plot = metab_plot,
       width = 25,
       height = 8,
       path = '../plots')

write_tsv(metab_results,
          '../stats/metabolomics.tsv')
```
