---
title: "obAnaerobe_investigation"
output: html_document
date: "2024-12-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(qiime2R)
library(tidyverse)
library(cowplot)
library(magrittr)
library(vegan)
library(viridis)
library(broom)
library(rstatix)
library(ggpubr)
library(ggh4x)
library(apppleplots)
library(viridis)
```

**Functions**
```{r}
## abundance file prep and filtering function
## combines all files into one large table
abun_file_prep <- function(metadata_fp,
                           tax_fp,
                           otu_table_fp,
                           tax_level,
                           wanted_tax){
  ## metadata
  metadata <- read_tsv(metadata_fp)
  names(metadata)[names(metadata) == '#SampleID'] <- 'sampleid'
  ## taxonomy
  taxonomy <- read_qza(tax_fp)$data %>% 
  parse_taxonomy() %>% 
  rownames_to_column('asv')
  ## otu table 
  otu_table <- read_qza(otu_table_fp)$data
  otu_table %>% 
    as_tibble(rownames = 'asv') %>% 
    gather(-asv, key = sampleid, value = abun) %>% 
    group_by(sampleid) %>% 
    mutate(rel_abun = abun/sum(abun)) %>% 
    mutate(rel_abun = rel_abun + 0.000001) -> otu_table
  ## joining all tables together 
  otu_table %>% 
    left_join(metadata, by = 'sampleid') %>% 
    left_join(taxonomy, by = 'asv') -> abun_table
  abun_table %>% 
    group_by(sampleid, day_post_inf, diet, mouse_id,
             .data[[tax_level]], Genus) %>% 
    summarise(rel_abund = sum(rel_abun)) %>% 
    filter(.data[[tax_level]] %in% wanted_tax) %>% 
    mutate(mouse_fact = as.factor(mouse_id),
           day_fact = as.factor(day_post_inf)) %>% 
    filter(!is.na(diet)) -> abun_filt
  ## creating a list for my outputs
  my_list <- list(Metadata = metadata,
                  Taxonomy = taxonomy,
                  OTUTable = otu_table,
                  AbundanceTable = abun_filt)
  return(my_list)
}

## plot for the top 5 genera in a family according to relative abundance 
sum_abun_plot <- function(input_table,
                          x_axis, 
                          y_axis,
                          bar_group_by,
                          bar_fill_by,
                          nudge_text_y,
                          x_axis_order,
                          x_name,
                          y_name,
                          title_content){
  plot <- input_table %>% 
    ggplot(aes(x = .data[[x_axis]], y = .data[[y_axis]])) +
    geom_bar(aes(group = .data[[bar_group_by]], fill = .data[[bar_fill_by]]), stat = 'identity', color = 'black') +
    geom_text(aes(label = .data[[y_axis]]), nudge_y = nudge_text_y, fontface = 'bold') +
    scale_x_discrete(limit = x_axis_order) +
    scale_y_continuous(expand = c(0.025, 0.025)) +
    scale_fill_discrete(breaks = x_axis_order) +
    theme_bw(base_size = 20) +
    theme(legend.title = element_blank(),
          axis.text.x = element_blank()) +
    labs(x = x_name,
         y = y_name,
         title = title_content)
  
  return(plot)
}

## plot of trend lines of genera per family only (no points/boxes)
smooth_abun_plot <- function(input_table,
                             x_axis,
                             y_axis,
                             line_group_by,
                             line_color_by,
                             viridis_palette,
                             legend_order,
                             legend_labels,
                             x_axis_breaks,
                             facet_by,
                             facet_labels,
                             x_name,
                             y_name,
                             title_content){
  plot <- input_table %>% 
    ggplot(aes(x = .data[[x_axis]], y = .data[[y_axis]])) +
    geom_smooth(aes(group = .data[[line_group_by]]), color = 'black',
                linewidth = 1.5, method = 'loess', se = FALSE, show.legend = T) +
    geom_smooth(aes(group = .data[[line_group_by]], color = .data[[line_color_by]]), 
                se = FALSE, method = 'loess', alpha = 0.7) +
    # guides(color = guide_legend(override.aes=list(shape = 22))) +
    scale_color_viridis(option = viridis_palette,
                        discrete = TRUE,
                        breaks = legend_order,
                        labels = legend_labels) +
    theme_bw(base_size = 20) +
    scale_y_continuous(transform = 'log10') +
    scale_x_continuous(breaks = x_axis_breaks,
                       labels = x_axis_breaks) +
    facet_wrap(~.data[[facet_by]],
               nrow = 1,
               labeller = labeller(.cols = facet_labels)) +
    labs(y = y_name,
         x = x_name,
         title = title_content)
  
  return(plot)
}

## preps dunns post hoc results for statistical visualization
stat_plot_prep <- function(filtered_table,
                           first_group,
                           second_group,
                           mean_value,
                           dunn_test){
  filtered_table %>% 
    group_by(filtered_table[first_group], filtered_table[second_group]) %>%
    summarise(mean = mean(.data[[mean_value]])) -> mean_table
  
  dunn_test %>% 
    merge(mean_table, 
          by.x = c('group1',
                   second_group),
          by.y = c(first_group,
                   second_group)) %>%
    rename_with(~paste0('group1_', mean_value, recycle0 = TRUE), contains('mean')) %>% 
    merge(mean_table,
          by.x = c('group2',
                   second_group),
          by.y = c(first_group,
                   second_group)) %>%
    rename_with(~paste0('group2_', mean_value, recycle0 = TRUE), contains('mean')) -> int_dunn
  
  group1_col <- paste0('group1_', mean_value)
  group2_col <- paste0('group2_', mean_value)
  
  int_dunn %>% 
    mutate(diff_means = (.data[[group1_col]] - .data[[group2_col]]),
           stat_diff_means = if_else(p.adj > 0.05, 0, diff_means)) -> new_dunn
  
  return(new_dunn)
}
```

**File paths**
```{r}
otu_table_FP <- '../data/comp_qiime/tax_s1_filt.qza'
tax_FP <- '../data/comp_qiime/taxonomy.qza'
metadata_FP <- '../data/misc/s1_filt_comp_metadata.tsv'

diet_labs <- 
  c('Chow', 
    'High Fat / High Fiber', 
    'High Fat / Low Fiber', 
    'Low Fat / High Fiber', 
    'Low Fat / Low Fiber')
names(diet_labs) <- c('Chow', 
                      'HF/HF', 
                      'HF/LF', 
                      'LF/HF', 
                      'LF/LF')

diet_y_labs <- c('HFt / HFb',
                 'HFt / LFb',
                 'LFt / HFb',
                 'LFt / LFb')

day_x_breaks <- c(-15, -8, -3, 0, 3)

wanted_level <- 'Family'

wanted_family <- c('Lachnospiraceae',
                   'Ruminococcaceae',
                   'Enterobacteriaceae',
                   'Peptostreptococcaceae',
                   'Enterococcaceae',
                   'Staphylococcaceae')

path_genus_labs <- c('Clostridioides',
                     'Enterococcus',
                     'Unclassified\nEnterobacteriaceae',
                     'Staphylococcus')
names(path_genus_labs) <- c(1, 
                            2, 
                            3, 
                            4)
```

**Data wrangled**
```{r}
abun_files <- abun_file_prep(metadata_fp = metadata_FP,
                             tax_fp = tax_FP,
                             otu_table_fp = otu_table_FP,
                             tax_level = wanted_level,
                             wanted_tax = wanted_family)

abun_table <- abun_files$AbundanceTable

abun_files$Taxonomy
abun_files$OTUTable
```
## **Pulling top genera per family**
```{r}
## lachnospiraceae
top_lachno <- abun_table %>% 
  filter(Family == 'Lachnospiraceae') %>% 
  group_by(Genus) %>% 
  summarise(sum_abund = round(sum(rel_abund), digits = 3)) %>% 
  slice_max(sum_abund, n = 15)

## ruminococcaceae
top_rumino <- abun_table %>% 
  filter(Family == 'Ruminococcaceae') %>% 
  group_by(Genus) %>% 
  summarise(sum_abund = round(sum(rel_abund), digits = 3)) %>% 
  slice_max(sum_abund, n = 15)
```

**Lachnospiraceae plot**
```{r, fig.width=12, fig.height=5}
lachno_order <- c("Lachnospiraceae_NK4A136_group",
                  NA,
                  "Lachnoclostridium",
                  "Blautia",
                  "Roseburia",
                  "uncultured",
                  "Marvinbryantia",
                  "[Eubacterium]_fissicatena_group",
                  "A2",
                  "Lachnospiraceae_UCG-001")

lachno_plot <- sum_abun_plot(input_table = top_lachno,
                             x_axis = 'Genus',
                             y_axis = 'sum_abund',
                             bar_group_by = 'Genus',
                             bar_fill_by = 'Genus',
                             nudge_text_y = 0.25,
                             x_axis_order = lachno_order,
                             x_name = 'Genus',
                             y_name = 'Sum Abundance',
                             title_content = 'Lachnospiraceae')

lachno_plot
```

**Ruminococcaceae plot**
```{r, fig.width=12, fig.height=5}
rumino_order <- c("Incertae_Sedis",
                  "Anaerotruncus",
                  "UBA1819",
                  "uncultured",
                  "Paludicola",
                  "Ruminococcaceae",
                  "CAG-352",
                  NA,
                  "Ruminococcus",
                  "Caproiciproducens")

rumino_plot <- sum_abun_plot(input_table = top_rumino,
                             x_axis = 'Genus',
                             y_axis = 'sum_abund',
                             bar_group_by = 'Genus',
                             bar_fill_by = 'Genus',
                             nudge_text_y = 0.05,
                             x_axis_order = rumino_order,
                             x_name = 'Genus',
                             y_name = 'Sum Abundance',
                             title_content = 'Ruminococccaceae')

rumino_plot
```

**Putting plots together**
```{r, fig.width=12, fig.height=10}
obAnaerobe_plots <- plot_grid(lachno_plot,
                              rumino_plot,
                              ncol = 1,
                              align = 'hv',
                              axis = 'tblr')

obAnaerobe_plots
```

## **Relative abundance plots**

**Data wrangled (part two)**
```{r}
## lachnospiraceae
wanted_lachno <- c('Lachnospiraceae_NK4A136_group', 
                   'unclassified',
                   'Lachnoclostridium', 
                   'Blautia', 
                   'Roseburia')

lachno_abun_table <- abun_table %>% 
  mutate(Genus = ifelse(is.na(Genus), 'unclassified', Genus)) %>% 
  filter(Family == 'Lachnospiraceae',
         Genus %in% wanted_lachno)

## ruminococcaceae
wanted_rumino <- c('Incertae_Sedis',
                   'Anaerotruncus',
                   'UBA1819',
                   'uncultured')

rumino_abun_table <- abun_table %>% 
  filter(Family == 'Ruminococcaceae',
         Genus %in% wanted_rumino)

## potential pathogens
path_families <- c('Enterobacteriaceae',
                   'Peptostreptococcaceae',
                   'Enterococcaceae',
                   'Staphylococcaceae')
wanted_pathogens <- c('Clostridioides',
                      'Enterococcus',
                      'Unclassified\nEnterobacteriaceae',
                      'Staphylococcus')

path_abun_table <- abun_table %>% 
  filter(Family %in% path_families) %>% 
  mutate(Genus = ifelse(is.na(Genus), paste0('Unclassified', '\n', Family), Genus)) %>% 
  filter(Genus %in% wanted_pathogens) %>% 
  mutate(genus_order = case_when(
    Genus == 'Clostridioides' ~ 1,
    Genus == 'Enterococcus' ~ 2,
    Genus == 'Unclassified\nEnterobacteriaceae' ~ 3,
    Genus == 'Staphylococcus' ~ 4
  ))
```

**Lachnospiraceae plot**
```{r, fig.width=20, fig.height=5}
## old lachnospiraceae plot (had no points, just geom smooth lines, need to include points moving forward)
lachno_legend_labs <- c('Lachnospiraceae\nNK4A136 group',
                        'Unclassified',
                        'Lachnoclostridium',
                        'Blautia',
                        'Roseburia')


(lachno_abun_plot <- smooth_abun_plot(input_table = lachno_abun_table,
                                     x_axis = 'day_post_inf',
                                     y_axis = 'rel_abund',
                                     line_group_by = 'Genus',
                                     line_color_by = 'Genus',
                                     viridis_palette = 'C',
                                     legend_order = wanted_lachno,
                                     legend_labels = lachno_legend_labs,
                                     x_axis_breaks = day_x_breaks,
                                     facet_by = 'diet',
                                     facet_labels = diet_labs,
                                     x_name = 'Days Relative to Infection',
                                     y_name = 'Relative Abundance (log10)',
                                     title_content = 'Lachnospiraceae'))
```

**Ruminococcaceae plot**
```{r, fig.width=20, fig.height=5}
rumino_legend_labs <- c('Incertae Sedis',
                        'Anaerotruncus',
                        'UBA1819',
                        'Uncultured')

(rumino_abun_plot <- smooth_abun_plot(input_table = rumino_abun_table,
                                     x_axis = 'day_post_inf',
                                     y_axis = 'rel_abund',
                                     line_group_by = 'Genus',
                                     line_color_by = 'Genus',
                                     viridis_palette = 'C',
                                     legend_order = wanted_rumino,
                                     legend_labels = rumino_legend_labs,
                                     x_axis_breaks = day_x_breaks,
                                     facet_by = 'diet',
                                     facet_labels = diet_labs,
                                     x_name = 'Days Relative to Infection',
                                     y_name = 'Relative Abundance (log10)',
                                     title_content = 'Ruminococcaceae'))
```

**Obligate anaerobes plot** 
done at family level since there are several genera in these families to get the general trend over time
```{r, fig.width=16.5, fig.height=6.5}
## new obligate anaerobes plot, put lachno and rumino in the same plot now
## have to sum the relative abundance per family since was broken down to genus level 
obAn_families <- c('Lachnospiraceae',
                   'Ruminococcaceae')

obAn_abun_table <- abun_table %>% 
  filter(Family %in% obAn_families) %>% 
  group_by(sampleid, day_post_inf, diet, mouse_id, Family) %>% 
  summarise(total_rel_abund = sum(rel_abund))

## plot actually includes points/boxplots now 
(obAn_abun_plot <- apppleplots::facet_twice_plots(input_table = obAn_abun_table,
                                                      x_value = 'day_post_inf',
                                                      y_value = 'total_rel_abund',
                                                      x_value_type = 'numeric',
                                                      y_transform = TRUE,
                                                      x_labels = c(-15, -8, -3, 0, 3),
                                                      box_group_by = 'day_post_inf',
                                                      line_group_by = 'mouse_id',
                                                      point_alpha = 0.4,
                                                      facet_rows = 'Family',
                                                      facet_cols = 'diet',
                                                      row_labs = NULL,
                                                      col_labs = diet_labs,
                                                      x_name = 'Days Relative to Infection',
                                                      y_name = 'Relative Abundance (log10)',
                                                      title_content = 'Obligate Anaerobe Relative Abundance') +
                          theme_bw(base_size = 20) +
                          theme(strip.text.y.right = element_text(angle = 0)))
```

**Potential pathogen plot**
unclassified enterobacteriaceae, enterococcus, staphylococcus, and clostridioides
```{r, fig.width=16, fig.height=8}
path_abun_plot <- apppleplots::facet_twice_plots(input_table = path_abun_table,
                                                 x_value = 'day_post_inf',
                                                 y_value = 'rel_abund',
                                                 x_value_type = 'numeric',
                                                 y_transform = TRUE,
                                                 x_labels = c(-15, -8, -3, 0, 3),
                                                 box_group_by = 'day_post_inf',
                                                 line_group_by = 'mouse_id',
                                                 point_alpha = 0.4,
                                                 facet_rows = 'genus_order',
                                                 facet_cols = 'diet',
                                                 row_labs = path_genus_labs,
                                                 col_labs = diet_labs,
                                                 x_name = 'Days Relative to Infection',
                                                 y_name = 'Relative Abundance (log10)',
                                                 title_content = 'Potential Pathogen Relative Abundance') +
  theme_bw(base_size = 20) +
  theme(strip.text.y.right = element_text(angle = 0))

path_abun_plot
```

**Putting relative abundance plots together**
old version 
```{r, fig.width=22, fig.height=21}
plots_together <- plot_grid(lachno_abun_plot +
                                  theme(axis.title.x = element_blank()) +
                                  labs(y = '',
                                       subtitle = 'Lachnospiraceae',
                                       title = 'Obligate Anaerobe Relative Abundance'),
                                rumino_abun_plot +
                                  theme(plot.title = element_blank()) +
                                  labs(x = 'Days Relative to Infection',
                                       y = '',
                                       subtitle = 'Ruminococcaceae'),
                                path_abun_plot,
                                ncol = 1,
                                labels = c('a)', ' ', 'b)'),
                                label_size = 20,
                                rel_heights = c(0.45, 0.45, 1),
                                align = 'hv',
                                axis = 'tblr')

genusAbun_plots_together_withLabs_old <- ggdraw(plots_together) +
  draw_label('Relative Abundace (log10)', x = 0, y = 0.765, angle = 90, size = 20, vjust = 1.5)


genusAbun_plots_together_withLabs_old
```

new version for manuscript (that we're moving forward with)
```{r, fig.width=17, fig.height=15}
(genusAbun_plots_together_withLabs <- plot_grid(obAn_abun_plot,
                                                path_abun_plot,
                                                ncol = 1,
                                                labels = c('a)', 'b)'),
                                                label_size = 20,
                                                rel_heights = c(0.7, 1),
                                                align = 'hv',
                                                axis = 'tblr'))
```

## **Stats**
**Data wrangled (part three)**
```{r}
diff_abun_tables <- rbind(lachno_abun_table,
                          rumino_abun_table,
                          path_abun_table)

diff_abun_tables <- diff_abun_tables %>% 
  mutate(log_rel_abund = log10(rel_abund),
         Genus = ifelse(Genus == 'unclassified' | Genus == 'uncultured', 
                        paste0(Genus, '\n', Family), Genus),
         genus_order = case_when(
           Genus == 'Lachnospiraceae_NK4A136_group' ~ 1, 
           Genus == 'unclassified\nLachnospiraceae' ~ 2,
           Genus == 'Lachnoclostridium' ~ 3, 
           Genus == 'Blautia' ~ 4, 
           Genus == 'Roseburia' ~ 5,
           Genus == 'Incertae_Sedis' ~ 6,
           Genus == 'Anaerotruncus' ~ 7,
           Genus == 'UBA1819' ~ 8,
           Genus == 'uncultured\nRuminococcaceae' ~ 9,
           Genus == 'Clostridioides' ~ 10,
           Genus == 'Enterococcus' ~ 11,
           Genus == 'Unclassified\nEnterobacteriaceae' ~ 12,
           Genus == 'Staphylococcus' ~ 13
         ))
```

**Kruskal-Wallis and Dunn's Post Hoc test**
```{r}
abun_krDunn <- apppleplots::kruskal_dunn_stats(input_table = diff_abun_tables,
                                               grouped_by = c('genus_order', 'day_post_inf'),
                                               adjust_method = 'BH',
                                               filter_adj_p_value = FALSE,
                                               formula_left = 'log_rel_abund',
                                               formula_right = 'diet')

abun_kruskal <- abun_krDunn$KruskalTest
abun_dunn <- abun_krDunn$DunnTest

## prepping dunn test for stat visualization
new_abun_dunn <- stat_plot_prep(filtered_table = diff_abun_tables,
                                first_group = 'diet',
                                second_group = c('genus_order', 'day_post_inf'),
                                mean_value = 'log_rel_abund',
                                dunn_test = abun_dunn)
```

**Stat visualization**
```{r, fig.width=16, fig.height=19}
genus_stat_labs <- c('Lachnospiraceae\nNK4A136 group',
                     'Unclassified\nLachnospiraceae',
                     'Lachnoclostridium',
                     'Blautia',
                     'Roseburia',
                     'Incertae Sedis',
                     'Anaerotruncus',
                     'UBA1819',
                     'Uncultured\nRuminococcaceae',
                     'Clostridioides',
                     'Enterococcus',
                     'Unclassified\nEnterobacteriaceae',
                     'Staphylococcus')
names(genus_stat_labs) <- c(1, 
                            2,
                            3, 
                            4, 
                            5,
                            6,
                            7,
                            8,
                            9,
                            10,
                            11,
                            12,
                            13)

abun_stat_plot <- apppleplots::stat_heat_plot(input_table = new_abun_dunn,
                                              x_value = 'day_post_inf',
                                              y_value = 'group2',
                                              x_value_type = 'continuous',
                                              y_value_type = 'discrete',
                                              pairwise = TRUE,
                                              correlation = FALSE,
                                              tile_fill = 'stat_diff_means',
                                              text_value = 'p.adj.signif',
                                              low_color = "#0D0887FF",
                                              high_color = "#CC4678FF",
                                              legend_name = 'Group 1 -\nGroup 2',
                                              x_labels = day_x_breaks,
                                              y_labels = diet_y_labs,
                                              facet_by = 'group1',
                                              facet_labels = diet_labs,
                                              x_name = 'Days Relative to Infection',
                                              y_name = 'Group 2',
                                              title_content = 'Figure 5 Genus Comparisons') +
                    theme_bw(base_size = 20) +
                    theme(strip.text.y.right = element_text(angle = 0),
                          plot.subtitle = element_text(hjust = 0.5)) +
                    facet_grid(rows = vars(genus_order),
                               cols = vars(group1),
                               labeller = labeller(group1 = diet_labs, genus_order = genus_stat_labs))

abun_stat_plot
```


## **Saving my outputs**
```{r}
ggsave('../plots/top_lachno_rumino.pdf',
       plot = obAnaerobe_plots,
       width = 12,
       height = 10)
ggsave('../plots/lachno_genera_by_day.pdf',
       plot = lachno_abun_plot,
       width = 20,
       height = 5)
ggsave('../plots/rumino_genera_by_day.pdf',
       plot = rumino_abun_plot,
       width = 20,
       height = 5)
ggsave('../plots/obAn_abun_by_day.pdf',
       plot = obAn_abun_plot,
       width = 16.5,
       height = 6.5)
ggsave('../plots/potential_new_fig5.pdf',
       plot = genusAbun_plots_together_withLabs,
       width = 17,
       height = 15)
ggsave('../plots/new_fig5_stats.pdf',
       plot = abun_stat_plot,
       width = 16,
       height = 19)
## save with all other figures too 
ggsave('../../../figures/figS5.pdf',
       plot = abun_stat_plot,
       width = 16,
       height = 19)


## as an .rdat object
save(genusAbun_plots_together_withLabs,
     file = '../../../figures/genusAbun_plots_together_withLabs.rdat')
save(obAn_abun_plot,
     file = '../../../figures/obAn_abun_plot.rdat')
save(lachno_abun_plot,
     file = '../../../figures/lachno_abun_plot.rdat')
save(rumino_abun_plot,
     file = '../../../figures/rumino_abun_plot.rdat')
save(path_abun_plot,
     file = '../../../figures/pathogen_abun_plot.rdat')
```


